---
title: "simulate-scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulate-scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
# Load Libraries
devtools::load_all()
library(patchwork) #
library(tidyverse)

directory <- "~/Documents/Research/Code/ECE/dev/"
export_power_file <- paste0(directory, "power-sims.csv")
export_est_file <- paste0(directory, "est-sims.csv")
export_mse_file <- paste0(directory, "mse-sims.csv")
```

# Simulate Scenarios
In this vignette, we will compare the _equivariant covariance estimator_ against some mean-removing alternatives. 

1. `demean`: estimate change points, estimate the segment means, remove means, and estimate Pearson correlation
2. `desmooth`: apply loess smoothing, remove estimated mean, estimate Pearson correlation

We consider a variety of simulations shown below, which have varying levels of mean structure complexity. All of them have minimum segment length of 4. 

```{R, echo=FALSE}
# Plot Scenarios
p1 <- plot_scenario(scenario(1, n = 1000))
p2 <- plot_scenario(scenario(2, n = 1000))
p3 <- plot_scenario(scenario(3, n = 1000))
p4 <- plot_scenario(scenario(4, n = 1000))
p5 <- plot_scenario(scenario(5, n = 1000))
p6 <- plot_scenario(scenario(6, n = 1000))
p7 <- plot_scenario(scenario(7, n = 1000))
p8 <- plot_scenario(scenario(8, n = 1000))
p9 <- plot_scenario(scenario(9, n = 1000))
p10 <- plot_scenario(scenario(10, n = 1000))
p11 <- plot_scenario(scenario(11, n = 1000))

# Show Results
p1 | p2
p3 | p4
p5 | p6
p7 | p8
p9 | p10
p11
```

## Simulate Data

We simulate the results and store them in a csv file. This way, we reduce our computational burden and we can conveniently call plots to provide different views of the data. 

```{R, echo=FALSE}
# Set Simulation Parameters
param_grid <- expand.grid(
  scenario = 1:11,
  method = c("ECE", "demean", "desmooth"),
  sxy = seq(0, 0.5, 0.05),
  sx = 1,
  sy = 1,
  n = c(200, 1000),
  n_sim = 1000
)
```

```{R, eval=FALSE, echo=FALSE}
# Simulate Power Results
progress_ct <- 0 # global counter
sim_power <- pmap_dfr(
  param_grid,
  function(scenario, method, sxy, sx, sy, n, signal, n_sim) {
    progress_ct <<- progress_ct + 1 # update counter
    message("Running: ", progress_ct, "/", nrow(param_grid))
    params <- scenario(
      scenario_num = scenario,
      sxy = sxy,
      sx = sx,
      sy = sy,
      n = n,
      signal = signal
    )
    power <- simulate_power(params, method = method, n_sim = n_sim)
    export_simulations(power, method, params, n_sims, export_power_file)
    tibble(scenario, method, n, n_sim, power)
  }
)
```

```{R, eval=FALSE, echo=FALSE}
# Simulate Estimate Results
progress_ct <- 0 # global counter
sim_est <- pmap_dfr(
  param_grid,
  function(scenario, method, sxy, sx, sy, n, signal, n_sim) {
    progress_ct <<- progress_ct + 1 # update counter
    message("Running: ", progress_ct, "/", nrow(param_grid))
    params <- scenario(
      scenario_num = scenario,
      sxy = sxy,
      sx = sx,
      sy = sy,
      n = n,
      signal = signal
    )
    est <- simulate_est(params, method = method, n_sim = n_sim)
    export_simulations(est, method, params, n_sims, export_est_file)
    tibble(scenario, method, n, n_sim, est)
  }
)
```

```{R, eval=FALSE, echo=FALSE}
# Simulate MSE Results
progress_ct <- 0 # global counter
sim_mse <- pmap_dfr(
  param_grid,
  function(scenario, method, sxy, sx, sy, n, signal, n_sim) {
    progress_ct <<- progress_ct + 1 # update counter
    message("Running: ", progress_ct, "/", nrow(param_grid))
    params <- scenario(
      scenario_num = scenario,
      sxy = sxy,
      sx = sx,
      sy = sy,
      n = n,
      signal = signal
    )
    mse <- simulate_mse(params, method = method, n_sim = n_sim)
    export_simulations(mse, method, params, n_sims, export_mse_file)
    tibble(scenario, method, n, n_sim, mse)
  }
)
```

## Analyze Results

```{R, message=FALSE, warning=FALSE}
# Read CSV
sim_power <- readr::read_csv(export_power_file)
sim_est <- readr::read_csv(export_est_file)
sim_mse <- readr::read_csv(export_mse_file)
```

```{R, echo=FALSE, warning=FALSE, mesage=FALSE}
# Plot Power
p1_power <- plot_power(sim_power, sample_size = 1000, scenario = 1, n_simulations = 1000)
p2_power <- plot_power(sim_power, sample_size = 1000, scenario = 2, n_simulations = 1000)
p3_power <- plot_power(sim_power, sample_size = 1000, scenario = 3, n_simulations = 1000)
p4_power <- plot_power(sim_power, sample_size = 1000, scenario = 4, n_simulations = 1000)
p5_power <- plot_power(sim_power, sample_size = 1000, scenario = 5, n_simulations = 1000)
p6_power <- plot_power(sim_power, sample_size = 1000, scenario = 6, n_simulations = 1000)
p7_power <- plot_power(sim_power, sample_size = 1000, scenario = 7, n_simulations = 1000)
p8_power <- plot_power(sim_power, sample_size = 1000, scenario = 8, n_simulations = 1000) #
p9_power <- plot_power(sim_power, sample_size = 1000, scenario = 9, n_simulations = 1000) #
p10_power <- plot_power(sim_power, sample_size = 1000, scenario = 10, n_simulations = 1000) #
p11_power <- plot_power(sim_power, sample_size = 1000, scenario = 11, n_simulations = 1000) #
```

```{R, echo=FALSE, warning=FALSE, mesage=FALSE}
```{R, echo=FALSE, warning=FALSE, message=FALSE}
# Plot Estimates
p1_est <- plot_est(sim_est, sample_size = 1000, scenario = 1, n_simulations = 1000)
p2_est <- plot_est(sim_est, sample_size = 1000, scenario = 2, n_simulations = 1000)
p3_est <- plot_est(sim_est, sample_size = 1000, scenario = 3, n_simulations = 1000)
p4_est <- plot_est(sim_est, sample_size = 1000, scenario = 4, n_simulations = 1000)
p5_est <- plot_est(sim_est, sample_size = 1000, scenario = 5, n_simulations = 1000)
p6_est <- plot_est(sim_est, sample_size = 1000, scenario = 6, n_simulations = 1000)
p7_est <- plot_est(sim_est, sample_size = 1000, scenario = 7, n_simulations = 1000)
p8_est <- plot_est(sim_est, sample_size = 1000, scenario = 8, n_simulations = 1000) #
p9_est <- plot_est(sim_est, sample_size = 1000, scenario = 9, n_simulations = 1000) #
p10_est <- plot_est(sim_est, sample_size = 1000, scenario = 10, n_simulations = 1000) #
p11_est <- plot_est(sim_est, sample_size = 1000, scenario = 11, n_simulations = 1000) #
```

```{R}
# Analyzing MSE
p1_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 1, n_simulations = 1000)
p2_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 2, n_simulations = 1000)
p3_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 3, n_simulations = 1000)
p4_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 4, n_simulations = 1000)
p5_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 5, n_simulations = 1000)
p6_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 6, n_simulations = 1000)
p7_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 7, n_simulations = 1000)
p8_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 8, n_simulations = 1000)
p9_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 9, n_simulations = 1000)
p10_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 10, n_simulations = 1000)
p11_mse <- plot_mse(sim_mse, sample_size = 1000, scenario = 11, n_simulations = 1000)
```

```{R, echo=FALSE, message=FALSE, warning=FALSE}
# Plot Instances Alongside
(p1 | p1_est) / (p1_power | p1_mse)
(p2 | p2_est) / (p2_power | p2_mse)
(p3 | p3_est) / (p3_power | p3_mse)
(p4 | p4_est) / (p4_power | p4_mse)
(p5 | p5_est) / (p5_power | p5_mse)
(p6 | p6_est) / (p6_power | p6_mse)
(p7 | p7_est) / (p7_power | p7_mse)
(p8 | p8_est) / (p8_power | p8_mse)
(p9 | p9_est) / (p9_power | p9_mse)
(p10 | p10_est) / (p10_power | p10_mse)
(p11 | p11_est) / (p11_power | p11_mse)
```

In all simulations considered, ECE is unbiased, even in misspecified cases like scenario 8 and 9, maximally frequent change point scenarios like 10, and in randomly distributed changes as in 11. While the mean-removing strategies are reasonably effective in many simple cases, they struggle once the changes become too erratic such as in scenario 8. 

```{R, echo=FALSE, message=FALSE, warning=FALSE}
plot_segment <- function(X, params) {
  # Combine
  df <- as.data.frame(X)
  colnames(df) <- c("X1", "X2")
  X_mean <- segment_mean(X)
  df$X1_mean <- X_mean[, 1]
  df$X2_mean <- X_mean[, 2]
  df$h1 <- params$h[, 1]
  df$h2 <- params$h[, 2]
  df$index <- seq_len(nrow(X))

  ggplot(df, aes(x = index)) +
    geom_line(aes(y = X1), color = "red", alpha = 0.5) +
    geom_line(aes(y = X2), color = "blue", alpha = 0.5) +
    geom_line(aes(y = h1), color = "red", linewidth = 1.5) +
    geom_line(aes(y = h2), color = "blue", linewidth = 1.5) +
    geom_line(aes(y = X1_mean), color = "red") +
    geom_line(aes(y = X2_mean), color = "blue") +
    theme_minimal()
}
```
