---
title: "Confidence Interval"
---

## Motivation

The naive confidence interval for $\rho$ takes the form

$$\hat\rho \pm z_{\alpha/2} \, SE(\hat\rho)$$

using the asymptotic normality of $\hat\rho$ directly. This works reasonably
well when $\rho$ is near $0$, but degrades near the boundary $\rho \to \pm 1$:
the sampling distribution of $\hat\rho$ becomes skewed, and the naive interval
can overshoot $[-1, 1]$ or have poor coverage.

::: {.callout-note collapse="true"}
## Remark: Classical Asymptotic Variance of $r$

For the ordinary (i.i.d., no change points) sample correlation $r$,
$$
\sqrt{n}(r-\rho) \xrightarrow{d} N(0,\, (1-\rho^2)^2).
$$
This variance already shrinks to $0$ as $\rho \to \pm1$, so the naive interval's
failure there isn't that the variance blows up — it's that $r$'s finite-sample
distribution grows markedly skewed near the boundary, and the normal
approximation breaks down well before $n$ is large enough to trust it. The
Fisher transform doesn't just shrink the variance further; it removes the
$\rho$-dependence entirely, making the normal approximation uniformly good
across $\rho$ rather than only where $|\rho|$ is already small.
:::

The classical fix is the **Fisher $z$-transform**,

$$\operatorname{arctanh}(\hat\rho) = \frac{1}{2}\ln\left(\frac{1+\hat\rho}{1-\hat\rho}\right),$$

whose sampling distribution is better-approximated by a normal
(variance-stabilizing, closer to symmetric) than that of $\hat\rho$ itself.

## Objective

Derive the asymptotic variance of $\operatorname{arctanh}(\hat\rho)$ via
the delta method, using the known (or plug-in) asymptotic variance of
$\hat\rho$ under the ECE framework. Use this to construct a confidence
interval for $\operatorname{arctanh}(\hat\rho)$, then map it back through
$\tanh$ to obtain a confidence interval for $\rho$ with better
finite-sample coverage than the naive interval.

::: callout-important
## Theorem: Asymptotic Distribution of $\operatorname{arctanh}(\hat\rho)$

Under the conditions of the [Z-Test theorem](z-test.qmd),
$$
\sqrt{n}\bigl(\operatorname{arctanh}(\hat\rho) - \operatorname{arctanh}(\rho)\bigr) \xrightarrow{d} N(0,\, \tau^2),
$$
$$
\tau^2 = \frac{1}{n}\left[\frac{\omega_x^2+\omega_y^2-2\rho\,\omega_{xy}}{1-\rho^2} + \frac{1}{(1-\rho^2)^2}\left(\frac{\rho^2}{4}(\kappa_{40}+2\kappa_{22}+\kappa_{04}) - \rho(\kappa_{31}+\kappa_{13}) + \kappa_{22}\right) + \frac{5}{2}\right].
$$

::: {.callout-note collapse="true"}
## Remark: Asymptotic Variance Under Gaussian Errors

Under the Gaussian assumption ($\kappa_{40}=\kappa_{04}=3$, $\kappa_{31}=\kappa_{13}=3\rho$,
$\kappa_{22}=1+2\rho^2$), $v^2$ reduces (per the Gaussian remark in
[z-test.qmd](z-test.qmd)) to
$$
v^2 = \frac{1}{n}(1-\rho^2)\left(\omega_x^2+\omega_y^2-2\rho\,\omega_{xy} + \frac{7}{2}(1-\rho^2)\right).
$$
Dividing by $(1-\rho^2)^2$ leaves a single power of $(1-\rho^2)$ in the denominator on the
$\omega$ terms, while $\frac{7}{2}(1-\rho^2)^2$ cancels to a constant:
$$
\tau^2_{\text{Gaussian}} = \frac{1}{n}\left[\frac{\omega_x^2+\omega_y^2-2\rho\,\omega_{xy}}{1-\rho^2} + \frac{7}{2}\right].
$$
This still depends on $\rho$ through the $\omega$ terms whenever change points are present.
:::

::: {.callout-note collapse="true"}
## Remark: Exact Variance Stabilization Requires No Jumps

Setting $\omega_x = \omega_y = \omega_{xy} = 0$ (no change points) in
$\tau^2_{\text{Gaussian}}$ above removes the only $\rho$-dependent term, leaving
$$
\tau^2 = \frac{7}{2n},
$$
constant in $\rho$ — the classical variance-stabilizing property of the Fisher transform
(up to a constant reflecting that ECE's estimator differs from the ordinary sample
correlation). With change points present, or under non-Gaussian errors, $\tau^2$ retains
genuine dependence on $\rho$.
:::
:::

**Step 1.** Recall the asymptotic distribution of $\hat\rho$ established in `z-test.qmd`:
$$
\sqrt{n}(\hat\rho - \rho) \xrightarrow{d} N(0,\, v^2).
$$

**Step 2.** Compute $\varphi'(\rho)$ for $\varphi(\rho) = \operatorname{arctanh}(\rho) = \frac{1}{2}\ln\left(\frac{1+\rho}{1-\rho}\right)$.

::: {.callout-note collapse="true"}
## Derivative: $\varphi'(\rho)$

Write
$$
\varphi(\rho) = \frac{1}{2}\ln(1+\rho) - \frac{1}{2}\ln(1-\rho).
$$

Differentiating term by term,
$$
\varphi'(\rho) = \frac{1}{2}\cdot\frac{1}{1+\rho} - \frac{1}{2}\cdot\frac{-1}{1-\rho}
= \frac{1}{2}\left(\frac{1}{1+\rho} + \frac{1}{1-\rho}\right).
$$

Combining over the common denominator $(1+\rho)(1-\rho) = 1-\rho^2$,
$$
\varphi'(\rho) = \frac{1}{2}\cdot\frac{(1-\rho) + (1+\rho)}{1-\rho^2} = \frac{1}{2}\cdot\frac{2}{1-\rho^2} = \frac{1}{1-\rho^2}.
$$
:::

**Step 3.** Apply the (univariate) delta method to obtain
$$
\tau^2 = [\varphi'(\rho)]^2\, v^2 = \frac{v^2}{(1-\rho^2)^2}.
$$

Substituting the general $v^2$ formula from `z-test.qmd` and dividing each term by
$(1-\rho^2)^2$: the $(1-\rho^2)(\omega_x^2+\omega_y^2-2\rho\omega_{xy})$ term picks up a
single power of $(1-\rho^2)$ in the denominator, the $\kappa$-cumulant terms and
$\kappa_{22}$ pick up the full $(1-\rho^2)^{-2}$, and $\frac{5}{2}(1-\rho^2)^2$ cancels
cleanly to the constant $\frac{5}{2}$, giving
$$
\tau^2 = \frac{1}{n}\left[\frac{\omega_x^2+\omega_y^2-2\rho\,\omega_{xy}}{1-\rho^2} + \frac{1}{(1-\rho^2)^2}\left(\frac{\rho^2}{4}(\kappa_{40}+2\kappa_{22}+\kappa_{04}) - \rho(\kappa_{31}+\kappa_{13}) + \kappa_{22}\right) + \frac{5}{2}\right].
$$

## Confidence Interval

**Plug-in estimate.** Following the convention used for $\hat\rho$ (`ece.cor.se(X)`
estimates $v$ directly from data), define
$$
\hat\tau = \frac{\hat v}{1-\hat\rho^2}.
$$

**CI on the $\operatorname{arctanh}$ scale.**
$$
\left[\,\operatorname{arctanh}(\hat\rho) - z_{\alpha/2}\hat\tau,\ \ \operatorname{arctanh}(\hat\rho) + z_{\alpha/2}\hat\tau\,\right].
$$

**Back-transform via $\tanh$.** Since $\tanh$ is monotonic, applying it to both
endpoints gives a valid CI for $\rho$:
$$
\left[\,\tanh\!\left(\operatorname{arctanh}(\hat\rho) - z_{\alpha/2}\hat\tau\right),\ \ \tanh\!\left(\operatorname{arctanh}(\hat\rho) + z_{\alpha/2}\hat\tau\right)\,\right].
$$

Note this interval is generally **not symmetric around $\hat\rho$** (unlike the naive
interval) — it is symmetric on the $\operatorname{arctanh}$ scale, then warped by
$\tanh$, which pulls it away from the boundary $\pm 1$ and gives it better coverage
properties there.

## Diagnostics

Variance stabilization is a claim about $\hat\tau$ (the SE on the
$\operatorname{arctanh}$ scale), not about the back-transformed $\rho$-scale
CI width — as established above, the back-transformed width shrinks like
$(1-\rho^2)$ regardless of stabilization, since that factor cancels between
$\hat\tau$ and $\tanh'$. To see stabilization directly, we compare the naive
variance $v^2$ against the transformed variance $\tau^2 = v^2/(1-\rho^2)^2$
as functions of $\rho \in [-0.9, 0.9]$, using the (deterministic) Gaussian
formula from the Remark above with $\omega_x^2, \omega_y^2, \omega_{xy}$
computed once via `ece.complexity()`. Scenario 1 (no change points) has
$\omega_x = \omega_y = \omega_{xy} = 0$, so $\tau^2$ is exactly constant in
$\rho$; Scenario 2 with $L = 4$ has change points, so $\tau^2$ retains
$\rho$-dependence but is still far flatter than the naive $v^2$.

```{r}
#| message: false
#| echo: false
devtools::load_all("..")
library(tidyverse)
```

```{r}
#| echo: false
variance_compare <- function(scenario_num, n = 200, rho_grid = seq(-0.9, 0.9, by = 0.1), ...) {
  params <- scenario(scenario_num, n = n, r = 0.5, ...)
  w   <- ece.complexity(params$h)
  wx2 <- w[1, 1]
  wy2 <- w[2, 2]
  wxy <- w[1, 2]

  tibble(rho = rho_grid) |>
    mutate(
      naive       = (1 - rho^2) * (wx2 + wy2 - 2 * rho * wxy + 7 / 2 * (1 - rho^2)),
      transformed = naive / (1 - rho^2)^2
    ) |>
    pivot_longer(c(naive, transformed), names_to = "scale", values_to = "variance")
}
```

```{r}
#| echo: false
vc <- bind_rows(
  variance_compare(1)          |> mutate(scenario = "Scenario 1"),
  variance_compare(2, L = 4)   |> mutate(scenario = "Scenario 2 (L=4)")
) |>
  mutate(scenario = factor(scenario, levels = c("Scenario 1", "Scenario 2 (L=4)")),
         scale    = factor(scale, levels = c("naive", "transformed")))
```

```{r}
#| echo: false
variance_plot <- function(vc_subset, title) {
  vc_subset |>
    ggplot(aes(x = rho, y = variance, color = scale)) +
    geom_line(linewidth = 1) +
    facet_wrap(~scenario) +
    scale_color_manual(values = c(naive = "black", transformed = "firebrick"),
                        breaks = c("naive", "transformed"),
                        labels = c(expression(paste("naive (", v^2, ")")),
                                   expression(paste("z-transformed (", tau^2, ")")))) +
    labs(x = expression(rho), y = "Asymptotic variance", color = NULL, title = title) +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

::: {.panel-tabset}
## $\rho > 0$

```{r}
#| echo: false
vc |> filter(rho >= 0) |> variance_plot("Naive vs. Fisher-transformed variance across ρ (ρ ≥ 0)")
```

## $\rho < 0$

```{r}
#| echo: false
vc |> filter(rho <= 0) |> variance_plot("Naive vs. Fisher-transformed variance across ρ (ρ ≤ 0)")
```
:::

## Coverage Simulation

The variance comparison above is a claim about the asymptotic formulas; it
says nothing about how the two intervals actually perform in finite samples.
Here we simulate empirical coverage and average width of the naive and
Fisher-transformed CIs, at nominal level $0.90$, across a grid of $\rho$ and
both scenarios, at $n = 200$ with `nsim = 2000` replicates per cell.

```{r z-test-fn}
#| echo: false
z_test <- function(X, conf.level = 0.95, kappa = "gaussian", rho = NULL, ztransform = FALSE) {
  z_crit <- qnorm((1 + conf.level) / 2)
  if (ncol(X) != 2) stop("type = 'z.test' requires a 2-column matrix")
  rxy <- ece.cor(X)[1, 2]
  se <- ece.cor.se(X[, 1], X[, 2], kappa = kappa, rho = rho)
  if (ztransform) {
    se <- se / (1 - rxy^2)
    ci <- (atanh(rxy) + c(-1, 1) * z_crit * se) |>
      tanh() # transform back into original scale
  } else {
    ci <- rxy + c(-1, 1) * z_crit * se
  }
  list(
    estimate = rxy,
    se       = se,
    conf.int = structure(ci, conf.level = conf.level),
    p.value  = 2 * pnorm(abs(rxy / se), lower.tail = FALSE)
  )
}
```

```{r coverage-grid-fn}
#| echo: false
coverage_grid <- function(rho_grid, scenario_num, L = NULL, n = 200,
                           nsim = 2000, conf.level = 0.90) {
  map_dfr(rho_grid, \(rho) {
    params <- if (is.null(L)) scenario(scenario_num, n = n, r = rho)
              else scenario(scenario_num, n = n, r = rho, L = L)
    n_warn <- 0

    sims <- map_dfr(1:nsim, ~ {
      X <- generate_data(params)
      map_dfr(c(FALSE, TRUE), \(zt) {
        fit <- withCallingHandlers(
          z_test(X, conf.level = conf.level, kappa = "gaussian", ztransform = zt),
          warning = function(w) { n_warn <<- n_warn + 1; invokeRestart("muffleWarning") }
        )
        tibble(method  = if (zt) "Fisher" else "Naive",
               covered = fit$conf.int[1] < rho & rho < fit$conf.int[2],
               width   = diff(fit$conf.int))
      })
    })

    sims |>
      summarise(coverage = mean(covered, na.rm = TRUE), width = mean(width, na.rm = TRUE), .by = method) |>
      mutate(se = sqrt(coverage * (1 - coverage) / nsim), rho = rho, n_warnings = n_warn)
  })
}
```

```{r ece-coverage-sim}
#| echo: false
#| cache: true
#| dependson: ["z-test-fn", "coverage-grid-fn"]
rho_grid <- c(-0.99, -0.95, seq(-0.9, 0.9, by = 0.1), 0.95, 0.99)
cov_s1 <- coverage_grid(rho_grid, 1, nsim = 2000) |> mutate(scenario = "Scenario 1")
cov_s2 <- coverage_grid(rho_grid, 2, L = 4, nsim = 2000) |> mutate(scenario = "Scenario 2 (L=4)")
cov <- bind_rows(cov_s1, cov_s2) |>
  pivot_longer(c(coverage, width), names_to = "metric", values_to = "value") |>
  mutate(method = factor(method, levels = c("Naive", "Fisher")))
```

```{r}
#| echo: false
coverage_plot <- function(cov_subset, title) {
  cov_subset |>
    ggplot(aes(x = rho, y = value, color = method)) +
    geom_ribbon(data = ~filter(.x, metric == "coverage"),
                aes(ymin = value - 1.96 * se, ymax = value + 1.96 * se, fill = method),
                alpha = 0.15, color = NA) +
    geom_hline(data = tibble(metric = "coverage", value = 0.90),
               aes(yintercept = value), linetype = "dashed", color = "gray50") +
    geom_line(linewidth = 1) +
    geom_point(size = 1.5) +
    facet_grid(metric ~ scenario, scales = "free_y") +
    scale_color_manual(values = c(Naive = "black", Fisher = "firebrick")) +
    scale_fill_manual(values = c(Naive = "black", Fisher = "firebrick"), guide = "none") +
    labs(x = expression(rho), y = NULL, color = NULL, title = title) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

n_warnings_in <- function(cov_long, rho_filter) {
  cov_long |>
    filter(rho_filter(rho)) |>
    distinct(scenario, rho, n_warnings) |>
    pull(n_warnings) |>
    sum()
}
```

::: {.panel-tabset}
## $\rho > 0$

```{r}
#| echo: false
coverage_plot(filter(cov, rho >= 0),
              "Coverage and width of naive vs. Fisher-transformed CIs (n = 200, nsim = 2000)")
cat(n_warnings_in(cov, \(r) r >= 0), "warnings suppressed during this simulation (rho >= 0).\n")
```

## $\rho < 0$

```{r}
#| echo: false
coverage_plot(filter(cov, rho <= 0),
              "Coverage and width of naive vs. Fisher-transformed CIs (n = 200, nsim = 2000)")
cat(n_warnings_in(cov, \(r) r <= 0), "warnings suppressed during this simulation (rho <= 0).\n")
```
:::

### Classical Pearson Correlation, for Comparison

Is the modest ECE gap above unusual, or is this just what a Fisher transform
buys you at $n = 200$ in general? Same $n$, same `rho_grid`, same `nsim`, same
nominal level — but now for ordinary i.i.d. bivariate normal data and the
textbook Pearson correlation, no ECE estimator involved: naive uses
$SE(r) = (1-r^2)/\sqrt{n}$; Fisher uses $SE(\operatorname{arctanh}(r)) = 1/\sqrt{n-3}$.

```{r pearson-coverage-grid-fn}
#| echo: false
pearson_coverage_grid <- function(rho_grid, n = 200, nsim = 2000, conf.level = 0.90) {
  z_crit <- qnorm((1 + conf.level) / 2)
  map_dfr(rho_grid, \(rho) {
    n_warn <- 0
    sims <- map_dfr(1:nsim, ~ {
      x <- rnorm(n)
      y <- rho * x + sqrt(1 - rho^2) * rnorm(n)
      withCallingHandlers({
        r <- cor(x, y)
        ci_naive  <- r + c(-1, 1) * z_crit * (1 - r^2) / sqrt(n)
        ci_fisher <- tanh(atanh(r) + c(-1, 1) * z_crit / sqrt(n - 3))
        bind_rows(
          tibble(method = "Naive",  covered = ci_naive[1] < rho & rho < ci_naive[2],  width = diff(ci_naive)),
          tibble(method = "Fisher", covered = ci_fisher[1] < rho & rho < ci_fisher[2], width = diff(ci_fisher))
        )
      }, warning = function(w) { n_warn <<- n_warn + 1; invokeRestart("muffleWarning") })
    })
    sims |>
      summarise(coverage = mean(covered, na.rm = TRUE), width = mean(width, na.rm = TRUE), .by = method) |>
      mutate(se = sqrt(coverage * (1 - coverage) / nsim), rho = rho, n_warnings = n_warn)
  })
}
```

```{r pearson-coverage-sim}
#| echo: false
#| cache: true
#| dependson: ["ece-coverage-sim", "pearson-coverage-grid-fn"]
set.seed(123)
cov_pearson <- pearson_coverage_grid(rho_grid, n = 200, nsim = 2000) |>
  pivot_longer(c(coverage, width), names_to = "metric", values_to = "value") |>
  mutate(method = factor(method, levels = c("Naive", "Fisher")))
```

```{r}
#| echo: false
pearson_coverage_plot <- function(cov_subset, title) {
  cov_subset |>
    ggplot(aes(x = rho, y = value, color = method)) +
    geom_ribbon(data = ~filter(.x, metric == "coverage"),
                aes(ymin = value - 1.96 * se, ymax = value + 1.96 * se, fill = method),
                alpha = 0.15, color = NA) +
    geom_hline(data = tibble(metric = "coverage", value = 0.90),
               aes(yintercept = value), linetype = "dashed", color = "gray50") +
    geom_line(linewidth = 1) +
    geom_point(size = 1.5) +
    facet_grid(metric ~ ., scales = "free_y") +
    scale_color_manual(values = c(Naive = "black", Fisher = "firebrick")) +
    scale_fill_manual(values = c(Naive = "black", Fisher = "firebrick"), guide = "none") +
    labs(x = expression(rho), y = NULL, color = NULL, title = title) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

n_warnings_in_pearson <- function(cov_long, rho_filter) {
  cov_long |>
    filter(rho_filter(rho)) |>
    distinct(rho, n_warnings) |>
    pull(n_warnings) |>
    sum()
}
```

::: {.panel-tabset}
## $\rho > 0$

```{r}
#| echo: false
pearson_coverage_plot(filter(cov_pearson, rho >= 0),
                       "Classical Pearson correlation: naive vs. Fisher-transformed CIs (n = 200, nsim = 2000)")
cat(n_warnings_in_pearson(cov_pearson, \(r) r >= 0), "warnings suppressed during this simulation (rho >= 0).\n")
```

## $\rho < 0$

```{r}
#| echo: false
pearson_coverage_plot(filter(cov_pearson, rho <= 0),
                       "Classical Pearson correlation: naive vs. Fisher-transformed CIs (n = 200, nsim = 2000)")
cat(n_warnings_in_pearson(cov_pearson, \(r) r <= 0), "warnings suppressed during this simulation (rho <= 0).\n")
```
:::

::: {.callout-note collapse="true"}
## Remark: Why Scenario 2 Doesn't Show a Larger Naive/Fisher Gap

One might expect Scenario 2 (change points present) to show *more* naive
undercoverage near the boundary than Scenario 1, since $\omega_x, \omega_y,
\omega_{xy} \neq 0$ there. The plot above shows the opposite: both scenarios
look qualitatively similar for the naive method.

The reason is specific to how `mean_square_wave()` builds the mean shift
(`R/sim-scenarios.R`): it applies the **same** square wave to both columns of
`h`, so $X$ and $Y$ jump at identical times with identical signs. This forces
$\omega_x = \omega_y = \omega_{xy}$ exactly, and the naive-variance term
reduces to
$$
\omega_x^2+\omega_y^2-2\rho\,\omega_{xy} = 2\omega_x^2(1-\rho),
$$
which itself vanishes as $\rho \to 1$ at the same order as the $(1-\rho^2)$
prefactor already multiplying it. So the "extra" variance from change points
does not act as a floor near the boundary — it decays just as fast. Numerically,
at $n=200$, $L=4$, the naive $v^2$ in Scenario 2 is only about $1.3\times$
Scenario 1's near $\rho = 0.9$–$0.99$, not dramatically larger.

This means Scenario 2 as currently defined is a weak test of the "change
points make the naive interval worse" hypothesis, since $X$ and $Y$ move in
lockstep. A sharper test would need independent jump patterns for $X$ and $Y$
(different timing, phase, or sign) so that $\omega_x^2+\omega_y^2-2\rho\,\omega_{xy}$
does not cancel near the boundary — left for future work.
:::

## Complexity Parameter Estimate

**How the estimate is produced.** `ece.complexity(x, y, L)` (via the internal
`ece_pair()`) estimates $\omega_{xy}$ by regression on lagged cross-products.
For lags $k = 1, \dots, L$, define
$$
R_k = \frac{T_k(X,Y)}{2n}, \qquad T_k(X,Y) = \sum_{i=1}^n (X_i - X_{i+k})(Y_i - Y_{i+k}),
$$
(indices taken circularly). Since $\mathbb{E}[T_k(X,Y)] = k\,W_{xy} + 2n\sigma_{xy}$
(established in `z-test.qmd`, with $W_{xy} = T_1(\theta_x,\theta_y)$ and
$\omega_{xy} = W_{xy}/n$), $R_k$ is linear in $k$ with intercept $\sigma_{xy}$
and slope $\omega_{xy}/2$:
$$
\mathbb{E}[R_k] = \sigma_{xy} + \frac{k}{2}\omega_{xy}.
$$
`ece_pair` regresses $R_1, \dots, R_L$ on $(1, k)$ by OLS and reports the
intercept as `cov` and twice the slope as `complexity`. The package default
(and the only case used elsewhere in these notes) is $L = 2$, for which the
regression uses exactly two points and reduces to a closed form: the slope of
a line through $(1, R_1)$ and $(2, R_2)$ is $R_2 - R_1$, so
$$
\hat\omega_{xy} = 2(R_2 - R_1) = \frac{T_2(X,Y) - T_1(X,Y)}{n}.
$$

::: callout-important
## Theorem: Mean and Variance of $\hat\omega_{xy}$ ($L = 2$)

$$
\mathbb{E}[\hat\omega_{xy}] = \omega_{xy},
$$
$$
\operatorname{Var}(\hat\omega_{xy}) = \frac{1}{n}\Big[4(\sigma_x^2\sigma_y^2 + \sigma_{xy}^2) + 2(\sigma_y^2\omega_x^2 + \sigma_x^2\omega_y^2) + 4\sigma_{xy}\,\omega_{xy}\Big].
$$

::: {.callout-note collapse="true"}
## Remark: Scale-Normalized Form

Setting $\sigma_x = \sigma_y = 1$, $\sigma_{xy} = \rho$,
$$
\operatorname{Var}(\hat\omega_{xy}) = \frac{1}{n}\Big[4(1+\rho^2) + 2(\omega_x^2+\omega_y^2) + 4\rho\,\omega_{xy}\Big].
$$
:::

::: {.callout-note collapse="true"}
## Remark: No Dependence on Fourth Cumulants

$\operatorname{Var}(T_k(X,Y))$ depends on $\kappa_{22}$ (see `z-test.qmd`), but
the $\kappa_{22}\sigma_x^2\sigma_y^2$ contribution is identical across
$\operatorname{Var}(T_1)$, $\operatorname{Var}(T_2)$, and
$\operatorname{Cov}(T_1,T_2)$ up to a fixed multiple of $n$, and cancels
exactly in $\operatorname{Var}(T_2 - T_1)$. So $\operatorname{Var}(\hat\omega_{xy})$
has no dependence on the error distribution's kurtosis — it only needs finite
second moments to be well-defined, unlike $v^2$ and $\tau^2$ above which both
involve $\kappa_{40}, \kappa_{22}, \kappa_{04}$, etc.
:::
:::

**Step 1.** By linearity of $\mathbb{E}[T_k(X,Y)] = k\,W_{xy} + 2n\sigma_{xy}$,
$$
\mathbb{E}[\hat\omega_{xy}] = \frac{\mathbb{E}[T_2(X,Y)] - \mathbb{E}[T_1(X,Y)]}{n} = \frac{(2W_{xy}+2n\sigma_{xy}) - (W_{xy}+2n\sigma_{xy})}{n} = \frac{W_{xy}}{n} = \omega_{xy}.
$$
The estimator is exactly unbiased for any $n$ — no asymptotics required.

**Step 2.** Compute $\operatorname{Var}(T_2(X,Y) - T_1(X,Y))$ using the
$\Sigma_Z$ building blocks from `z-test.qmd` (`var_Qh_est`, `cov_Qh_Qk_est`
in `R/theory-estimators.R`):
$$
\operatorname{Var}(T_2 - T_1) = \operatorname{Var}(T_2) + \operatorname{Var}(T_1) - 2\operatorname{Cov}(T_1, T_2).
$$
Substituting the formulas for $\operatorname{Var}(T_k(X,Y))$ and
$\operatorname{Cov}(T_1(X,Y), T_2(X,Y))$ and collecting terms, every
$\kappa_{22}\sigma_x^2\sigma_y^2$ contribution cancels, leaving
$$
\operatorname{Var}(T_2 - T_1) = 4n(\sigma_x^2\sigma_y^2+\sigma_{xy}^2) + 2n(\sigma_y^2\omega_x^2+\sigma_x^2\omega_y^2) + 4n\sigma_{xy}\,\omega_{xy}.
$$
Dividing by $n^2$ gives $\operatorname{Var}(\hat\omega_{xy})$ as stated in the Theorem.

::: {.callout-note collapse="true"}
## Consistency Check: Diagonal Case $\hat\omega_x^2$

Setting $y = x$ (so $\sigma_y = \sigma_x$, $\sigma_{xy} = \sigma_x^2$,
$\omega_y = \omega_x$, $\omega_{xy} = \omega_x^2$) in the Theorem's variance
formula gives
$$
\operatorname{Var}(\hat\omega_x^2) = \frac{1}{n}\Big[4(\sigma_x^4+\sigma_x^4) + 2(\sigma_x^2\omega_x^2+\sigma_x^2\omega_x^2) + 4\sigma_x^2\omega_x^2\Big] = \frac{8\sigma_x^2(\sigma_x^2+\omega_x^2)}{n},
$$
which matches deriving $\operatorname{Var}(T_2(X,X)-T_1(X,X))$ directly from
`var_Tk_est`/`cov_Th_Tk_est` (the auto-complexity building blocks), a
useful check that the general cross-term derivation specializes correctly.
:::

### Diagnostics

```{r}
#| message: false
#| echo: false
devtools::load_all("..")
library(tidyverse)
```

We first verify the closed-form variance algebraically against the sum of
building blocks, for arbitrary (not necessarily scenario-derived) inputs, and
confirm it does not depend on the value of $\kappa_{22}$ passed in:

```{r}
#| message: false
library(testthat)
n <- 200; sx <- 1.3; sy <- 0.8; sxy <- 0.4; wx <- 2.1; wy <- 1.4; wxy <- 0.9

var_omega_xy <- function(n, sx, sy, sxy, wx, wy, wxy) {
  (4 * (sx^2 * sy^2 + sxy^2) + 2 * (sy^2 * wx + sx^2 * wy) + 4 * sxy * wxy) / n
}

test_that("Var(omega_hat_xy) matches sum of building blocks, any k22", {
  for (k22 in c(-3, 0, 1, 5)) {
    V1  <- var_Qh_est(n, sx, sy, sxy, wx, wy, wxy, k22, k = 1)
    V2  <- var_Qh_est(n, sx, sy, sxy, wx, wy, wxy, k22, k = 2)
    C12 <- cov_Qh_Qk_est(n, sx, sy, sxy, wx, wy, wxy, k22)
    expect_equal((V1 + V2 - 2 * C12) / n^2, var_omega_xy(n, sx, sy, sxy, wx, wy, wxy))
  }
})
```

Next, a Monte Carlo check: simulate Scenario 2 ($n = 200$, $L = 4$, $\rho =
0.5$), repeatedly estimate $\hat\omega_{xy}$ via `ece.complexity(X, L = 2)`,
and compare the empirical mean and variance of the residual
$\hat\omega_{xy} - \omega_{xy}$ against the theoretical values above (the
residual should center at zero with the theoretical variance).

```{r complexity-bias-sim}
#| echo: false
#| cache: true
params <- scenario(2, n = 200, r = 0.5, L = 4)
n   <- params$n
sx  <- sqrt(params$S[1, 1]); sy <- sqrt(params$S[2, 2]); sxy <- params$S[1, 2]
wx  <- lag_diff(params$h[, 1]) / n
wy  <- lag_diff(params$h[, 2]) / n
wxy <- lag_diff(params$h[, 1], params$h[, 2]) / n

omega_hat <- map_dbl(1:2000, ~ ece.complexity(generate_data(params), L = 2)[1, 2],
                      .progress = TRUE)
theoretical_var <- var_omega_xy(n, sx, sy, sxy, wx, wy, wxy)
```

```{r}
#| echo: false
tibble(residual = omega_hat - wxy) |>
  ggplot(aes(x = residual)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "firebrick", linewidth = 0.8) +
  geom_vline(xintercept = mean(omega_hat) - wxy, color = "orange", linewidth = 0.8, linetype = "dashed") +
  labs(x = expression(hat(omega)[xy] - omega[xy]), y = NULL,
       title = "Bias of the complexity estimator",
       subtitle = paste0("Red = 0 (theoretical bias); orange = sample mean; t-test p = ",
                          signif(t.test(omega_hat - wxy)$p.value, 3))) +
  theme_minimal()
```

```{r}
#| echo: false
tibble(
  quantity  = c("mean", "variance"),
  empirical = c(mean(omega_hat), var(omega_hat)),
  theoretical = c(wxy, theoretical_var)
)
```

