---
title: "Confidence Interval"
---

## Motivation

The naive confidence interval for $\rho$ takes the form

$$\hat\rho \pm z_{\alpha/2} \, SE(\hat\rho)$$

using the asymptotic normality of $\hat\rho$ directly. This works reasonably
well when $\rho$ is near $0$, but degrades near the boundary $\rho \to \pm 1$:
the sampling distribution of $\hat\rho$ becomes skewed, and the naive interval
can overshoot $[-1, 1]$ or have poor coverage.

The classical fix is the **Fisher $z$-transform**,

$$\operatorname{arctanh}(\hat\rho) = \frac{1}{2}\ln\left(\frac{1+\hat\rho}{1-\hat\rho}\right),$$

whose sampling distribution is better-approximated by a normal
(variance-stabilizing, closer to symmetric) than that of $\hat\rho$ itself.

## Objective

Derive the asymptotic variance of $\operatorname{arctanh}(\hat\rho)$ via
the delta method, using the known (or plug-in) asymptotic variance of
$\hat\rho$ under the ECE framework. Use this to construct a confidence
interval for $\operatorname{arctanh}(\hat\rho)$, then map it back through
$\tanh$ to obtain a confidence interval for $\rho$ with better
finite-sample coverage than the naive interval.

::: callout-important
## Theorem: Asymptotic Distribution of $\operatorname{arctanh}(\hat\rho)$

Under the conditions of the [Z-Test theorem](z-test.qmd),
$$
\sqrt{n}\bigl(\operatorname{arctanh}(\hat\rho) - \operatorname{arctanh}(\rho)\bigr) \xrightarrow{d} N(0,\, \tau^2),
$$
$$
\tau^2 = \frac{1}{n}\left[\frac{\omega_x^2+\omega_y^2-2\rho\,\omega_{xy}}{1-\rho^2} + \frac{1}{(1-\rho^2)^2}\left(\frac{\rho^2}{4}(\kappa_{40}+2\kappa_{22}+\kappa_{04}) - \rho(\kappa_{31}+\kappa_{13}) + \kappa_{22}\right) + \frac{5}{2}\right].
$$

::: {.callout-note collapse="true"}
## Remark: Asymptotic Variance Under Gaussian Errors

Under the Gaussian assumption ($\kappa_{40}=\kappa_{04}=3$, $\kappa_{31}=\kappa_{13}=3\rho$,
$\kappa_{22}=1+2\rho^2$), $v^2$ reduces (per the Gaussian remark in
[z-test.qmd](z-test.qmd)) to
$$
v^2 = \frac{1}{n}(1-\rho^2)\left(\omega_x^2+\omega_y^2-2\rho\,\omega_{xy} + \frac{7}{2}(1-\rho^2)\right).
$$
Dividing by $(1-\rho^2)^2$ leaves a single power of $(1-\rho^2)$ in the denominator on the
$\omega$ terms, while $\frac{7}{2}(1-\rho^2)^2$ cancels to a constant:
$$
\tau^2_{\text{Gaussian}} = \frac{1}{n}\left[\frac{\omega_x^2+\omega_y^2-2\rho\,\omega_{xy}}{1-\rho^2} + \frac{7}{2}\right].
$$
This still depends on $\rho$ through the $\omega$ terms whenever change points are present.
:::

::: {.callout-note collapse="true"}
## Remark: Exact Variance Stabilization Requires No Jumps

Setting $\omega_x = \omega_y = \omega_{xy} = 0$ (no change points) in
$\tau^2_{\text{Gaussian}}$ above removes the only $\rho$-dependent term, leaving
$$
\tau^2 = \frac{7}{2n},
$$
constant in $\rho$ — the classical variance-stabilizing property of the Fisher transform
(up to a constant reflecting that ECE's estimator differs from the ordinary sample
correlation). With change points present, or under non-Gaussian errors, $\tau^2$ retains
genuine dependence on $\rho$.
:::
:::

**Step 1.** Recall the asymptotic distribution of $\hat\rho$ established in `z-test.qmd`:
$$
\sqrt{n}(\hat\rho - \rho) \xrightarrow{d} N(0,\, v^2).
$$

**Step 2.** Compute $\varphi'(\rho)$ for $\varphi(\rho) = \operatorname{arctanh}(\rho) = \frac{1}{2}\ln\left(\frac{1+\rho}{1-\rho}\right)$.

::: {.callout-note collapse="true"}
## Derivative: $\varphi'(\rho)$

Write
$$
\varphi(\rho) = \frac{1}{2}\ln(1+\rho) - \frac{1}{2}\ln(1-\rho).
$$

Differentiating term by term,
$$
\varphi'(\rho) = \frac{1}{2}\cdot\frac{1}{1+\rho} - \frac{1}{2}\cdot\frac{-1}{1-\rho}
= \frac{1}{2}\left(\frac{1}{1+\rho} + \frac{1}{1-\rho}\right).
$$

Combining over the common denominator $(1+\rho)(1-\rho) = 1-\rho^2$,
$$
\varphi'(\rho) = \frac{1}{2}\cdot\frac{(1-\rho) + (1+\rho)}{1-\rho^2} = \frac{1}{2}\cdot\frac{2}{1-\rho^2} = \frac{1}{1-\rho^2}.
$$
:::

**Step 3.** Apply the (univariate) delta method to obtain
$$
\tau^2 = [\varphi'(\rho)]^2\, v^2 = \frac{v^2}{(1-\rho^2)^2}.
$$

Substituting the general $v^2$ formula from `z-test.qmd` and dividing each term by
$(1-\rho^2)^2$: the $(1-\rho^2)(\omega_x^2+\omega_y^2-2\rho\omega_{xy})$ term picks up a
single power of $(1-\rho^2)$ in the denominator, the $\kappa$-cumulant terms and
$\kappa_{22}$ pick up the full $(1-\rho^2)^{-2}$, and $\frac{5}{2}(1-\rho^2)^2$ cancels
cleanly to the constant $\frac{5}{2}$, giving
$$
\tau^2 = \frac{1}{n}\left[\frac{\omega_x^2+\omega_y^2-2\rho\,\omega_{xy}}{1-\rho^2} + \frac{1}{(1-\rho^2)^2}\left(\frac{\rho^2}{4}(\kappa_{40}+2\kappa_{22}+\kappa_{04}) - \rho(\kappa_{31}+\kappa_{13}) + \kappa_{22}\right) + \frac{5}{2}\right].
$$

## Confidence Interval

**Plug-in estimate.** Following the convention used for $\hat\rho$ (`ece.cor.se(X)`
estimates $v$ directly from data), define
$$
\hat\tau = \frac{\hat v}{1-\hat\rho^2}.
$$

**CI on the $\operatorname{arctanh}$ scale.**
$$
\left[\,\operatorname{arctanh}(\hat\rho) - z_{\alpha/2}\hat\tau,\ \ \operatorname{arctanh}(\hat\rho) + z_{\alpha/2}\hat\tau\,\right].
$$

**Back-transform via $\tanh$.** Since $\tanh$ is monotonic, applying it to both
endpoints gives a valid CI for $\rho$:
$$
\left[\,\tanh\!\left(\operatorname{arctanh}(\hat\rho) - z_{\alpha/2}\hat\tau\right),\ \ \tanh\!\left(\operatorname{arctanh}(\hat\rho) + z_{\alpha/2}\hat\tau\right)\,\right].
$$

Note this interval is generally **not symmetric around $\hat\rho$** (unlike the naive
interval) — it is symmetric on the $\operatorname{arctanh}$ scale, then warped by
$\tanh$, which pulls it away from the boundary $\pm 1$ and gives it better coverage
properties there.

## Diagnostics

Variance stabilization is a claim about $\hat\tau$ (the SE on the
$\operatorname{arctanh}$ scale), not about the back-transformed $\rho$-scale
CI width — as established above, the back-transformed width shrinks like
$(1-\rho^2)$ regardless of stabilization, since that factor cancels between
$\hat\tau$ and $\tanh'$. To see stabilization directly, we plot $\hat\tau$
itself against $\rho$ for each $\rho \in \{0, 0.2, 0.4, 0.6, 0.8, 0.9, 0.95\}$
($n = 1000$), comparing Scenario 1 (no change points, where the Gaussian
remark predicts $\tau^2 = 7/(2n)$, flat in $\rho$) against Scenario 2 with
$L = 4$ (change points present, where the $\omega$ terms should make
$\hat\tau$ grow as $\rho \to \pm 1$). The theoretical curve
$\tau = \texttt{ece.cor.se.formula(params)} / (1-\rho^2)$ is overlaid in red.
The number of replicates is kept small for now during development; this will
be increased once the pipeline is verified.

```{r}
#| message: false
devtools::load_all("..")
library(tidyverse)
```

```{r}
#| echo: false
tau_hat_once <- function(params) {
  X   <- generate_data(params)
  rxy <- ece.cor(X)[1, 2]
  se  <- ece.cor.se(X, kappa = "gaussian")
  se / (1 - rxy^2)
}

sim_tau <- function(scenario_num, L, rho_grid, n = 1000, nsim = 20) {
  map_dfr(rho_grid, \(rho) {
    params <- scenario(scenario_num, n = n, r = rho, L = L)
    tibble(
      rho           = rho,
      tau_hat       = map_dbl(1:nsim, ~ tau_hat_once(params)),
      tau_theoretical = ece.cor.se.formula(params) / (1 - rho^2)
    )
  })
}
```

```{r}
#| echo: false
#| cache: true
rho_grid <- c(0, 0.2, 0.4, 0.6, 0.8, 0.9, 0.95)
tau_s1 <- sim_tau(1, L = 2, rho_grid = rho_grid)
tau_s2 <- sim_tau(2, L = 4, rho_grid = rho_grid)
```

```{r}
#| echo: false
tau_plot <- function(tau_sim) {
  theoretical <- tau_sim |> distinct(rho, tau_theoretical)
  tau_sim |>
    ggplot(aes(x = factor(rho), y = tau_hat)) +
    geom_boxplot(fill = "steelblue", alpha = 0.6) +
    geom_point(data = theoretical, aes(x = factor(rho), y = tau_theoretical),
               color = "firebrick", size = 2) +
    labs(x = expression(rho), y = expression(hat(tau)),
         title = "Estimated tau (arctanh-scale SE) vs. rho",
         subtitle = "Red = theoretical tau; flat = stabilized, rising = not stabilized") +
    theme_minimal()
}
```

::: {.panel-tabset}
## Scenario 1

```{r}
#| echo: false
tau_plot(tau_s1)
```

## Scenario 2 (L=4)

```{r}
#| echo: false
tau_plot(tau_s2)
```
:::
