---
title: "Covariance of $Z$"
---

## Setup

Define the vector
$$
Z = \bigl(T_1(X,X),\; T_2(X,X),\; T_1(Y,Y),\; T_2(Y,Y),\; T_1(X,Y),\; T_2(X,Y)\bigr)^T.
$$
Our goal is to compute the $6\times 6$ covariance matrix $\Sigma_Z = \operatorname{Cov}(Z)$ exactly.

**Moment notation.** Let $(\varepsilon_{x,i}, \varepsilon_{y,i})$ denote a generic error pair. We define standardized joint moments
$$
\kappa_{ab} = \frac{E[\varepsilon_{x,i}^a \varepsilon_{y,i}^b]}{\sigma_x^a \sigma_y^b}, \qquad a + b = 4,
$$
and $\rho = \sigma_{xy}/(\sigma_x\sigma_y)$. Under bivariate normality, $\kappa_{40} = \kappa_{04} = 3$, $\kappa_{22} = 1+2\rho^2$, $\kappa_{31} = \kappa_{13} = 3\rho$.

**Signal summary statistics.** For $k \leq L/2$, define
$$
W_x = T_1(\theta_x, \theta_x), \quad W_y = T_1(\theta_y, \theta_y), \quad W_{xy} = T_1(\theta_x, \theta_y).
$$

## Variance of $T_k(X, Y)$

We decompose the variance into diagonal and off-diagonal terms:
\begin{align}
\operatorname{Var}(T_k(X,Y))
&= \operatorname{Var}\!\left(\sum_{i=1}^n (X_i - X_{i+k})(Y_i - Y_{i+k})\right) \\
&= \sum_{i=1}^n \operatorname{Var}\!\bigl((X_i-X_{i+k})(Y_i-Y_{i+k})\bigr) \\
&\quad + \sum_{i \neq i'} \operatorname{Cov}\!\bigl((X_i-X_{i+k})(Y_i-Y_{i+k}),\,(X_{i'}-X_{i'+k})(Y_{i'}-Y_{i'+k})\bigr).
\end{align}

We evaluate these two types of terms in Lemmas 1 and 2 below.

**Lemma 1** (Single-index variance).
$$
\operatorname{Var}\bigl[(X_i - X_{i+k})(Y_i - Y_{i+k})\bigr]
= 2\sigma_y^2(\theta_{x,i}-\theta_{x,i+k})^2 + 2\sigma_x^2(\theta_{y,i}-\theta_{y,i+k})^2
+ 4\sigma_{xy}(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k})
+ 2(\kappa_{22}+1)\sigma_x^2\sigma_y^2.
$$

*Proof.* Substituting $X_i = \theta_{x,i} + \varepsilon_{x,i}$ and centering,
\begin{align}
(X_i-X_{i+k})(Y_i-Y_{i+k}) - E[(X_i-X_{i+k})(Y_i-Y_{i+k})]
&= (\theta_{x,i}-\theta_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k}) \\
&\quad + (\theta_{y,i}-\theta_{y,i+k})(\varepsilon_{x,i}-\varepsilon_{x,i+k}) \\
&\quad + (\varepsilon_{x,i}-\varepsilon_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k}) - 2\sigma_{xy}.
\end{align}
Taking the variance of this centered expression and expanding:
\begin{align}
&\operatorname{Var}\bigl[(X_i-X_{i+k})(Y_i-Y_{i+k})\bigr] \\
&= (\theta_{x,i}-\theta_{x,i+k})^2 \operatorname{Var}(\varepsilon_{y,i}-\varepsilon_{y,i+k})
+ (\theta_{y,i}-\theta_{y,i+k})^2 \operatorname{Var}(\varepsilon_{x,i}-\varepsilon_{x,i+k}) \\
&\quad + \operatorname{Var}\bigl[(\varepsilon_{x,i}-\varepsilon_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k})\bigr] \\
&\quad + 2(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k})\operatorname{Cov}(\varepsilon_{y,i}-\varepsilon_{y,i+k},\,\varepsilon_{x,i}-\varepsilon_{x,i+k}) \\
&\quad + 2(\theta_{x,i}-\theta_{x,i+k})\operatorname{Cov}\bigl(\varepsilon_{y,i}-\varepsilon_{y,i+k},\,(\varepsilon_{x,i}-\varepsilon_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k})\bigr) \\
&\quad + 2(\theta_{y,i}-\theta_{y,i+k})\operatorname{Cov}\bigl(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,(\varepsilon_{x,i}-\varepsilon_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k})\bigr).
\end{align}
We evaluate each component. Since errors are i.i.d. with zero mean:
$$
\operatorname{Var}(\varepsilon_{y,i} - \varepsilon_{y,i+k}) = 2\sigma_y^2, \quad
\operatorname{Var}(\varepsilon_{x,i} - \varepsilon_{x,i+k}) = 2\sigma_x^2, \quad
\operatorname{Cov}(\varepsilon_{y,i}-\varepsilon_{y,i+k},\, \varepsilon_{x,i}-\varepsilon_{x,i+k}) = 2\sigma_{xy}.
$$
For the pure noise variance, expanding and using independence of times $i$ and $i+k$:
$$
E\bigl[(\varepsilon_{x,i}-\varepsilon_{x,i+k})^2(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2\bigr]
= 2\kappa_{22}\sigma_x^2\sigma_y^2 + 2\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2,
$$
so $\operatorname{Var}[(\varepsilon_{x,i}-\varepsilon_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k})] = 2(\kappa_{22}+1)\sigma_x^2\sigma_y^2$.

For the last two covariances, consider the first:
$$
\operatorname{Cov}\bigl(\varepsilon_{y,i}-\varepsilon_{y,i+k},\,(\varepsilon_{x,i}-\varepsilon_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k})\bigr)
= E\bigl[(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2(\varepsilon_{x,i}-\varepsilon_{x,i+k})\bigr].
$$
Expanding and using independence of times $i$ and $i+k$, this equals $E[\varepsilon_{x,i}\varepsilon_{y,i}^2] - E[\varepsilon_{x,i+k}\varepsilon_{y,i+k}^2] = \kappa_{12}\sigma_x\sigma_y^2 - \kappa_{12}\sigma_x\sigma_y^2 = 0$. The last covariance vanishes by the same argument. Substituting:
$$
\operatorname{Var}\bigl[(X_i-X_{i+k})(Y_i-Y_{i+k})\bigr]
= 2\sigma_y^2(\theta_{x,i}-\theta_{x,i+k})^2 + 2\sigma_x^2(\theta_{y,i}-\theta_{y,i+k})^2
+ 4\sigma_{xy}(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k})
+ 2(\kappa_{22}+1)\sigma_x^2\sigma_y^2. \quad\square
$$

**Lemma 2** (Adjacent-index covariance). For three different indices $i$, $i'$, $i''$,
\begin{align}
&\operatorname{Cov}\bigl((X_i - X_{i'})(Y_i - Y_{i'}),\,(X_{i'} - X_{i''})(Y_{i'} - Y_{i''})\bigr) \\
&= (\kappa_{22} - \rho^2)\sigma_x^2\sigma_y^2
- (\theta_{x,i}-\theta_{x,i'})(\theta_{x,i'}-\theta_{x,i''})\sigma_y^2
- (\theta_{y,i}-\theta_{y,i'})(\theta_{y,i'}-\theta_{y,i''})\sigma_x^2 \\
&\quad - \bigl[(\theta_{x,i}-\theta_{x,i'})(\theta_{y,i'}-\theta_{y,i''}) + (\theta_{y,i}-\theta_{y,i'})(\theta_{x,i'}-\theta_{x,i''})\bigr]\sigma_{xy} \\
&\quad - (\theta_{x,i} - 2\theta_{x,i'} + \theta_{x,i''})\kappa_{12}\sigma_x\sigma_y^2
- (\theta_{y,i} - 2\theta_{y,i'} + \theta_{y,i''})\kappa_{21}\sigma_x^2\sigma_y.
\end{align}

*Proof.* Substituting $X_j = \theta_{x,j} + \varepsilon_{x,j}$ and centering, each factor decomposes as
$$
(X_j - X_{j'})(Y_j - Y_{j'}) - E[\cdots]
= (\theta_{x,j}-\theta_{x,j'})(\varepsilon_{y,j}-\varepsilon_{y,j'})
+ (\theta_{y,j}-\theta_{y,j'})(\varepsilon_{x,j}-\varepsilon_{x,j'})
+ (\varepsilon_{x,j}-\varepsilon_{x,j'})(\varepsilon_{y,j}-\varepsilon_{y,j'}) - 2\sigma_{xy}.
$$
Expanding the covariance as a sum of nine terms, those involving only variables at times $i$ and $i''$ are zero by independence from $i'$. The surviving terms, evaluated using the shared index $i'$, give:

- $(\theta_{x,i}-\theta_{x,i'})\cdot(\theta_{x,i'}-\theta_{x,i''})\operatorname{Cov}(\varepsilon_{y,i}-\varepsilon_{y,i'},\,\varepsilon_{y,i'}-\varepsilon_{y,i''}) = -(\theta_{x,i}-\theta_{x,i'})(\theta_{x,i'}-\theta_{x,i''})\sigma_y^2$
- $(\theta_{y,i}-\theta_{y,i'})\cdot(\theta_{y,i'}-\theta_{y,i''})\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i'},\,\varepsilon_{x,i'}-\varepsilon_{x,i''}) = -(\theta_{y,i}-\theta_{y,i'})(\theta_{y,i'}-\theta_{y,i''})\sigma_x^2$
- $(\theta_{x,i}-\theta_{x,i'})\cdot(\theta_{y,i'}-\theta_{y,i''})\operatorname{Cov}(\varepsilon_{y,i}-\varepsilon_{y,i'},\,\varepsilon_{x,i'}-\varepsilon_{x,i''}) = -(\theta_{x,i}-\theta_{x,i'})(\theta_{y,i'}-\theta_{y,i''})\sigma_{xy}$
- $(\theta_{y,i}-\theta_{y,i'})\cdot(\theta_{x,i'}-\theta_{x,i''})\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i'},\,\varepsilon_{y,i'}-\varepsilon_{y,i''}) = -(\theta_{y,i}-\theta_{y,i'})(\theta_{x,i'}-\theta_{x,i''})\sigma_{xy}$
- $(\theta_{x,i}-\theta_{x,i'})\operatorname{Cov}\bigl(\varepsilon_{y,i}-\varepsilon_{y,i'},\,(\varepsilon_{x,i'}-\varepsilon_{x,i''})(\varepsilon_{y,i'}-\varepsilon_{y,i''})\bigr) = -(\theta_{x,i}-\theta_{x,i'})\kappa_{12}\sigma_x\sigma_y^2$
- $(\theta_{y,i}-\theta_{y,i'})\operatorname{Cov}\bigl(\varepsilon_{x,i}-\varepsilon_{x,i'},\,(\varepsilon_{x,i'}-\varepsilon_{x,i''})(\varepsilon_{y,i'}-\varepsilon_{y,i''})\bigr) = -(\theta_{y,i}-\theta_{y,i'})\kappa_{21}\sigma_x^2\sigma_y$
- $(\theta_{x,i'}-\theta_{x,i''})\operatorname{Cov}\bigl((\varepsilon_{x,i}-\varepsilon_{x,i'})(\varepsilon_{y,i}-\varepsilon_{y,i'}),\,\varepsilon_{y,i'}-\varepsilon_{y,i''}\bigr) = (\theta_{x,i'}-\theta_{x,i''})\kappa_{12}\sigma_x\sigma_y^2$
- $(\theta_{y,i'}-\theta_{y,i''})\operatorname{Cov}\bigl((\varepsilon_{x,i}-\varepsilon_{x,i'})(\varepsilon_{y,i}-\varepsilon_{y,i'}),\,\varepsilon_{x,i'}-\varepsilon_{x,i''}\bigr) = (\theta_{y,i'}-\theta_{y,i''})\kappa_{21}\sigma_x^2\sigma_y$
- $\operatorname{Cov}\bigl((\varepsilon_{x,i}-\varepsilon_{x,i'})(\varepsilon_{y,i}-\varepsilon_{y,i'}),\,(\varepsilon_{x,i'}-\varepsilon_{x,i''})(\varepsilon_{y,i'}-\varepsilon_{y,i''})\bigr) = (\kappa_{22}-\rho^2)\sigma_x^2\sigma_y^2$

where for the last three, we condition on $(\varepsilon_{x,i'}, \varepsilon_{y,i'})$ and use $E[(\varepsilon_{x,i}-\varepsilon_{x,i'})(\varepsilon_{y,i}-\varepsilon_{y,i'}) \mid \varepsilon_{x,i'}, \varepsilon_{y,i'}] = \sigma_{xy} + \varepsilon_{x,i'}\varepsilon_{y,i'}$, giving $E[(\sigma_{xy}+\varepsilon_{x,i'}\varepsilon_{y,i'})^2] = 3\sigma_{xy}^2 + \kappa_{22}\sigma_x^2\sigma_y^2$ for the last term. Collecting and using $(\theta_{x,i'}-\theta_{x,i''}) - (\theta_{x,i}-\theta_{x,i'}) = -(\theta_{x,i} - 2\theta_{x,i'} + \theta_{x,i''})$ gives the result. $\square$

**Proof of $\operatorname{Var}(T_k(X,Y))$.** The summands are nonzero only when $i = i'$, $i = i' + k$, $i + k = i'$, or $i + k = i' + k$. The last case reduces to $i = i'$, and the cases $i = i' + k$ and $i + k = i'$ are both at offset $k$, contributing $2n$ ordered off-diagonal terms in total. Therefore:
\begin{align}
\operatorname{Var}(T_k(X,Y))
&= \sum_{i=1}^n \operatorname{Var}\bigl((X_i-X_{i+k})(Y_i-Y_{i+k})\bigr) \\
&\quad + 2\sum_{i=1}^n \operatorname{Cov}\bigl((X_i-X_{i+k})(Y_i-Y_{i+k}),\,(X_{i+k}-X_{i+2k})(Y_{i+k}-Y_{i+2k})\bigr).
\end{align}

Applying Lemma 1 to the diagonal:
$$
\sum_{i=1}^n \operatorname{Var}\bigl[(X_i-X_{i+k})(Y_i-Y_{i+k})\bigr]
= 2\sigma_y^2 \sum_i(\theta_{x,i}-\theta_{x,i+k})^2 + 2\sigma_x^2\sum_i(\theta_{y,i}-\theta_{y,i+k})^2
+ 4\sigma_{xy}\sum_i(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k}) + 2n(\kappa_{22}+1)\sigma_x^2\sigma_y^2.
$$

Applying Lemma 2 with $(i, i', i'') = (i, i+k, i+2k)$ to the off-diagonal:
\begin{align}
&2\sum_{i=1}^n \operatorname{Cov}\bigl((X_i-X_{i+k})(Y_i-Y_{i+k}),\,(X_{i+k}-X_{i+2k})(Y_{i+k}-Y_{i+2k})\bigr) \\
&= 2\sum_{i=1}^n \Bigl[
(\kappa_{22}-\rho^2)\sigma_x^2\sigma_y^2
- (\theta_{x,i}-\theta_{x,i+k})(\theta_{x,i+k}-\theta_{x,i+2k})\sigma_y^2
- (\theta_{y,i}-\theta_{y,i+k})(\theta_{y,i+k}-\theta_{y,i+2k})\sigma_x^2 \\
&\qquad - \bigl[(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i+k}-\theta_{y,i+2k}) + (\theta_{y,i}-\theta_{y,i+k})(\theta_{x,i+k}-\theta_{x,i+2k})\bigr]\sigma_{xy} \\
&\qquad - (\theta_{x,i}-2\theta_{x,i+k}+\theta_{x,i+2k})\kappa_{12}\sigma_x\sigma_y^2
- (\theta_{y,i}-2\theta_{y,i+k}+\theta_{y,i+2k})\kappa_{21}\sigma_x^2\sigma_y
\Bigr].
\end{align}

The sums simplify for $k \leq L/2$: since no two change points lie within distance $2k$, any product $(\theta_{x,i}-\theta_{x,i+k})(\theta_{x,i+k}-\theta_{x,i+2k}) = 0$ for all $i$, and by circular telescoping $\sum_i(\theta_{x,i}-2\theta_{x,i+k}+\theta_{x,i+2k}) = 0$. The off-diagonal therefore reduces to $2n(\kappa_{22}-\rho^2)\sigma_x^2\sigma_y^2$.

Adding diagonal and off-diagonal and using $\sum_i(\theta_{x,i}-\theta_{x,i+k})^2 = kW_x$, $\sum_i(\theta_{y,i}-\theta_{y,i+k})^2 = kW_y$, and $\sum_i(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k}) = kW_{xy}$:
$$
\operatorname{Var}(T_k(X,Y)) = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1-\rho^2) + 2k\sigma_y^2 W_x + 2k\sigma_x^2 W_y + 4k\sigma_{xy}W_{xy}.
$$

## Variance of $T_k(X,X)$ and $T_k(Y,Y)$

Setting $Y = X$ in the above (so $\sigma_y^2 \to \sigma_x^2$, $\sigma_{xy} \to \sigma_x^2$, $\rho \to 1$, $\kappa_{22} \to \kappa_{40}$, $W_y \to W_x$, $W_{xy} \to W_x$):
$$
\operatorname{Var}(T_k(X,X)) = 2n\sigma_x^4(2\kappa_{40}+1-1) + 2k\sigma_x^2 W_x + 2k\sigma_x^2 W_x + 4k\sigma_x^2 W_x = 4n\kappa_{40}\sigma_x^4 + 8k\sigma_x^2 W_x.
$$

By the $X \leftrightarrow Y$ symmetry:
$$
\operatorname{Var}(T_k(Y,Y)) = 4n\kappa_{04}\sigma_y^4 + 8k\sigma_y^2 W_y.
$$

## Summary of diagonal entries

For $k = 1, 2$ and $k \leq L/2$:

| Entry | Formula |
|---|---|
| $\operatorname{Var}(T_k(X,X))$ | $4n\kappa_{40}\sigma_x^4 + 8k\sigma_x^2 W_x$ |
| $\operatorname{Var}(T_k(Y,Y))$ | $4n\kappa_{04}\sigma_y^4 + 8k\sigma_y^2 W_y$ |
| $\operatorname{Var}(T_k(X,Y))$ | $2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1-\rho^2) + 2k\sigma_y^2 W_x + 2k\sigma_x^2 W_y + 4k\sigma_{xy}W_{xy}$ |

## Covariance of $T_h(X,Y)$ and $T_k(X,Y)$

**Proposition.** For $h \neq k$ with $h, k \leq L/2$, let $m = \min(h,k)$. Then
$$
\operatorname{Cov}(T_h(X,Y),\,T_k(X,Y))
= 4n\sigma_x^2\sigma_y^2(\kappa_{22} - \rho^2)
+ 2m\sigma_y^2 W_x + 2m\sigma_x^2 W_y + 4m\sigma_{xy}W_{xy}.
$$

*Proof.* Expanding as a double sum and using independence, only pairs of summands whose index sets share an element contribute. For $h \neq k$, the four overlapping families are:

| Case | Shared index | Pair 1 | Pair 2 |
|---|---|---|---|
| A | $i$ (first of both) | $(i,\,i+h)$ | $(i,\,i+k)$ |
| B | $i$ (last of 2, first of 1) | $(i,\,i+h)$ | $(i-k,\,i)$ |
| C | $i+h$ (last of 1, first of 2) | $(i,\,i+h)$ | $(i+h,\,i+h+k)$ |
| D | $i+h$ (last of both) | $(i,\,i+h)$ | $(i+h-k,\,i+h)$ |

**Cases B and C** match the adjacent-index pattern of Lemma 2 (shared middle index). Applying Lemma 2 and summing over $i$: the signal cross-products vanish because the two index windows are disjoint (e.g., in Case B, $\Delta_{x,i-k,i} \neq 0$ and $\Delta_{x,i,i+h} \neq 0$ simultaneously would require two changes within distance $k+h \leq 2k$, violating the separation assumption). Each case therefore contributes only the pure-noise term $n(\kappa_{22}-\rho^2)\sigma_x^2\sigma_y^2$.

**Cases A and D** share a first (resp. last) index $a$. Decomposing signal and noise at the shared index, the covariance for a single $i$ is (writing $\Delta_{x,a,b} = \theta_{x,a}-\theta_{x,b}$ and similarly for $y$):
\begin{align}
&\operatorname{Cov}\bigl((X_i-X_{i+h})(Y_i-Y_{i+h}),\,(X_i-X_{i+k})(Y_i-Y_{i+k})\bigr) \notag\\
&= \Delta_{x,i,i+h}\Delta_{x,i,i+k}\sigma_y^2
 + \Delta_{y,i,i+h}\Delta_{y,i,i+k}\sigma_x^2
 + \bigl(\Delta_{x,i,i+h}\Delta_{y,i,i+k}+\Delta_{y,i,i+h}\Delta_{x,i,i+k}\bigr)\sigma_{xy}
 + (\kappa_{22}-\rho^2)\sigma_x^2\sigma_y^2,
\end{align}
where odd-order cross-moments vanish by symmetry of the error distribution. The same formula holds for Case D by an identical calculation at the shared last index.

Summing the signal terms over $i$: for $m = h < k$, the product $\Delta_{x,i,i+h}\Delta_{x,i,i+k}$ is nonzero only when a change point lies in $(i, i+h]$ (the smaller window forces the constraint), so $\sum_i \Delta_{x,i,i+h}\Delta_{x,i,i+k} = m\,W_x$ by the same telescoping argument as in the variance proof. Case D contributes the same sum. Combined signal from Cases A and D:
$$
2m\sigma_y^2 W_x + 2m\sigma_x^2 W_y + 4m\sigma_{xy}W_{xy}.
$$

Adding all four cases' noise contributions ($4n(\kappa_{22}-\rho^2)\sigma_x^2\sigma_y^2$) and signal contributions gives the result. $\square$

## Diagnostics

To verify the formulas, we simulate scenario 2 ($n = 1000$, $L = 4$, $r = 0.3$) under two error distributions and compare the empirical covariance matrix of $Z$ to the theoretical $\Sigma_Z$.

For each error type, we run 500 outer replicates. In each replicate, we draw 10 independent datasets, compute $Z$ for each, and take their sample covariance as one estimate of $\Sigma_Z$. Each panel in the grid below shows the distribution of residuals (estimated minus theoretical) for one entry of $\Sigma_Z$ across the 500 replicates. The red line marks zero (no bias) and the orange line marks the sample mean. The p-value is from a one-sample t-test of whether the mean residual is zero.

```{r}
#| message: false
#| echo: false
devtools::load_all("~/Documents/Research/Code/ECE")
library(tidyverse)
```

```{r}
#| echo: false
cov_Z <- function(params) {
  n   <- params$n
  sx  <- sqrt(params$S[1, 1])
  sy  <- sqrt(params$S[2, 2])
  sxy <- params$S[1, 2]
  hx  <- params$h[, 1]
  hy  <- params$h[, 2]
  wx  <- lag_diff(hx) / n
  wy  <- lag_diff(hy) / n
  wxy <- lag_diff(hx, hy) / n
  rxy <- sxy / (sx * sy)

  k <- kappa_from_params(params)
  k22 <- k$k22; k40 <- k$k40; k04 <- k$k04; k31 <- k$k31; k13 <- k$k13

  S11 <- var_Tk_est(n, sx, wx, k40, k = 1)
  S12 <- cov_Th_Tk_est(n, sx, wx, k40)
  S13 <- cov_Tk_Rk_est(n, sx, sy, sxy, wxy, k22, k = 1)
  S14 <- cov_Th_Rk_est(n, sx, sy, sxy, wxy, k22)
  S15 <- cov_Tk_Qk_est(n, sx, sy, sxy, wx, wxy, k31, k = 1)
  S16 <- cov_Th_Qk_est(n, sx, sy, sxy, wx, wxy, k31)
  S22 <- var_Tk_est(n, sx, wx, k40, k = 2)
  S23 <- cov_Th_Rk_est(n, sx, sy, sxy, wxy, k22)
  S24 <- cov_Tk_Rk_est(n, sx, sy, sxy, wxy, k22, k = 2)
  S25 <- cov_Th_Qk_est(n, sx, sy, sxy, wx, wxy, k31)
  S26 <- cov_Tk_Qk_est(n, sx, sy, sxy, wx, wxy, k31, k = 2)
  S33 <- var_Tk_est(n, sy, wy, k04, k = 1)
  S34 <- cov_Th_Tk_est(n, sy, wy, k04)
  S35 <- cov_Tk_Qk_est(n, sy, sx, sxy, wy, wxy, k13, k = 1)
  S36 <- cov_Th_Qk_est(n, sy, sx, sxy, wy, wxy, k13)
  S44 <- var_Tk_est(n, sy, wy, k04, k = 2)
  S45 <- cov_Th_Qk_est(n, sy, sx, sxy, wy, wxy, k13)
  S46 <- cov_Tk_Qk_est(n, sy, sx, sxy, wy, wxy, k13, k = 2)
  S55 <- var_Qh_est(n, sx, sy, sxy, wx, wy, wxy, k22, k = 1)
  S56 <- cov_Qh_Qk_est(n, sx, sy, sxy, wx, wy, wxy, k22)
  S66 <- var_Qh_est(n, sx, sy, sxy, wx, wy, wxy, k22, k = 2)

  S_u_upper <- (1 / (4 * n^2)) * c(
    S11,
    S12, S22,
    S13, S23, S33,
    S14, S24, S34, S44,
    S15, S25, S35, S45, S55,
    S16, S26, S36, S46, S56, S66
  )
  S_u <- matrix(0, nrow = 6, ncol = 6)
  S_u[upper.tri(S_u, diag = TRUE)] <- S_u_upper
  S_u + t(S_u) - diag(diag(S_u))
}

sim_Z_once <- function(params) {
  map_dfr(1:10, ~{
    X <- generate_data(params)
    matrix(c(
      lag_diff(X[,1], X[,1], k = 1),
      lag_diff(X[,1], X[,1], k = 2),
      lag_diff(X[,2], X[,2], k = 1),
      lag_diff(X[,2], X[,2], k = 2),
      lag_diff(X[,1], X[,2], k = 1),
      lag_diff(X[,1], X[,2], k = 2)
    ), nrow = 1) |> as.data.frame()
  }) |> cov()
}
```

```{r}
#| echo: false
#| cache: true
err_types <- c("normal", "exponential")

sim_results <- map(err_types, \(et) {
  params <- scenario(2, n = 1000, L = 4, r = 0.3, err_type = et)
  list(
    params        = params,
    cov_estimates = map(1:500, ~sim_Z_once(params)),
    S_z           = 4 * params$n^2 * cov_Z(params)
  )
}) |> setNames(err_types)
```

```{r}
#| echo: false
z_labels <- c("T1(X,X)", "T2(X,X)", "T1(Y,Y)", "T2(Y,Y)", "T1(X,Y)", "T2(X,Y)")
idx <- expand.grid(i = 1:6, j = 1:6) |> filter(i <= j)

diag_plots <- imap(sim_results, \(res, et) {
  cov_long <- imap_dfr(res$cov_estimates, ~{
    mutate(idx, rep = .y, value = .x[cbind(i, j)] - res$S_z[cbind(i, j)])
  }) |> mutate(
    row_label = factor(z_labels[i], levels = z_labels),
    col_label = factor(z_labels[j], levels = z_labels)
  )

  panel_stats <- cov_long |>
    summarise(
      mean  = mean(value),
      p_val = t.test(value)$p.value,
      .by   = c(row_label, col_label)
    ) |>
    mutate(label = paste0("p=", formatC(p_val, format = "g", digits = 2)))

  cov_long |>
    ggplot(aes(x = value)) +
    geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
    geom_vline(data = distinct(cov_long, row_label, col_label),
               aes(xintercept = 0), color = "firebrick", linewidth = 0.8) +
    geom_vline(data = panel_stats, aes(xintercept = mean),
               color = "orange", linewidth = 0.8) +
    geom_text(data = panel_stats, aes(x = Inf, y = Inf, label = label),
              hjust = 1.1, vjust = 1.5, size = 2.5, color = "black") +
    facet_grid(row_label ~ col_label, scales = "free") +
    labs(x = NULL, y = NULL,
         title = "Residuals: estimated - theoretical (n_inner=10, n_outer=500)",
         subtitle = "Red = theoretical (0), orange = sample mean, p = t-test for bias") +
    theme_minimal(base_size = 9) +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
})
```

```{r}
#| echo: false
#| results: asis
cat("::: {.panel-tabset}\n\n")
iwalk(diag_plots, \(p, et) {
  cat(paste0("## ", et, "\n\n"))
  print(p)
  cat("\n\n")
})
cat(":::\n")
```
