---
title: "Independence Test"
---

## Classical Theory

For jointly Gaussian $(X, Y)$, zero correlation and independence are equivalent, so testing $H_0: \rho = 0$ is also a test of $H_0: X \perp Y$. The statistic $T = \sqrt{n-2}\,\hat\rho / \sqrt{1 - \hat\rho^2}$ follows an exact $t_{n-2}$ null distribution.

Without the Gaussian assumption, $\rho = 0$ no longer implies independence — for example, $X \sim N(0,1)$ and $Y = X^2$ satisfy $\operatorname{Cov}(X, Y) = 0$ but $X \not\perp Y$. Two approaches handle the non-Gaussian case:

- **Permutation test.** Permuting $Y$ breaks any dependence on $X$ while preserving marginals, giving an exact test of $H_0: X \perp Y$ under any distribution.
- **Asymptotic test of independence.** Under $H_0: X \perp Y$, the errors are independent, so $\kappa_{22} = 1$ and $\sqrt{n}\,\hat\rho \xrightarrow{d} N(0, 1)$.

For the weaker null $H_0: \rho = 0$ without assuming independence, $\kappa_{22}$ need not equal 1, and the asymptotic variance is
$$
v^2 = \frac{1}{n} \left(\omega_x^2 + \omega_y^2 + \kappa_{22} + \frac{5}{2}\right).
$$
A robust test plugs in $\hat\kappa_{22}$.

{{< include kappa22-v.qmd >}}

{{< include kappa22-moments.qmd >}}

---

## Method of Moments via $S_i$ Terms

The method-of-moments estimator for $\kappa_{22} = E[\varepsilon_x^2\varepsilon_y^2]/(\sigma_x^2\sigma_y^2)$ plugs in squared residuals directly:
$$\hat\kappa_{22} = \frac{\frac{1}{n}\sum_i \varepsilon_{x,i}^2\,\varepsilon_{y,i}^2}{\hat\sigma_x^2\,\hat\sigma_y^2}.$$
When there are no change points, $\varepsilon_{x,i}$ and $\varepsilon_{y,i}$ are observable as demeaned residuals, and this estimator is consistent.

Under change points, the mean $\theta_{x,i}$ varies over time, so $X_i - \bar X$ is contaminated by jumps and $\varepsilon_{x,i}$ is not directly observable. We need surrogates for $\varepsilon_{x,i}^2$ and $\varepsilon_{y,i}^2$ that have constant expectation $\sigma_x^2$ and $\sigma_y^2$ at every index regardless of change-point amplitudes.

Recall from the multiplier bootstrap that
$$S_i = \frac{s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)}}{2}, \qquad s_i^{(k)} = (X_i - X_{i+k})(Y_i - Y_{i+k}),$$
satisfies $E[S_i] = \sigma_{xy}$ regardless of change-point amplitudes. Applying the same construction to each series separately gives
$$S_{xx,i} = \frac{(X_i-X_{i+1})^2 + (X_{i+1}-X_{i+2})^2 - (X_i-X_{i+2})^2}{2}, \qquad S_{yy,i} = \frac{(Y_i-Y_{i+1})^2 + (Y_{i+1}-Y_{i+2})^2 - (Y_i-Y_{i+2})^2}{2},$$
with $E[S_{xx,i}] = \sigma_x^2$ and $E[S_{yy,i}] = \sigma_y^2$ at every index, change points included. Substituting these as surrogates for $\varepsilon_{x,i}^2$ and $\varepsilon_{y,i}^2$ gives the ECE MoM estimator:
$$\hat\kappa_{22}^{\mathrm{MoM}} = \frac{\frac{1}{n}\sum_i S_{xx,i}\,S_{yy,i}}{\bar S_{xx}\,\bar S_{yy}}.$$

The $S_i$ terms are autocorrelated at lags 1 and 2 (consecutive terms share observations), so $\frac{1}{n}\sum S_{xx,i}S_{yy,i}$ has inflated variance relative to an i.i.d. average. The estimator is consistent for $\kappa_{22}$ under the null $\sigma_{xy}=0$ — the regime relevant to the independence test (proof below). For $\sigma_{xy}\neq 0$ the surrogate product carries an $O(\rho^2)$ bias, so $\hat\kappa_{22}^{\mathrm{MoM}}$ targets $\kappa_{22}+3\rho^2+\rho\,\bar B/(\sigma_x^2\sigma_y^2)$ rather than $\kappa_{22}$.

::: {.callout-note collapse="true"}
## Consistency under $\sigma_{xy}=0$: $\frac{1}{n}\sum_i S_{xx,i}S_{yy,i} \xrightarrow{p} E[\varepsilon_x^2\varepsilon_y^2]$

**Definitions.** Write $X_i - X_{i+k} = h_{x,i}^{(k)} + e_{x,i}^{(k)}$, where $h_{x,i}^{(k)} = \theta_{x,i} - \theta_{x,i+k}$ is the signal difference and $e_{x,i}^{(k)} = \varepsilon_{x,i} - \varepsilon_{x,i+k}$ is the noise difference. Define the squared differences
$$s_{x,i}^{(k)} = \bigl(X_i - X_{i+k}\bigr)^2 = \bigl(h_{x,i}^{(k)} + e_{x,i}^{(k)}\bigr)^2,$$
and likewise $s_{y,i}^{(k)} = (Y_i - Y_{i+k})^2 = (h_{y,i}^{(k)} + e_{y,i}^{(k)})^2$. Then
$$S_{xx,i} = \frac{s_{x,i}^{(1)} + s_{x,i+1}^{(1)} - s_{x,i}^{(2)}}{2}, \qquad S_{yy,i} = \frac{s_{y,i}^{(1)} + s_{y,i+1}^{(1)} - s_{y,i}^{(2)}}{2}.$$

**Expanding the product.** Multiplying out gives
$$4\,S_{xx,i}\,S_{yy,i} = s_{x,i}^{(1)}s_{y,i}^{(1)} + s_{x,i}^{(1)}s_{y,i+1}^{(1)} - s_{x,i}^{(1)}s_{y,i}^{(2)} + s_{x,i+1}^{(1)}s_{y,i}^{(1)} + s_{x,i+1}^{(1)}s_{y,i+1}^{(1)} - s_{x,i+1}^{(1)}s_{y,i}^{(2)} - s_{x,i}^{(2)}s_{y,i}^{(1)} - s_{x,i}^{(2)}s_{y,i+1}^{(1)} + s_{x,i}^{(2)}s_{y,i}^{(2)}.$$

**General term.** It suffices to analyze $E[s_{x,i}^{(k)}\,s_{y,j}^{(r)}]$ for general starting indices $i,j$ and lags $k,r$. Substituting the signal–noise decomposition and expanding each square gives nine terms:
$$
s_{x,i}^{(k)}\,s_{y,j}^{(r)} = \begin{array}{lll}
(h_{x,i}^{(k)})^2(h_{y,j}^{(r)})^2 & +\; 2(h_{x,i}^{(k)})^2 h_{y,j}^{(r)} e_{y,j}^{(r)} & +\; (h_{x,i}^{(k)})^2(e_{y,j}^{(r)})^2 \\[6pt]
+\; 2h_{x,i}^{(k)} e_{x,i}^{(k)} (h_{y,j}^{(r)})^2 & +\; 4h_{x,i}^{(k)} h_{y,j}^{(r)} e_{x,i}^{(k)} e_{y,j}^{(r)} & +\; 2h_{x,i}^{(k)} e_{x,i}^{(k)}(e_{y,j}^{(r)})^2 \\[6pt]
+\; (e_{x,i}^{(k)})^2(h_{y,j}^{(r)})^2 & +\; 2(e_{x,i}^{(k)})^2 h_{y,j}^{(r)} e_{y,j}^{(r)} & +\; (e_{x,i}^{(k)})^2(e_{y,j}^{(r)})^2.
\end{array}
$$
Taking expectations and using $E[e_{x,i}^{(k)}] = E[e_{y,j}^{(r)}] = 0$ to drop the terms in position (1,2) and (2,1) of the grid, seven terms remain:
\begin{align}
E[s_{x,i}^{(k)}\,s_{y,j}^{(r)}]
&= (h_{x,i}^{(k)})^2(h_{y,j}^{(r)})^2 \\
&\quad+ (h_{x,i}^{(k)})^2\,E[(e_{y,j}^{(r)})^2] + (h_{y,j}^{(r)})^2\,E[(e_{x,i}^{(k)})^2] \\
&\quad+ 4h_{x,i}^{(k)}h_{y,j}^{(r)}\,E[e_{x,i}^{(k)}\,e_{y,j}^{(r)}] \\
&\quad+ 2h_{x,i}^{(k)}\,E[e_{x,i}^{(k)}(e_{y,j}^{(r)})^2] + 2h_{y,j}^{(r)}\,E[(e_{x,i}^{(k)})^2 e_{y,j}^{(r)}] \\
&\quad+ E[(e_{x,i}^{(k)})^2(e_{y,j}^{(r)})^2].
\end{align}

**Case 1: $i = j$, $k = r$.** Both increments span the same pair of time indices, so $e_{x,i}^{(k)}$ and $e_{y,i}^{(k)}$ involve $(\varepsilon_{x,i}, \varepsilon_{y,i})$ and $(\varepsilon_{x,i+k}, \varepsilon_{y,i+k})$. By iid:
\begin{align}
E[(e_{x,i}^{(k)})^2] &= 2\sigma_x^2, \qquad E[(e_{y,i}^{(k)})^2] = 2\sigma_y^2, \\
E[e_{x,i}^{(k)}\,e_{y,i}^{(k)}] &= 2\sigma_{xy}, \\
E[e_{x,i}^{(k)}(e_{y,i}^{(k)})^2] &= 0, \qquad E[(e_{x,i}^{(k)})^2\,e_{y,i}^{(k)}] = 0, \\
E[(e_{x,i}^{(k)})^2(e_{y,i}^{(k)})^2] &= 2\kappa_{22}\sigma_x^2\sigma_y^2 + 2\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2.
\end{align}
Substituting into the general formula:
$$E[s_{x,i}^{(k)}\,s_{y,i}^{(k)}] = (h_{x,i}^{(k)})^2(h_{y,i}^{(k)})^2 + 2\sigma_y^2(h_{x,i}^{(k)})^2 + 2\sigma_x^2(h_{y,i}^{(k)})^2 + 8\sigma_{xy}\,h_{x,i}^{(k)}\,h_{y,i}^{(k)} + 2\kappa_{22}\sigma_x^2\sigma_y^2 + 2\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2.$$

**Case 2: $j = i+1$, $k = r = 1$.** The increments $e_{x,i}^{(1)} = \varepsilon_{x,i} - \varepsilon_{x,i+1}$ and $e_{y,i+1}^{(1)} = \varepsilon_{y,i+1} - \varepsilon_{y,i+2}$ share only time index $i+1$. By iid:
\begin{align}
E[(e_{x,i}^{(1)})^2] &= 2\sigma_x^2, \qquad E[(e_{y,i+1}^{(1)})^2] = 2\sigma_y^2, \\
E[e_{x,i}^{(1)}\,e_{y,i+1}^{(1)}] &= -\sigma_{xy}, \\
E[e_{x,i}^{(1)}(e_{y,i+1}^{(1)})^2] &= -E[\varepsilon_x\varepsilon_y^2], \qquad E[(e_{x,i}^{(1)})^2\,e_{y,i+1}^{(1)}] = E[\varepsilon_x^2\varepsilon_y], \\
E[(e_{x,i}^{(1)})^2(e_{y,i+1}^{(1)})^2] &= \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_x^2\sigma_y^2.
\end{align}
Substituting into the general formula:
\begin{align}
E[s_{x,i}^{(1)}\,s_{y,i+1}^{(1)}] &= (h_{x,i}^{(1)})^2(h_{y,i+1}^{(1)})^2 + 2\sigma_y^2(h_{x,i}^{(1)})^2 + 2\sigma_x^2(h_{y,i+1}^{(1)})^2 - 4\sigma_{xy}\,h_{x,i}^{(1)}\,h_{y,i+1}^{(1)} \\
&\quad - 2h_{x,i}^{(1)}\,E[\varepsilon_x\varepsilon_y^2] + 2h_{y,i+1}^{(1)}\,E[\varepsilon_x^2\varepsilon_y] + \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_x^2\sigma_y^2.
\end{align}

**Case 3: $i = j$, $k = 1$, $r = 2$.** The increments $e_{x,i}^{(1)} = \varepsilon_{x,i} - \varepsilon_{x,i+1}$ and $e_{y,i}^{(2)} = \varepsilon_{y,i} - \varepsilon_{y,i+2}$ share only time index $i$. By iid:
\begin{align}
E[(e_{x,i}^{(1)})^2] &= 2\sigma_x^2, \qquad E[(e_{y,i}^{(2)})^2] = 2\sigma_y^2, \\
E[e_{x,i}^{(1)}\,e_{y,i}^{(2)}] &= \sigma_{xy}, \\
E[e_{x,i}^{(1)}(e_{y,i}^{(2)})^2] &= E[\varepsilon_x\varepsilon_y^2], \qquad E[(e_{x,i}^{(1)})^2\,e_{y,i}^{(2)}] = E[\varepsilon_x^2\varepsilon_y], \\
E[(e_{x,i}^{(1)})^2(e_{y,i}^{(2)})^2] &= \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_x^2\sigma_y^2.
\end{align}
Substituting into the general formula:
\begin{align}
E[s_{x,i}^{(1)}\,s_{y,i}^{(2)}] &= (h_{x,i}^{(1)})^2(h_{y,i}^{(2)})^2 + 2\sigma_y^2(h_{x,i}^{(1)})^2 + 2\sigma_x^2(h_{y,i}^{(2)})^2 + 4\sigma_{xy}\,h_{x,i}^{(1)}\,h_{y,i}^{(2)} \\
&\quad + 2h_{x,i}^{(1)}\,E[\varepsilon_x\varepsilon_y^2] + 2h_{y,i}^{(2)}\,E[\varepsilon_x^2\varepsilon_y] + \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_x^2\sigma_y^2.
\end{align}

**Case 4: $j = i+1$ (i.e. $j \neq i$), $k = 1$, $r = 2$.** The increments $e_{x,i+1}^{(1)} = \varepsilon_{x,i+1} - \varepsilon_{x,i+2}$ and $e_{y,i}^{(2)} = \varepsilon_{y,i} - \varepsilon_{y,i+2}$ share only time index $i+2$, which appears with a $-$ sign in both. By iid:
\begin{align}
E[(e_{x,i+1}^{(1)})^2] &= 2\sigma_x^2, \qquad E[(e_{y,i}^{(2)})^2] = 2\sigma_y^2, \\
E[e_{x,i+1}^{(1)}\,e_{y,i}^{(2)}] &= \sigma_{xy}, \\
E[e_{x,i+1}^{(1)}(e_{y,i}^{(2)})^2] &= -E[\varepsilon_x\varepsilon_y^2], \qquad E[(e_{x,i+1}^{(1)})^2\,e_{y,i}^{(2)}] = -E[\varepsilon_x^2\varepsilon_y], \\
E[(e_{x,i+1}^{(1)})^2(e_{y,i}^{(2)})^2] &= \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_x^2\sigma_y^2.
\end{align}
Substituting into the general formula:
\begin{align}
E[s_{x,i+1}^{(1)}\,s_{y,i}^{(2)}] &= (h_{x,i+1}^{(1)})^2(h_{y,i}^{(2)})^2 + 2\sigma_y^2(h_{x,i+1}^{(1)})^2 + 2\sigma_x^2(h_{y,i}^{(2)})^2 + 4\sigma_{xy}\,h_{x,i+1}^{(1)}\,h_{y,i}^{(2)} \\
&\quad - 2h_{x,i+1}^{(1)}\,E[\varepsilon_x\varepsilon_y^2] - 2h_{y,i}^{(2)}\,E[\varepsilon_x^2\varepsilon_y] + \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_x^2\sigma_y^2.
\end{align}

**Assembling the sum.** Substituting each case result into $4E[S_{xx,i}S_{yy,i}]$ and grouping by term type:
\begin{align}
4E[S_{xx,i}S_{yy,i}]
&= \bigl[(h_{x,i}^{(1)})^2 + (h_{x,i+1}^{(1)})^2 - (h_{x,i}^{(2)})^2\bigr]\bigl[(h_{y,i}^{(1)})^2 + (h_{y,i+1}^{(1)})^2 - (h_{y,i}^{(2)})^2\bigr] \\
&\quad + 2\sigma_y^2\bigl[(h_{x,i}^{(1)})^2 + (h_{x,i+1}^{(1)})^2 - (h_{x,i}^{(2)})^2\bigr] + 2\sigma_x^2\bigl[(h_{y,i}^{(1)})^2 + (h_{y,i+1}^{(1)})^2 - (h_{y,i}^{(2)})^2\bigr] \\
&\quad + \sigma_{xy}\bigl[8h_{x,i}^{(1)}h_{y,i}^{(1)} - 4h_{x,i}^{(1)}h_{y,i+1}^{(1)} - 4h_{x,i}^{(1)}h_{y,i}^{(2)} - 4h_{x,i+1}^{(1)}h_{y,i}^{(1)} + 8h_{x,i+1}^{(1)}h_{y,i+1}^{(1)} \\
&\qquad\qquad - 4h_{x,i+1}^{(1)}h_{y,i}^{(2)} - 4h_{x,i}^{(2)}h_{y,i}^{(1)} - 4h_{x,i}^{(2)}h_{y,i+1}^{(1)} + 8h_{x,i}^{(2)}h_{y,i}^{(2)}\bigr] \\
&\quad + E[\varepsilon_x\varepsilon_y^2]\bigl[-2h_{x,i}^{(1)} - 2h_{x,i}^{(1)} + 2h_{x,i+1}^{(1)} + 2h_{x,i+1}^{(1)} - 2h_{x,i}^{(2)} + 2h_{x,i}^{(2)}\bigr] \\
&\quad + E[\varepsilon_x^2\varepsilon_y]\bigl[2h_{y,i+1}^{(1)} - 2h_{y,i}^{(2)} - 2h_{y,i}^{(1)} + 2h_{y,i}^{(2)} - 2h_{y,i}^{(1)} + 2h_{y,i+1}^{(1)}\bigr] \\
&\quad + 3(2\kappa_{22}\sigma_x^2\sigma_y^2 + 2\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2) - 2(\kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_x^2\sigma_y^2).
\end{align}

**Simplification.** The first two lines each contain the factor $(h_{x,i}^{(1)})^2 + (h_{x,i+1}^{(1)})^2 - (h_{x,i}^{(2)})^2$. We claim
$$\boxed{(h_{x,i}^{(1)})^2 + (h_{x,i+1}^{(1)})^2 - (h_{x,i}^{(2)})^2 = 0.}$$
From the linear identity $h_{x,i}^{(1)} + h_{x,i+1}^{(1)} - h_{x,i}^{(2)} = 0$ we have $h_{x,i}^{(2)} = h_{x,i}^{(1)} + h_{x,i+1}^{(1)}$, so
$$(h_{x,i}^{(1)})^2 + (h_{x,i+1}^{(1)})^2 - (h_{x,i}^{(2)})^2 = (h_{x,i}^{(1)})^2 + (h_{x,i+1}^{(1)})^2 - (h_{x,i}^{(1)} + h_{x,i+1}^{(1)})^2 = -2h_{x,i}^{(1)}h_{x,i+1}^{(1)}.$$
This is not yet zero, but under the minimum segment length 2 assumption consecutive lag-1 differences cannot both be nonzero, so $h_{x,i}^{(1)}h_{x,i+1}^{(1)} = 0$ at every index. The same argument applies to $y$, so the first two lines both vanish. Collecting the remaining terms:
\begin{align}
4E[S_{xx,i}S_{yy,i}]
&= 4\sigma_{xy}\bigl[2h_{x,i}^{(1)}h_{y,i}^{(1)} - h_{x,i}^{(1)}h_{y,i+1}^{(1)} - h_{x,i}^{(1)}h_{y,i}^{(2)} - h_{x,i+1}^{(1)}h_{y,i}^{(1)} + 2h_{x,i+1}^{(1)}h_{y,i+1}^{(1)} \\
&\qquad\quad - h_{x,i+1}^{(1)}h_{y,i}^{(2)} - h_{x,i}^{(2)}h_{y,i}^{(1)} - h_{x,i}^{(2)}h_{y,i+1}^{(1)} + 2h_{x,i}^{(2)}h_{y,i}^{(2)}\bigr] \\
&\quad + 4E[\varepsilon_x\varepsilon_y^2]\bigl(h_{x,i+1}^{(1)} - h_{x,i}^{(1)}\bigr) + 4E[\varepsilon_x^2\varepsilon_y]\bigl(h_{y,i+1}^{(1)} - h_{y,i}^{(1)}\bigr) \\
&\quad + 4\kappa_{22}\sigma_x^2\sigma_y^2 + 12\sigma_{xy}^2.
\end{align}
Applying the linear identities $h_{x,i}^{(1)} + h_{x,i+1}^{(1)} - h_{x,i}^{(2)} = 0$ and $h_{y,i}^{(1)} + h_{y,i+1}^{(1)} - h_{y,i}^{(2)} = 0$ to substitute out $h_{x,i}^{(2)}$ and $h_{y,i}^{(2)}$ reduces the nine-term bracket to four terms:
\begin{align}
4E[S_{xx,i}S_{yy,i}]
&= 4\sigma_{xy}\bigl[2h_{x,i}^{(1)}h_{y,i}^{(1)} - h_{x,i}^{(1)}h_{y,i+1}^{(1)} - h_{x,i+1}^{(1)}h_{y,i}^{(1)} + 2h_{x,i+1}^{(1)}h_{y,i+1}^{(1)}\bigr] \\
&\quad + 4E[\varepsilon_x\varepsilon_y^2]\bigl(h_{x,i+1}^{(1)} - h_{x,i}^{(1)}\bigr) + 4E[\varepsilon_x^2\varepsilon_y]\bigl(h_{y,i+1}^{(1)} - h_{y,i}^{(1)}\bigr) \\
&\quad + 4\kappa_{22}\sigma_x^2\sigma_y^2 + 12\sigma_{xy}^2.
\end{align}

**Mixed-lag products vanish.** The two cross terms $h_{x,i}^{(1)}h_{y,i+1}^{(1)}$ and $h_{x,i+1}^{(1)}h_{y,i}^{(1)}$ are identically zero at every index. Take the first: $h_{x,i}^{(1)} = \theta_{x,i} - \theta_{x,i+1}$ is nonzero only if a change point falls in the gap $(i, i+1)$, while $h_{y,i+1}^{(1)} = \theta_{y,i+1} - \theta_{y,i+2}$ is nonzero only if one falls in the adjacent gap $(i+1, i+2)$. Under common change-point locations, minimum segment length 2 forbids change points in adjacent gaps, so the two factors are never simultaneously nonzero and their product is $0$. The same argument applies to $h_{x,i+1}^{(1)}h_{y,i}^{(1)}$. The bracket collapses to two terms, and dividing by 4 gives
\begin{align}
E[S_{xx,i}S_{yy,i}]
&= 2\sigma_{xy}\bigl(h_{x,i}^{(1)}h_{y,i}^{(1)} + h_{x,i+1}^{(1)}h_{y,i+1}^{(1)}\bigr) \\
&\quad + E[\varepsilon_x\varepsilon_y^2]\bigl(h_{x,i+1}^{(1)} - h_{x,i}^{(1)}\bigr) + E[\varepsilon_x^2\varepsilon_y]\bigl(h_{y,i+1}^{(1)} - h_{y,i}^{(1)}\bigr) \\
&\quad + \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_{xy}^2.
\end{align}

The signal-dependent terms make this index-varying, so $E[S_{xx,i}S_{yy,i}] \neq \kappa_{22}\sigma_x^2\sigma_y^2$ pointwise; it is the average $\frac{1}{n}\sum_i S_{xx,i}S_{yy,i}$ that matters. Summing over $i$, use that $\sum_i h_{\cdot,i+1}^{(1)} = \sum_i h_{\cdot,i}^{(1)}$, since the two are rotations of each other. The coskewness terms then cancel,
$$
\sum_i \bigl(h_{x,i+1}^{(1)} - h_{x,i}^{(1)}\bigr) = 0,
$$
and the two covariance terms coincide, giving
$$
\frac{1}{n}\sum_i E[S_{xx,i}S_{yy,i}] = \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_{xy}^2 + 4\sigma_{xy}\omega_{xy}.
$$
Both bias terms $3\sigma_{xy}^2$ and $4\sigma_{xy}\omega_{xy}$ carry a factor of $\sigma_{xy}$, so under $H_0:\sigma_{xy}=0$ they vanish and $\hat\kappa_{22}^{\mathrm{MoM}}\xrightarrow{p}\kappa_{22}$.
:::


::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{22}^{\mathrm{MoM}}$

Since $S_{xx,i}$ and $S_{yy,i}$ are constructed from $(X,X)$ and $(Y,Y)$ separately, the estimator captures dependence between $\varepsilon_x^2$ and $\varepsilon_y^2$ without requiring $\varepsilon_x \perp \varepsilon_y$. We verify this across three settings using scenario 2 with $L = 4$ change points and Gaussian marginals.

```{r}
#| echo: false
#| cache: true
#| message: false
k22_mom <- function(x, y) {
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  mean(Sxx * Syy) / (mean(Sxx) * mean(Syy))
}

params <- scenario(2, L = 4, err_type = "normal")

k22_indep_mom <- map_dbl(1:1000, ~ {
  X <- generate_data(params)
  k22_mom(X[, 1], X[, 2])
})

k22_dep_mom <- map_dbl(1:1000, ~ {
  ex <- rnorm(params$n)
  ey <- (ex^2 - 1) / sqrt(2)
  X  <- cbind(params$h[, 1] + ex, params$h[, 2] + ey)
  k22_mom(X[, 1], X[, 2])
})

lambda_vals <- c(0, 0.5, 1/sqrt(2), 1)
k22_true    <- 1 + 4 * lambda_vals^2

sim_results_mom <- map(lambda_vals, \(lam) {
  map_dbl(1:1000, ~ {
    ex <- rnorm(params$n)
    ey <- lam * (ex^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params$n)
    X  <- cbind(params$h[, 1] + ex, params$h[, 2] + ey)
    k22_mom(X[, 1], X[, 2])
  })
})
```

::: {.panel-tabset}

## Independent ($\kappa_{22} = 1$)

Scenario 2, $L = 4$, Gaussian independent errors. Under independence $\kappa_{22} = 1$ regardless of marginals.

```{r}
#| echo: false
make_sim_hist(k22_indep_mom, true = params$kappa$k22, expression(hat(kappa)[22]^MoM))
```

## Dependent ($\kappa_{22} = 5$)

Using $\varepsilon_y = (\varepsilon_x^2 - 1)/\sqrt{2}$ with $\varepsilon_x \sim N(0,1)$, giving $\sigma_{xy} = 0$ and $\kappa_{22} = 5$.

```{r}
#| echo: false
make_sim_hist(k22_dep_mom, true = 5, expression(hat(kappa)[22]^MoM))
```

## $\lambda$-Mixing

Using $\varepsilon_y = \lambda(\varepsilon_x^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,\eta$ with $\varepsilon_x, \eta \stackrel{iid}{\sim} N(0,1)$, giving $\sigma_{xy} = 0$ and $\kappa_{22} = 1 + 4\lambda^2$.

```{r}
#| echo: false
sim_df_mom <- tibble::tibble(
  k22   = unlist(sim_results_mom),
  group = factor(rep(round(k22_true, 1), each = 1000),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1)))
)
ref_df_mom <- tibble::tibble(
  group = factor(round(k22_true, 1),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1))),
  true  = k22_true
)
mean_df_mom <- sim_df_mom |>
  dplyr::summarise(mean_k22 = mean(k22), .by = group)

ggplot(sim_df_mom, aes(x = k22, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_mom, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_mom, aes(xintercept = mean_k22, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[22]^MoM), y = "density") +
  theme_minimal()
```

:::

:::

### Estimating $\kappa_{40}$

Since $\kappa_{22}(X,X) = E[\varepsilon_x^4]/\sigma_x^4 = \kappa_{40}$, the $\kappa_{22}$ MoM estimator applies directly with $Y = X$. Unlike the $\kappa_{22}$ case where $\sigma_{xy}$ is small under the null, here $\sigma_{xy}\vert_{Y=X} = \sigma_x^2 \neq 0$, so the bias correction is non-trivial. Substituting $\sigma_{xy} = \sigma_x^2$ into the $\kappa_{22}$ consistency result gives
$$\frac{1}{n}\sum_i E\bigl[S_{xx,i}^2\bigr] = \kappa_{40}\sigma_x^4 + 3\sigma_x^4 + 4\sigma_x^2\omega_{xx},$$
so
$$\hat\kappa_{40}^{\mathrm{MoM}} = \frac{\frac{1}{n}\sum_i S_{xx,i}^2 - 3\hat\sigma_x^4 - 4\hat\sigma_x^2\hat\omega_{xx}}{\hat\sigma_x^4}.$$

::: {.callout-note collapse="true"}
## Consistency: $\frac{1}{n}\sum_i S_{xx,i}^2 \xrightarrow{p} \kappa_{40}\sigma_x^4 + 3\sigma_x^4 + 4\sigma_x^2\omega_{xx}$

Because $\kappa_{40} = \kappa_{22}(X,X)$, we do not redo the four-case computation: we specialize the per-index master result of the $\kappa_{22}$ consistency note to $Y = X$. Under that substitution $\sigma_y^2 \to \sigma_x^2$, $\sigma_{xy} \to \sigma_x^2$, $\kappa_{22} \to \kappa_{40}$, and both coskewnesses collapse to the skewness, $E[\varepsilon_x\varepsilon_y^2], E[\varepsilon_x^2\varepsilon_y] \to \mu_3 := E[\varepsilon_x^3]$.

**Pointwise mean.** Setting $Y = X$ in the per-index result
$$E[S_{xx,i}S_{yy,i}] = 2\sigma_{xy}\bigl(h_{x,i}^{(1)}h_{y,i}^{(1)} + h_{x,i+1}^{(1)}h_{y,i+1}^{(1)}\bigr) + E[\varepsilon_x\varepsilon_y^2]\bigl(h_{x,i+1}^{(1)} - h_{x,i}^{(1)}\bigr) + E[\varepsilon_x^2\varepsilon_y]\bigl(h_{y,i+1}^{(1)} - h_{y,i}^{(1)}\bigr) + \kappa_{22}\sigma_x^2\sigma_y^2 + 3\sigma_{xy}^2$$
collapses the covariance bracket into a sum of squares:
$$E\bigl[S_{xx,i}^2\bigr] = 2\sigma_x^2\bigl((h_i^{(1)})^2 + (h_{i+1}^{(1)})^2\bigr) + 2\mu_3\bigl(h_{i+1}^{(1)} - h_i^{(1)}\bigr) + (\kappa_{40} + 3)\sigma_x^4,$$
writing $h_i^{(k)} = \theta_{x,i} - \theta_{x,i+k}$.

**Averaging over $i$.** Three pieces:

- The skewness term telescopes. With circular indexing $\sum_i h_{i+1}^{(1)} = \sum_i h_i^{(1)}$, so $\sum_i 2\mu_3(h_{i+1}^{(1)} - h_i^{(1)}) = 0$.
- The squared-jump terms each average to the signal-complexity functional $\omega_{xx} := \frac{1}{n}\sum_i (h_i^{(1)})^2 \ge 0$ (again $\sum_i (h_{i+1}^{(1)})^2 = \sum_i (h_i^{(1)})^2$ by rotation), contributing $4\sigma_x^2\omega_{xx}$.
- The constant survives as $(\kappa_{40}+3)\sigma_x^4$.

Hence
$$\frac{1}{n}\sum_i E\bigl[S_{xx,i}^2\bigr] = \kappa_{40}\sigma_x^4 + 3\sigma_x^4 + 4\sigma_x^2\omega_{xx},$$
which is the formula quoted above, now derived.

**Why the bias does not vanish.** In the $\kappa_{22}$ case the bias $3\sigma_{xy}^2 + 4\sigma_{xy}\omega_{xy}$ carried a factor of $\sigma_{xy}$ and disappeared under $H_0:\sigma_{xy}=0$. With $Y=X$ there is no null to invoke: $\sigma_{xy} = \sigma_x^2 \neq 0$. The $3\sigma_x^4$ piece — arising because $S_{xx,i}$ shares $\varepsilon_{x,i+1}$ across its two factors — is an $O(1)$ bias present at every sample size, while $4\sigma_x^2\omega_{xx}$ is an $O(L/n)$ signal bias (only the $L$ change-point gaps contribute nonzero $h_i^{(1)}$). The estimator removes both explicitly through the $-3\hat\sigma_x^4 - 4\hat\sigma_x^2\hat\omega_{xx}$ correction.

**Convergence in probability.** Each $S_{xx,i}^2$ depends only on $(\varepsilon_{x,i}, \varepsilon_{x,i+1}, \varepsilon_{x,i+2})$, so the sequence is $2$-dependent; assuming finite eighth moments $E[\varepsilon_x^8] < \infty$, the average has variance $O(1/n) \to 0$, and by Chebyshev
$$\frac{1}{n}\sum_i S_{xx,i}^2 \xrightarrow{p} \kappa_{40}\sigma_x^4 + 3\sigma_x^4 + 4\sigma_x^2\omega_{xx}.$$
With $\hat\sigma_x^2 \xrightarrow{p} \sigma_x^2$ and $\hat\omega_{xx} \xrightarrow{p} \omega_{xx}$, the continuous mapping theorem gives $\hat\kappa_{40}^{\mathrm{MoM}} \xrightarrow{p} \kappa_{40}$ — unconditionally, with no null hypothesis required.
:::

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{40}$

```{r}
#| echo: false
#| cache: true
#| message: false
k40_regress <- function(x) {
  n   <- length(x)
  sx2 <- ece.cov(x, x)
  (2 * lag_diff2(x, x, k = 1) - lag_diff2(x, x, k = 2)) / (2 * n * sx2^2) - 3
}

params_k40_gauss <- scenario(2, L = 4, err_type = "normal")
params_k40_exp   <- scenario(2, L = 4, err_type = "exponential")

k40_gauss_reg <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k40_gauss)
  k40_regress(X[, 1])
})

k40_exp_reg <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k40_exp)
  k40_regress(X[, 1])
})

lambda_vals_k40 <- c(0, 0.5, 1/sqrt(2), 1)
k40_true_reg    <- 12 * lambda_vals_k40^4 + 3

sim_results_k40_reg <- map(lambda_vals_k40, \(lam) {
  map_dbl(1:1000, \(.) {
    z  <- rnorm(params_k40_gauss$n)
    ex <- lam * (z^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params_k40_gauss$n)
    k40_regress(params_k40_gauss$h[, 1] + ex)
  })
})
```

::: {.panel-tabset}

## Gaussian ($\kappa_{40} = 3$)

Scenario 2, $L = 4$, Gaussian errors. Under Gaussianity $\kappa_{40} = 3$.

```{r}
#| echo: false
make_sim_hist(k40_gauss_reg, true = params_k40_gauss$kappa$k40, expression(hat(kappa)[40]))
```

## Exponential ($\kappa_{40} = 9$)

Scenario 2, $L = 4$, exponential errors ($\kappa_{40} = 9$).

```{r}
#| echo: false
make_sim_hist(k40_exp_reg, true = params_k40_exp$kappa$k40, expression(hat(kappa)[40]))
```

## $\lambda$-Mixing

Using $\varepsilon_x = \lambda(z^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,w$ with $z, w \stackrel{iid}{\sim} N(0,1)$, which gives $\kappa_{40} = 12\lambda^4 + 3$.

```{r}
#| echo: false
sim_df_k40_reg <- tibble::tibble(
  k40   = unlist(sim_results_k40_reg),
  group = factor(rep(round(k40_true_reg, 1), each = 1000),
                 labels = paste0("κ₄₀ = ", round(k40_true_reg, 1)))
)
ref_df_k40_reg <- tibble::tibble(
  group = factor(round(k40_true_reg, 1),
                 labels = paste0("κ₄₀ = ", round(k40_true_reg, 1))),
  true  = k40_true_reg
)
mean_df_k40_reg <- sim_df_k40_reg |>
  dplyr::summarise(mean_k40 = mean(k40), .by = group)

ggplot(sim_df_k40_reg, aes(x = k40, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k40_reg, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k40_reg, aes(xintercept = mean_k40, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[40]), y = "density") +
  theme_minimal()
```

:::

:::

::: {.callout-note collapse="true"}
## Regression estimator and connection to quartic ECE terms

Define the quartic ECE terms
$$S_{xx,i}^{(4)} = \frac{(X_i - X_{i+1})^4 + (X_{i+1} - X_{i+2})^4 - (X_i - X_{i+2})^4}{2}.$$

Write $X_i - X_{i+k} = h_i^{(k)} + e_i^{(k)}$ where $h_i^{(k)} = \theta_{x,i} - \theta_{x,i+k}$ and $e_i^{(k)} = \varepsilon_{x,i} - \varepsilon_{x,i+k}$. Expanding:
$$(h_i^{(k)} + e_i^{(k)})^4 = (h_i^{(k)})^4 + 4(h_i^{(k)})^3 e_i^{(k)} + 6(h_i^{(k)})^2(e_i^{(k)})^2 + 4h_i^{(k)}(e_i^{(k)})^3 + (e_i^{(k)})^4.$$
Taking expectations, the odd cross-terms $4(h_i^{(k)})^3 e_i^{(k)}$ and $4h_i^{(k)}(e_i^{(k)})^3$ vanish: $E[e_i^{(k)}] = 0$ trivially, and by iid (even for asymmetric distributions),
$$E\bigl[(e_i^{(k)})^3\bigr] = E[\varepsilon_x^3] - E[\varepsilon_x^3] = 0.$$
The remaining three groups each cancel:

**Pure signal:**
$$(h_i^{(1)})^4 + (h_{i+1}^{(1)})^4 - (h_i^{(2)})^4 = 0$$
by the change-point argument (min segment length 2 forces $h_i^{(1)} = h_i^{(2)}$ and $h_{i+1}^{(1)} = 0$, or vice versa).

**Signal–noise cross term:** $E[6h^2 e^2] = 12\sigma_x^2(h_i^{(k)})^2$, so the combination gives
$$(h_i^{(1)})^2 + (h_{i+1}^{(1)})^2 - (h_i^{(2)})^2 = 0$$
by the quadratic ECE cancellation.

**Pure noise:** For any $k \geq 1$, $E[(e_i^{(k)})^4] = (2\kappa_{40}+6)\sigma_x^4$ by iid, so the combination gives
$$\tfrac{1}{2}\bigl[(2\kappa_{40}+6) + (2\kappa_{40}+6) - (2\kappa_{40}+6)\bigr]\sigma_x^4 = (\kappa_{40}+3)\sigma_x^4.$$

Hence $S_{xx,i}^{(4)}$ has **constant mean at every index**:
$$E\bigl[S_{xx,i}^{(4)}\bigr] = (\kappa_{40}+3)\sigma_x^4.$$

The residual $+3$ arises because $S_{xx,i}$ shares $\varepsilon_{x,i+1}$ across its two factors. Since $S_{xx,i}$ and $S_{xx,i+3}$ use non-overlapping observations, $E[S_{xx,i}\cdot S_{xx,i+3}] = \sigma_x^4$ at every index. Subtracting three times this lag-3 product gives constant mean $\kappa_{40}\sigma_x^4$:
$$\hat\kappa_{40}^{\mathrm{MoM}} = \frac{\frac{1}{n}\sum_i \bigl[S_{xx,i}^{(4)} - 3\,S_{xx,i}\,S_{xx,i+3}\bigr]}{\bar S_{xx}^2}.$$

With circular indexing $\sum_i S_{xx,i}^{(4)} = (2T_1^{(2)}(X,X) - T_2^{(2)}(X,X))/2$, so this is algebraically identical to the regression formula above when $\tfrac{1}{n}\sum S_{xx,i}S_{xx,i+3}$ is replaced by $\bar S_{xx}^2$.
:::

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{40}^{\mathrm{MoM}}$

```{r}
#| echo: false
#| cache: true
#| message: false
k40_mom <- function(x) {
  Sxx <- ece_terms(cbind(x, x))
  sx2 <- mean(Sxx)
  wxx <- ece.complexity(x, x)
  (mean(Sxx^2) - 3 * sx2^2 - 4 * sx2 * wxx) / sx2^2
}

k40_gauss_mom <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k40_gauss)
  k40_mom(X[, 1])
})

k40_exp_mom <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k40_exp)
  k40_mom(X[, 1])
})

sim_results_k40_mom <- map(lambda_vals_k40, \(lam) {
  map_dbl(1:1000, \(.) {
    z  <- rnorm(params_k40_gauss$n)
    ex <- lam * (z^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params_k40_gauss$n)
    k40_mom(params_k40_gauss$h[, 1] + ex)
  })
})
```

::: {.panel-tabset}

## Gaussian ($\kappa_{40} = 3$)

```{r}
#| echo: false
make_sim_hist(k40_gauss_mom, true = params_k40_gauss$kappa$k40, expression(hat(kappa)[40]^MoM))
```

## Exponential ($\kappa_{40} = 9$)

```{r}
#| echo: false
make_sim_hist(k40_exp_mom, true = params_k40_exp$kappa$k40, expression(hat(kappa)[40]^MoM))
```

## $\lambda$-Mixing

```{r}
#| echo: false
sim_df_k40_mom <- tibble::tibble(
  k40   = unlist(sim_results_k40_mom),
  group = factor(rep(round(k40_true_reg, 1), each = 1000),
                 labels = paste0("κ₄₀ = ", round(k40_true_reg, 1)))
)
ref_df_k40_mom <- tibble::tibble(
  group = factor(round(k40_true_reg, 1),
                 labels = paste0("κ₄₀ = ", round(k40_true_reg, 1))),
  true  = k40_true_reg
)
mean_df_k40_mom <- sim_df_k40_mom |>
  dplyr::summarise(mean_k40 = mean(k40), .by = group)

ggplot(sim_df_k40_mom, aes(x = k40, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k40_mom, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k40_mom, aes(xintercept = mean_k40, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[40]^MoM), y = "density") +
  theme_minimal()
```

:::

:::

::: {.callout-tip collapse="true"}
## Numerical Results: Comparing $\hat\kappa_{40}$ Estimators

We compare bias and variance of the regression and MoM estimators across scenarios.

```{r}
#| echo: false
simulate_k40 <- function(params) {
  map(1:1000, ~{
    x <- generate_data(params)[, 1]
    list(
      regress = k40_regress(x),
      mom     = k40_mom(x)
    )
  })
}
```

```{r}
#| cache: true
#| echo: false
param_grid_k40 <- tribble(
  ~label,                       ~scen, ~err_type,      ~amp, ~L,
  "Scenario 1, Normal",          1,     "normal",       1,    NA,
  "Scenario 1, Exponential",     1,     "exponential",  1,    NA,
  "Scenario 2, Normal",          2,     "normal",       1,    4,
  "Scenario 2, Exponential",     2,     "exponential",  1,    4,
  "Scenario 2 (amp = 2), Exp.", 2,     "exponential",  2,    4
)

k40_grid <- pmap(param_grid_k40, function(label, scen, err_type, amp, L) {
  params <- if (is.na(L)) {
    scenario(scen, err_type = err_type, amp = amp)
  } else {
    scenario(scen, err_type = err_type, amp = amp, L = L)
  }
  sim <- simulate_k40(params)
  list(
    label       = label,
    true        = params$kappa$k40,
    var_regress = var(map_dbl(sim, "regress")),
    var_mom     = var(map_dbl(sim, "mom")),
    sim         = sim
  )
})

ci_df_k40 <- map_dfr(k40_grid, \(g) {
  tibble(
    scenario = g$label,
    method   = c("Regression", "MoM"),
    bias     = c(mean(map_dbl(g$sim, "regress")),
                 mean(map_dbl(g$sim, "mom"))) - g$true,
    sd       = c(sd(map_dbl(g$sim, "regress")),
                 sd(map_dbl(g$sim, "mom")))
  )
}) |>
  mutate(
    lo       = bias - 1.96 * sd,
    hi       = bias + 1.96 * sd,
    scenario = factor(scenario, levels = map_chr(k40_grid, "label")),
    method   = factor(method, levels = c("Regression", "MoM"))
  )

ggplot(ci_df_k40, aes(x = scenario, y = bias, color = method, group = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(aes(ymin = lo, ymax = hi),
                  position = position_dodge(width = 0.4)) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = NULL, y = expression(hat(kappa)[40] - kappa[40]),
       title = expression("Bias and variance of " * hat(kappa)[40] * " estimators")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

::: {.panel-tabset}

## Scenario 1, Normal

```{r}
#| echo: false
make_overlap_hist(
  sims = list(
    Regression = map_dbl(k40_grid[[1]]$sim, "regress"),
    MoM        = map_dbl(k40_grid[[1]]$sim, "mom")
  ),
  xlab   = expression(hat(kappa)[40]),
  colors = c(Regression = "#E41A1C", MoM = "#4DAF4A")
) + ggplot2::labs(title = "Scenario 1, normal errors, n = 200")
```

## Scenario 2, Exponential (amp = 2)

```{r}
#| echo: false
make_overlap_hist(
  sims = list(
    Regression = map_dbl(k40_grid[[5]]$sim, "regress"),
    MoM        = map_dbl(k40_grid[[5]]$sim, "mom")
  ),
  xlab   = expression(hat(kappa)[40]),
  colors = c(Regression = "#E41A1C", MoM = "#4DAF4A")
) + ggplot2::labs(title = "Scenario 2, exponential errors, n = 200, L = 4, amp = 2")
```

:::

:::

### Estimating $\kappa_{31}$

Define the cubic-linear ECE terms
$$T_i = \frac{t_i^{(1)} + t_{i+1}^{(1)} - t_i^{(2)}}{2}, \qquad t_i^{(k)} = (X_i - X_{i+k})^3(Y_i - Y_{i+k}).$$

The same change-point cancellation gives constant mean at every index: $E[T_i] = \kappa_{31}\sigma_x^3\sigma_y + 3\sigma_x^2\sigma_{xy}$. Solving,
$$\hat\kappa_{31} = \frac{\bar T - 3\hat\sigma_x^2\hat\sigma_{xy}}{\bar S_{xx}^{3/2}\,\bar S_{yy}^{1/2}}.$$

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{31}$

```{r}
#| echo: false
#| cache: true
#| message: false
cubic_linear_terms <- function(x, y) {
  d1x <- x - rotate(x, 1)
  d2x <- x - rotate(x, 2)
  d1y <- y - rotate(y, 1)
  d2y <- y - rotate(y, 2)
  (d1x^3 * d1y + rotate(d1x, 1)^3 * rotate(d1y, 1) - d2x^3 * d2y) / 2
}

k31_mom <- function(x, y) {
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  Sxy <- ece_terms(cbind(x, y))
  sx2 <- mean(Sxx); sy2 <- mean(Syy); sxy <- mean(Sxy)
  (mean(cubic_linear_terms(x, y)) - 3 * sx2 * sxy) / (sx2^(3/2) * sy2^(1/2))
}

params_k31 <- scenario(2, L = 4, err_type = "normal")

k31_indep_mom <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k31)
  k31_mom(X[, 1], X[, 2])
})

k31_dep_mom <- map_dbl(1:1000, ~ {
  ex <- rnorm(params_k31$n)
  X  <- cbind(params_k31$h[, 1] + ex, params_k31$h[, 2] + ex)
  k31_mom(X[, 1], X[, 2])
})

lambda_vals_k31_mom <- c(0, 0.5, 1/sqrt(2), 1)
k31_true_mom        <- 3 * lambda_vals_k31_mom

sim_results_k31_mom <- map(lambda_vals_k31_mom, \(lam) {
  map_dbl(1:1000, \(.) {
    ex <- rnorm(params_k31$n)
    ey <- lam * ex + sqrt(1 - lam^2) * rnorm(params_k31$n)
    X  <- cbind(params_k31$h[, 1] + ex, params_k31$h[, 2] + ey)
    k31_mom(X[, 1], X[, 2])
  })
})
```

::: {.panel-tabset}

## Independent ($\kappa_{31} = 0$)

Scenario 2, $L = 4$, Gaussian independent errors. Under independence $\kappa_{31} = 0$.

```{r}
#| echo: false
make_sim_hist(k31_indep_mom, true = params_k31$kappa$k31, expression(hat(kappa)[31]))
```

## Dependent ($\kappa_{31} = 3$)

Using $\varepsilon_y = \varepsilon_x$ with $\varepsilon_x \sim N(0,1)$, giving $\sigma_{xy} = \sigma_x^2 = 1$ and $\kappa_{31} = 3$.

```{r}
#| echo: false
make_sim_hist(k31_dep_mom, true = 3, expression(hat(kappa)[31]))
```

## $\lambda$-Mixing

Using $\varepsilon_y = \lambda\varepsilon_x + \sqrt{1-\lambda^2}\,\eta$ with $\varepsilon_x, \eta \stackrel{iid}{\sim} N(0,1)$, giving $\kappa_{31} = 3\lambda$.

```{r}
#| echo: false
sim_df_k31_mom <- tibble::tibble(
  k31   = unlist(sim_results_k31_mom),
  group = factor(rep(round(k31_true_mom, 2), each = 1000),
                 labels = paste0("κ₃₁ = ", round(k31_true_mom, 2)))
)
ref_df_k31_mom <- tibble::tibble(
  group = factor(round(k31_true_mom, 2),
                 labels = paste0("κ₃₁ = ", round(k31_true_mom, 2))),
  true  = k31_true_mom
)
mean_df_k31_mom <- sim_df_k31_mom |>
  dplyr::summarise(mean_k31 = mean(k31), .by = group)

ggplot(sim_df_k31_mom, aes(x = k31, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k31_mom, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k31_mom, aes(xintercept = mean_k31, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[31]), y = "density") +
  theme_minimal()
```

:::

:::

---

## Numerical Results: Estimating $\kappa_{22}$

The MoM estimator has lower variance than both the regression and covariance estimators across scenarios. Between regression and covariance, neither dominates: regression wins in low-amplitude settings, covariance wins when change-point amplitude is large.

```{R}
#| echo: false

# Compare k22 Estimators
ece.k22.regress <- function(x, y){
  n <- length(x) # WARN: be careful if x is a matrix
  sx2 <- ece.cov(x, x)
  sy2 <- ece.cov(y, y)
  sxy <- ece.cov(x, y)
  (2 * lag_diff2(x, y, k=1) - lag_diff2(x, y, k=2)) / (2 * n * sx2 * sy2) - 
    2 * sxy^2 / (sx2 * sy2) - 1
}

ece.k22.matrix <- function(x, y = NULL, rho = NULL) {
  if (is.matrix(x)) {
    if (ncol(x) != 2) stop("matrix input must have exactly 2 columns")
    y <- x[, 2]
    x <- x[, 1]
  }

  n <- length(x)
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  Sxy <- ece_terms(cbind(x, y))

  sx2 <- mean(Sxx)
  sy2 <- mean(Syy)
  sxy <- if (!is.null(rho) && rho == 0) 0 else mean(Sxy)

  S <- cbind(Sxx - sx2, Syy - sy2, Sxy - sxy)
  V <- (1 / n^2) * (
    t(S) %*% S +
      t(rotate(S)) %*% S + t(S) %*% rotate(S) +
      t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
  )

  wx2 <- ece.complexity(x, x)
  wy2 <- ece.complexity(y, y)
  wxy <- ece.complexity(x, y)

  k22 <- n * V[3, 3] / (sx2 * sy2) -
    3 / 2 * sxy^2 / (sx2 * sy2) -
    2 * sxy * wxy / (sx2 * sy2) -
    wx2 / sx2 -
    wy2 / sy2 - 5 / 2
  return(k22)
}

ece.k22.mom <- function(x, y) {
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))

  sx2 <- mean(Sxx)
  sy2 <- mean(Syy)
  sxy <- ece.cov(x, y)
  wxy <- ece.complexity(x, y)

  (mean(Sxx * Syy) - 3 * sxy^2 - 4 * sxy * wxy) / (sx2 * sy2)
}
```

```{R}
#| cache: true
#| echo: false

# Simulation
simulate_k22 <- function(params){
  map(1:1000, ~{
    X <- generate_data(params) 
    x <- X[,1]; y <- X[,2]
    list(
      regress = ece.k22.regress(x, y),
      matrix  = ece.k22.matrix(x, y),
      mom     = ece.k22.mom(x, y)
    )
  })
}
```

```{R}
#| cache: true
#| echo: false
param_grid <- tribble(
  ~label,                        ~scen, ~r,   ~err_type,      ~amp, ~L,
  "Scenario 1, Normal",           1,     0,    "normal",       1,    NA,
  "Scenario 1, Exponential",      1,     0,    "exponential",  1,    NA,
  "Scenario 1 (r = 0.5), Exp.",  1,     0.5,  "exponential",  1,    NA,
  "Scenario 2, Normal",           2,     0,    "normal",       1,    4,
  "Scenario 2, Exponential",      2,     0,    "exponential",  1,    4,
  "Scenario 2 (amp = 2), Exp.",  2,     0,    "exponential",  2,    4
)

k22_grid <- pmap(param_grid, function(label, scen, r, err_type, amp, L) {
  params <- if (is.na(L)) {
    scenario(scen, r = r, err_type = err_type, amp = amp)
  } else {
    scenario(scen, r = r, err_type = err_type, amp = amp, L = L)
  }
  sim <- simulate_k22(params)
  list(
    label = label,
    true  = params$kappa$k22,
    sim   = sim
  )
})
```

```{R}
#| echo: false
ci_df <- map_dfr(k22_grid, \(g) {
  tibble(
    scenario = g$label,
    method   = c("Regression", "Covariance", "MoM"),
    bias     = c(mean(map_dbl(g$sim, "regress")),
                 mean(map_dbl(g$sim, "matrix")),
                 mean(map_dbl(g$sim, "mom"))) - g$true,
    sd       = c(sd(map_dbl(g$sim, "regress")),
                 sd(map_dbl(g$sim, "matrix")),
                 sd(map_dbl(g$sim, "mom")))
  )
}) |>
  mutate(
    lo       = bias - 1.96 * sd,
    hi       = bias + 1.96 * sd,
    scenario = factor(scenario, levels = map_chr(k22_grid, "label")),
    method   = factor(method, levels = c("Regression", "Covariance", "MoM"))
  )

ggplot(ci_df, aes(x = scenario, y = bias, color = method, group = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(aes(ymin = lo, ymax = hi),
                  position = position_dodge(width = 0.4)) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = NULL, y = expression(hat(kappa)[22] - kappa[22]),
       title = expression("Bias and variance of " * hat(kappa)[22] * " estimators")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

::: {.panel-tabset}

## Regression Dominates

```{R}
#| echo: false
make_overlap_hist(
  sims = list(
    Regression = map_dbl(k22_grid[[1]]$sim, "regress"),
    Covariance = map_dbl(k22_grid[[1]]$sim, "matrix"),
    MoM        = map_dbl(k22_grid[[1]]$sim, "mom")
  ),
  xlab   = expression(hat(kappa)[22]),
  colors = c(Regression = "#E41A1C", Covariance = "#377EB8", MoM = "#4DAF4A")
) + ggplot2::labs(title = "Scenario 1, normal errors, n = 1000, ρ = 0") +
  ggplot2::geom_vline(xintercept = k22_grid[[1]]$true, linetype = "solid")
```

## Covariance Dominates

```{R}
#| echo: false
make_overlap_hist(
  sims = list(
    Regression = map_dbl(k22_grid[[6]]$sim, "regress"),
    Covariance = map_dbl(k22_grid[[6]]$sim, "matrix"),
    MoM        = map_dbl(k22_grid[[6]]$sim, "mom")
  ),
  xlab   = expression(hat(kappa)[22]),
  colors = c(Regression = "#E41A1C", Covariance = "#377EB8", MoM = "#4DAF4A")
) + ggplot2::labs(title = "Scenario 2, exponential errors, n = 1000, ρ = 0, L = 4, amp = 2") +
  ggplot2::geom_vline(xintercept = k22_grid[[6]]$true, linetype = "solid")
```

:::

::: {.callout-warning}
## Why does MoM have lower variance than the regression estimator?

The regression estimator computes a single fourth-order joint statistic $T_k^{(2)} = \sum_i (X_i - X_{i+k})^2(Y_i - Y_{i+k})^2$, whereas the MoM builds separate second-order estimates $S_{xx,i} \approx \sigma_x^2$ and $S_{yy,i} \approx \sigma_y^2$ and multiplies them. Since $\kappa_{22}\sigma_x^2\sigma_y^2 = E[\varepsilon_x^2\varepsilon_y^2]$ factors, this separation pays off: for independent estimates $\hat A, \hat B$ of $A, B$,
$$\operatorname{Var}[\hat A \hat B] \approx A^2\operatorname{Var}[\hat B] + B^2\operatorname{Var}[\hat A] \leq \operatorname{Var}[\widehat{AB}].$$

Near a joint change point the advantage is even larger: $(X_i - X_{i+k})^2(Y_i - Y_{i+k})^2 \sim h_x^2 h_y^2$ is fourth-order in jump amplitudes, while $S_{xx,i} S_{yy,i}$ is only quadratic (the signal's squared term cancels inside each ECE term), so the regression numerator is far noisier in high-amplitude scenarios.
:::


---

## Kappa Table
```{R} 
#| echo: false
#| cache: true

# Create Table
rho <- 0
sample_size <- 1000

set.seed(2)
params_norm <- scenario(1, n=sample_size, r=rho, err_type="normal")
params_exp <- scenario(1, n=sample_size, r=rho, err_type="exponential")
params_lognorm <- scenario(1, n=sample_size, r=rho, err_type="lognormal")
params_t <- scenario(1, n=sample_size, r=rho, err_type="t")
params <- list(params_norm, params_exp, params_lognorm, params_t)

# Create Table
dist <- c("Normal", "Exponential", "Log-Normal", "t (df = 5)")
kappa_tbl <- map_dfr(params, function(p) unlist(p$kappa)) |> round(4)
se <- map_dbl(params, ece.cor.se.formula) |> round(4)
tbl <- cbind(dist, kappa_tbl, se)
knitr::kable(tbl,
  col.names = c("Distribution", "$\\kappa_{40}$", "$\\kappa_{04}$",
                "$\\kappa_{22}$", "$\\kappa_{31}$", "$\\kappa_{13}$", "$\\operatorname{SE}(\\hat{\\rho})$"))
```



For $\rho = 0.5$, we get

```{R} 
#| echo: false
#| cache: true

# Create Table
rho <- 0.5
sample_size <- 1000

set.seed(2)
params_norm <- scenario(1, n=sample_size, r=rho, err_type="normal")
params_exp <- scenario(1, n=sample_size, r=rho, err_type="exponential")
params_lognorm <- scenario(1, n=sample_size, r=rho, err_type="lognormal")
params_t <- scenario(1, n=sample_size, r=rho, err_type="t")
params2 <- list(params_norm, params_exp, params_lognorm, params_t)

# Create Table
dist <- c("Normal", "Exponential", "Log-Normal", "t (df = 5)")
kappa_tbl <- map_dfr(params2, function(p) unlist(p$kappa)) |> round(4)
se <- map_dbl(params2, ece.cor.se.formula) |> round(4)
tbl <- cbind(dist, kappa_tbl, se)
knitr::kable(tbl,
  col.names = c("Distribution", "$\\kappa_{40}$", "$\\kappa_{04}$",
                "$\\kappa_{22}$", "$\\kappa_{31}$", "$\\kappa_{13}$", "$\\operatorname{SE}(\\hat{\\rho})$"))
```


```{R} 
#| echo: false

# Plot All Data
dist_labels <- c("Normal", "Exponential", "Log-Normal", "t")
X_lst <- params |>
  map(generate_data) |>
  map(function(X) c(X[, 1], X[, 2]))

map2_dfr(X_lst, dist_labels, ~tibble(value = .x, Distribution = .y)) |>
  mutate(Distribution = factor(Distribution, levels = dist_labels)) |>
  ggplot2::ggplot(ggplot2::aes(x = value, y = ggplot2::after_stat(density))) +
  ggplot2::geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
  ggplot2::facet_wrap(~Distribution, nrow = 2) +
  ggplot2::coord_cartesian(xlim = c(-4, 4)) +
  ggplot2::labs(x = NULL, y = "density") +
  ggplot2::theme_minimal()
```
