---
title: "Independence Test"
---

## Classical Theory

For jointly Gaussian $(X, Y)$, zero correlation and independence are equivalent, so testing $H_0: \rho = 0$ is also a test of $H_0: X \perp Y$. The statistic $T = \sqrt{n-2}\,\hat\rho / \sqrt{1 - \hat\rho^2}$ follows an exact $t_{n-2}$ null distribution.

Without the Gaussian assumption, $\rho = 0$ no longer implies independence — for example, $X \sim N(0,1)$ and $Y = X^2$ satisfy $\operatorname{Cov}(X, Y) = 0$ but $X \not\perp Y$. Two approaches handle the non-Gaussian case:

- **Permutation test.** Permuting $Y$ breaks any dependence on $X$ while preserving marginals, giving an exact test of $H_0: X \perp Y$ under any distribution.
- **Asymptotic test of independence.** Under $H_0: X \perp Y$, the errors are independent, so $\kappa_{22} = 1$ and $\sqrt{n}\,\hat\rho \xrightarrow{d} N(0, 1)$.

For the weaker null $H_0: \rho = 0$ without assuming independence, $\kappa_{22}$ need not equal 1, and the asymptotic variance is
$$
v^2 = \frac{1}{n} \left(\omega_x^2 + \omega_y^2 + \kappa_{22} + \frac{5}{2}\right).
$$
A robust test plugs in $\hat\kappa_{22}$.

{{< include kappa22-v.qmd >}}

{{< include kappa22-moments.qmd >}}

---

## Method of Moments via $S_i$ Terms

Recall from the multiplier bootstrap that
$$S_i = \frac{s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)}}{2}, \qquad s_i^{(k)} = (X_i - X_{i+k})(Y_i - Y_{i+k}),$$
satisfies $E[S_i] = \sigma_{xy}$ regardless of change-point amplitudes. The same construction applied to each series separately gives
$$S_{xx,i} = \frac{(X_i-X_{i+1})^2 + (X_{i+1}-X_{i+2})^2 - (X_i-X_{i+2})^2}{2}, \qquad S_{yy,i} = \frac{(Y_i-Y_{i+1})^2 + (Y_{i+1}-Y_{i+2})^2 - (Y_i-Y_{i+2})^2}{2},$$
with $E[S_{xx,i}] = \sigma_x^2$ and $E[S_{yy,i}] = \sigma_y^2$ at every index, change points included.

These terms play the role of $\varepsilon_{x,i}^2$ and $\varepsilon_{y,i}^2$ with change-point-robust constant means. The classical method-of-moments estimator for $\kappa_{22} = E[\varepsilon_x^2\varepsilon_y^2]/(\sigma_x^2\sigma_y^2)$ is
$$\hat\kappa_{22} = \frac{\frac{1}{n}\sum_i \varepsilon_{x,i}^2\,\varepsilon_{y,i}^2}{\hat\sigma_x^2\,\hat\sigma_y^2},$$
where the numerator estimates $E[\varepsilon_x^2\varepsilon_y^2]$ and the denominator estimates $\sigma_x^2\sigma_y^2$. Replacing $\varepsilon_{x,i}^2$ with $S_{xx,i}$ and $\varepsilon_{y,i}^2$ with $S_{yy,i}$ gives the ECE analog:
$$\hat\kappa_{22}^{\mathrm{MoM}} = \frac{\frac{1}{n}\sum_i S_{xx,i}\,S_{yy,i}}{\bar S_{xx}\,\bar S_{yy}}.$$

The $S_i$ terms are autocorrelated at lags 1 and 2 (consecutive terms share observations), so $\frac{1}{n}\sum S_{xx,i}S_{yy,i}$ has inflated variance relative to an i.i.d. average. The estimator remains consistent.

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{22}^{\mathrm{MoM}}$

Since $S_{xx,i}$ and $S_{yy,i}$ are constructed from $(X,X)$ and $(Y,Y)$ separately, the estimator captures dependence between $\varepsilon_x^2$ and $\varepsilon_y^2$ without requiring $\varepsilon_x \perp \varepsilon_y$. We verify this across three settings using scenario 2 with $L = 4$ change points and Gaussian marginals.

```{r}
#| echo: false
#| cache: true
#| message: false
k22_mom <- function(x, y) {
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  mean(Sxx * Syy) / (mean(Sxx) * mean(Syy))
}

params <- scenario(2, L = 4, err_type = "normal")

k22_indep_mom <- map_dbl(1:1000, ~ {
  X <- generate_data(params)
  k22_mom(X[, 1], X[, 2])
})

k22_dep_mom <- map_dbl(1:1000, ~ {
  ex <- rnorm(params$n)
  ey <- (ex^2 - 1) / sqrt(2)
  X  <- cbind(params$h[, 1] + ex, params$h[, 2] + ey)
  k22_mom(X[, 1], X[, 2])
})

lambda_vals <- c(0, 0.5, 1/sqrt(2), 1)
k22_true    <- 1 + 4 * lambda_vals^2

sim_results_mom <- map(lambda_vals, \(lam) {
  map_dbl(1:1000, ~ {
    ex <- rnorm(params$n)
    ey <- lam * (ex^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params$n)
    X  <- cbind(params$h[, 1] + ex, params$h[, 2] + ey)
    k22_mom(X[, 1], X[, 2])
  })
})
```

::: {.panel-tabset}

## Independent ($\kappa_{22} = 1$)

Scenario 2, $L = 4$, Gaussian independent errors. Under independence $\kappa_{22} = 1$ regardless of marginals.

```{r}
#| echo: false
make_sim_hist(k22_indep_mom, true = params$kappa$k22, expression(hat(kappa)[22]^MoM))
```

## Dependent ($\kappa_{22} = 5$)

Using $\varepsilon_y = (\varepsilon_x^2 - 1)/\sqrt{2}$ with $\varepsilon_x \sim N(0,1)$, giving $\sigma_{xy} = 0$ and $\kappa_{22} = 5$.

```{r}
#| echo: false
make_sim_hist(k22_dep_mom, true = 5, expression(hat(kappa)[22]^MoM))
```

## $\lambda$-Mixing

Using $\varepsilon_y = \lambda(\varepsilon_x^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,\eta$ with $\varepsilon_x, \eta \stackrel{iid}{\sim} N(0,1)$, giving $\sigma_{xy} = 0$ and $\kappa_{22} = 1 + 4\lambda^2$.

```{r}
#| echo: false
sim_df_mom <- tibble::tibble(
  k22   = unlist(sim_results_mom),
  group = factor(rep(round(k22_true, 1), each = 1000),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1)))
)
ref_df_mom <- tibble::tibble(
  group = factor(round(k22_true, 1),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1))),
  true  = k22_true
)
mean_df_mom <- sim_df_mom |>
  dplyr::summarise(mean_k22 = mean(k22), .by = group)

ggplot(sim_df_mom, aes(x = k22, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_mom, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_mom, aes(xintercept = mean_k22, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[22]^MoM), y = "density") +
  theme_minimal()
```

:::

:::

### Estimating $\kappa_{40}$

Define $T_k^{(2)}(X,X) = \sum_i (X_i - X_{i+k})^4$. Expanding the expectation and applying the change-point cancellation $2T_1^{(m)}(\theta_x) = T_2^{(m)}(\theta_x)$ at both the second and fourth powers gives
$$2E[T_1^{(2)}(X,X)] - E[T_2^{(2)}(X,X)] = 2n(\kappa_{40}+3)\sigma_x^4,$$
so
$$\hat\kappa_{40} = \frac{2T_1^{(2)}(X,X) - T_2^{(2)}(X,X)}{2n\hat\sigma_x^4} - 3.$$

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{40}$

```{r}
#| echo: false
#| cache: true
#| message: false
k40_regress <- function(x) {
  n   <- length(x)
  sx2 <- ece.cov(x, x)
  (2 * lag_diff2(x, x, k = 1) - lag_diff2(x, x, k = 2)) / (2 * n * sx2^2) - 3
}

params_k40_gauss <- scenario(2, L = 4, err_type = "normal")
params_k40_exp   <- scenario(2, L = 4, err_type = "exponential")

k40_gauss_reg <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k40_gauss)
  k40_regress(X[, 1])
})

k40_exp_reg <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k40_exp)
  k40_regress(X[, 1])
})

lambda_vals_k40 <- c(0, 0.5, 1/sqrt(2), 1)
k40_true_reg    <- 12 * lambda_vals_k40^4 + 3

sim_results_k40_reg <- map(lambda_vals_k40, \(lam) {
  map_dbl(1:1000, \(.) {
    z  <- rnorm(params_k40_gauss$n)
    ex <- lam * (z^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params_k40_gauss$n)
    k40_regress(params_k40_gauss$h[, 1] + ex)
  })
})
```

::: {.panel-tabset}

## Gaussian ($\kappa_{40} = 3$)

Scenario 2, $L = 4$, Gaussian errors. Under Gaussianity $\kappa_{40} = 3$.

```{r}
#| echo: false
make_sim_hist(k40_gauss_reg, true = params_k40_gauss$kappa$k40, expression(hat(kappa)[40]))
```

## Exponential ($\kappa_{40} = 9$)

Scenario 2, $L = 4$, exponential errors ($\kappa_{40} = 9$).

```{r}
#| echo: false
make_sim_hist(k40_exp_reg, true = params_k40_exp$kappa$k40, expression(hat(kappa)[40]))
```

## $\lambda$-Mixing

Using $\varepsilon_x = \lambda(z^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,w$ with $z, w \stackrel{iid}{\sim} N(0,1)$, which gives $\kappa_{40} = 12\lambda^4 + 3$.

```{r}
#| echo: false
sim_df_k40_reg <- tibble::tibble(
  k40   = unlist(sim_results_k40_reg),
  group = factor(rep(round(k40_true_reg, 1), each = 1000),
                 labels = paste0("κ₄₀ = ", round(k40_true_reg, 1)))
)
ref_df_k40_reg <- tibble::tibble(
  group = factor(round(k40_true_reg, 1),
                 labels = paste0("κ₄₀ = ", round(k40_true_reg, 1))),
  true  = k40_true_reg
)
mean_df_k40_reg <- sim_df_k40_reg |>
  dplyr::summarise(mean_k40 = mean(k40), .by = group)

ggplot(sim_df_k40_reg, aes(x = k40, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k40_reg, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k40_reg, aes(xintercept = mean_k40, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[40]), y = "density") +
  theme_minimal()
```

:::

:::

::: {.callout-note collapse="true"}
## Alternative: Term-by-Term MoM via Quartic ECE Terms

Define the quartic ECE terms
$$S_{xx,i}^{(4)} = \frac{(X_i - X_{i+1})^4 + (X_{i+1} - X_{i+2})^4 - (X_i - X_{i+2})^4}{2}.$$

By the same change-point cancellation as `ece_terms` — minimum segment length 2 forces $\Delta\theta_i\Delta\theta_{i+1} = 0$, so both the $(\Delta\theta)^4$ terms and the $12\sigma_x^2(\Delta\theta)^2$ cross-terms vanish — $S_{xx,i}^{(4)}$ has **constant mean at every index**:
$$E[S_{xx,i}^{(4)}] = (\kappa_{40} + 3)\sigma_x^4.$$

The residual $+3$ arises because $S_{xx,i} = -(X_i - X_{i+1})(X_{i+1} - X_{i+2})$ shares $\varepsilon_{x,i+1}$ across its two factors, inflating the squared expectation. Since $S_{xx,i}$ and $S_{xx,i+3}$ use non-overlapping observations they are independent, so $E[S_{xx,i} \cdot S_{xx,i+3}] = \sigma_x^4$ at every index. Subtracting three times this lag-3 product gives constant mean $\kappa_{40}\sigma_x^4$:
$$\hat\kappa_{40}^{\mathrm{MoM}} = \frac{\frac{1}{n}\sum_i \bigl[S_{xx,i}^{(4)} - 3\,S_{xx,i}\,S_{xx,i+3}\bigr]}{\bar S_{xx}^2}.$$

With circular indexing $\sum_i S_{xx,i}^{(4)} = (2T_1^{(2)}(X,X) - T_2^{(2)}(X,X))/2$, so this is algebraically identical to the regression formula above when $\frac{1}{n}\sum S_{xx,i}S_{xx,i+3}$ is replaced by $\bar S_{xx}^2$.
:::

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{40}^{\mathrm{MoM}}$

```{r}
#| echo: false
#| cache: true
#| message: false
quartic_terms <- function(x) {
  d1 <- (x - rotate(x, 1))^4
  d2 <- (x - rotate(x, 2))^4
  (d1 + rotate(d1, 1) - d2) / 2
}

k40_mom <- function(x) {
  Sxx  <- ece_terms(cbind(x, x))
  Sxx4 <- quartic_terms(x)
  mean(Sxx4 - 3 * Sxx * rotate(Sxx, 3)) / mean(Sxx)^2
}

k40_gauss_mom <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k40_gauss)
  k40_mom(X[, 1])
})

k40_exp_mom <- map_dbl(1:1000, ~ {
  X <- generate_data(params_k40_exp)
  k40_mom(X[, 1])
})

sim_results_k40_mom <- map(lambda_vals_k40, \(lam) {
  map_dbl(1:1000, \(.) {
    z  <- rnorm(params_k40_gauss$n)
    ex <- lam * (z^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params_k40_gauss$n)
    k40_mom(params_k40_gauss$h[, 1] + ex)
  })
})
```

::: {.panel-tabset}

## Gaussian ($\kappa_{40} = 3$)

```{r}
#| echo: false
make_sim_hist(k40_gauss_mom, true = params_k40_gauss$kappa$k40, expression(hat(kappa)[40]^MoM))
```

## Exponential ($\kappa_{40} = 9$)

```{r}
#| echo: false
make_sim_hist(k40_exp_mom, true = params_k40_exp$kappa$k40, expression(hat(kappa)[40]^MoM))
```

## $\lambda$-Mixing

```{r}
#| echo: false
sim_df_k40_mom <- tibble::tibble(
  k40   = unlist(sim_results_k40_mom),
  group = factor(rep(round(k40_true_reg, 1), each = 1000),
                 labels = paste0("κ₄₀ = ", round(k40_true_reg, 1)))
)
ref_df_k40_mom <- tibble::tibble(
  group = factor(round(k40_true_reg, 1),
                 labels = paste0("κ₄₀ = ", round(k40_true_reg, 1))),
  true  = k40_true_reg
)
mean_df_k40_mom <- sim_df_k40_mom |>
  dplyr::summarise(mean_k40 = mean(k40), .by = group)

ggplot(sim_df_k40_mom, aes(x = k40, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k40_mom, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k40_mom, aes(xintercept = mean_k40, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[40]^MoM), y = "density") +
  theme_minimal()
```

:::

:::

---

## Numerical Results: Estimating $\kappa_{22}$

No single estimator dominates: the regression estimator has lower variance in most scenarios, but the matrix estimator wins when change-point amplitude is large.

```{R}
#| echo: false

# Compare k22 Estimators
ece.k22.regress <- function(x, y){
  n <- length(x) # WARN: be careful if x is a matrix
  sx2 <- ece.cov(x, x)
  sy2 <- ece.cov(y, y)
  sxy <- ece.cov(x, y)
  (2 * lag_diff2(x, y, k=1) - lag_diff2(x, y, k=2)) / (2 * n * sx2 * sy2) - 
    2 * sxy^2 / (sx2 * sy2) - 1
}

ece.k22.matrix <- function(x, y = NULL, rho = NULL) {
  if (is.matrix(x)) {
    if (ncol(x) != 2) stop("matrix input must have exactly 2 columns")
    y <- x[, 2]
    x <- x[, 1]
  }

  n <- length(x)
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  Sxy <- ece_terms(cbind(x, y))

  sx2 <- mean(Sxx)
  sy2 <- mean(Syy)
  sxy <- if (!is.null(rho) && rho == 0) 0 else mean(Sxy)

  S <- cbind(Sxx - sx2, Syy - sy2, Sxy - sxy)
  V <- (1 / n^2) * (
    t(S) %*% S +
      t(rotate(S)) %*% S + t(S) %*% rotate(S) +
      t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
  )

  wx2 <- ece.complexity(x, x)
  wy2 <- ece.complexity(y, y)
  wxy <- ece.complexity(x, y)

  k22 <- n * V[3, 3] / (sx2 * sy2) -
    3 / 2 * sxy^2 / (sx2 * sy2) -
    2 * sxy * wxy / (sx2 * sy2) -
    wx2 / sx2 -
    wy2 / sy2 - 5 / 2
  return(k22)
}

ece.k22.mom <- function(x, y) {
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  mean(Sxx * Syy) / (mean(Sxx) * mean(Syy))
}
```

```{R}
#| cache: true
#| echo: false

# Simulation
simulate_k22 <- function(params){
  map(1:1000, ~{
    X <- generate_data(params) 
    x <- X[,1]; y <- X[,2]
    list(
      regress = ece.k22.regress(x, y),
      matrix  = ece.k22.matrix(x, y),
      mom     = ece.k22.mom(x, y)
    )
  })
}

```

```{R}
#| cache: true
#| echo: false
param_grid <- tribble(
  ~label,                        ~scen, ~r,   ~err_type,      ~amp, ~L,
  "Scenario 1, Normal",           1,     0,    "normal",       1,    NA,
  "Scenario 1, Exponential",      1,     0,    "exponential",  1,    NA,
  "Scenario 1 (r = 0.5), Exp.",  1,     0.5,  "exponential",  1,    NA,
  "Scenario 2, Normal",           2,     0,    "normal",       1,    4,
  "Scenario 2, Exponential",      2,     0,    "exponential",  1,    4,
  "Scenario 2 (amp = 2), Exp.",  2,     0,    "exponential",  2,    4
)

k22_grid <- pmap(param_grid, function(label, scen, r, err_type, amp, L) {
  params <- if (is.na(L)) {
    scenario(scen, r = r, err_type = err_type, amp = amp)
  } else {
    scenario(scen, r = r, err_type = err_type, amp = amp, L = L)
  }
  sim    <- simulate_k22(params)
  list(
    label       = label,
    var_regress = var(map_dbl(sim, "regress")),
    var_matrix  = var(map_dbl(sim, "matrix")),
    var_mom     = var(map_dbl(sim, "mom")),
    sim         = sim
  )
})

k22_var_tbl <- map_dfr(k22_grid, \(g) tibble(
  Scenario               = g$label,
  `Var(k22-Regression)`  = round(g$var_regress, 2),
  `Var(k22-Covariance)`  = round(g$var_matrix,  2),
  `Var(k22-MoM)`         = round(g$var_mom,      2)
))

```

```{R}
#| echo: false
knitr::kable(k22_var_tbl)
```

::: {.panel-tabset}

## Regression Dominates

```{R}
#| echo: false
make_overlap_hist(
  sims = list(
    Regression = map_dbl(k22_grid[[1]]$sim, "regress"),
    Covariance = map_dbl(k22_grid[[1]]$sim, "matrix"),
    MoM        = map_dbl(k22_grid[[1]]$sim, "mom")
  ),
  xlab   = expression(hat(kappa)[22]),
  colors = c(Regression = "#E41A1C", Covariance = "#377EB8", MoM = "#4DAF4A")
) + ggplot2::labs(title = "Scenario 1, normal errors, n = 200, ρ = 0")
```

## Covariance Dominates

```{R}
#| echo: false
make_overlap_hist(
  sims = list(
    Regression = map_dbl(k22_grid[[4]]$sim, "regress"),
    Covariance = map_dbl(k22_grid[[4]]$sim, "matrix"),
    MoM        = map_dbl(k22_grid[[4]]$sim, "mom")
  ),
  xlab   = expression(hat(kappa)[22]),
  colors = c(Regression = "#E41A1C", Covariance = "#377EB8", MoM = "#4DAF4A")
) + ggplot2::labs(title = "Scenario 2, normal errors, n = 200, ρ = 0, L = 4")
```

:::


---

## Kappa Table
```{R} 
#| echo: false
#| cache: true

# Create Table
rho <- 0
sample_size <- 1000

set.seed(2)
params_norm <- scenario(1, n=sample_size, r=rho, err_type="normal")
params_exp <- scenario(1, n=sample_size, r=rho, err_type="exponential")
params_lognorm <- scenario(1, n=sample_size, r=rho, err_type="lognormal")
params_t <- scenario(1, n=sample_size, r=rho, err_type="t")
params <- list(params_norm, params_exp, params_lognorm, params_t)

# Create Table
dist <- c("Normal", "Exponential", "Log-Normal", "t (df = 5)")
kappa_tbl <- map_dfr(params, function(p) unlist(p$kappa)) |> round(4)
se <- map_dbl(params, ece.cor.se.formula) |> round(4)
tbl <- cbind(dist, kappa_tbl, se)
knitr::kable(tbl,
  col.names = c("Distribution", "$\\kappa_{40}$", "$\\kappa_{04}$",
                "$\\kappa_{22}$", "$\\kappa_{31}$", "$\\kappa_{13}$", "$\\operatorname{SE}(\\hat{\\rho})$"))
```



For $\rho = 0.5$, we get

```{R} 
#| echo: false
#| cache: true

# Create Table
rho <- 0.5
sample_size <- 1000

set.seed(2)
params_norm <- scenario(1, n=sample_size, r=rho, err_type="normal")
params_exp <- scenario(1, n=sample_size, r=rho, err_type="exponential")
params_lognorm <- scenario(1, n=sample_size, r=rho, err_type="lognormal")
params_t <- scenario(1, n=sample_size, r=rho, err_type="t")
params2 <- list(params_norm, params_exp, params_lognorm, params_t)

# Create Table
dist <- c("Normal", "Exponential", "Log-Normal", "t (df = 5)")
kappa_tbl <- map_dfr(params2, function(p) unlist(p$kappa)) |> round(4)
se <- map_dbl(params2, ece.cor.se.formula) |> round(4)
tbl <- cbind(dist, kappa_tbl, se)
knitr::kable(tbl,
  col.names = c("Distribution", "$\\kappa_{40}$", "$\\kappa_{04}$",
                "$\\kappa_{22}$", "$\\kappa_{31}$", "$\\kappa_{13}$", "$\\operatorname{SE}(\\hat{\\rho})$"))
```


```{R} 
#| echo: false

# Plot All Data
dist_labels <- c("Normal", "Exponential", "Log-Normal", "t")
X_lst <- params |>
  map(generate_data) |>
  map(function(X) c(X[, 1], X[, 2]))

map2_dfr(X_lst, dist_labels, ~tibble(value = .x, Distribution = .y)) |>
  mutate(Distribution = factor(Distribution, levels = dist_labels)) |>
  ggplot2::ggplot(ggplot2::aes(x = value, y = ggplot2::after_stat(density))) +
  ggplot2::geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
  ggplot2::facet_wrap(~Distribution, nrow = 2) +
  ggplot2::coord_cartesian(xlim = c(-4, 4)) +
  ggplot2::labs(x = NULL, y = "density") +
  ggplot2::theme_minimal()
```
