---
title: "Independence Test"
---

## Classical Theory

For jointly Gaussian $(X, Y)$, zero correlation and independence are equivalent, so testing $H_0: \rho = 0$ is also a test of $H_0: X \perp Y$. The statistic $T = \sqrt{n-2}\,\hat\rho / \sqrt{1 - \hat\rho^2}$ follows an exact $t_{n-2}$ null distribution.

Without the Gaussian assumption, $\rho = 0$ no longer implies independence — for example, $X \sim N(0,1)$ and $Y = X^2$ satisfy $\operatorname{Cov}(X, Y) = 0$ but $X \not\perp Y$. Two approaches handle the non-Gaussian case:

- **Permutation test.** Permuting $Y$ breaks any dependence on $X$ while preserving marginals, giving an exact test of $H_0: X \perp Y$ under any distribution.
- **Asymptotic test of independence.** Under $H_0: X \perp Y$, the errors are independent, so $\kappa_{22} = 1$ and $\sqrt{n}\,\hat\rho \xrightarrow{d} N(0, 1)$.

For the weaker null $H_0: \rho = 0$ without assuming independence, $\kappa_{22}$ need not equal 1, and the asymptotic variance is
$$
v^2 = \frac{1}{n} \left(\omega_x^2 + \omega_y^2 + \kappa_{22} + \frac{5}{2}\right).
$$
A robust test plugs in $\hat\kappa_{22}$.

## Estimating $\kappa_{22}$ via $V$

### Estimating $\kappa_{22}$ via $V_{xy}$

Let $s_i^{(k)} = (X_i - X_{i+k})(Y_i - Y_{i+k})$ and $S_i = \frac{s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)}}{2}$ as in the Multiplier Bootstrap chapter, so that
$$
\boxed{\bar{S} = \frac{1}{n}\sum_{i=1}^n S_i = \frac{2T_1(X,Y) - T_2(X,Y)}{2n} = \hat\sigma_{xy}.}
$$ Since $S_i$ depends only on indices $i-2, i-1, i$, terms at lags $\ge 3$ are independent, so
$$
\operatorname{Var}(\bar{S}) = \frac{1}{n^2}\!\left[\sum_i \operatorname{Var}(S_i) + 2\sum_i \operatorname{Cov}(S_i, S_{i+1}) + 2\sum_i \operatorname{Cov}(S_i, S_{i+2})\right].
$$
Replacing each population quantity with its sample analog gives the lag-2 variance estimate
$$
V := \widehat{\operatorname{Var}}(\bar{S}) = \frac{1}{n^2}\!\left(\sum_{i=1}^n S_i^2 + \sum_{i=1}^n (S_i S_{i+1} + S_{i+1} S_i) + \sum_{i=1}^n (S_i S_{i+2} + S_{i+2} S_i)\right),
$$
or in matrix form, writing $\mathbf{S} = (S_1, \ldots, S_n)^\top$,
$$
\boxed{V = \frac{1}{n^2}\,\mathbf{S}^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\mathbf{S}.}
$$

Since $\bar S = (2T_1 - T_2)/(2n)$, we have $V \approx \operatorname{Var}(2T_1 - T_2)/(4n^2)$. The variance on the right has a general closed form; under $H_0$ ($\sigma_{xy} = 0$) it simplifies so we can solve for $\kappa_{22}$ from $V$.

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Var}(\bar{S})$ from $\Sigma_Z$

Since $\bar{S} = (2T_1 - T_2)/(2n)$,
$$
\operatorname{Var}(\bar{S}) = \frac{1}{4n^2}\bigl(4[\Sigma_Z]_{55} - 4[\Sigma_Z]_{56} + [\Sigma_Z]_{66}\bigr).
$$
The general $\Sigma_Z$ entries from the Z-Test chapter are
$$[\Sigma_Z]_{55} = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) - 2n\sigma_{xy}^2 + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2 + 4n\sigma_{xy}\omega_{xy},$$
$$[\Sigma_Z]_{56} = 4n(\sigma_x^2\sigma_y^2\kappa_{22} - \sigma_{xy}^2) + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2 + 4n\sigma_{xy}\omega_{xy},$$
$$[\Sigma_Z]_{66} = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) - 2n\sigma_{xy}^2 + 4n\sigma_y^2\omega_x^2 + 4n\sigma_x^2\omega_y^2 + 8n\sigma_{xy}\omega_{xy}.$$
Substituting and collecting terms gives the general expression
$$\operatorname{Var}(\bar{S}) = \frac{1}{2n}\bigl[(2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 3\sigma_{xy}^2 + 4\sigma_{xy}\omega_{xy} + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2\bigr].$$
Under $H_0$ ($\sigma_{xy} = 0$) the last two terms vanish, yielding
$$\operatorname{Var}(\bar{S}) = \frac{1}{2n}\bigl[(2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2\bigr]. \qquad\square$$
:::

Since $V \approx \operatorname{Var}(\bar{S})$, substituting the general expression and solving gives
$$\boxed{\hat\kappa_{22} = \frac{nV}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{3}{2}\frac{\hat\sigma_{xy}^2}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{2\hat\sigma_{xy}\hat\omega_{xy}}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{\hat\omega_x^2}{\hat\sigma_x^2} - \frac{\hat\omega_y^2}{\hat\sigma_y^2} - \frac{5}{2}.}$$
Under $H_0$ ($\hat\sigma_{xy} = 0$) the middle two terms vanish, yielding
$$\hat\kappa_{22} = \frac{nV}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{\hat\omega_x^2}{\hat\sigma_x^2} - \frac{\hat\omega_y^2}{\hat\sigma_y^2} - \frac{5}{2}.$$

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{22}$ via $V_{xy}$

```{r}
#| echo: false
#| message: false
devtools::load_all("..")
library(purrr)
library(ggplot2)
```

```{r}
#| echo: false
#| cache: true
#| message: false
params <- scenario(2, L = 4, err_type = "normal")

k22_indep <- map_dbl(1:1000, ~ {
  X <- generate_data(params)
  ece.kappa(X)$k22
})

k22_dep <- map_dbl(1:1000, ~ {
  ex <- rnorm(params$n)
  ey <- (ex^2 - 1) / sqrt(2)
  X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
  ece.kappa(X)$k22
})

lambda_vals <- c(0, 0.5, 1/sqrt(2), 1)
k22_true    <- 1 + 4 * lambda_vals^2

sim_results <- map(lambda_vals, \(lam) {
  map_dbl(1:1000, ~ {
    ex <- rnorm(params$n)
    ey <- lam * (ex^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params$n)
    X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
    ece.kappa(X)$k22
  })
})
```

::: {.panel-tabset}

## Independent ($\kappa_{22} = 1$)

Scenario 2, $L = 4$, Gaussian independent errors. Under independence $\kappa_{22} = 1$ regardless of marginals.

```{r}
#| echo: false
make_sim_hist(k22_indep, true = params$kappa$k22, expression(hat(kappa)[22]))
```

## Dependent ($\kappa_{22} = 5$)

Using $\varepsilon_y = (\varepsilon_x^2 - 1)/\sqrt{2}$ with $\varepsilon_x \sim N(0,1)$, giving $\sigma_{xy} = 0$ and $\kappa_{22} = 5$.

```{r}
#| echo: false
make_sim_hist(k22_dep, true = 5, expression(hat(kappa)[22]))
```

## $\lambda$-Mixing

Using $\varepsilon_y = \lambda(\varepsilon_x^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,\eta$ with $\varepsilon_x, \eta \stackrel{iid}{\sim} N(0,1)$, which gives $\sigma_{xy} = 0$ and $\kappa_{22} = 1 + 4\lambda^2$.

```{r}
#| echo: false
sim_df <- tibble::tibble(
  k22   = unlist(sim_results),
  group = factor(rep(round(k22_true, 1), each = 1000),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1)))
)
ref_df <- tibble::tibble(
  group = factor(round(k22_true, 1),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1))),
  true  = k22_true
)
mean_df <- sim_df |>
  dplyr::summarise(mean_k22 = mean(k22), .by = group)

ggplot(sim_df, aes(x = k22, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df, aes(xintercept = mean_k22, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[22]), y = "density") +
  theme_minimal()
```

:::

:::

### Estimating $\kappa_{40}$ via $V_{xx}$

Applying the same construction to $(X, X)$ in place of $(X, Y)$: let $s_i^{(k)} = (X_i - X_{i+k})^2$ and $S_i^{xx} = (s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)})/2$, so that $\bar S^{xx} = (2T_1(X,X) - T_2(X,X))/(2n) = \hat\sigma_x^2$. Define
$$
V_{xx} := \widehat{\operatorname{Var}}(\bar S^{xx}) = \frac{1}{n^2}\,(\tilde{\mathbf{S}}^{xx})^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\tilde{\mathbf{S}}^{xx}, \qquad \tilde S_i^{xx} = S_i^{xx} - \bar S^{xx}.
$$

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Var}(\bar{S}^{xx})$ from $\Sigma_Z$

Since $\bar{S}^{xx} = (2T_1(X,X) - T_2(X,X))/(2n)$,
$$
\operatorname{Var}(\bar{S}^{xx}) = \frac{1}{4n^2}\bigl(4[\Sigma_Z]_{11} - 4[\Sigma_Z]_{12} + [\Sigma_Z]_{22}\bigr).
$$
Substituting the entries from the Z-Test chapter:
$$[\Sigma_Z]_{11} = 4n\kappa_{40}\sigma_x^4 + 8n\sigma_x^2\omega_x^2,$$
$$[\Sigma_Z]_{12} = 4n(\kappa_{40}-1)\sigma_x^4 + 8n\sigma_x^2\omega_x^2,$$
$$[\Sigma_Z]_{22} = 4n\kappa_{40}\sigma_x^4 + 16n\sigma_x^2\omega_x^2.$$
Collecting terms:
$$\operatorname{Var}(\bar{S}^{xx}) = \frac{1}{n}\bigl[(\kappa_{40}+4)\sigma_x^4 + 4\sigma_x^2\omega_x^2\bigr]. \qquad\square$$
:::

Since $V_{xx} \approx \operatorname{Var}(\bar{S}^{xx})$, substituting and solving gives
$$
\boxed{\hat\kappa_{40} = \frac{nV_{xx}}{\hat\sigma_x^4} - \frac{4\hat\omega_x^2}{\hat\sigma_x^2} - 4.}
$$
By symmetry ($x \leftrightarrow y$), the same construction applied to $(Y, Y)$ yields $\hat\kappa_{04} = nV_{yy}/\hat\sigma_y^4 - 4\hat\omega_y^2/\hat\sigma_y^2 - 4$.

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{40}$

```{r}
#| echo: false
#| cache: true
#| message: false
params     <- scenario(2, L = 4, err_type = "normal")
params_exp <- scenario(2, L = 4, err_type = "exponential")

k40_gaussian <- map_dbl(1:1000, ~ ece.kappa(generate_data(params))$k40)
k40_exp      <- map_dbl(1:1000, ~ ece.kappa(generate_data(params_exp))$k40)

lambda_vals  <- c(0, 0.5, 1/sqrt(2), 1)
k40_true     <- 12 * lambda_vals^4 + 3

sim_results_k40 <- map(lambda_vals, \(lam) {
  map_dbl(1:1000, \(.) {
    z  <- rnorm(params$n)
    ex <- lam * (z^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params$n)
    ey <- rnorm(params$n)
    X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
    ece.kappa(X)$k40
  })
})
```

::: {.panel-tabset}

## Gaussian ($\kappa_{40} = 3$)

Scenario 2, $L = 4$, Gaussian errors ($\kappa_{40} = 3$).

```{r}
#| echo: false
make_sim_hist(k40_gaussian, true = params$kappa$k40, expression(hat(kappa)[40]))
```

## Exponential ($\kappa_{40} = 9$)

Scenario 2, $L = 4$, exponential errors ($\kappa_{40} = 9$).

```{r}
#| echo: false
make_sim_hist(k40_exp, true = params_exp$kappa$k40, expression(hat(kappa)[40]))
```

## $\lambda$-Mixing

Using $\varepsilon_x = \lambda(z^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,w$ with $z, w \stackrel{iid}{\sim} N(0,1)$, which gives $\kappa_{40} = 12\lambda^4 + 3$.

```{r}
#| echo: false
sim_df_k40 <- tibble::tibble(
  k40   = unlist(sim_results_k40),
  group = factor(rep(round(k40_true, 1), each = 1000),
                 labels = paste0("κ₄₀ = ", round(k40_true, 1)))
)
ref_df_k40 <- tibble::tibble(
  group = factor(round(k40_true, 1),
                 labels = paste0("κ₄₀ = ", round(k40_true, 1))),
  true  = k40_true
)
mean_df_k40 <- sim_df_k40 |>
  dplyr::summarise(mean_k40 = mean(k40), .by = group)

ggplot(sim_df_k40, aes(x = k40, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k40, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k40, aes(xintercept = mean_k40, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[40]), y = "density") +
  theme_minimal()
```

:::

:::

### Estimating $\kappa_{31}$ via $C_{xx,xy}$

The cumulant $\kappa_{31}$ appears in the cross-covariances between the $(X,X)$ and $(X,Y)$ blocks of $\Sigma_Z$. Define the lag-2 cross-covariance
$$
C_{xx,xy} := \frac{1}{n^2}\,(\tilde{\mathbf{S}}^{xx})^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\tilde{\mathbf{S}}^{xy}, \qquad \tilde S_i^{xx} = S_i^{xx} - \bar S^{xx},\quad \tilde S_i^{xy} = S_i^{xy} - \bar S^{xy},
$$
which estimates $\operatorname{Cov}(\bar S^{xx}, \bar S^{xy})$.

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Cov}(\bar{S}^{xx}, \bar{S}^{xy})$ from $\Sigma_Z$

Since $\bar{S}^{xx} = (2T_1(X,X) - T_2(X,X))/(2n)$ and $\bar{S}^{xy} = (2T_1(X,Y) - T_2(X,Y))/(2n)$,
$$
\operatorname{Cov}(\bar{S}^{xx}, \bar{S}^{xy}) = \frac{1}{4n^2}\bigl(4[\Sigma_Z]_{15} - 2[\Sigma_Z]_{16} - 2[\Sigma_Z]_{25} + [\Sigma_Z]_{26}\bigr).
$$
The relevant entries from the Z-Test chapter are
$$[\Sigma_Z]_{15} = 4n\sigma_x^3\sigma_y\kappa_{31} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{16} = 4n\sigma_x^3\sigma_y\kappa_{31} - 4n\sigma_x^2\sigma_{xy} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{25} = 4n\sigma_x^3\sigma_y\kappa_{31} - 4n\sigma_x^2\sigma_{xy} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{26} = 4n\sigma_x^3\sigma_y\kappa_{31} + 8n\sigma_{xy}\omega_x^2 + 8n\sigma_x^2\omega_{xy}.$$
Collecting terms:
$$\operatorname{Cov}(\bar{S}^{xx}, \bar{S}^{xy}) = \frac{1}{n}\bigl[\sigma_x^3\sigma_y\kappa_{31} + 4\sigma_x^2\sigma_{xy} + 2\sigma_{xy}\omega_x^2 + 2\sigma_x^2\omega_{xy}\bigr]. \qquad\square$$
:::

Since $C_{xx,xy} \approx \operatorname{Cov}(\bar{S}^{xx}, \bar{S}^{xy})$, substituting and solving gives
$$
\hat\kappa_{31} = \frac{nC_{xx,xy}}{\hat\sigma_x^3\hat\sigma_y} - \frac{4\hat\sigma_{xy}}{\hat\sigma_x\hat\sigma_y} - \frac{2\hat\sigma_{xy}\hat\omega_x^2}{\hat\sigma_x^3\hat\sigma_y} - \frac{2\hat\omega_{xy}}{\hat\sigma_x\hat\sigma_y}.
$$
By symmetry ($x \leftrightarrow y$), the same construction with $C_{yy,xy}$ yields $\hat\kappa_{13}$.

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{31}$

```{r}
#| echo: false
#| cache: true
#| message: false
params <- scenario(2, L = 4, err_type = "normal")

k31_indep <- map_dbl(1:1000, ~ {
  X <- generate_data(params)
  ece.kappa(X)$k31
})

k31_dep <- map_dbl(1:1000, ~ {
  ex <- rnorm(params$n)
  X  <- cbind(params$h[,1] + ex, params$h[,2] + ex)
  ece.kappa(X)$k31
})

lambda_vals_k31 <- c(0, 0.5, 1/sqrt(2), 1)
k31_true        <- 3 * lambda_vals_k31

sim_results_k31 <- map(lambda_vals_k31, \(lam) {
  map_dbl(1:1000, \(.) {
    ex <- rnorm(params$n)
    ey <- lam * ex + sqrt(1 - lam^2) * rnorm(params$n)
    X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
    ece.kappa(X)$k31
  })
})
```

::: {.panel-tabset}

## Independent ($\kappa_{31} = 0$)

Scenario 2, $L = 4$, Gaussian independent errors. Under independence $\kappa_{31} = 0$.

```{r}
#| echo: false
make_sim_hist(k31_indep, true = params$kappa$k31, expression(hat(kappa)[31]))
```

## Dependent ($\kappa_{31} = 3$)

Using $\varepsilon_y = \varepsilon_x$ with $\varepsilon_x \sim N(0,1)$, giving $\sigma_{xy} = \sigma_x^2$ and $\kappa_{31} = \kappa_{40} = 3$.

```{r}
#| echo: false
make_sim_hist(k31_dep, true = 3, expression(hat(kappa)[31]))
```

## $\lambda$-Mixing

Using $\varepsilon_y = \lambda\varepsilon_x + \sqrt{1-\lambda^2}\,\eta$ with $\varepsilon_x, \eta \stackrel{iid}{\sim} N(0,1)$, which gives $\kappa_{31} = 3\lambda$.

```{r}
#| echo: false
sim_df_k31 <- tibble::tibble(
  k31   = unlist(sim_results_k31),
  group = factor(rep(round(k31_true, 2), each = 1000),
                 labels = paste0("κ₃₁ = ", round(k31_true, 2)))
)
ref_df_k31 <- tibble::tibble(
  group = factor(round(k31_true, 2),
                 labels = paste0("κ₃₁ = ", round(k31_true, 2))),
  true  = k31_true
)
mean_df_k31 <- sim_df_k31 |>
  dplyr::summarise(mean_k31 = mean(k31), .by = group)

ggplot(sim_df_k31, aes(x = k31, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k31, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k31, aes(xintercept = mean_k31, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[31]), y = "density") +
  theme_minimal()
```

:::

:::


## Estimating $\kappa_{22}$ via Moments



Under the ECE model $X_i = \theta_{x,i} + \varepsilon_{x,i}$, $Y_i = \theta_{y,i} + \varepsilon_{y,i}$ with iid mean-zero errors, define $\kappa_{22} = E[\varepsilon_{x,i}^2\varepsilon_{y,i}^2]/(\sigma_x^2\sigma_y^2)$, which equals 1 under independence. Write $T_k(\theta_x, \theta_y) = \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\theta_{y,i} - \theta_{y,i+k})$ for the signal-difference inner product and
$$
\boxed{T_k^{(2)}(X, Y) = \sum_{i=1}^n (X_i - X_{i+k})^2(Y_i - Y_{i+k})^2}
$$
for its squared analog (applicable to any pair of sequences).

The strategy is to compute $E[T_k^{(2)}(X, Y)]$ exactly and solve for $\kappa_{22}$.

::: {.callout-note collapse="true"}
## Lemma: $\operatorname{Cov}\bigl((X_i - X_{i+k})^2,\,(Y_i - Y_{i+k})^2\bigr)$

Write $X_i - X_{i+k} = (\theta_{x,i} - \theta_{x,i+k}) + (\varepsilon_{x,i} - \varepsilon_{x,i+k})$ and $Y_i - Y_{i+k} = (\theta_{y,i} - \theta_{y,i+k}) + (\varepsilon_{y,i} - \varepsilon_{y,i+k})$. Since the signal differences are constants,
$$
\begin{aligned}
&\operatorname{Cov}\bigl((X_i-X_{i+k})^2,\,(Y_i-Y_{i+k})^2\bigr) \\
&\quad= 4(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k})\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,\varepsilon_{y,i}-\varepsilon_{y,i+k}) \\
&\quad\phantom{=}\; + 2(\theta_{x,i}-\theta_{x,i+k})\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2) \\
&\quad\phantom{=}\; + 2(\theta_{y,i}-\theta_{y,i+k})\operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,\varepsilon_{y,i}-\varepsilon_{y,i+k}) \\
&\quad\phantom{=}\; + \operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2).
\end{aligned}
$$
The four pieces:

- $\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,\varepsilon_{y,i}-\varepsilon_{y,i+k}) = \sigma_{xy} + \sigma_{xy} = 2\sigma_{xy}$.

- $\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2) = 0$: since $E[\varepsilon_{x,i}-\varepsilon_{x,i+k}] = 0$, this equals $E[(\varepsilon_{x,i}-\varepsilon_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2]$. Swapping $i \leftrightarrow i+k$ negates $\varepsilon_{x,i}-\varepsilon_{x,i+k}$ but fixes $(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2$; since the pairs are iid the expectation is unchanged, so it equals its own negative and must be 0. By symmetry $\operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,\varepsilon_{y,i}-\varepsilon_{y,i+k}) = 0$.

- For $\operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2)$: expanding using independence between time $i$ and time $i+k$ (and mean-zero errors),
$$
E[(\varepsilon_{x,i}-\varepsilon_{x,i+k})^2(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2] = 2E[\varepsilon_{x}^2\varepsilon_{y}^2] + 2\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2 = 2\kappa_{22}\sigma_x^2\sigma_y^2 + 2\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2,
$$
while $E[(\varepsilon_{x,i}-\varepsilon_{x,i+k})^2]E[(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2] = 4\sigma_x^2\sigma_y^2$, so $\operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2) = 2(\kappa_{22}-1)\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2$.

Combining:
$$
\operatorname{Cov}\bigl((X_i-X_{i+k})^2,\,(Y_i-Y_{i+k})^2\bigr) = 8\sigma_{xy}(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k}) + 2(\kappa_{22}-1)\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2. \qquad\square
$$
:::

::: {.callout-note collapse="true"}
## Lemma: $E[T_k^{(2)}(X, Y)]$

**Exact expectation.** Applying $E[AB] = \operatorname{Cov}(A,B) + E[A]E[B]$ to each term and summing, using the lemma and $E\!\left[\sum_i(X_i-X_{i+k})^2\right] = T_k(\theta_x,\theta_x) + 2n\sigma_x^2$:
$$
\begin{aligned}
E[T_k^{(2)}(X, Y)]
&= 2n(\kappa_{22}-1)\sigma_x^2\sigma_y^2 + 8\sigma_{xy}T_k(\theta_x,\theta_y) + 4n\sigma_{xy}^2 \\
&\quad + T_k^{(2)}(\theta_x, \theta_y) + 2\sigma_y^2 T_k(\theta_x,\theta_x) + 2\sigma_x^2 T_k(\theta_y,\theta_y) + 4n\sigma_x^2\sigma_y^2.
\end{aligned}
$$

::: {.callout-tip collapse="true"}
## Simulation: Verifying $E[T_k^{(2)}(X, Y)]$

We simulate 1000 datasets from scenario 2 with $n = 500$ observations, $L = 4$ change points (frequent relative to $n$), exponential errors, and heterogeneous marginal variances ($\sigma_x^2 = 4$, $\sigma_y^2 = 0.25$). For each dataset we compute $T_2^{(2)}(X, Y)$ and compare its distribution to the formula value (red line).

```{r}
#| echo: false
#| cache: true
#| message: false
params <- scenario(2, L = 4, err_type = "exponential", sigma = c(2, 0.5))
k   <- 2
n   <- params$n
hx  <- params$h[, 1]; hy <- params$h[, 2]
k22 <- params$kappa$k22
sx  <- sqrt(params$S[1, 1]); sy <- sqrt(params$S[2, 2]); sxy <- params$S[1, 2]

target <- 2 * n * (k22 - 1) * sx^2 * sy^2 +
  8 * sxy * lag_diff(hx, hy, k = k) + 4 * n * sxy^2 +
  lag_diff2(hx, hy, k = k) +
  2 * sy^2 * lag_diff(hx, hx, k = k) +
  2 * sx^2 * lag_diff(hy, hy, k = k) +
  4 * n * sx^2 * sy^2

lag2 <- map_dbl(1:1000, ~ {
  X <- generate_data(params)
  lag_diff2(X[, 1], X[, 2], k = k)
})

```
```{R}
#| echo: false
tibble::tibble(val = lag2) |>
  ggplot(aes(x = val)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = target, color = "firebrick") +
  geom_vline(xintercept = mean(lag2), linetype = "dashed") +
  labs(x = expression(T[k]^{(2)}(X, Y)), y = "density",
       subtitle = "Red = formula, dashed = sample mean") +
  theme_minimal()
```
:::

:::

::: {.callout-warning}
## Remark
As with $2T_1(\theta_x,\theta_y) - T_2(\theta_x,\theta_y) = 0$, we have $2T_1^{(2)}(\theta_x,\theta_y) - T_2^{(2)}(\theta_x,\theta_y) = 0$.
:::

**Linear combination.** Applying the exact expectation at $k=1$ and $k=2$ and forming $2E[T_1^{(2)}(X,Y)] - E[T_2^{(2)}(X,Y)]$, all $\theta$-dependent terms vanish by the remark and the standard property applied to $(X,X)$ and $(Y,Y)$:
$$
2E[T_1^{(2)}(X,Y)] - E[T_2^{(2)}(X,Y)] = 
\boxed{2n(\kappa_{22}+1)\sigma_x^2\sigma_y^2 + 4n\sigma_{xy}^2.}
$$

Solving for $\kappa_{22}$:
$$
\kappa_{22} = \frac{2E[T_1^{(2)}(X,Y)] - E[T_2^{(2)}(X,Y)]}{2n\sigma_x^2\sigma_y^2} - \frac{2\sigma_{xy}^2}{\sigma_x^2\sigma_y^2} - 1.
$$

Plugging in estimates:
$$
\boxed{
\hat\kappa_{22} = \frac{2T_1^{(2)}(X,Y) - T_2^{(2)}(X,Y)}{2n\hat\sigma_x^2\hat\sigma_y^2} - \frac{2\hat\sigma_{xy}^2}{\hat\sigma_x^2\hat\sigma_y^2} - 1.
}
$$

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{22}$ (Lag Product)

We simulate 10,000 datasets from scenario 2 with $L = 4$ change points, exponential errors, and heterogeneous marginal variances ($\sigma_x^2 = 4$, $\sigma_y^2 = 0.25$). The sample mean of $\hat\kappa_{22}$ (dashed) should fall on the true value (red).

```{r}
#| echo: false
#| cache: true
#| message: false
params <- scenario(2, L = 4, err_type = "exponential", sigma = c(2, 0.5))
k22    <- params$kappa$k22

k22_est <- function(x, y) {
  n   <- length(x)
  sx2 <- ece.cov(x, x)
  sy2 <- ece.cov(y, y)
  sxy <- ece.cov(x, y)
  (2 * lag_diff2(x, y, k = 1) - lag_diff2(x, y, k = 2)) / (2 * n * sx2 * sy2) -
    2 * sxy^2 / (sx2 * sy2) - 1
}

k22_sim <- map_dbl(1:10000, ~ {
  X <- generate_data(params)
  k22_est(X[, 1], X[, 2])
})
```
```{r}
#| echo: false
tibble::tibble(val = k22_sim) |>
  ggplot(aes(x = val)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = k22, color = "firebrick") +
  geom_vline(xintercept = mean(k22_sim), linetype = "dashed") +
  labs(x = expression(hat(kappa)[22]), y = "density",
       subtitle = "Red = true, dashed = sample mean") +
  theme_minimal()
```

:::
