---
title: "Independence Test"
---

## Classical Theory

For jointly Gaussian $(X, Y)$, zero correlation and independence are equivalent, so testing $H_0: \rho = 0$ is also a test of $H_0: X \perp Y$. The statistic $T = \sqrt{n-2}\,\hat\rho / \sqrt{1 - \hat\rho^2}$ follows an exact $t_{n-2}$ null distribution.

Without the Gaussian assumption, $\rho = 0$ no longer implies independence — for example, $X \sim N(0,1)$ and $Y = X^2$ satisfy $\operatorname{Cov}(X, Y) = 0$ but $X \not\perp Y$. Two approaches handle the non-Gaussian case:

- **Permutation test.** Permuting $Y$ breaks any dependence on $X$ while preserving marginals, giving an exact test of $H_0: X \perp Y$ under any distribution.
- **Asymptotic test of independence.** Under $H_0: X \perp Y$, the errors are independent, so $\kappa_{22} = 1$ and $\sqrt{n}\,\hat\rho \xrightarrow{d} N(0, 1)$.

For the weaker null $H_0: \rho = 0$ without assuming independence, $\kappa_{22}$ need not equal 1, and the asymptotic variance is
$$
v^2 = \frac{1}{n} \left(\omega_x^2 + \omega_y^2 + \kappa_{22} + \frac{5}{2}\right).
$$
A robust test plugs in $\hat\kappa_{22}$.

## Estimating $\kappa_{22}$ via $V$

Let $s_i^{(k)} = (X_i - X_{i+k})(Y_i - Y_{i+k})$ and $S_i = \frac{s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)}}{2}$ as in the Multiplier Bootstrap chapter, so that
$$
\bar{S} = \frac{1}{n}\sum_{i=1}^n S_i = \frac{2T_1(X,Y) - T_2(X,Y)}{2n} = \hat\sigma_{xy}.
$$ Define the lag-2 variance estimate
$$
V := \widehat{\operatorname{Var}}(\bar{S}) = \frac{1}{n^2}\!\left(\sum_{i=1}^n S_i^2 + \sum_{i=1}^n (S_i S_{i+1} + S_{i+1} S_i) + \sum_{i=1}^n (S_i S_{i+2} + S_{i+2} S_i)\right),
$$
or in matrix form, writing $\mathbf{S} = (S_1, \ldots, S_n)^\top$,
$$
V = \frac{1}{n^2}\,\mathbf{S}^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\mathbf{S}.
$$

Since $\bar S = (2T_1 - T_2)/(2n)$, we have $V \approx \operatorname{Var}(2T_1 - T_2)/(4n^2)$. The variance on the right has a general closed form; under $H_0$ ($\sigma_{xy} = 0$) it simplifies so we can solve for $\kappa_{22}$ from $V$.

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Var}(2T_1 - T_2)$ from $\Sigma_Z$

The variance of a linear combination of correlated statistics is
$$
\operatorname{Var}(2T_1 - T_2) = 4[\Sigma_Z]_{55} - 4[\Sigma_Z]_{56} + [\Sigma_Z]_{66}.
$$
The general $\Sigma_Z$ entries from the Z-Test chapter are
$$[\Sigma_Z]_{55} = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) - 2n\sigma_{xy}^2 + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2 + 4n\sigma_{xy}\omega_{xy},$$
$$[\Sigma_Z]_{56} = 4n(\sigma_x^2\sigma_y^2\kappa_{22} - \sigma_{xy}^2) + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2 + 4n\sigma_{xy}\omega_{xy},$$
$$[\Sigma_Z]_{66} = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) - 2n\sigma_{xy}^2 + 4n\sigma_y^2\omega_x^2 + 4n\sigma_x^2\omega_y^2 + 8n\sigma_{xy}\omega_{xy}.$$
Substituting and collecting terms gives the general expression
$$\operatorname{Var}(2T_1 - T_2) = 2n\bigl[(2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 3\sigma_{xy}^2 + 4\sigma_{xy}\omega_{xy} + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2\bigr].$$
Under $H_0$ ($\sigma_{xy} = 0$) the last two terms vanish, yielding
$$\operatorname{Var}(2T_1 - T_2) = 2n\bigl[(2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2\bigr]. \qquad\square$$
:::

Since $4n^2 V \approx \operatorname{Var}(2T_1-T_2)$, substituting the general expression and solving gives
$$\hat\kappa_{22} = \frac{nV}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{3}{2}\frac{\hat\sigma_{xy}^2}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{2\hat\sigma_{xy}\hat\omega_{xy}}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{\hat\omega_x^2}{\hat\sigma_x^2} - \frac{\hat\omega_y^2}{\hat\sigma_y^2} - \frac{5}{2}.$$
Under $H_0$ ($\hat\sigma_{xy} = 0$) the middle two terms vanish, yielding
$$\hat\kappa_{22} = \frac{nV}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{\hat\omega_x^2}{\hat\sigma_x^2} - \frac{\hat\omega_y^2}{\hat\sigma_y^2} - \frac{5}{2}.$$

## Estimating $\kappa_{40}$ via $V_{xx}$

Applying the same construction to $(X, X)$ in place of $(X, Y)$: let $s_i^{(k)} = (X_i - X_{i+k})^2$ and $S_i^{xx} = (s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)})/2$, so that $\bar S^{xx} = (2T_1(X,X) - T_2(X,X))/(2n) = \hat\sigma_x^2$. Define
$$
V_{xx} := \widehat{\operatorname{Var}}(\bar S^{xx}) = \frac{1}{n^2}\,(\tilde{\mathbf{S}}^{xx})^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\tilde{\mathbf{S}}^{xx}, \qquad \tilde S_i^{xx} = S_i^{xx} - \bar S^{xx}.
$$

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Var}(2T_1(X,X) - T_2(X,X))$ from $\Sigma_Z$

$$
\operatorname{Var}(2T_1 - T_2) = 4[\Sigma_Z]_{11} - 4[\Sigma_Z]_{12} + [\Sigma_Z]_{22}.
$$
Substituting the entries from the Z-Test chapter:
$$[\Sigma_Z]_{11} = 4n\kappa_{40}\sigma_x^4 + 8n\sigma_x^2\omega_x^2,$$
$$[\Sigma_Z]_{12} = 4n(\kappa_{40}-1)\sigma_x^4 + 8n\sigma_x^2\omega_x^2,$$
$$[\Sigma_Z]_{22} = 4n\kappa_{40}\sigma_x^4 + 16n\sigma_x^2\omega_x^2.$$
Collecting terms:
$$\operatorname{Var}(2T_1(X,X) - T_2(X,X)) = 4n(\kappa_{40}+4)\sigma_x^4 + 16n\sigma_x^2\omega_x^2. \qquad\square$$
:::

Since $4n^2 V_{xx} \approx \operatorname{Var}(2T_1(X,X) - T_2(X,X))$, substituting and solving gives
$$
\hat\kappa_{40} = \frac{nV_{xx}}{\hat\sigma_x^4} - \frac{4\hat\omega_x^2}{\hat\sigma_x^2} - 4.
$$
By symmetry ($x \leftrightarrow y$), the same construction applied to $(Y, Y)$ yields $\hat\kappa_{04} = nV_{yy}/\hat\sigma_y^4 - 4\hat\omega_y^2/\hat\sigma_y^2 - 4$.

### Simulation: Does $\hat\kappa_{40}$ Recover the Truth?

```{r}
#| message: false
#| echo: false
devtools::load_all("..")
library(purrr)
library(ggplot2)

estimate_k40 <- function(X, params) {
  n    <- params$n
  Sxx  <- ece_terms(cbind(X[,1], X[,1]))
  Sxx  <- Sxx - mean(Sxx)
  Vxx  <- (1 / n^2) * as.numeric(
    t(Sxx) %*% Sxx +
      t(rotate(Sxx)) %*% Sxx + t(Sxx) %*% rotate(Sxx) +
      t(rotate(Sxx, 2)) %*% Sxx + t(Sxx) %*% rotate(Sxx, 2)
  )
  sx2 <- ece.cov(X[,1])
  wx2 <- ece.complexity(X[,1])
  n * Vxx / sx2^2 - 4 * wx2 / sx2 - 4
}
```

```{r}
#| cache: true
#| echo: false
params     <- scenario(2, L = 4, err_type = "normal")
params_exp <- scenario(2, L = 4, err_type = "exponential")


k40_gaussian <- map_dbl(1:1000, \(.) estimate_k40(generate_data(params), params))
k40_exp      <- map_dbl(1:1000, \(.) estimate_k40(generate_data(params_exp), params_exp))

lambda_vals  <- c(0, 0.5, 1/sqrt(2), 1)
k40_true     <- 12 * lambda_vals^4 + 3

sim_results_k40 <- map(lambda_vals, \(lam) {
  map_dbl(1:1000, \(.) {
    z  <- rnorm(params$n)
    ex <- lam * (z^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params$n)
    ey <- rnorm(params$n)
    X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
    estimate_k40(X, params)
  })
})
```

```{r}
#| echo: false
make_k40_plot <- function(sim, true_k40) {
  tibble::tibble(k40 = sim) |>
    ggplot(aes(x = k40)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
    geom_vline(xintercept = true_k40, color = "firebrick") +
    geom_vline(xintercept = mean(sim), linetype = "dashed") +
    labs(x = expression(hat(kappa)[40]), y = "density",
         subtitle = "Red = true, dashed = sample mean") +
    theme_minimal()
}
```

To check the formula for $\hat\kappa_{40}$, we simulate 1000 datasets and compare the distribution of $\hat\kappa_{40}$ to the true value. If the formula is correct, the sample mean (dashed line) should be close to the true value (red line). We consider three cases: Gaussian errors ($\kappa_{40} = 3$), exponential errors ($\kappa_{40} = 9$), and a one-parameter family interpolating between the two.

::: {.panel-tabset}

## Gaussian ($\kappa_{40} = 3$)

```{r}
#| echo: false
make_k40_plot(k40_gaussian, true_k40 = params$kappa$k40)
```

## Exponential ($\kappa_{40} = 9$)

```{r}
#| echo: false
make_k40_plot(k40_exp, true_k40 = params_exp$kappa$k40)
```

## $\lambda$-Mixing

Using $\varepsilon_x = \lambda(z^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,w$ with $z, w \stackrel{iid}{\sim} N(0,1)$, which gives $\kappa_{40} = 12\lambda^4 + 3$.

```{r}
#| echo: false
sim_df_k40 <- tibble::tibble(
  k40   = unlist(sim_results_k40),
  group = factor(rep(round(k40_true, 1), each = 1000),
                 labels = paste0("κ₄₀ = ", round(k40_true, 1)))
)
ref_df_k40 <- tibble::tibble(
  group = factor(round(k40_true, 1),
                 labels = paste0("κ₄₀ = ", round(k40_true, 1))),
  true  = k40_true
)
mean_df_k40 <- sim_df_k40 |>
  dplyr::summarise(mean_k40 = mean(k40), .by = group)

ggplot(sim_df_k40, aes(x = k40, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k40, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k40, aes(xintercept = mean_k40, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[40]), y = "density") +
  theme_minimal()
```

:::

## Estimating $\kappa_{31}$ via $C_{xx,xy}$

The cumulant $\kappa_{31}$ appears in the cross-covariances between the $(X,X)$ and $(X,Y)$ blocks of $\Sigma_Z$. Define the lag-2 cross-covariance
$$
C_{xx,xy} := \frac{1}{n^2}\,(\mathbf{S}^{xx})^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\mathbf{S}^{xy},
$$
which estimates $\operatorname{Cov}(\bar S^{xx}, \bar S^{xy}) = \operatorname{Cov}(2T_1(X,X) - T_2(X,X),\; 2T_1(X,Y) - T_2(X,Y))/(4n^2)$.

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Cov}(2T_1(X,X) - T_2(X,X),\; 2T_1(X,Y) - T_2(X,Y))$ from $\Sigma_Z$

$$
\operatorname{Cov}(2T_1(X,X) - T_2(X,X),\; 2T_1(X,Y) - T_2(X,Y)) = 4[\Sigma_Z]_{15} - 2[\Sigma_Z]_{16} - 2[\Sigma_Z]_{25} + [\Sigma_Z]_{26}.
$$
The relevant entries from the Z-Test chapter are
$$[\Sigma_Z]_{15} = 4n\sigma_x^3\sigma_y\kappa_{31} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{16} = 4n\sigma_x^3\sigma_y\kappa_{31} - 4n\sigma_x^2\sigma_{xy} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{25} = 4n\sigma_x^3\sigma_y\kappa_{31} - 4n\sigma_x^2\sigma_{xy} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{26} = 4n\sigma_x^3\sigma_y\kappa_{31} + 8n\sigma_{xy}\omega_x^2 + 8n\sigma_x^2\omega_{xy}.$$
Collecting terms:
$$\operatorname{Cov}(2T_1(X,X) - T_2(X,X),\; 2T_1(X,Y) - T_2(X,Y)) = 4n\sigma_x^3\sigma_y\kappa_{31} + 16n\sigma_x^2\sigma_{xy} + 8n\sigma_{xy}\omega_x^2 + 8n\sigma_x^2\omega_{xy}. \qquad\square$$
:::

Since $4n^2 C_{xx,xy} \approx \operatorname{Cov}(2T_1(X,X) - T_2(X,X),\; 2T_1(X,Y) - T_2(X,Y))$, substituting and solving gives
$$
\hat\kappa_{31} = \frac{nC_{xx,xy}}{\hat\sigma_x^3\hat\sigma_y} - \frac{4\hat\sigma_{xy}}{\hat\sigma_x\hat\sigma_y} - \frac{2\hat\sigma_{xy}\hat\omega_x^2}{\hat\sigma_x^3\hat\sigma_y} - \frac{2\hat\omega_{xy}}{\hat\sigma_x\hat\sigma_y}.
$$
By symmetry ($x \leftrightarrow y$), the same construction with $C_{yy,xy}$ yields $\hat\kappa_{13}$.

### Simulation: Does $\hat\kappa_{22}$ Recover the Truth?

```{r}
#| message: false
#| echo: false
devtools::load_all("..")
library(purrr)
library(ggplot2)
```

```{r}
#| cache: true
#| echo: false
params <- scenario(2, L = 4, err_type = "exponential")

estimate_k22 <- function(X, params) {
  n   <- params$n
  S   <- ece_terms(X)
  W   <- (1 / n^2) * (
    t(S) %*% S +
      t(rotate(S)) %*% S + t(S) %*% rotate(S) +
      t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
  )
  sx2 <- ece.cov(X[,1])
  sy2 <- ece.cov(X[,2])
  sxy <- ece.cov(X[,1], X[,2])
  wx2 <- ece.complexity(X[,1])
  wy2 <- ece.complexity(X[,2])
  wxy <- ece.complexity(X[,1], X[,2])
  n * as.numeric(W) / (sx2 * sy2) - (3/2) * sxy^2 / (sx2 * sy2) - 2 * sxy * wxy / (sx2 * sy2) - wx2 / sx2 - wy2 / sy2 - 5/2
}

k22_indep <- map_dbl(1:1000, \(.) {
  estimate_k22(generate_data(params), params)
})

k22_dep <- map_dbl(1:1000, \(.) {
  ex <- rnorm(params$n)
  ey <- (ex^2 - 1) / sqrt(2)
  X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
  estimate_k22(X, params)
})

lambda_vals <- c(0, 0.5, 1/sqrt(2), 1)
k22_true    <- 1 + 4 * lambda_vals^2

sim_results <- map(lambda_vals, \(lam) {
  map_dbl(1:1000, \(.) {
    ex <- rnorm(params$n)
    ey <- lam * (ex^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params$n)
    X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
    estimate_k22(X, params)
  })
})
```

```{r}
#| echo: false
make_k22_plot <- function(sim, true_k22) {
  tibble::tibble(k22 = sim) |>
    ggplot(aes(x = k22)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
    geom_vline(xintercept = true_k22, color = "firebrick") +
    geom_vline(xintercept = mean(sim), linetype = "dashed") +
    labs(x = expression(hat(kappa)[22]), y = "density",
         subtitle = "Red = true, dashed = sample mean") +
    theme_minimal()
}
```

To check the formula for $\hat\kappa_{22}$, we simulate 1000 datasets and compare the distribution of $\hat\kappa_{22}$ to the true value. If the formula is correct, the sample mean (dashed line) should be close to the true value (red line). We consider three cases: independent errors ($\kappa_{22} = 1$), a strongly dependent case ($\kappa_{22} = 5$), and a one-parameter family interpolating between the two.

::: {.panel-tabset}

## Independent ($\kappa_{22} = 1$)

```{r}
#| echo: false
make_k22_plot(k22_indep, true_k22 = params$kappa$k22)
```

## Dependent ($\kappa_{22} = 5$)

```{r}
#| echo: false
make_k22_plot(k22_dep, true_k22 = 5)
```

## $\lambda$-Mixing

Using $\varepsilon_y = \lambda(\varepsilon_x^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,\eta$ with $\varepsilon_x, \eta \stackrel{iid}{\sim} N(0,1)$, which gives $\sigma_{xy} = 0$ and $\kappa_{22} = 1 + 4\lambda^2$.

```{r}
#| echo: false
sim_df <- tibble::tibble(
  k22   = unlist(sim_results),
  group = factor(rep(round(k22_true, 1), each = 1000),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1)))
)
ref_df <- tibble::tibble(
  group = factor(round(k22_true, 1),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1))),
  true  = k22_true
)

mean_df <- sim_df |>
  dplyr::summarise(mean_k22 = mean(k22), .by = group)

ggplot(sim_df, aes(x = k22, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df, aes(xintercept = mean_k22, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[22]), y = "density") +
  theme_minimal()
```

:::



