---
title: "Independence Test"
---

# Independence Test

The asymptotic variance of $\hat{\rho}$ under the null ($\rho = 0$) is
$$
v^2 = \frac{1}{n} \left(\omega_x^2 + \omega_y^2 + \kappa_{22} + \frac{5}{2}\right). 
$$
If we assume $\varepsilon_x$ and $\varepsilon_y$ are independent, then $\kappa_{22} = 1$. 

```{R} 
pearson_permutation_test <- function(x, y, B = 10000, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  n <- length(x)
  r_obs <- cor(x, y)
  
  r_perm <- numeric(B)
  for (b in seq_len(B)) {
    r_perm[b] <- cor(x, sample(y, size = n, replace = FALSE))
  }
  
  pval <- (1 + sum(abs(r_perm) >= abs(r_obs))) / (B + 1)
  
  list(r = r_obs, p.value = pval, B = B)
}

pearson_robust_zero_corr_test <- function(x, y) {
  n <- length(x)
  
  xc <- x - mean(x)
  yc <- y - mean(y)
  
  sx2 <- mean(xc^2)
  sy2 <- mean(yc^2)
  
  r <- mean(xc * yc) / sqrt(sx2 * sy2)
  
  kappa22_hat <- mean(xc^2 * yc^2) / (sx2 * sy2)
  
  Z <- sqrt(n) * r / sqrt(kappa22_hat)
  pval <- 2 * pnorm(-abs(Z))
  
  list(
    r = r,
    kappa22_hat = kappa22_hat,
    statistic = Z,
    p.value = pval
  )
}
```


| Setting                  | Null hypothesis                    | Finite-sample calibration                  | Asymptotic calibration                  | p-value                                   |   |    |
| ------------------------ | ---------------------------------- | ------------------------------------------ | --------------------------------------- | ----------------------------------------- | - | -- |
| Gaussian                 | (\rho=0), equivalent to (X\perp Y) | Exact (t_{n-2})                            | (\sqrt n r\Rightarrow N(0,1))           | (2P(t_{n-2}\ge                            | T | )) |
| Independent non-Gaussian | (X\perp Y)                         | Permutation exact; classical (t) not exact | (\sqrt n r\Rightarrow N(0,1))           | permutation or normal approximation       |   |    |
| Merely uncorrelated      | (\rho=0), dependence allowed       | No distribution-free exact law             | (\sqrt n r\Rightarrow N(0,\kappa_{22})) | robust normal using (\widehat\kappa_{22}) |   |    |

Let $X \sim N(0, 1)$ and $Y = X^2$. Then, $Cov(X,Y) = E(X^3) - E(X) E(X^2) = 0$. That is, $X$ and $Y$ have zero correlation, but they are clearly no independent.

## Estimating $\kappa_{22}$ via $W$

Since `lag_terms(X) |> mean() / 2 == ece.cov(X[,1], X[,2])`, the sample mean of the $S_i$ terms satisfies

$$\bar S = \frac{2T_1(X,Y) - T_2(X,Y)}{n},$$

so the parametric bootstrap variance $W = \widehat{\operatorname{Var}}(\bar S)$ estimates $\operatorname{Var}((2T_1-T_2)/n)$. This variance has a closed form containing $\kappa_{22}$, so we can solve for $\kappa_{22}$ directly from $W$.

Under $H_0$ ($\sigma_{xy} = 0$), applying $\operatorname{Var}(2Z_5 - Z_6) = 4[\Sigma_Z]_{55} - 4[\Sigma_Z]_{56} + [\Sigma_Z]_{66}$ to the entries of $\Sigma_Z$ gives

$$\operatorname{Var}(2T_1 - T_2) = 2n\!\left[(2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2\right].$$

Since $W = \operatorname{Var}(2T_1 - T_2)/n^2$, solving for $\kappa_{22}$:

$$\hat\kappa_{22} = \frac{nW}{4\hat\sigma_x^2\hat\sigma_y^2} - \frac{\hat\omega_x^2}{\hat\sigma_x^2} - \frac{\hat\omega_y^2}{\hat\sigma_y^2} - \frac{5}{2},$$

where $\hat\omega_x^2 = (T_2(X,X) - T_1(X,X))/n$ and $\hat\omega_y^2 = (T_2(Y,Y) - T_1(Y,Y))/n$.

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Var}(2T_1 - T_2)$ from $\Sigma_Z$

From the $\Sigma_Z$ table, under $H_0$ ($\sigma_{xy} = \omega_{xy} = 0$):

$$[\Sigma_Z]_{55} = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2,$$
$$[\Sigma_Z]_{56} = 4n\sigma_x^2\sigma_y^2\kappa_{22} + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2,$$
$$[\Sigma_Z]_{66} = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) + 4n\sigma_y^2\omega_x^2 + 4n\sigma_x^2\omega_y^2.$$

Substituting into $4[\Sigma_Z]_{55} - 4[\Sigma_Z]_{56} + [\Sigma_Z]_{66}$:

- **$\sigma_x^2\sigma_y^2$ coefficient:** $4\cdot 2(2\kappa_{22}+1) - 4\cdot 4\kappa_{22} + 2(2\kappa_{22}+1) = 4\kappa_{22} + 10 = 2(2\kappa_{22}+5)$.
- **$\sigma_y^2\omega_x^2$ coefficient:** $4\cdot 2 - 4\cdot 2 + 4 = 4$.
- **$\sigma_x^2\omega_y^2$ coefficient:** $4\cdot 2 - 4\cdot 2 + 4 = 4$.

Therefore $\operatorname{Var}(2T_1 - T_2) = 2n[(2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2]$.

Since $W = \operatorname{Var}(2T_1-T_2)/n^2$, we have $nW/2 = (2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2$. Solving:

$$\kappa_{22} = \frac{nW/2 - 2\sigma_y^2\omega_x^2 - 2\sigma_x^2\omega_y^2}{2\sigma_x^2\sigma_y^2} - \frac{5}{2} \cdot \frac{1}{1} = \frac{nW}{4\sigma_x^2\sigma_y^2} - \frac{\omega_x^2}{\sigma_x^2} - \frac{\omega_y^2}{\sigma_y^2} - \frac{5}{2}. \quad\square$$
:::

In code, using the parametric bootstrap $W$ already computed:

```{r}
#| eval: false
wx2 <- (lag_diff(X[,1], X[,1], k=2) - lag_diff(X[,1], X[,1], k=1)) / n
wy2 <- (lag_diff(X[,2], X[,2], k=2) - lag_diff(X[,2], X[,2], k=1)) / n

k22 <- n * as.numeric(W) / (4 * sx^2 * sy^2) - wx2 / sx^2 - wy2 / sy^2 - 5/2
```

### Simulation: Does $\hat\kappa_{22}$ Recover the Truth?

We run two scenarios with exponential mean structure ($L = 4$):

- **Independent:** errors drawn from $\text{Exp}(1) - 1$, independent across series. True $\kappa_{22} = E[\varepsilon_x^2]E[\varepsilon_y^2]/(\sigma_x^2\sigma_y^2) = 1$.
- **Dependent:** $\varepsilon_x \sim N(0,1)$, $\varepsilon_y = (\varepsilon_x^2 - 1)/\sqrt{2}$. Zero correlation but strong dependence. True $\kappa_{22} = E[\varepsilon_x^2 \varepsilon_y^2] = \tfrac{1}{2}E[\varepsilon_x^2(\varepsilon_x^2-1)^2] = 5$.

```{r}
#| cache: true
#| message: false
devtools::load_all("..")
library(purrr)

params <- scenario(2, L = 4, err_type = "exponential")

run_k22_sim <- function(data_gen, n_sim = 1000) {
  map_dbl(seq_len(n_sim), \(.) {
    n   <- params$n
    X   <- data_gen(params)
    S   <- lag_terms(X)
    W   <- (1 / n^2) * (
      t(S) %*% S +
        t(rotate(S)) %*% S + t(S) %*% rotate(S) +
        t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
    )
    sx  <- sqrt(ece_pair(X[,1], X[,1])$cov)
    sy  <- sqrt(ece_pair(X[,2], X[,2])$cov)
    wx2 <- ece.complexity(X[,1])
    wy2 <- ece.complexity(X[,2])
    n * as.numeric(W) / (4 * sx^2 * sy^2) - wx2 / sx^2 - wy2 / sy^2 - 5/2
  }, .progress = TRUE)
}

gen_independent <- function(params) generate_data(params)

gen_dependent <- function(params) {
  ex <- rnorm(params$n)
  ey <- (ex^2 - 1) / sqrt(2)
  cbind(params$h[,1] + ex, params$h[,2] + ey)
}

k22_indep <- run_k22_sim(gen_independent)
k22_dep   <- run_k22_sim(gen_dependent)
```

```{r}
# True k22 for dependent case: (1/2) E[ex^2 (ex^2 - 1)^2] = (1/2)(15 - 6 + 1) = 5
ex_large <- rnorm(1e6)
ey_large <- (ex_large^2 - 1) / sqrt(2)
k22_dep_true <- mean(ex_large^2 * ey_large^2)

tibble::tribble(
  ~scenario,    ~true_k22,             ~estimated_k22,
  "independent", params$kappa$k22,     mean(k22_indep),
  "dependent",   k22_dep_true,         mean(k22_dep)
)
```



