---
title: "Independence Test"
---

## Classical Theory

For jointly Gaussian $(X, Y)$, zero correlation and independence are equivalent, so testing $H_0: \rho = 0$ is also a test of $H_0: X \perp Y$. The statistic $T = \sqrt{n-2}\,\hat\rho / \sqrt{1 - \hat\rho^2}$ follows an exact $t_{n-2}$ null distribution.

Without the Gaussian assumption, $\rho = 0$ no longer implies independence — for example, $X \sim N(0,1)$ and $Y = X^2$ satisfy $\operatorname{Cov}(X, Y) = 0$ but $X \not\perp Y$. Two approaches handle the non-Gaussian case:

- **Permutation test.** Permuting $Y$ breaks any dependence on $X$ while preserving marginals, giving an exact test of $H_0: X \perp Y$ under any distribution.
- **Asymptotic test of independence.** Under $H_0: X \perp Y$, the errors are independent, so $\kappa_{22} = 1$ and $\sqrt{n}\,\hat\rho \xrightarrow{d} N(0, 1)$.

For the weaker null $H_0: \rho = 0$ without assuming independence, $\kappa_{22}$ need not equal 1, and the asymptotic variance is
$$
v^2 = \frac{1}{n} \left(\omega_x^2 + \omega_y^2 + \kappa_{22} + \frac{5}{2}\right).
$$
A robust test plugs in $\hat\kappa_{22}$.

{{< include kappa22-v.qmd >}}

{{< include kappa22-moments.qmd >}}

---

## Method of Moments via $S_i$ Terms

Recall from the multiplier bootstrap that
$$S_i = \frac{s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)}}{2}, \qquad s_i^{(k)} = (X_i - X_{i+k})(Y_i - Y_{i+k}),$$
satisfies $E[S_i] = \sigma_{xy}$ regardless of change-point amplitudes. The same construction applied to each series separately gives
$$S_{xx,i} = \frac{(X_i-X_{i+1})^2 + (X_{i+1}-X_{i+2})^2 - (X_i-X_{i+2})^2}{2}, \qquad S_{yy,i} = \frac{(Y_i-Y_{i+1})^2 + (Y_{i+1}-Y_{i+2})^2 - (Y_i-Y_{i+2})^2}{2},$$
with $E[S_{xx,i}] = \sigma_x^2$ and $E[S_{yy,i}] = \sigma_y^2$ at every index, change points included.

These terms play the role of $\varepsilon_{x,i}^2$ and $\varepsilon_{y,i}^2$ with change-point-robust constant means. The classical method-of-moments estimator for $\kappa_{22} = E[\varepsilon_x^2\varepsilon_y^2]/(\sigma_x^2\sigma_y^2)$ is
$$\hat\kappa_{22} = \frac{\frac{1}{n}\sum_i \varepsilon_{x,i}^2\,\varepsilon_{y,i}^2}{\hat\sigma_x^2\,\hat\sigma_y^2},$$
where the numerator estimates $E[\varepsilon_x^2\varepsilon_y^2]$ and the denominator estimates $\sigma_x^2\sigma_y^2$. Replacing $\varepsilon_{x,i}^2$ with $S_{xx,i}$ and $\varepsilon_{y,i}^2$ with $S_{yy,i}$ gives the ECE analog:
$$\hat\kappa_{22}^{\mathrm{MoM}} = \frac{\frac{1}{n}\sum_i S_{xx,i}\,S_{yy,i}}{\bar S_{xx}\,\bar S_{yy}}.$$

The $S_i$ terms are autocorrelated at lags 1 and 2 (consecutive terms share observations), so $\frac{1}{n}\sum S_{xx,i}S_{yy,i}$ has inflated variance relative to an i.i.d. average. The estimator remains consistent.

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{22}^{\mathrm{MoM}}$

We simulate 5,000 datasets across four scenarios varying the error distribution and change-point structure. The sample mean of $\hat\kappa_{22}^{\mathrm{MoM}}$ (dashed) should fall on the true value (red) in each panel.

```{r}
#| echo: false
#| cache: true
#| message: false
k22_mom <- function(x, y) {
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  mean(Sxx * Syy) / (mean(Sxx) * mean(Syy))
}

scens <- list(
  list(params = scenario(1, err_type = "normal"),                            label = "Sc. 1, normal"),
  list(params = scenario(1, err_type = "exponential"),                       label = "Sc. 1, exponential"),
  list(params = scenario(2, L = 4, err_type = "exponential"),                label = "Sc. 2, exponential"),
  list(params = scenario(2, L = 4, err_type = "exponential", sigma = c(2, 0.5)), label = "Sc. 2, exp., het. var.")
)

mom_verify <- map(scens, \(s) {
  sim <- map_dbl(1:5000, ~ {
    X <- generate_data(s$params)
    k22_mom(X[, 1], X[, 2])
  })
  tibble(val = sim, truth = s$params$kappa$k22, label = s$label)
}) |> list_rbind()
```

```{r}
#| echo: false
mom_verify |>
  ggplot(aes(x = val)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = truth), color = "firebrick") +
  geom_vline(aes(xintercept = mean(val)), linetype = "dashed") +
  facet_wrap(~label, scales = "free") +
  labs(x = expression(hat(kappa)[22]^MoM), y = "density",
       subtitle = "Red = true, dashed = sample mean") +
  theme_minimal()
```

:::

---

## Numerical Results: Estimating $\kappa_{22}$

No single estimator dominates: the regression estimator has lower variance in most scenarios, but the matrix estimator wins when change-point amplitude is large.

```{R}
#| echo: false

# Compare k22 Estimators
ece.k22.regress <- function(x, y){
  n <- length(x) # WARN: be careful if x is a matrix
  sx2 <- ece.cov(x, x)
  sy2 <- ece.cov(y, y)
  sxy <- ece.cov(x, y)
  (2 * lag_diff2(x, y, k=1) - lag_diff2(x, y, k=2)) / (2 * n * sx2 * sy2) - 
    2 * sxy^2 / (sx2 * sy2) - 1
}

ece.k22.matrix <- function(x, y = NULL, rho = NULL) {
  if (is.matrix(x)) {
    if (ncol(x) != 2) stop("matrix input must have exactly 2 columns")
    y <- x[, 2]
    x <- x[, 1]
  }

  n <- length(x)
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  Sxy <- ece_terms(cbind(x, y))

  sx2 <- mean(Sxx)
  sy2 <- mean(Syy)
  sxy <- if (!is.null(rho) && rho == 0) 0 else mean(Sxy)

  S <- cbind(Sxx - sx2, Syy - sy2, Sxy - sxy)
  V <- (1 / n^2) * (
    t(S) %*% S +
      t(rotate(S)) %*% S + t(S) %*% rotate(S) +
      t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
  )

  wx2 <- ece.complexity(x, x)
  wy2 <- ece.complexity(y, y)
  wxy <- ece.complexity(x, y)

  k22 <- n * V[3, 3] / (sx2 * sy2) -
    3 / 2 * sxy^2 / (sx2 * sy2) -
    2 * sxy * wxy / (sx2 * sy2) -
    wx2 / sx2 -
    wy2 / sy2 - 5 / 2
  return(k22)
}

ece.k22.mom <- function(x, y) {
  Sxx <- ece_terms(cbind(x, x))
  Syy <- ece_terms(cbind(y, y))
  mean(Sxx * Syy) / (mean(Sxx) * mean(Syy))
}
```

```{R}
#| cache: true
#| echo: false

# Simulation
simulate_k22 <- function(params){
  map(1:1000, ~{
    X <- generate_data(params) 
    x <- X[,1]; y <- X[,2]
    list(
      regress = ece.k22.regress(x, y),
      matrix  = ece.k22.matrix(x, y),
      mom     = ece.k22.mom(x, y)
    )
  })
}

```

```{R}
#| cache: true
#| echo: false
param_grid <- tribble(
  ~label,                        ~scen, ~r,   ~err_type,      ~amp, ~L,
  "Scenario 1, Normal",           1,     0,    "normal",       1,    NA,
  "Scenario 1, Exponential",      1,     0,    "exponential",  1,    NA,
  "Scenario 1 (r = 0.5), Exp.",  1,     0.5,  "exponential",  1,    NA,
  "Scenario 2, Normal",           2,     0,    "normal",       1,    4,
  "Scenario 2, Exponential",      2,     0,    "exponential",  1,    4,
  "Scenario 2 (amp = 2), Exp.",  2,     0,    "exponential",  2,    4
)

k22_grid <- pmap(param_grid, function(label, scen, r, err_type, amp, L) {
  params <- if (is.na(L)) {
    scenario(scen, r = r, err_type = err_type, amp = amp)
  } else {
    scenario(scen, r = r, err_type = err_type, amp = amp, L = L)
  }
  sim    <- simulate_k22(params)
  list(
    label       = label,
    var_regress = var(map_dbl(sim, "regress")),
    var_matrix  = var(map_dbl(sim, "matrix")),
    var_mom     = var(map_dbl(sim, "mom")),
    sim         = sim
  )
})

k22_var_tbl <- map_dfr(k22_grid, \(g) tibble(
  Scenario               = g$label,
  `Var(k22-Regression)`  = round(g$var_regress, 2),
  `Var(k22-Covariance)`  = round(g$var_matrix,  2),
  `Var(k22-MoM)`         = round(g$var_mom,      2)
))

```

```{R}
#| echo: false
knitr::kable(k22_var_tbl)
```

::: {.panel-tabset}

## Regression Dominates

```{R}
#| echo: false
make_overlap_hist(
  sims = list(
    Regression = map_dbl(k22_grid[[1]]$sim, "regress"),
    Covariance = map_dbl(k22_grid[[1]]$sim, "matrix"),
    MoM        = map_dbl(k22_grid[[1]]$sim, "mom")
  ),
  xlab   = expression(hat(kappa)[22]),
  colors = c(Regression = "#E41A1C", Covariance = "#377EB8", MoM = "#4DAF4A")
) + ggplot2::labs(title = "Scenario 1, normal errors, n = 200, ρ = 0")
```

## Covariance Dominates

```{R}
#| echo: false
make_overlap_hist(
  sims = list(
    Regression = map_dbl(k22_grid[[4]]$sim, "regress"),
    Covariance = map_dbl(k22_grid[[4]]$sim, "matrix"),
    MoM        = map_dbl(k22_grid[[4]]$sim, "mom")
  ),
  xlab   = expression(hat(kappa)[22]),
  colors = c(Regression = "#E41A1C", Covariance = "#377EB8", MoM = "#4DAF4A")
) + ggplot2::labs(title = "Scenario 2, normal errors, n = 200, ρ = 0, L = 4")
```

:::


---

## Estimating $\kappa_{22}$ via $V$

Since $S_{xx}$, $S_{yy}$, $S_{xy}$ estimate $\sigma_x^2 = E[\varepsilon_x^2]$, $\sigma_y^2 = E[\varepsilon_y^2]$, $\sigma_{xy} = E[\varepsilon_x\varepsilon_y]$, the CLT for sample averages gives

$$
n\,\mathrm{Cov}(S_{xx}, S_{yy}) \;\to\; \mathrm{Cov}(\varepsilon_x^2,\, \varepsilon_y^2) = \sigma_x^2\sigma_y^2(\kappa_{22} - 1),
$$
$$
n\,\mathrm{Var}(S_{xy}) \;\to\; \mathrm{Var}(\varepsilon_x\varepsilon_y) = \sigma_x^2\sigma_y^2\kappa_{22} - \sigma_{xy}^2.
$$

Both $V_{12}$ and $V_{33}$ therefore contain $\kappa_{22}$. Their population counterparts differ by a known quantity:

$$
\mathrm{Var}(\varepsilon_x\varepsilon_y) - \mathrm{Cov}(\varepsilon_x^2,\varepsilon_y^2) = \sigma_x^2\sigma_y^2(1 - \rho^2),
$$

so the two estimators target the same $\kappa_{22}$ with a fixed, estimable offset — the natural setting for a GLS combination with weights determined by their asymptotic variances.

Under $H_0: \rho = 0$, $\sigma_{xy} = 0$ and the $\omega_{xy}$ corrections vanish from $V_{12}$, while $V_{33}$ retains $\omega_x^2/\sigma_x^2$ and $\omega_y^2/\sigma_y^2$ corrections. The $V_{12}$-based estimator is therefore cleaner under the independence null.

### Relationship between $\mathrm{Var}(S_{xy})$ and $\mathrm{Cov}(S_{xx}, S_{yy})$

The two entries of $V$ used to estimate $\kappa_{22}$ expand as

$$
n\,\mathrm{Var}(S_{xy}) = \sigma_x^2\sigma_y^2\kappa_{22} + \tfrac{5}{2}\sigma_x^2\sigma_y^2 + \tfrac{3}{2}\sigma_{xy}^2 + \sigma_y^2\omega_x^2 + \sigma_x^2\omega_y^2 + 2\sigma_{xy}\omega_{xy},
$$
$$
n\,\mathrm{Cov}(S_{xx}, S_{yy}) = \sigma_x^2\sigma_y^2(\kappa_{22}-1) + 2\sigma_{xy}^2 + 4\sigma_{xy}\omega_{xy}.
$$

Their difference is free of $\kappa_{22}$ at any $\rho$:

$$
n\bigl[\mathrm{Var}(S_{xy}) - \mathrm{Cov}(S_{xx}, S_{yy})\bigr] = \frac{7-\rho^2}{2}\,\sigma_x^2\sigma_y^2 + \sigma_y^2\omega_x^2 + \sigma_x^2\omega_y^2 - 2\sigma_{xy}\omega_{xy}.
$$

Dividing by $\sigma_x^2\sigma_y^2$, this gives a natural decomposition of $n\,\mathrm{Var}(S_{xy})/(\sigma_x^2\sigma_y^2)$:

$$
\underbrace{\frac{n\,\mathrm{Cov}(S_{xx},S_{yy})}{\sigma_x^2\sigma_y^2}}_{\kappa_{22} - 1 + 2\rho^2 + 4\rho\,\omega_{xy}/(\sigma_x\sigma_y)} + \underbrace{\frac{n\bigl[\mathrm{Var}(S_{xy}) - \mathrm{Cov}(S_{xx},S_{yy})\bigr]}{\sigma_x^2\sigma_y^2}}_{\frac{7-\rho^2}{2} + \omega_x^2/\sigma_x^2 + \omega_y^2/\sigma_y^2 - 2\rho\,\omega_{xy}/(\sigma_x\sigma_y)}.
$$

$\kappa_{22}$ is isolated in the first term at all $\rho$. The second term is the Gaussian-plus-change-point contribution, with its constant shifting from $7/2$ at $\rho = 0$ to $(7-\rho^2)/2$ in general. Under Gaussian errors ($\kappa_{22} = 1 + 2\rho^2$), $\mathrm{Cov}(S_{xx}, S_{yy})$ carries only the $\rho$-induced correlation and the entire excess over Gaussian is captured by its deviation from $2\rho^2\sigma_x^2\sigma_y^2/n$.

---

## Kappa Table
```{R} 
#| echo: false
#| cache: true

# Create Table
rho <- 0
sample_size <- 1000

set.seed(2)
params_norm <- scenario(1, n=sample_size, r=rho, err_type="normal")
params_exp <- scenario(1, n=sample_size, r=rho, err_type="exponential")
params_lognorm <- scenario(1, n=sample_size, r=rho, err_type="lognormal")
params_t <- scenario(1, n=sample_size, r=rho, err_type="t")
params <- list(params_norm, params_exp, params_lognorm, params_t)

# Create Table
dist <- c("Normal", "Exponential", "Log-Normal", "t (df = 5)")
kappa_tbl <- map_dfr(params, function(p) unlist(p$kappa)) |> round(4)
se <- map_dbl(params, ece.cor.se.formula) |> round(4)
tbl <- cbind(dist, kappa_tbl, se)
knitr::kable(tbl,
  col.names = c("Distribution", "$\\kappa_{40}$", "$\\kappa_{04}$",
                "$\\kappa_{22}$", "$\\kappa_{31}$", "$\\kappa_{13}$", "$\\operatorname{SE}(\\hat{\\rho})$"))
```



For $\rho = 0.5$, we get

```{R} 
#| echo: false
#| cache: true

# Create Table
rho <- 0.5
sample_size <- 1000

set.seed(2)
params_norm <- scenario(1, n=sample_size, r=rho, err_type="normal")
params_exp <- scenario(1, n=sample_size, r=rho, err_type="exponential")
params_lognorm <- scenario(1, n=sample_size, r=rho, err_type="lognormal")
params_t <- scenario(1, n=sample_size, r=rho, err_type="t")
params2 <- list(params_norm, params_exp, params_lognorm, params_t)

# Create Table
dist <- c("Normal", "Exponential", "Log-Normal", "t (df = 5)")
kappa_tbl <- map_dfr(params2, function(p) unlist(p$kappa)) |> round(4)
se <- map_dbl(params2, ece.cor.se.formula) |> round(4)
tbl <- cbind(dist, kappa_tbl, se)
knitr::kable(tbl,
  col.names = c("Distribution", "$\\kappa_{40}$", "$\\kappa_{04}$",
                "$\\kappa_{22}$", "$\\kappa_{31}$", "$\\kappa_{13}$", "$\\operatorname{SE}(\\hat{\\rho})$"))
```


```{R} 
#| echo: false

# Plot All Data
dist_labels <- c("Normal", "Exponential", "Log-Normal", "t")
X_lst <- params |>
  map(generate_data) |>
  map(function(X) c(X[, 1], X[, 2]))

map2_dfr(X_lst, dist_labels, ~tibble(value = .x, Distribution = .y)) |>
  mutate(Distribution = factor(Distribution, levels = dist_labels)) |>
  ggplot2::ggplot(ggplot2::aes(x = value, y = ggplot2::after_stat(density))) +
  ggplot2::geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
  ggplot2::facet_wrap(~Distribution, nrow = 2) +
  ggplot2::coord_cartesian(xlim = c(-4, 4)) +
  ggplot2::labs(x = NULL, y = "density") +
  ggplot2::theme_minimal()
```
