## Estimating $\kappa_{22}$ via Moments

Under the ECE model $X_i = \theta_{x,i} + \varepsilon_{x,i}$, $Y_i = \theta_{y,i} + \varepsilon_{y,i}$ with iid mean-zero errors, define $\kappa_{22} = E[\varepsilon_{x,i}^2\varepsilon_{y,i}^2]/(\sigma_x^2\sigma_y^2)$, which equals 1 under independence. Write $T_k(\theta_x, \theta_y) = \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\theta_{y,i} - \theta_{y,i+k})$ for the signal-difference inner product and
$$
\boxed{T_k^{(2)}(X, Y) = \sum_{i=1}^n (X_i - X_{i+k})^2(Y_i - Y_{i+k})^2}
$$
for its squared analog (applicable to any pair of sequences).

The strategy is to compute $E[T_k^{(2)}(X, Y)]$ exactly and solve for $\kappa_{22}$.

::: {.callout-note collapse="true"}
## Lemma: $\operatorname{Cov}\bigl((X_i - X_{i+k})^2,\,(Y_i - Y_{i+k})^2\bigr)$

Write $X_i - X_{i+k} = (\theta_{x,i} - \theta_{x,i+k}) + (\varepsilon_{x,i} - \varepsilon_{x,i+k})$ and $Y_i - Y_{i+k} = (\theta_{y,i} - \theta_{y,i+k}) + (\varepsilon_{y,i} - \varepsilon_{y,i+k})$. Since the signal differences are constants,
$$
\begin{aligned}
&\operatorname{Cov}\bigl((X_i-X_{i+k})^2,\,(Y_i-Y_{i+k})^2\bigr) \\
&\quad= 4(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k})\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,\varepsilon_{y,i}-\varepsilon_{y,i+k}) \\
&\quad\phantom{=}\; + 2(\theta_{x,i}-\theta_{x,i+k})\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2) \\
&\quad\phantom{=}\; + 2(\theta_{y,i}-\theta_{y,i+k})\operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,\varepsilon_{y,i}-\varepsilon_{y,i+k}) \\
&\quad\phantom{=}\; + \operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2).
\end{aligned}
$$
The four pieces:

- $\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,\varepsilon_{y,i}-\varepsilon_{y,i+k}) = \sigma_{xy} + \sigma_{xy} = 2\sigma_{xy}$.

- $\operatorname{Cov}(\varepsilon_{x,i}-\varepsilon_{x,i+k},\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2) = 0$: since $E[\varepsilon_{x,i}-\varepsilon_{x,i+k}] = 0$, this equals $E[(\varepsilon_{x,i}-\varepsilon_{x,i+k})(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2]$. Swapping $i \leftrightarrow i+k$ negates $\varepsilon_{x,i}-\varepsilon_{x,i+k}$ but fixes $(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2$; since the pairs are iid the expectation is unchanged, so it equals its own negative and must be 0. By symmetry $\operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,\varepsilon_{y,i}-\varepsilon_{y,i+k}) = 0$.

- For $\operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2)$: expanding using independence between time $i$ and time $i+k$ (and mean-zero errors),
$$
E[(\varepsilon_{x,i}-\varepsilon_{x,i+k})^2(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2] = 2E[\varepsilon_{x}^2\varepsilon_{y}^2] + 2\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2 = 2\kappa_{22}\sigma_x^2\sigma_y^2 + 2\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2,
$$
while $E[(\varepsilon_{x,i}-\varepsilon_{x,i+k})^2]E[(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2] = 4\sigma_x^2\sigma_y^2$, so $\operatorname{Cov}((\varepsilon_{x,i}-\varepsilon_{x,i+k})^2,\,(\varepsilon_{y,i}-\varepsilon_{y,i+k})^2) = 2(\kappa_{22}-1)\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2$.

Combining:
$$
\operatorname{Cov}\bigl((X_i-X_{i+k})^2,\,(Y_i-Y_{i+k})^2\bigr) = 8\sigma_{xy}(\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k}) + 2(\kappa_{22}-1)\sigma_x^2\sigma_y^2 + 4\sigma_{xy}^2. \qquad\square
$$
:::

::: {.callout-note collapse="true"}
## Lemma: $E[T_k^{(2)}(X, Y)]$

**Exact expectation.** Applying $E[AB] = \operatorname{Cov}(A,B) + E[A]E[B]$ to each term and summing, using the lemma and $E\!\left[\sum_i(X_i-X_{i+k})^2\right] = T_k(\theta_x,\theta_x) + 2n\sigma_x^2$:
$$
\begin{aligned}
E[T_k^{(2)}(X, Y)]
&= 2n(\kappa_{22}-1)\sigma_x^2\sigma_y^2 + 8\sigma_{xy}T_k(\theta_x,\theta_y) + 4n\sigma_{xy}^2 \\
&\quad + T_k^{(2)}(\theta_x, \theta_y) + 2\sigma_y^2 T_k(\theta_x,\theta_x) + 2\sigma_x^2 T_k(\theta_y,\theta_y) + 4n\sigma_x^2\sigma_y^2.
\end{aligned}
$$

::: {.callout-tip collapse="true"}
## Simulation: Verifying $E[T_k^{(2)}(X, Y)]$

We simulate 1000 datasets from scenario 2 with $n = 500$ observations, $L = 4$ change points (frequent relative to $n$), exponential errors, and heterogeneous marginal variances ($\sigma_x^2 = 4$, $\sigma_y^2 = 0.25$). For each dataset we compute $T_2^{(2)}(X, Y)$ and compare its distribution to the formula value (red line).

```{r}
#| echo: false
#| cache: true
#| message: false
params <- scenario(2, L = 4, err_type = "exponential", sigma = c(2, 0.5))
k   <- 2
n   <- params$n
hx  <- params$h[, 1]; hy <- params$h[, 2]
k22 <- params$kappa$k22
sx  <- sqrt(params$S[1, 1]); sy <- sqrt(params$S[2, 2]); sxy <- params$S[1, 2]

target <- 2 * n * (k22 - 1) * sx^2 * sy^2 +
  8 * sxy * lag_diff(hx, hy, k = k) + 4 * n * sxy^2 +
  lag_diff2(hx, hy, k = k) +
  2 * sy^2 * lag_diff(hx, hx, k = k) +
  2 * sx^2 * lag_diff(hy, hy, k = k) +
  4 * n * sx^2 * sy^2

lag2 <- map_dbl(1:1000, ~ {
  X <- generate_data(params)
  lag_diff2(X[, 1], X[, 2], k = k)
})

```
```{R}
#| echo: false
tibble::tibble(val = lag2) |>
  ggplot(aes(x = val)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = target, color = "firebrick") +
  geom_vline(xintercept = mean(lag2), linetype = "dashed") +
  labs(x = expression(T[k]^{(2)}(X, Y)), y = "density",
       subtitle = "Red = formula, dashed = sample mean") +
  theme_minimal()
```
:::

:::

::: {.callout-warning}
## Remark
As with $2T_1(\theta_x,\theta_y) - T_2(\theta_x,\theta_y) = 0$, we have $2T_1^{(2)}(\theta_x,\theta_y) - T_2^{(2)}(\theta_x,\theta_y) = 0$.
:::

**Linear combination.** Applying the exact expectation at $k=1$ and $k=2$ and forming $2E[T_1^{(2)}(X,Y)] - E[T_2^{(2)}(X,Y)]$, all $\theta$-dependent terms vanish by the remark and the standard property applied to $(X,X)$ and $(Y,Y)$:
$$
2E[T_1^{(2)}(X,Y)] - E[T_2^{(2)}(X,Y)] =
\boxed{2n(\kappa_{22}+1)\sigma_x^2\sigma_y^2 + 4n\sigma_{xy}^2.}
$$

Solving for $\kappa_{22}$:
$$
\kappa_{22} = \frac{2E[T_1^{(2)}(X,Y)] - E[T_2^{(2)}(X,Y)]}{2n\sigma_x^2\sigma_y^2} - \frac{2\sigma_{xy}^2}{\sigma_x^2\sigma_y^2} - 1.
$$

Plugging in estimates:
$$
\boxed{
\hat\kappa_{22} = \frac{2T_1^{(2)}(X,Y) - T_2^{(2)}(X,Y)}{2n\hat\sigma_x^2\hat\sigma_y^2} - \frac{2\hat\sigma_{xy}^2}{\hat\sigma_x^2\hat\sigma_y^2} - 1.
}
$$

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{22}$ (Lag Product)

We simulate 10,000 datasets from scenario 2 with $L = 4$ change points, exponential errors, and heterogeneous marginal variances ($\sigma_x^2 = 4$, $\sigma_y^2 = 0.25$). The sample mean of $\hat\kappa_{22}$ (dashed) should fall on the true value (red).

```{r}
#| echo: false
#| cache: true
#| message: false
params <- scenario(2, L = 4, err_type = "exponential", sigma = c(2, 0.5))
k22    <- params$kappa$k22

k22_est <- function(x, y) {
  n   <- length(x)
  sx2 <- ece.cov(x, x)
  sy2 <- ece.cov(y, y)
  sxy <- ece.cov(x, y)
  (2 * lag_diff2(x, y, k = 1) - lag_diff2(x, y, k = 2)) / (2 * n * sx2 * sy2) -
    2 * sxy^2 / (sx2 * sy2) - 1
}

k22_sim <- map_dbl(1:10000, ~ {
  X <- generate_data(params)
  k22_est(X[, 1], X[, 2])
})
```
```{r}
#| echo: false
tibble::tibble(val = k22_sim) |>
  ggplot(aes(x = val)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = k22, color = "firebrick") +
  geom_vline(xintercept = mean(k22_sim), linetype = "dashed") +
  labs(x = expression(hat(kappa)[22]), y = "density",
       subtitle = "Red = true, dashed = sample mean") +
  theme_minimal()
```

:::
