## Estimating $\kappa_{22}$ via $V$

### Estimating $\kappa_{22}$ via $V_{xy}$

Let $s_i^{(k)} = (X_i - X_{i+k})(Y_i - Y_{i+k})$ and $S_i = \frac{s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)}}{2}$ as in the Multiplier Bootstrap chapter, so that
$$
\boxed{\bar{S} = \frac{1}{n}\sum_{i=1}^n S_i = \frac{2T_1(X,Y) - T_2(X,Y)}{2n} = \hat\sigma_{xy}.}
$$ Since $S_i$ depends only on indices $i-2, i-1, i$, terms at lags $\ge 3$ are independent, so
$$
\operatorname{Var}(\bar{S}) = \frac{1}{n^2}\!\left[\sum_i \operatorname{Var}(S_i) + 2\sum_i \operatorname{Cov}(S_i, S_{i+1}) + 2\sum_i \operatorname{Cov}(S_i, S_{i+2})\right].
$$
Replacing each population quantity with its sample analog gives the lag-2 variance estimate
$$
V := \widehat{\operatorname{Var}}(\bar{S}) = \frac{1}{n^2}\!\left(\sum_{i=1}^n S_i^2 + \sum_{i=1}^n (S_i S_{i+1} + S_{i+1} S_i) + \sum_{i=1}^n (S_i S_{i+2} + S_{i+2} S_i)\right),
$$
or in matrix form, writing $\mathbf{S} = (S_1, \ldots, S_n)^\top$,
$$
\boxed{V = \frac{1}{n^2}\,\mathbf{S}^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\mathbf{S}.}
$$

Since $\bar S = (2T_1 - T_2)/(2n)$, we have $V \approx \operatorname{Var}(2T_1 - T_2)/(4n^2)$. The variance on the right has a general closed form; under $H_0$ ($\sigma_{xy} = 0$) it simplifies so we can solve for $\kappa_{22}$ from $V$.

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Var}(\bar{S})$ from $\Sigma_Z$

Since $\bar{S} = (2T_1 - T_2)/(2n)$,
$$
\operatorname{Var}(\bar{S}) = \frac{1}{4n^2}\bigl(4[\Sigma_Z]_{55} - 4[\Sigma_Z]_{56} + [\Sigma_Z]_{66}\bigr).
$$
The general $\Sigma_Z$ entries from the Z-Test chapter are
$$[\Sigma_Z]_{55} = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) - 2n\sigma_{xy}^2 + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2 + 4n\sigma_{xy}\omega_{xy},$$
$$[\Sigma_Z]_{56} = 4n(\sigma_x^2\sigma_y^2\kappa_{22} - \sigma_{xy}^2) + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2 + 4n\sigma_{xy}\omega_{xy},$$
$$[\Sigma_Z]_{66} = 2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) - 2n\sigma_{xy}^2 + 4n\sigma_y^2\omega_x^2 + 4n\sigma_x^2\omega_y^2 + 8n\sigma_{xy}\omega_{xy}.$$
Substituting and collecting terms gives the general expression
$$\operatorname{Var}(\bar{S}) = \frac{1}{2n}\bigl[(2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 3\sigma_{xy}^2 + 4\sigma_{xy}\omega_{xy} + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2\bigr].$$
Under $H_0$ ($\sigma_{xy} = 0$) the last two terms vanish, yielding
$$\operatorname{Var}(\bar{S}) = \frac{1}{2n}\bigl[(2\kappa_{22}+5)\sigma_x^2\sigma_y^2 + 2\sigma_y^2\omega_x^2 + 2\sigma_x^2\omega_y^2\bigr]. \qquad\square$$
:::

Since $V \approx \operatorname{Var}(\bar{S})$, substituting the general expression and solving gives
$$\boxed{\hat\kappa_{22} = \frac{nV}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{3}{2}\frac{\hat\sigma_{xy}^2}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{2\hat\sigma_{xy}\hat\omega_{xy}}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{\hat\omega_x^2}{\hat\sigma_x^2} - \frac{\hat\omega_y^2}{\hat\sigma_y^2} - \frac{5}{2}.}$$
Under $H_0$ ($\hat\sigma_{xy} = 0$) the middle two terms vanish, yielding
$$\hat\kappa_{22} = \frac{nV}{\hat\sigma_x^2\hat\sigma_y^2} - \frac{\hat\omega_x^2}{\hat\sigma_x^2} - \frac{\hat\omega_y^2}{\hat\sigma_y^2} - \frac{5}{2}.$$

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{22}$ via $V_{xy}$

```{r}
#| echo: false
#| message: false
devtools::load_all("..")
library(purrr)
library(ggplot2)
```

```{r}
#| echo: false
#| cache: true
#| message: false
params <- scenario(2, L = 4, err_type = "normal")

k22_indep <- map_dbl(1:1000, ~ {
  X <- generate_data(params)
  ece.kappa(X, method="matrix")$k22
})

k22_dep <- map_dbl(1:1000, ~ {
  ex <- rnorm(params$n)
  ey <- (ex^2 - 1) / sqrt(2)
  X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
  ece.kappa(X, method="matrix")$k22
})

lambda_vals <- c(0, 0.5, 1/sqrt(2), 1)
k22_true    <- 1 + 4 * lambda_vals^2

sim_results <- map(lambda_vals, \(lam) {
  map_dbl(1:1000, ~ {
    ex <- rnorm(params$n)
    ey <- lam * (ex^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params$n)
    X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
    ece.kappa(X, method="matrix")$k22
  })
})
```

::: {.panel-tabset}

## Independent ($\kappa_{22} = 1$)

Scenario 2, $L = 4$, Gaussian independent errors. Under independence $\kappa_{22} = 1$ regardless of marginals.

```{r}
#| echo: false
make_sim_hist(k22_indep, true = params$kappa$k22, expression(hat(kappa)[22]))
```

## Dependent ($\kappa_{22} = 5$)

Using $\varepsilon_y = (\varepsilon_x^2 - 1)/\sqrt{2}$ with $\varepsilon_x \sim N(0,1)$, giving $\sigma_{xy} = 0$ and $\kappa_{22} = 5$.

```{r}
#| echo: false
make_sim_hist(k22_dep, true = 5, expression(hat(kappa)[22]))
```

## $\lambda$-Mixing

Using $\varepsilon_y = \lambda(\varepsilon_x^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,\eta$ with $\varepsilon_x, \eta \stackrel{iid}{\sim} N(0,1)$, which gives $\sigma_{xy} = 0$ and $\kappa_{22} = 1 + 4\lambda^2$.

```{r}
#| echo: false
sim_df <- tibble::tibble(
  k22   = unlist(sim_results),
  group = factor(rep(round(k22_true, 1), each = 1000),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1)))
)
ref_df <- tibble::tibble(
  group = factor(round(k22_true, 1),
                 labels = paste0("κ₂₂ = ", round(k22_true, 1))),
  true  = k22_true
)
mean_df <- sim_df |>
  dplyr::summarise(mean_k22 = mean(k22), .by = group)

ggplot(sim_df, aes(x = k22, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df, aes(xintercept = mean_k22, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[22]), y = "density") +
  theme_minimal()
```

:::

:::

### Estimating $\kappa_{22}$ via $V_{12}$

An alternative estimator uses the cross-covariance $V_{12} = \widehat{\mathrm{Cov}}(\bar S^{xx}, \bar S^{yy})$ in place of $V_{xy}$. Since $S_{xx} = (2T_1(X,X) - T_2(X,X))/(2n)$ and $S_{yy} = (2T_1(Y,Y) - T_2(Y,Y))/(2n)$,

$$
\mathrm{Cov}(S_{xx}, S_{yy}) = \frac{1}{4n^2}\,\mathrm{Cov}\!\bigl(2T_1(X,X) - T_2(X,X),\; 2T_1(Y,Y) - T_2(Y,Y)\bigr).
$$

Expanding by bilinearity gives a linear combination of the four cross-diagonal $\Sigma_Z$ entries:

$$
4\,[\Sigma_Z]_{T_1(X,X),T_1(Y,Y)} - 2\,[\Sigma_Z]_{T_1(X,X),T_2(Y,Y)} - 2\,[\Sigma_Z]_{T_2(X,X),T_1(Y,Y)} + [\Sigma_Z]_{T_2(X,X),T_2(Y,Y)}.
$$

All four entries share the coefficient $4n\sigma_x^2\sigma_y^2$ on $(\kappa_{22}-1)$; the $\omega_{xy}$ terms cancel to $16n\sigma_{xy}\omega_{xy}$ and the $\sigma_{xy}^2$ terms collect to $8n\sigma_{xy}^2$. Dividing through by $4n^2$ gives the closed form

$$
\mathrm{Cov}(S_{xx}, S_{yy}) = \frac{1}{n}\!\left[\sigma_x^2\sigma_y^2(\kappa_{22}-1) + 2\sigma_{xy}^2 + 4\sigma_{xy}\omega_{xy}\right].
$$

Solving for $\kappa_{22}$,

$$
\boxed{\hat\kappa_{22} = \frac{nV_{12} - 2\hat\sigma_{xy}^2 - 4\hat\sigma_{xy}\hat\omega_{xy}}{\hat\sigma_x^2\hat\sigma_y^2} + 1,}
$$

which under $H_0: \rho = 0$ reduces to $\hat\kappa_{22} = nV_{12}/(\hat\sigma_x^2\hat\sigma_y^2) + 1$, free of all $\omega$ corrections.

### Estimating $\kappa_{40}$ via $V_{xx}$

Applying the same construction to $(X, X)$ in place of $(X, Y)$: let $s_i^{(k)} = (X_i - X_{i+k})^2$ and $S_i^{xx} = (s_i^{(1)} + s_{i+1}^{(1)} - s_i^{(2)})/2$, so that $\bar S^{xx} = (2T_1(X,X) - T_2(X,X))/(2n) = \hat\sigma_x^2$. Define
$$
V_{xx} := \widehat{\operatorname{Var}}(\bar S^{xx}) = \frac{1}{n^2}\,(\tilde{\mathbf{S}}^{xx})^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\tilde{\mathbf{S}}^{xx}, \qquad \tilde S_i^{xx} = S_i^{xx} - \bar S^{xx}.
$$

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Var}(\bar{S}^{xx})$ from $\Sigma_Z$

Since $\bar{S}^{xx} = (2T_1(X,X) - T_2(X,X))/(2n)$,
$$
\operatorname{Var}(\bar{S}^{xx}) = \frac{1}{4n^2}\bigl(4[\Sigma_Z]_{11} - 4[\Sigma_Z]_{12} + [\Sigma_Z]_{22}\bigr).
$$
Substituting the entries from the Z-Test chapter:
$$[\Sigma_Z]_{11} = 4n\kappa_{40}\sigma_x^4 + 8n\sigma_x^2\omega_x^2,$$
$$[\Sigma_Z]_{12} = 4n(\kappa_{40}-1)\sigma_x^4 + 8n\sigma_x^2\omega_x^2,$$
$$[\Sigma_Z]_{22} = 4n\kappa_{40}\sigma_x^4 + 16n\sigma_x^2\omega_x^2.$$
Collecting terms:
$$\operatorname{Var}(\bar{S}^{xx}) = \frac{1}{n}\bigl[(\kappa_{40}+4)\sigma_x^4 + 4\sigma_x^2\omega_x^2\bigr]. \qquad\square$$
:::

Since $V_{xx} \approx \operatorname{Var}(\bar{S}^{xx})$, substituting and solving gives
$$
\boxed{\hat\kappa_{40} = \frac{nV_{xx}}{\hat\sigma_x^4} - \frac{4\hat\omega_x^2}{\hat\sigma_x^2} - 4.}
$$
By symmetry ($x \leftrightarrow y$), the same construction applied to $(Y, Y)$ yields $\hat\kappa_{04} = nV_{yy}/\hat\sigma_y^4 - 4\hat\omega_y^2/\hat\sigma_y^2 - 4$.

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{40}$

```{r}
#| echo: false
#| cache: true
#| message: false
params     <- scenario(2, L = 4, err_type = "normal")
params_exp <- scenario(2, L = 4, err_type = "exponential")

k40_gaussian <- map_dbl(1:1000, ~ ece.kappa(generate_data(params))$k40)
k40_exp      <- map_dbl(1:1000, ~ ece.kappa(generate_data(params_exp))$k40)

lambda_vals  <- c(0, 0.5, 1/sqrt(2), 1)
k40_true     <- 12 * lambda_vals^4 + 3

sim_results_k40 <- map(lambda_vals, \(lam) {
  map_dbl(1:1000, \(.) {
    z  <- rnorm(params$n)
    ex <- lam * (z^2 - 1) / sqrt(2) + sqrt(1 - lam^2) * rnorm(params$n)
    ey <- rnorm(params$n)
    X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
    ece.kappa(X)$k40
  })
})
```

::: {.panel-tabset}

## Gaussian ($\kappa_{40} = 3$)

Scenario 2, $L = 4$, Gaussian errors ($\kappa_{40} = 3$).

```{r}
#| echo: false
make_sim_hist(k40_gaussian, true = params$kappa$k40, expression(hat(kappa)[40]))
```

## Exponential ($\kappa_{40} = 9$)

Scenario 2, $L = 4$, exponential errors ($\kappa_{40} = 9$).

```{r}
#| echo: false
make_sim_hist(k40_exp, true = params_exp$kappa$k40, expression(hat(kappa)[40]))
```

## $\lambda$-Mixing

Using $\varepsilon_x = \lambda(z^2-1)/\sqrt{2} + \sqrt{1-\lambda^2}\,w$ with $z, w \stackrel{iid}{\sim} N(0,1)$, which gives $\kappa_{40} = 12\lambda^4 + 3$.

```{r}
#| echo: false
sim_df_k40 <- tibble::tibble(
  k40   = unlist(sim_results_k40),
  group = factor(rep(round(k40_true, 1), each = 1000),
                 labels = paste0("κ₄₀ = ", round(k40_true, 1)))
)
ref_df_k40 <- tibble::tibble(
  group = factor(round(k40_true, 1),
                 labels = paste0("κ₄₀ = ", round(k40_true, 1))),
  true  = k40_true
)
mean_df_k40 <- sim_df_k40 |>
  dplyr::summarise(mean_k40 = mean(k40), .by = group)

ggplot(sim_df_k40, aes(x = k40, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k40, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k40, aes(xintercept = mean_k40, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[40]), y = "density") +
  theme_minimal()
```

:::

:::

### Estimating $\kappa_{31}$ via $C_{xx,xy}$

The cumulant $\kappa_{31}$ appears in the cross-covariances between the $(X,X)$ and $(X,Y)$ blocks of $\Sigma_Z$. Define the lag-2 cross-covariance
$$
C_{xx,xy} := \frac{1}{n^2}\,(\tilde{\mathbf{S}}^{xx})^\top(I + C_1 + C_1^\top + C_2 + C_2^\top)\tilde{\mathbf{S}}^{xy}, \qquad \tilde S_i^{xx} = S_i^{xx} - \bar S^{xx},\quad \tilde S_i^{xy} = S_i^{xy} - \bar S^{xy},
$$
which estimates $\operatorname{Cov}(\bar S^{xx}, \bar S^{xy})$.

::: {.callout-note collapse="true"}
## Derivation: $\operatorname{Cov}(\bar{S}^{xx}, \bar{S}^{xy})$ from $\Sigma_Z$

Since $\bar{S}^{xx} = (2T_1(X,X) - T_2(X,X))/(2n)$ and $\bar{S}^{xy} = (2T_1(X,Y) - T_2(X,Y))/(2n)$,
$$
\operatorname{Cov}(\bar{S}^{xx}, \bar{S}^{xy}) = \frac{1}{4n^2}\bigl(4[\Sigma_Z]_{15} - 2[\Sigma_Z]_{16} - 2[\Sigma_Z]_{25} + [\Sigma_Z]_{26}\bigr).
$$
The relevant entries from the Z-Test chapter are
$$[\Sigma_Z]_{15} = 4n\sigma_x^3\sigma_y\kappa_{31} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{16} = 4n\sigma_x^3\sigma_y\kappa_{31} - 4n\sigma_x^2\sigma_{xy} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{25} = 4n\sigma_x^3\sigma_y\kappa_{31} - 4n\sigma_x^2\sigma_{xy} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy},$$
$$[\Sigma_Z]_{26} = 4n\sigma_x^3\sigma_y\kappa_{31} + 8n\sigma_{xy}\omega_x^2 + 8n\sigma_x^2\omega_{xy}.$$
Collecting terms:
$$\operatorname{Cov}(\bar{S}^{xx}, \bar{S}^{xy}) = \frac{1}{n}\bigl[\sigma_x^3\sigma_y\kappa_{31} + 4\sigma_x^2\sigma_{xy} + 2\sigma_{xy}\omega_x^2 + 2\sigma_x^2\omega_{xy}\bigr]. \qquad\square$$
:::

Since $C_{xx,xy} \approx \operatorname{Cov}(\bar{S}^{xx}, \bar{S}^{xy})$, substituting and solving gives
$$
\hat\kappa_{31} = \frac{nC_{xx,xy}}{\hat\sigma_x^3\hat\sigma_y} - \frac{4\hat\sigma_{xy}}{\hat\sigma_x\hat\sigma_y} - \frac{2\hat\sigma_{xy}\hat\omega_x^2}{\hat\sigma_x^3\hat\sigma_y} - \frac{2\hat\omega_{xy}}{\hat\sigma_x\hat\sigma_y}.
$$
By symmetry ($x \leftrightarrow y$), the same construction with $C_{yy,xy}$ yields $\hat\kappa_{13}$.

::: {.callout-tip collapse="true"}
## Simulation: Verifying $\hat\kappa_{31}$

```{r}
#| echo: false
#| cache: true
#| message: false
params <- scenario(2, L = 4, err_type = "normal")

k31_indep <- map_dbl(1:1000, ~ {
  X <- generate_data(params)
  ece.kappa(X)$k31
})

k31_dep <- map_dbl(1:1000, ~ {
  ex <- rnorm(params$n)
  X  <- cbind(params$h[,1] + ex, params$h[,2] + ex)
  ece.kappa(X)$k31
})

lambda_vals_k31 <- c(0, 0.5, 1/sqrt(2), 1)
k31_true        <- 3 * lambda_vals_k31

sim_results_k31 <- map(lambda_vals_k31, \(lam) {
  map_dbl(1:1000, \(.) {
    ex <- rnorm(params$n)
    ey <- lam * ex + sqrt(1 - lam^2) * rnorm(params$n)
    X  <- cbind(params$h[,1] + ex, params$h[,2] + ey)
    ece.kappa(X)$k31
  })
})
```

::: {.panel-tabset}

## Independent ($\kappa_{31} = 0$)

Scenario 2, $L = 4$, Gaussian independent errors. Under independence $\kappa_{31} = 0$.

```{r}
#| echo: false
make_sim_hist(k31_indep, true = params$kappa$k31, expression(hat(kappa)[31]))
```

## Dependent ($\kappa_{31} = 3$)

Using $\varepsilon_y = \varepsilon_x$ with $\varepsilon_x \sim N(0,1)$, giving $\sigma_{xy} = \sigma_x^2$ and $\kappa_{31} = \kappa_{40} = 3$.

```{r}
#| echo: false
make_sim_hist(k31_dep, true = 3, expression(hat(kappa)[31]))
```

## $\lambda$-Mixing

Using $\varepsilon_y = \lambda\varepsilon_x + \sqrt{1-\lambda^2}\,\eta$ with $\varepsilon_x, \eta \stackrel{iid}{\sim} N(0,1)$, which gives $\kappa_{31} = 3\lambda$.

```{r}
#| echo: false
sim_df_k31 <- tibble::tibble(
  k31   = unlist(sim_results_k31),
  group = factor(rep(round(k31_true, 2), each = 1000),
                 labels = paste0("κ₃₁ = ", round(k31_true, 2)))
)
ref_df_k31 <- tibble::tibble(
  group = factor(round(k31_true, 2),
                 labels = paste0("κ₃₁ = ", round(k31_true, 2))),
  true  = k31_true
)
mean_df_k31 <- sim_df_k31 |>
  dplyr::summarise(mean_k31 = mean(k31), .by = group)

ggplot(sim_df_k31, aes(x = k31, fill = group, color = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, position = "identity", alpha = 0.4, linewidth = 0) +
  geom_vline(data = ref_df_k31, aes(xintercept = true, color = group), linewidth = 0.9) +
  geom_vline(data = mean_df_k31, aes(xintercept = mean_k31, color = group), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = expression(hat(kappa)[31]), y = "density") +
  theme_minimal()
```

:::

:::
