---
title: "Mean Removal"
---

## Segmentation

Assume each time series has a piecewise-constant mean:
$$X_t = \mu_t + \varepsilon_t, \qquad \mu_t \text{ piecewise constant.}$$
Segmentation estimates $\mu_t$ by the within-segment sample mean and subtracts it:
$$\hat{\varepsilon}_t = X_t - \hat{\mu}_t.$$
The residuals $\hat{\varepsilon}_t$ are approximately stationary, so Pearson correlation applied to $(\hat{\varepsilon}_t^{(1)}, \hat{\varepsilon}_t^{(2)})$ estimates contemporaneous correlation between the two series.

## Degrees of Freedom Correction

To do inference on Pearson correlation, we first estimate the correlation $\hat{r}$. Then, we find that
$$t_{\text{naive}} = \frac{r \sqrt{n-2}}{\sqrt{1-r^2}} \sim t_{n-2}.$$
That is, the standard Pearson correlation has $df = n-2$, because we use two degrees of freedom to estimate the mean and variance.

However, with segmentation, we must estimate multiple means. If there are $K$ change points, then we must estimate $K + 1$ sample means. In all, we must estimate $K + 2$ parameters, so the correct statistic is
$$t_{\text{correct}} = \frac{r\sqrt{n-K-2}}{\sqrt{1-r^2}} \sim t_{n-K-2}.$$
Since both statistics share the same $r$, they are related by
$$t_{\text{naive}} = \lambda \cdot t_{\text{correct}}, \qquad \lambda = \sqrt{\frac{n-2}{n-K-2}} > 1.$$

The naive test rejects when
$$|t_{\text{naive}}| > t_{n-2,\,\alpha/2}.$$
Substituting $t_{\text{naive}} = \lambda \cdot t_{\text{correct}}$, this is equivalent to
$$|t_{\text{correct}}| > \frac{t_{n-2,\,\alpha/2}}{\lambda}.$$
So the actual type I error of the naive test is
$$P\!\left(|t_{\text{correct}}| > \frac{t_{n-2,\,\alpha/2}}{\lambda}\right), \qquad t_{\text{correct}} \sim t_{n-K-2}.$$
Since $\lambda > 1$, the threshold $t_{n-2,\,\alpha/2}/\lambda < t_{n-K-2,\,\alpha/2}$, so the naive test rejects more often than it should.

**Theoretical type I error.** Given $n$, $K$, and $\alpha$, the exact inflation is computable in closed form. Let $c = t_{n-2,\,\alpha/2}$ be the naive critical value. The actual type I error is
$$P\!\left(|T| > \frac{c}{\lambda}\right), \qquad T \sim t_{n-K-2}.$$
In R: `2 * pt(-qt(1 - alpha/2, df = n-2) / lambda, df = n - K - 2)`.

**Example.** The table below uses $\alpha = 0.05$ and $K/n \approx 0.25$ (as in `scenario(10, n)`), showing that the inflation is determined by the ratio $K/n$, not $n$ alone.

| $n$ | $K$ | $\lambda$ | Naive critical value | Effective threshold | Actual type I error |
|-----|-----|-----------|----------------------|---------------------|---------------------|
| 100 | 25 | $\sqrt{98/73} \approx 1.159$ | $t_{98,\,0.025} \approx 1.984$ | $1.984/1.159 \approx 1.712$ | $P(\lvert t_{73}\rvert > 1.712) \approx 0.091$ |
| 500 | 125 | $\sqrt{498/373} \approx 1.155$ | $t_{498,\,0.025} \approx 1.965$ | $1.965/1.155 \approx 1.701$ | $P(\lvert t_{373}\rvert > 1.701) \approx 0.090$ |
| 1000 | 249 | $\sqrt{998/749} \approx 1.155$ | $t_{998,\,0.025} \approx 1.962$ | $1.962/1.155 \approx 1.699$ | $P(\lvert t_{749}\rvert > 1.699) \approx 0.090$ |

The simulation below confirms this. The histogram shows the corrected t-statistics alongside both candidate densities. When $n$ and $n - K$ are both large, the two $t$-curves are visually indistinguishable — the inflation does not appear in the bulk of the distribution, only in the rejection rate.

```{r}
#| message: false
#| echo: false
devtools::load_all("~/Documents/Research/Code/ECE")

make_config <- function(n, K, s12 = 0, signal = 1) {
  seg_len <- ceiling(n / (K + 1))
  vals <- ifelse(seq_len(K + 1) %% 2 == 1, 1, -1)
  h_vec <- rep(vals, each = seg_len)[seq_len(n)]
  h <- signal * cbind(h_vec, h_vec)
  S <- matrix(c(1, s12, s12, 1), ncol = 2, byrow = TRUE)
  list(scenario = 0, n = n, K = K, S = S, h = h, signal = signal)
}

segment.test <- function(X, config) {
  n <- nrow(X)
  K <- length(find_cp(config$h))

  X_resid <- X - segment_mean_oracle(X, config)
  r <- cor(X_resid[, 1], X_resid[, 2])

  df <- n - K - 2
  t_stat <- r * sqrt(df) / sqrt(1 - r^2)
  pval <- 2 * pt(-abs(t_stat), df = df)

  list(statistic = t_stat, p.value = pval, df = df, r = r, K = K)
}
```

### Distribution of T-Statistic
After applying segmentation, this is the distribution of t-statistics

```{R}
#| cache: true
#| echo: false
N <- 1000
n <- 1000
K <- n / 2

set.seed(1)
t_naive <- replicate(N, {
  # Generate Data
  config <- make_config(n=n, K=K)
  X <- generate_data(config)

  # Remove Mean and Estimate Correlation
  X_resid <- X - segment_mean_oracle(X, config)
  r <- cor(X_resid[, 1], X_resid[, 2])

  # Calculate t-Statistic
  df <- n - 2
  t_stat <- r * sqrt(df) / sqrt(1 - r^2)
  pval <- 2 * pt(-abs(t_stat), df = df)
  t_stat
})

set.seed(1)
t_correct <- replicate(N, {
  # Generate Data
  config <- make_config(n=n, K=K)
  X <- generate_data(config)

  # Remove Mean and Estimate Correlation
  X_resid <- X - segment_mean_oracle(X, config)
  r <- cor(X_resid[, 1], X_resid[, 2])

  # Calculate t-Statistic
  df <- n - K - 2
  t_stat <- r * sqrt(df) / sqrt(1 - r^2)
  pval <- 2 * pt(-abs(t_stat), df = df)
  t_stat
})

```

```{R}
#| echo: false
xlim <- range(c(t_naive, t_correct))
par(mfrow = c(1, 2))

  hist(t_naive, breaks = 40, freq = FALSE,
       col = "lightgray", border = "white",
       main = expression("Naive:" ~ frac(r * sqrt(n-2), sqrt(1-r^2))),
       xlab = "t-statistic", xlim = xlim)
  xx <- seq(xlim[1], xlim[2], length.out = 200)
  lines(xx, dt(xx, df = n - 2),     col = "#E41A1C", lwd = 2)
  lines(xx, dt(xx, df = n - K - 2), col = "#377EB8", lwd = 2)
  legend("topright", legend = c(sprintf("t[%d]", n-2), sprintf("t[%d]", n-K-2)),
         col = c("#E41A1C", "#377EB8"), lwd = 2, bty = "n", cex = 0.8)

  hist(t_correct, breaks = 40, freq = FALSE,
       col = "lightgray", border = "white",
       main = expression("Corrected:" ~ frac(r * sqrt(n - K - 2), sqrt(1 - r^2))),
       xlab = "t-statistic", xlim = xlim)
  lines(xx, dt(xx, df = n - 2),     col = "#E41A1C", lwd = 2)
  lines(xx, dt(xx, df = n - K - 2), col = "#377EB8", lwd = 2)
  legend("topright", legend = c(sprintf("t[%d]", n-2), sprintf("t[%d]", n-K-2)),
         col = c("#E41A1C", "#377EB8"), lwd = 2, bty = "n", cex = 0.8)

par(mfrow = c(1, 1))
```

We see that the correction fits the curve much better. While the t-distributions are nearly identical (they're stacked in the figure), the distribution of the t-statistic just doesn't fit the t-distribution correctly. 

```{R}
#| echo: false
alpha <- 0.05
lambda <- sqrt((n - 2) / (n - K - 2))
theoretical <- 2 * pt(-qt(1 - alpha / 2, df = n - 2) / lambda, df = n - K - 2)

pval_naive <- 2 * pt(-abs(t_naive), df = n - 2)
pval_correct <- 2 * pt(-abs(t_correct), df = n - K - 2)

data.frame(
  Test = c("Theoretical (naive)", "Naive", "Corrected"),
  `Type I Error` = c(
    theoretical,
    mean(pval_naive < alpha), 
    mean(pval_correct < alpha)
  ),
  check.names = FALSE
) |> knitr::kable(digits = 3)
```

We also see that the theoretical type I error is empirically prominent.

## Type I Error vs Number of Change Points

With $n = 1000$ fixed, we sweep over $K \in \{2, 100, 250, 500\}$ and compare the theoretical prediction against the empirical naive and corrected rejection rates.

```{r}
#| cache: true
#| echo: false
n <- 1000
K_vals <- c(2, 100, 250, 500)
alpha <- 0.05
N <- 1000

results <- map(K_vals, function(K) {
  config <- make_config(n = n, K = K)
  lambda <- sqrt((n - 2) / (n - K - 2))
  theoretical <- 2 * pt(-qt(1 - alpha / 2, df = n - 2) / lambda, df = n - K - 2)

  set.seed(1)
  pvals <- replicate(N, {
    X <- generate_data(config)
    X_resid <- X - segment_mean_oracle(X, config)
    r <- cor(X_resid[, 1], X_resid[, 2])
    c(
      naive   = 2 * pt(-abs(r * sqrt(n - 2)     / sqrt(1 - r^2)), df = n - 2),
      correct = 2 * pt(-abs(r * sqrt(n - K - 2) / sqrt(1 - r^2)), df = n - K - 2)
    )
  })

  data.frame(
    K           = K,
    Theoretical = theoretical,
    Naive       = mean(pvals["naive", ]   < alpha),
    Corrected   = mean(pvals["correct", ] < alpha)
  )
})

list_rbind(results) |> knitr::kable(digits = 3)
```

