---
title: "Model Misspecification"
---

## Key Concepts

**The signal $\theta$.** A vector $(\theta_0, \ldots, \theta_{n-1})$ where $\theta_i$ is the true mean of the time series at time $i$. The ECE is designed for piecewise-constant $\theta$; misspecification means $\theta$ has some other shape.

**Frequency index $j$.** An integer from $0$ to $n-1$ labeling a sinusoidal component. Specifically, $j$ counts how many complete oscillations the sinusoid $\cos(2\pi j i / n)$ makes as $i$ runs from $0$ to $n-1$:

- $j = 0$: flat constant — no oscillations
- $j = 1$: one slow wave across the whole window
- $j = n/2$: the fastest possible — alternates sign at every step (Nyquist)

The plot below shows what each of these looks like concretely.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
n <- 20
i <- 0:(n - 1)

par(mfrow = c(1, 4), mar = c(4, 2.5, 3, 1))
for (f in c(1, 3, 7, 10)) {
  plot(i, cos(2 * pi * f * i / n),
    type = "b", pch = 16, cex = 0.7, lwd = 1.5, col = "steelblue",
    xlab = "i", ylab = "", ylim = c(-1.4, 1.4),
    main = bquote(j == .(f) ~ "(" * .(f) ~ "cycles)")
  )
  abline(h = 0, col = "grey70", lty = 2)
}
```

**The DFT.** The Discrete Fourier Transform rewrites $\theta$ as a sum of sinusoids — one at each frequency $j$. The complex coefficient $\tilde\theta_j$ says how much of $\theta$ oscillates at rate $j$.

**Energy $|\tilde\theta_j|^2/n$.** How much of $\theta$'s total variation comes from frequency $j$. A tall bar at $j = 5$ means $\theta$ has a strong 5-cycle-per-window oscillation. By Parseval's theorem, the energies sum to $\|\theta\|^2/n$.

**Eigenvalues $\lambda_j$.** The ECE matrix $A$ acts differently on each frequency: it scales sinusoidal components by $\lambda_j$. Small $|\lambda_j|$ means the ECE ignores energy at that frequency; large $|\lambda_j|$ means energy there leaks into the bias.

**The bias formula.** Because of the eigendecomposition $A = Q\Lambda Q^*$, the bias simplifies to a frequency-by-frequency weighted sum:

$$\theta^T A \theta = \sum_{j=0}^{n-1} \lambda_j \cdot \frac{|\tilde\theta_j|^2}{n}$$

This is the central formula for understanding misspecification: the bias is the dot product between the signal's energy profile and the ECE's frequency sensitivity. Wherever $\lambda_j \approx 0$, that frequency contributes nothing to the bias regardless of how much energy $\theta$ places there.

## Circulant Matrices and the DFT

Every symmetric circulant $A = \operatorname{circ}(a)$ is diagonalized by the DFT matrix $F \in \mathbb{C}^{n \times n}$,

$$
A = F^* \Lambda F,
$$

where $F_{jk} = \frac{1}{\sqrt{n}} \omega^{jk}$, $\omega = e^{-2\pi i/n}$, and $\Lambda = \operatorname{diag}(\lambda_0, \ldots, \lambda_{n-1})$. The eigenvalues are the DFT of the first row:

$$
\lambda_j = \sum_{k=0}^{n-1} a_k\, \omega^{jk} = (\mathcal{F} a)_j.
$$

Since $A$ is real and symmetric, the eigenvalues are real and satisfy $\lambda_j = \lambda_{n-j}$.

Because $F$ is unitary ($F F^* = I$), the decomposition $A = F^* \Lambda F$ is exactly the eigendecomposition $A = Q \Lambda Q^{-1}$ with $Q = F^*$. The columns of $F^*$ are the eigenvectors of $A$: the $j$-th eigenvector is $f_j = \frac{1}{\sqrt{n}}(1, \omega^{-j}, \omega^{-2j}, \ldots, \omega^{-(n-1)j})^T$, a complex sinusoid at frequency $j/n$.

::: {.callout-note collapse="true"}
## DFT and eigendecomposition are the same thing for circulants

```{r}
#| message: false
#| warning: false
devtools::load_all("~/Documents/Research/Code/ECE")
library(testthat)

n <- 8
a <- ece_first_row(n, L = 2)
A <- circ(a) / n

# Eigendecomposition
ed  <- eigen(A)

# DFT matrix: F[j,k] = (1/sqrt(n)) * exp(-2*pi*i*j*k/n), 0-indexed
j <- 0:(n - 1)
k <- 0:(n - 1)
F_mat <- exp(-2i * pi * outer(j, k) / n) / sqrt(n)

test_that("columns of F* are eigenvectors of A", {
  # A = F* Lambda F  =>  A F* = F* Lambda
  # Each column of F* should satisfy A v = lambda v
  Q  <- Conj(t(F_mat))           # F*, columns are eigenvectors
  AQ <- A %*% Q
  QD <- Q %*% diag(fft(a) / n)  # eigenvalues = fft(a)/n for scaled A
  expect_equal(Re(AQ), Re(QD), tolerance = 1e-10)
  expect_equal(Im(AQ), Im(QD), tolerance = 1e-10)
})

test_that("eigenvalues of A equal fft(a)/n", {
  lambda_fft  <- sort(Re(fft(a) / n))
  lambda_eig  <- sort(Re(ed$values))
  expect_equal(lambda_fft, lambda_eig, tolerance = 1e-10)
})
```
:::

**Spectral representation of the bias.** Using the eigendecomposition $A = Q\Lambda Q^*$ with $Q = F^*$,

$$
\theta_x^T A \theta_y = \theta_x^T Q \Lambda Q^* \theta_y = (Q^*\theta_x)^* \Lambda (Q^*\theta_y) = \sum_{j=0}^{n-1} \lambda_j\, \tilde{\theta}_{x,j}^*\, \tilde{\theta}_{y,j},
$$

where $\tilde{\theta} = Q^*\theta$ are the coordinates of $\theta$ in the eigenbasis.

Under correct specification ($\theta \in \Theta_L$), the $j = 0$ term vanishes because $\lambda_0 = 0$ always — the DC eigenvalue is zero since $\sum_k a_k = 0$ by the block-sum constraints. For $j > 0$, neither $\lambda_j$ nor $\tilde{\theta}_j$ is zero in general; the sum cancels through structured pairwise cancellation across frequencies, a consequence of the block-sum constraints rather than anything visible in $\lambda$ or $\tilde{\theta}$ individually.

The spectral representation is most useful under misspecification. When $\theta \notin \Theta_L$, the pairwise cancellation breaks down and the bias

$$
\theta_x^T A \theta_y = \sum_{j=0}^{n-1} \lambda_j\, \tilde{\theta}_{x,j}^*\, \tilde{\theta}_{y,j}
$$

is nonzero. Its magnitude depends on which eigenvalues $\lambda_j$ are large and how much energy $\tilde{\theta}_x$, $\tilde{\theta}_y$ place at the corresponding frequencies.

## Closed-Form Bias

Since $A = \sum_{k=1}^L c_k A_k$ with $c_k = H_{1,k} = \frac{2L+1-3k}{nL(L-1)}$ (see Other Mathematical Results), the bias decomposes as

$$
\theta_x^T A \theta_y = \sum_{k=1}^L c_k\, T_k(\theta_x, \theta_y),
$$

where $T_k(\theta_x, \theta_y) = \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\theta_{y,i} - \theta_{y,i+k})$ is the $k$-lagged cross-product of the mean vectors. For $L = 2$ this reduces to

$$
\theta_x^T A \theta_y = \frac{1}{2n}\bigl(2T_1(\theta_x, \theta_y) - T_2(\theta_x, \theta_y)\bigr).
$$

**Spectral form of $\lambda_j$.** Applying the eigendecomposition of $A_k$ — whose eigenvalues are $4\sin^2(\pi jk/n)$ — gives a closed form for the eigenvalues of $A$:

$$
\lambda_j = \sum_{k=1}^L c_k \cdot 4\sin^2\!\left(\frac{\pi jk}{n}\right) = \frac{4}{nL(L-1)}\sum_{k=1}^L (2L+1-3k)\sin^2\!\left(\frac{\pi jk}{n}\right).
$$

For $L = 2$ this simplifies to

$$
\lambda_j = -\frac{4}{n}\cos\!\left(\frac{2\pi j}{n}\right)\sin^2\!\left(\frac{\pi j}{n}\right).
$$

**Bias bound.** By Cauchy–Schwarz in the eigenbasis, and using $\|\tilde{\theta}\|_2 = \|\theta\|_2$ (unitarity of $Q$),

$$
|\theta_x^T A \theta_y| \leq \max_j |\lambda_j| \cdot \|\theta_x\|_2 \|\theta_y\|_2.
$$

The bias is therefore controlled by $\max_j |\lambda_j|$, which from the $L=2$ formula is largest near the Nyquist frequency $j = n/2$. Misspecification is most harmful when the mean vectors have energy concentrated at high frequencies.

## Closed-Form Variance

The variance of $X^T A Y$ is (see the Bilinear Form chapter):

$$
\operatorname{Var}(X^T A Y) = \sigma_y^2\,\theta_x^T A^2 \theta_x + \sigma_x^2\,\theta_y^T A^2 \theta_y + 2\sigma_{xy}\,\theta_x^T A^2 \theta_y + (\sigma_x^2\sigma_y^2 + \sigma_{xy}^2)\operatorname{tr}(A^2).
$$

The noise term $(\sigma_x^2\sigma_y^2 + \sigma_{xy}^2)\operatorname{tr}(A^2) = O(1/n)$ regardless of the mean. The signal-dependent terms involve $\theta_x^T A^2 \theta_y$, which mirrors the bias with $A$ replaced by $A^2$.

**Spectral form.** Since $A^2 = Q\Lambda^2 Q^*$,

$$
\theta_x^T A^2 \theta_y = \sum_{j=0}^{n-1} \lambda_j^2\, \tilde{\theta}_{x,j}^*\, \tilde{\theta}_{y,j},
$$

where $\tilde{\theta} = Q^*\theta$. This is the spectral bias formula with $\lambda_j$ replaced by $\lambda_j^2$.

**Spectral form of $\lambda_j^2$.** For $L = 2$, squaring $\lambda_j = -\frac{4}{n}\cos(2\pi j/n)\sin^2(\pi j/n)$:

$$
\lambda_j^2 = \frac{16}{n^2}\cos^2\!\left(\frac{2\pi j}{n}\right)\sin^4\!\left(\frac{\pi j}{n}\right).
$$

**Variance bound.** By Cauchy–Schwarz in the eigenbasis:

$$
|\theta_x^T A^2 \theta_y| \leq \max_j \lambda_j^2 \cdot \|\theta_x\|_2 \|\theta_y\|_2 = \left(\max_j |\lambda_j|\right)^2 \|\theta_x\|_2 \|\theta_y\|_2.
$$

The signal-dependent variance is controlled by the same Nyquist-dominated $\max_j|\lambda_j|$ as the bias.

## Case Study: Sinusoidal Mean

Suppose $\theta_x = \theta_y = \theta$ with $\theta_i = A\sin(2\pi fi/n)$ for integer frequency $1 \leq f < n/2$ and amplitude $A > 0$. This is misspecified since a sinusoid is not piecewise constant.

### Bias

**Computing $T_k(\theta, \theta)$.** Applying the sum-to-product identity $\sin\alpha - \sin\beta = 2\cos\!\left(\frac{\alpha+\beta}{2}\right)\sin\!\left(\frac{\alpha-\beta}{2}\right)$,

$$
\theta_i - \theta_{i+k} = -2A\sin\!\left(\frac{\pi fk}{n}\right)\cos\!\left(\frac{2\pi fi}{n} + \frac{\pi fk}{n}\right).
$$

Squaring and summing over $i$, and using $\sum_{i=0}^{n-1}\cos^2(\alpha + 2\pi fi/n) = n/2$ for $f \neq 0$,

$$
T_k(\theta, \theta) = 2nA^2\sin^2\!\left(\frac{\pi fk}{n}\right).
$$

**Bias for $L = 2$.** Substituting into $\frac{1}{2n}(2T_1 - T_2)$,

$$
\theta^T A \theta = 2A^2\sin^2\!\left(\frac{\pi f}{n}\right) - A^2\sin^2\!\left(\frac{2\pi f}{n}\right).
$$

Using $\sin^2(2\phi) = 4\sin^2(\phi)\cos^2(\phi)$ and the double-angle identity $1 - 2\cos^2\phi = -\cos(2\phi)$,

$$
\theta^T A \theta = -2A^2\sin^2\!\left(\frac{\pi f}{n}\right)\cos\!\left(\frac{2\pi f}{n}\right).
$$

**Interpretation.** The bias depends on frequency and amplitude but not on phase. It is zero at $f = 0$ (constant — well specified) and at $f = n/4$ (where $\cos(2\pi f/n) = 0$), negative for $f < n/4$, and positive for $f > n/4$. It grows as $A^2$ and is largest near the Nyquist frequency $f = n/2$, consistent with the general bound $|\text{bias}| \leq \max_j|\lambda_j| \cdot \|\theta\|_2^2$.

```{r}
#| message: false
#| warning: false
devtools::load_all("~/Documents/Research/Code/ECE")
library(testthat)

n <- 40; L <- 2
a     <- ece_reg_row(n, L)
A_mat <- circ(a) / n
freqs <- 1:(n/2 - 1)

bias_numerical <- sapply(freqs, function(f) {
  theta <- sin(2 * pi * f * (0:(n - 1)) / n)
  as.numeric(t(theta) %*% A_mat %*% theta)
})

bias_formula <- -2 * sin(pi * freqs / n)^2 * cos(2 * pi * freqs / n)

test_that("sinusoidal bias matches closed form", {
  expect_equal(bias_numerical, bias_formula, tolerance = 1e-10)
})

plot(freqs, bias_formula,
  type = "l", lwd = 2, col = "steelblue",
  xlab = "Frequency f", ylab = "Bias",
  main = "Bias for sinusoidal mean (A = 1, L = 2)"
)
abline(h = 0, col = "grey60", lty = 2)
```

**Application to scenario 14.** Scenario 14 sets $\theta_i = \sin(2\pi i / 365)$ — a yearly cycle sampled at daily resolution. With $n = 100$ observations, the signal has effective DFT bin $j \approx n/365 \approx 0.27$, well below $j = 1$. Since $\lambda_j \approx 0$ near $j = 0$, the formula predicts negligible bias: $|\text{bias}| \approx 2\sin^2(\pi/365) \approx 1.5 \times 10^{-4}$ per unit amplitude squared. Numerically the bias is of order $10^{-5}$, confirming that the ECE is effectively unbiased in this scenario despite the misspecification — the yearly cycle is too slow relative to the observation window to contaminate the estimate.

With $n = 1000$ observations, the window covers approximately 2.7 full periods and the effective bin shifts to $j \approx 2.74$. The bias is now $\approx -1.6 \times 10^{-4}$ — negative since $j < n/4 = 250$, and consistent with the formula prediction of $-1.5 \times 10^{-4}$. Despite the larger signal ($\|\theta\|^2 \approx 499$), the bias remains negligible: the yearly cycle is still low-frequency relative to $n$.

### Variance

With $\theta_x = \theta_y = \theta$, the closed-form variance simplifies to

$$\operatorname{Var}(X^T A Y) = (\sigma_x^2 + \sigma_y^2 + 2\sigma_{xy})\,\theta^T A^2 \theta + (\sigma_x^2\sigma_y^2 + \sigma_{xy}^2)\operatorname{tr}(A^2).$$

**Computing $\theta^T A^2 \theta$.** Since $\theta_i = A\sin(2\pi fi/n)$ lies in the eigenspace of $A$ at frequency $f$ and $\lambda_f = \lambda_{n-f}$, we have $A\theta = \lambda_f\,\theta$. Therefore $\theta^T A^2 \theta = \lambda_f^2\|\theta\|^2 = \frac{nA^2}{2}\lambda_f^2$.

**For $L = 2$.** Substituting the $L=2$ eigenvalue:

$$\operatorname{Var}(X^T A Y) = \frac{8A^2}{n}(\sigma_x^2 + \sigma_y^2 + 2\sigma_{xy})\cos^2\!\left(\frac{2\pi f}{n}\right)\sin^4\!\left(\frac{\pi f}{n}\right) + (\sigma_x^2\sigma_y^2 + \sigma_{xy}^2)\operatorname{tr}(A^2).$$

Both terms are $O(1/n)$. The signal-dependent term vanishes at the same frequencies as the bias ($f = 0$ and $f = n/4$).

```{r}
#| message: false
#| warning: false
devtools::load_all("~/Documents/Research/Code/ECE")
library(testthat)

n <- 40; L <- 2
a      <- ece_reg_row(n, L)
A_mat  <- circ(a) / n
lambda <- Re(fft(a) / n)
freqs  <- 1:(n/2 - 1)

test_that("A * theta = lambda_f * theta for sinusoidal theta", {
  for (f in freqs) {
    theta <- sin(2 * pi * f * (0:(n - 1)) / n)
    expect_equal(as.numeric(A_mat %*% theta), lambda[f + 1] * theta,
                 tolerance = 1e-10)
  }
})

var_signal_numerical <- sapply(freqs, function(f) {
  theta <- sin(2 * pi * f * (0:(n - 1)) / n)
  as.numeric(t(theta) %*% A_mat %*% A_mat %*% theta)
})
var_signal_formula <- (n / 2) * lambda[freqs + 1]^2

test_that("signal variance term matches closed form", {
  expect_equal(var_signal_numerical, var_signal_formula, tolerance = 1e-10)
})
```

## Case Study: Random Walk Mean

Suppose $\theta_x = \theta_y = \theta$ where $\theta_i = \sum_{j=1}^i \xi_j$ is a random walk with i.i.d. increments $\xi_j \sim (0, \sigma^2)$. This corresponds to scenario 13.

**Expected bias is $O(1/n)$.** Since $\theta_i - \theta_{i+k} = \xi_{i+1} + \cdots + \xi_{i+k}$ is a sum of $k$ independent increments,

$$
E\bigl[(\theta_i - \theta_{i+k})^2\bigr] = k\sigma^2,
$$

::: {.callout-note collapse="true"}
## Why $E[(\theta_i - \theta_{i+k})^2] = k\sigma^2$

The lag-$k$ difference telescopes into a sum of exactly $k$ consecutive increments:

$$\theta_i - \theta_{i+k} = -(\xi_{i+1} + \cdots + \xi_{i+k})$$

Since the $\xi_t$ are i.i.d. with mean 0 and variance $\sigma^2$, squaring and taking expectations gives:

$$E[(\theta_i - \theta_{i+k})^2] = \sum_{s=i+1}^{i+k} \sum_{t=i+1}^{i+k} E[\xi_s \xi_t] = \sum_{s=i+1}^{i+k} \sigma^2 = k\sigma^2$$

where the cross terms $E[\xi_s \xi_t] = 0$ for $s \neq t$ by independence.

Summing over $i = 1, \ldots, n-k$ (the valid pairs) then gives $E[T_k] = (n-k)k\sigma^2$. For $k \ll n$ this is approximately $nk\sigma^2$, linear in $k$.
:::

so $E[T_k(\theta,\theta)] = (n-k)k\sigma^2$, which is approximately linear in $k$ for $k \ll n$. Since the ECE is constructed to cancel exactly a linear trend in $k \mapsto T_k$,

$$
E[\theta^T A \theta] = \frac{1}{2n}\bigl(2(n-1)\sigma^2 - 2(n-2)\sigma^2\bigr) = \frac{\sigma^2}{n} \to 0.
$$

The expected bias vanishes asymptotically — the ECE is nearly unbiased in expectation for random walk means, even though random walks are not piecewise constant.

**Spectral interpretation.** The power spectral density of a random walk concentrates at low frequencies: $E[|\tilde{\theta}_j|^2] \propto 1/\sin^2(\pi j/n)$. Combined with $\lambda_j \propto -\cos(2\pi j/n)\sin^2(\pi j/n)$, the expected bias is proportional to $\sum_j \cos(2\pi j/n) = 0$, confirming the result.

**Individual realizations.** While $E[\theta^T A \theta] \approx 0$, each realization of the random walk produces a nonzero bias. The distribution of realized biases has mean $\approx \sigma^2/n$ and standard deviation that grows with $\sigma^2$.

```{r}
#| message: false
#| warning: false
devtools::load_all("~/Documents/Research/Code/ECE")
library(testthat)

n <- 100; L <- 2
a     <- ece_reg_row(n, L)
A_mat <- circ(a) / n

set.seed(42)
B      <- 5000
biases <- replicate(B, {
  theta <- cumsum(rnorm(n))
  as.numeric(t(theta) %*% A_mat %*% theta)
})

test_that("expected bias is close to sigma^2/n = 1/100", {
  expect_equal(mean(biases), 1 / n, tolerance = 0.05)
})

hist(biases,
  breaks = 60, col = "steelblue", border = "white",
  main = "Distribution of realized bias (random walk mean, n = 100)",
  xlab = "Bias"
)
abline(v = 0, col = "red", lwd = 2, lty = 2)
abline(v = mean(biases), col = "orange", lwd = 2)
legend("topright",
  legend = c("zero", "sample mean"),
  col = c("red", "orange"), lwd = 2, lty = 2:1
)
```

## Eigenvalue Visualization

Each eigenvalue $\lambda_j = \sum_{k=0}^{n-1} a_k \omega^{jk}$ is a sum of $n$ complex numbers. The plot below shows the individual terms $a_k \omega^{jk}$ for $k = 0, \ldots, n-1$ as arrows in the complex plane, for each $j$. The red star marks the sum $\lambda_j$, which is always real.

```{r}
#| message: false
#| warning: false
devtools::load_all("~/Documents/Research/Code/ECE")

n <- 10
L <- 2
a <- ece_first_row(n, L)
a[abs(a) < 1e-10] <- 0
omega <- exp(-2i * pi / n)

par(mfrow = c(2, 5), mar = c(3, 3, 2, 1))

for (j in 0:(n - 1)) {
  k     <- 0:(n - 1)
  terms <- a * omega^(j * k)
  lim   <- max(abs(c(Re(terms), Im(terms))), 0.5) * 1.3

  plot(Re(terms), Im(terms),
    xlim = c(-lim, lim), ylim = c(-lim, lim),
    xlab = "Re", ylab = "Im", asp = 1,
    main = bquote(j == .(j) ~ "," ~ lambda[.(j)] == .(round(Re(sum(terms)), 3))),
    pch = 16, col = "steelblue"
  )
  abline(h = 0, v = 0, col = "grey80", lty = 2)
  arrows(0, 0, Re(terms), Im(terms), length = 0.07, col = "steelblue", lwd = 1.2)
  points(Re(sum(terms)), Im(sum(terms)), pch = 8, col = "red", cex = 1.5, lwd = 2)
}
```
