---
title: "Piecewise Linear"
---

Recall that for any $\theta_x$,
$$
\mathbb{E}[T_k(X)] = 2n\sigma_x^2 + T_k(\theta_x).
$$
$\sigma_x^2$ is recovered once $T_k(\theta_x)$ is characterized as a function of $k$. Throughout, $T_k$ is the package's actual circular statistic (`lag_diff()`, built on `rotate()`): pairs $(i, i+k)$ with indices taken mod $n$, so index $n$ wraps to index $1$. Treat the index set as a cycle of length $n$ rather than a line — there is no distinguished start or end, only the mean's actual breakpoints.

```{r setup}
#| echo: false
#| message: false
devtools::load_all("~/Documents/Research/Code/ECE")

fit_poly <- function(Rk, L, M, n) {
  K <- outer(1:L, 0:M, `^`)
  beta <- solve(t(K) %*% K) %*% t(K) %*% Rk
  beta[1] / (2 * n)
}
```

## Piecewise Constant

As shown before, $T_k(\theta_x) = kT_1(\theta_x)$ — linear in $k$. This is the original ECE: regress $R_k=T_k(X)$ on $(1,k)$ at $k=1,2$ and read off the intercept.

```{R fig-pwc}
#| echo: false
n <- 8
h <- c(rep(-1, n/2), rep(1, n/2))
e <- rnorm(n)
X <- h + e
plot(h)
segments(1, -1, 4, -1)
segments(5,  1, 8,  1)
k <- 2; i <- 1
segments(i, h[i], i + k, h[i +k], lwd=2, col="red", lty="dashed")
segments(i, h[i], i + k, h[i], lwd=2, col="red", lty="dashed")
segments(i + k, h[i], i + k, h[i +k], lwd=2, col="red")
i <- 3
segments(i, h[i], i + k, h[i +k], lwd=2, col="blue", lty="dashed")
segments(i, h[i], i + k, h[i], lwd=2, col="blue", lty="dashed")
segments(i + k, h[i], i + k, h[i +k], lwd=2, col="blue")
i <- 4
segments(i, h[i], i + k, h[i +k], lwd=2, col="blue", lty="dashed")
segments(i, h[i], i + k, h[i], lwd=2, col="blue", lty="dashed")
segments(i + k, h[i], i + k, h[i +k], lwd=2, col="blue")
```

## Constant Slope, Any Number of Jumps

Let $\theta_x$ have one global slope $\beta_x$, with $J\ge0$ breakpoints (any positions, any jumps), consecutive breakpoints spaced $\ge L$ apart around the cycle. Every pair $(i,i+k)$ is either non-crossing, contributing $k^2\beta_x^2$ (slope only), or crossing, contributing $(k\beta_x-\delta)^2$ for whatever jump $\delta$ it straddles. Summed over $i$, both pair types give a polynomial in $k$ of degree $\le3$ with zero constant term — but the two $k^3$ coefficients are always $\mp\beta_x^2$, driven by the one shared slope, and they cancel exactly. What's left is degree $\le2$.

::: callout-important
## Theorem: Quadratic-Regression Estimator (Constant Slope)

Let $\theta_x$ have global slope $\beta_x$ and $J\ge0$ breakpoints, consecutive breakpoints spaced $\ge L\ge3$ apart around the cycle (the wrap itself counts as one gap). Fit
$$
R_k = \gamma_0+\gamma_1k+\gamma_2k^2,\qquad k=1,\dots,L,
$$
by OLS. Then $\mathbb{E}[\hat\gamma_0]=2n\sigma_x^2$ — regardless of $J$, the individual jumps, or their positions.
:::

::: {.callout-note collapse="true"}
## Proof

Non-crossing pairs in a segment of length $\ell$ contribute $(\ell-k)k^2\beta_x^2 = \ell k^2\beta_x^2 - k^3\beta_x^2$, summed once per segment; the $-k^3\beta_x^2$ coefficient is the same in every segment, independent of $J$. A crossing pair at breakpoint $j$ contributes $k(k\beta_x-\delta_j)^2 = k^3\beta_x^2 - 2k^2\beta_x\delta_j + k\delta_j^2$: a $+k^3\beta_x^2$ that exactly cancels one unit of the non-crossing $-k^3\beta_x^2$ deficit. With $J$ real breakpoints plus the wrap (which, spaced $\ge L$ from the rest, behaves exactly like one more crossing point with some effective jump $\delta_0$), there are $J+1$ crossings and the $k^3$ terms cancel completely, leaving an exact degree-$\le2$, zero-constant-term polynomial in $k$. Since $R_k=2n\sigma_x^2+T_k(\theta_x)$ exactly, the quadratic design matrix correctly specifies the mean model and OLS is unbiased for $\hat\gamma_0$. $\qquad\square$

```{r pwlin-quad-proof-check}
#| message: false
library(testthat)

n <- 200
s <- -0.7
breakpoints <- c(40, 90, 150)
deltas <- c(5, -8, 12)
theta <- s * (1:n)
for (j in seq_along(breakpoints)) theta[(breakpoints[j] + 1):n] <- theta[(breakpoints[j] + 1):n] + deltas[j]

sigma2 <- 1
L <- 8
Tk_exact <- sapply(1:L, function(k) lag_diff(theta, k = k))
Rk_exact <- 2 * n * sigma2 + Tk_exact

test_that("T_k(theta_x) is exactly quadratic (zero constant term) with three breakpoints, constant slope", {
  K2 <- outer(1:L, 1:2, `^`)
  beta2 <- solve(t(K2) %*% K2) %*% t(K2) %*% Tk_exact
  expect_equal(as.numeric(K2 %*% beta2), Tk_exact, tolerance = 1e-6)
})

test_that("the quadratic-regression intercept recovers 2n sigma_x^2 exactly, no noise", {
  expect_equal(fit_poly(Rk_exact, L, 2, n), sigma2)
})
```
:::

### Simulation

$n=200$, breakpoints at $40,90,150$, slope $\beta_x=-0.7$, jumps $(5,-8,12)$, $\sigma_x^2=1$, $L=4$, $1000$ replicates. Comparing Linear ($M=1$) against the Theorem's Quadratic ($M=2$); Cubic would also be unbiased here but wastes a degree of freedom on a $k^3$ term that isn't needed.

```{r pwlin-quad-setup}
#| echo: false
#| cache: true
#| dependson: ["setup"]
n <- 200
breakpoints <- c(40, 90, 150)
deltas <- c(5, -8, 12)
s <- -0.7
sigma <- 1
L <- 4
reps <- 1000

theta <- s * (1:n)
for (j in seq_along(breakpoints)) theta[(breakpoints[j] + 1):n] <- theta[(breakpoints[j] + 1):n] + deltas[j]

fit_labs2 <- c("Linear (M=1)", "Quadratic (M=2)")
fit_cols2 <- c("Linear (M=1)" = "black", "Quadratic (M=2)" = "steelblue")
fit_ltys2 <- c("Linear (M=1)" = "solid", "Quadratic (M=2)" = "dashed")

set.seed(1)
X_example <- theta + rnorm(n, sd = sigma)

set.seed(1)
Tk_mat <- t(sapply(1:reps, function(r) {
  X <- theta + rnorm(n, sd = sigma)
  sapply(1:L, function(k) lag_diff(X, k = k))
}))

ests <- t(apply(Tk_mat, 1, function(Tk) c(fit_poly(Tk, L, 1, n), fit_poly(Tk, L, 2, n))))
colnames(ests) <- fit_labs2
```

```{r pwlin-quad-mean-noise}
#| echo: false
#| dependson: ["pwlin-quad-setup"]
tibble::tibble(i = 1:n, theta = theta, x = X_example) |>
  ggplot2::ggplot(ggplot2::aes(x = i)) +
  ggplot2::geom_line(ggplot2::aes(y = x), color = "grey80") +
  ggplot2::geom_line(ggplot2::aes(y = theta), color = "steelblue", linewidth = 1) +
  ggplot2::labs(x = "Index i", y = "Value") +
  ggplot2::theme_minimal()
```

```{r pwlin-quad-regression}
#| echo: false
#| dependson: ["pwlin-quad-setup"]
Rk_noiseless <- sapply(1:L, function(k) lag_diff(theta, k = k)) # X = theta exactly, no noise: true sigma_x^2 = 0

K_seq <- seq(0, L, length.out = 200)
fits <- lapply(1:2, function(M) {
  K <- outer(1:L, 0:M, `^`)
  beta <- solve(t(K) %*% K) %*% t(K) %*% Rk_noiseless
  Kseq <- outer(K_seq, 0:M, `^`)
  as.numeric(Kseq %*% beta) / (2 * n)
})
names(fits) <- fit_labs2

lines_df <- tibble::tibble(k = rep(K_seq, 2), fit = rep(fit_labs2, each = length(K_seq)), val = unlist(fits))
points_df <- tibble::tibble(k = 1:L, val = Rk_noiseless / (2 * n))

ggplot2::ggplot() +
  ggplot2::geom_hline(yintercept = 0, color = "grey40") +
  ggplot2::geom_vline(xintercept = 0, color = "grey40") +
  ggplot2::geom_line(data = lines_df, ggplot2::aes(x = k, y = val, color = fit, linetype = fit), linewidth = 0.9) +
  ggplot2::geom_point(data = points_df, ggplot2::aes(x = k, y = val), size = 2) +
  ggplot2::geom_point(ggplot2::aes(x = 0, y = 0), color = "red", size = 3) +
  ggplot2::scale_color_manual(values = fit_cols2, name = NULL) +
  ggplot2::scale_linetype_manual(values = fit_ltys2, name = NULL) +
  ggplot2::scale_x_continuous(breaks = 0:L) +
  ggplot2::labs(x = "Lag k", y = expression(R[k] / (2 * n) ~ "(no noise:" ~ X == theta[x] * ")")) +
  ggplot2::theme_minimal()
```

No noise: $X=\theta_x$ exactly, true $\sigma_x^2=0$ (red point). Quadratic passes through every point and through $0$; Linear extrapolates to a nonzero, misspecification-driven value.

```{r pwlin-quad-histogram}
#| echo: false
#| dependson: ["pwlin-quad-setup"]
hist_df <- tibble::tibble(val = as.vector(ests), fit = rep(fit_labs2, each = nrow(ests)))
means_df <- tibble::tibble(fit = fit_labs2, mean_val = colMeans(ests))

ggplot2::ggplot(hist_df, ggplot2::aes(x = val, fill = fit)) +
  ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)), bins = 40, color = "white", alpha = 0.6, position = "identity") +
  ggplot2::geom_vline(xintercept = sigma^2, color = "red") +
  ggplot2::geom_vline(data = means_df, ggplot2::aes(xintercept = mean_val, color = fit), linetype = "dashed") +
  ggplot2::scale_fill_manual(values = fit_cols2, name = NULL) +
  ggplot2::scale_color_manual(values = fit_cols2, guide = "none") +
  ggplot2::labs(x = expression(hat(sigma)[x]^2), y = "density") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r pwlin-quad-convergence}
#| echo: false
#| cache: true
#| dependson: ["pwlin-quad-setup"]
n_grid <- c(20, 50, 100, 200, 500, 1000, 2000)
reps_n <- 500
bp_frac <- c(0.2, 0.45, 0.75)

sweep_df2 <- purrr::map_dfr(n_grid, function(n_val) {
  breakpoints_val <- pmax(1, pmin(n_val - 1, round(bp_frac * n_val)))
  theta_val <- s * (1:n_val)
  for (j in seq_along(breakpoints_val)) theta_val[(breakpoints_val[j] + 1):n_val] <- theta_val[(breakpoints_val[j] + 1):n_val] + deltas[j]
  set.seed(1)
  ests_val <- t(sapply(1:reps_n, function(r) {
    X <- theta_val + rnorm(n_val, sd = sigma)
    Tk <- sapply(1:L, function(k) lag_diff(X, k = k))
    c(fit_poly(Tk, L, 1, n_val), fit_poly(Tk, L, 2, n_val))
  }))
  colnames(ests_val) <- fit_labs2
  tibble::tibble(n = n_val, fit = fit_labs2, est = colMeans(ests_val), mse = colMeans((ests_val - sigma^2)^2))
})
```

```{r pwlin-quad-convergence-plot}
#| echo: false
#| dependson: ["pwlin-quad-convergence"]
p1 <- ggplot2::ggplot(sweep_df2, ggplot2::aes(x = n, y = est, color = fit)) +
  ggplot2::geom_hline(yintercept = sigma^2, color = "red", linetype = "dashed") +
  ggplot2::geom_line() + ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(trans = "log10", breaks = n_grid) +
  ggplot2::scale_color_manual(values = fit_cols2, name = NULL) +
  ggplot2::labs(x = "Sample size n (log scale)", y = expression(hat(sigma)[x]^2)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

p2 <- ggplot2::ggplot(sweep_df2, ggplot2::aes(x = n, y = mse, color = fit)) +
  ggplot2::geom_line() + ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(trans = "log10", breaks = n_grid) +
  ggplot2::scale_y_continuous(trans = "log10") +
  ggplot2::scale_color_manual(values = fit_cols2, name = NULL) +
  ggplot2::labs(x = "Sample size n (log scale)", y = "MSE (log scale)") +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

patchwork::wrap_plots(p1, p2, guides = "collect") & ggplot2::theme(legend.position = "bottom")
```

Quadratic sits on the truth at every $n$ (guaranteed by the Theorem); Linear's bias never vanishes, since even the leading term of $T_k(\theta_x)$ is quadratic in $k$. Quadratic's MSE also falls with $n$; Linear's is dominated by squared bias and barely moves.

## Varying Slopes (General Piecewise Linear)

Let each segment have its own slope $\beta_x^{(0)},\dots,\beta_x^{(J)}$. The crossing-pair argument still gives a degree-$\le3$, zero-constant polynomial per breakpoint (now a sum of squares of an affine function of position within the crossing window), but the cancellation above no longer applies: the non-crossing $k^3$ coefficient is a mix of every segment's own $\beta_x^{(j)2}$, while the wrap's crossing term only involves the two segments touching the seam. These don't match in general — not even when the two boundary segments happen to share a slope, since interior segments still contribute uncancelled $k^3$ terms. Cubic regression is needed, exactly as it would be without circular indexing.

::: callout-important
## Theorem: Cubic-Regression Estimator (General Piecewise-Linear)

Let $\theta_x$ have breakpoints $m_1,\dots,m_J$ with arbitrary per-segment slopes and jumps, consecutive breakpoints (including the wrap) spaced $\ge L\ge4$ apart around the cycle. Fit
$$
R_k = \gamma_0+\gamma_1k+\gamma_2k^2+\gamma_3k^3,\qquad k=1,\dots,L,
$$
by OLS. Then $\mathbb{E}[\hat\gamma_0]=2n\sigma_x^2$ — regardless of $J$ or how slopes and jumps vary segment to segment.
:::

::: {.callout-note collapse="true"}
## Proof

Each of the $J+1$ crossings (the $J$ real breakpoints, plus the wrap treated as one more) contributes a degree-$\le3$, zero-constant-term polynomial in $k$ — non-crossing pairs by the Linear-case argument, crossing pairs by the sum-of-squares-of-an-affine-function argument used for non-constant slope. A finite sum of such polynomials is itself one, so $T_k(\theta_x)$ is exactly cubic in $k$ with zero constant term. $R_k=2n\sigma_x^2+T_k(\theta_x)$ exactly, so the cubic design matrix correctly specifies the mean model and OLS is unbiased for $\hat\gamma_0$. $\qquad\square$

```{r pwlin-cubic-proof-check}
#| message: false
library(testthat)

build_theta <- function(n, breakpoints, slopes, jumps) {
  step_idx <- findInterval(1:(n - 1), c(0, breakpoints, n), rightmost.closed = TRUE)
  theta <- c(0, cumsum(slopes[step_idx]))
  for (j in seq_along(breakpoints)) theta[(breakpoints[j] + 1):n] <- theta[(breakpoints[j] + 1):n] + jumps[j]
  theta
}

n <- 200
L <- 8
breakpoints <- c(40, 90, 150)
slopes <- c(0.3, -0.9, 1.4, 0.1)
jumps <- c(6, -10, 15)
sigma2 <- 1

theta <- build_theta(n, breakpoints, slopes, jumps)
Tk_exact <- sapply(1:L, function(k) lag_diff(theta, k = k))
Rk_exact <- 2 * n * sigma2 + Tk_exact

test_that("T_k(theta_x) needs the full cubic once slopes vary by segment", {
  K2 <- outer(1:L, 1:2, `^`)
  beta2 <- solve(t(K2) %*% K2) %*% t(K2) %*% Tk_exact
  expect_true(max(abs(Tk_exact - K2 %*% beta2)) > 1)
})

test_that("the cubic-regression intercept recovers 2n sigma_x^2 exactly, no noise", {
  expect_equal(fit_poly(Rk_exact, L, 3, n), sigma2)
})
```
:::

### Simulation

Same breakpoints ($n=200$, at $40,90,150$), each segment its own slope ($0.3,-0.9,1.4,0.1$), each breakpoint its own jump ($6,-10,15$), $\sigma_x^2=1$, $L=4$, $1000$ replicates. Comparing Linear ($M=1$), Quadratic ($M=2$) — now genuinely biased — and Cubic ($M=3$).

```{r pwlin-cubic-setup}
#| echo: false
#| cache: true
#| dependson: ["setup"]
n <- 200
breakpoints <- c(40, 90, 150)
slopes <- c(0.3, -0.9, 1.4, 0.1)
jumps <- c(6, -10, 15)
sigma <- 1
L <- 4
reps <- 1000

build_theta <- function(n, breakpoints, slopes, jumps) {
  step_idx <- findInterval(1:(n - 1), c(0, breakpoints, n), rightmost.closed = TRUE)
  theta <- c(0, cumsum(slopes[step_idx]))
  for (j in seq_along(breakpoints)) theta[(breakpoints[j] + 1):n] <- theta[(breakpoints[j] + 1):n] + jumps[j]
  theta
}
theta <- build_theta(n, breakpoints, slopes, jumps)

fit_labs3 <- c("Linear (M=1)", "Quadratic (M=2)", "Cubic (M=3)")
fit_cols3 <- c("Linear (M=1)" = "black", "Quadratic (M=2)" = "steelblue", "Cubic (M=3)" = "firebrick")
fit_ltys3 <- c("Linear (M=1)" = "solid", "Quadratic (M=2)" = "dashed", "Cubic (M=3)" = "dotted")

set.seed(1)
X_example <- theta + rnorm(n, sd = sigma)

set.seed(1)
Tk_mat <- t(sapply(1:reps, function(r) {
  X <- theta + rnorm(n, sd = sigma)
  sapply(1:L, function(k) lag_diff(X, k = k))
}))

ests <- t(apply(Tk_mat, 1, function(Tk) c(fit_poly(Tk, L, 1, n), fit_poly(Tk, L, 2, n), fit_poly(Tk, L, 3, n))))
colnames(ests) <- fit_labs3
```

```{r pwlin-cubic-mean-noise}
#| echo: false
#| dependson: ["pwlin-cubic-setup"]
tibble::tibble(i = 1:n, theta = theta, x = X_example) |>
  ggplot2::ggplot(ggplot2::aes(x = i)) +
  ggplot2::geom_line(ggplot2::aes(y = x), color = "grey80") +
  ggplot2::geom_line(ggplot2::aes(y = theta), color = "steelblue", linewidth = 1) +
  ggplot2::labs(x = "Index i", y = "Value") +
  ggplot2::theme_minimal()
```

```{r pwlin-cubic-regression}
#| echo: false
#| dependson: ["pwlin-cubic-setup"]
Rk_noiseless <- sapply(1:L, function(k) lag_diff(theta, k = k)) # X = theta exactly, no noise: true sigma_x^2 = 0

K_seq <- seq(0, L, length.out = 200)
fits <- lapply(1:3, function(M) {
  K <- outer(1:L, 0:M, `^`)
  beta <- solve(t(K) %*% K) %*% t(K) %*% Rk_noiseless
  Kseq <- outer(K_seq, 0:M, `^`)
  as.numeric(Kseq %*% beta) / (2 * n)
})
names(fits) <- fit_labs3

lines_df <- tibble::tibble(k = rep(K_seq, 3), fit = rep(fit_labs3, each = length(K_seq)), val = unlist(fits))
points_df <- tibble::tibble(k = 1:L, val = Rk_noiseless / (2 * n))

ggplot2::ggplot() +
  ggplot2::geom_hline(yintercept = 0, color = "grey40") +
  ggplot2::geom_vline(xintercept = 0, color = "grey40") +
  ggplot2::geom_line(data = lines_df, ggplot2::aes(x = k, y = val, color = fit, linetype = fit), linewidth = 0.9) +
  ggplot2::geom_point(data = points_df, ggplot2::aes(x = k, y = val), size = 2) +
  ggplot2::geom_point(ggplot2::aes(x = 0, y = 0), color = "red", size = 3) +
  ggplot2::scale_color_manual(values = fit_cols3, name = NULL) +
  ggplot2::scale_linetype_manual(values = fit_ltys3, name = NULL) +
  ggplot2::scale_x_continuous(breaks = 0:L) +
  ggplot2::labs(x = "Lag k", y = expression(R[k] / (2 * n) ~ "(no noise:" ~ X == theta[x] * ")")) +
  ggplot2::theme_minimal()
```

No noise, true $\sigma_x^2=0$. Cubic passes through every point and through $0$; Linear and Quadratic both extrapolate to spurious nonzero values.

```{r pwlin-cubic-histogram}
#| echo: false
#| dependson: ["pwlin-cubic-setup"]
hist_df <- tibble::tibble(val = as.vector(ests), fit = rep(fit_labs3, each = nrow(ests)))
means_df <- tibble::tibble(fit = fit_labs3, mean_val = colMeans(ests))

ggplot2::ggplot(hist_df, ggplot2::aes(x = val, fill = fit)) +
  ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)), bins = 40, color = "white", alpha = 0.6, position = "identity") +
  ggplot2::geom_vline(xintercept = sigma^2, color = "red") +
  ggplot2::geom_vline(data = means_df, ggplot2::aes(xintercept = mean_val, color = fit), linetype = "dashed") +
  ggplot2::scale_fill_manual(values = fit_cols3, name = NULL) +
  ggplot2::scale_color_manual(values = fit_cols3, guide = "none") +
  ggplot2::labs(x = expression(hat(sigma)[x]^2), y = "density") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r pwlin-cubic-convergence}
#| echo: false
#| cache: true
#| dependson: ["pwlin-cubic-setup"]
n_grid <- c(20, 50, 100, 200, 500, 1000, 2000)
reps_n <- 500
bp_frac <- c(0.2, 0.45, 0.75)

sweep_df3 <- purrr::map_dfr(n_grid, function(n_val) {
  breakpoints_val <- pmax(1, pmin(n_val - 1, round(bp_frac * n_val)))
  theta_val <- build_theta(n_val, breakpoints_val, slopes, jumps)
  set.seed(1)
  ests_val <- t(sapply(1:reps_n, function(r) {
    X <- theta_val + rnorm(n_val, sd = sigma)
    Tk <- sapply(1:L, function(k) lag_diff(X, k = k))
    c(fit_poly(Tk, L, 1, n_val), fit_poly(Tk, L, 2, n_val), fit_poly(Tk, L, 3, n_val))
  }))
  colnames(ests_val) <- fit_labs3
  tibble::tibble(n = n_val, fit = fit_labs3, est = colMeans(ests_val), mse = colMeans((ests_val - sigma^2)^2))
})
```

```{r pwlin-cubic-convergence-plot}
#| echo: false
#| dependson: ["pwlin-cubic-convergence"]
p1 <- ggplot2::ggplot(sweep_df3, ggplot2::aes(x = n, y = est, color = fit)) +
  ggplot2::geom_hline(yintercept = sigma^2, color = "red", linetype = "dashed") +
  ggplot2::geom_line() + ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(trans = "log10", breaks = n_grid) +
  ggplot2::scale_color_manual(values = fit_cols3, name = NULL) +
  ggplot2::labs(x = "Sample size n (log scale)", y = expression(hat(sigma)[x]^2)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

p2 <- ggplot2::ggplot(sweep_df3, ggplot2::aes(x = n, y = mse, color = fit)) +
  ggplot2::geom_line() + ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(trans = "log10", breaks = n_grid) +
  ggplot2::scale_y_continuous(trans = "log10") +
  ggplot2::scale_color_manual(values = fit_cols3, name = NULL) +
  ggplot2::labs(x = "Sample size n (log scale)", y = "MSE (log scale)") +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

patchwork::wrap_plots(p1, p2, guides = "collect") & ggplot2::theme(legend.position = "bottom")
```

Cubic sits on the truth at every $n$; Linear and Quadratic don't. At $L=4$, Cubic pays for its correctness with the highest MSE of the three (zero residual degrees of freedom) — resolved in the next section by letting $L$ grow.

## Larger L: Quadratic vs. Cubic at Their Best

$L=4$ is the minimum that identifies Cubic at all, maximizing its variance, while it's also small enough that Quadratic's missing $k^3$ term hasn't had room to matter. Fixed $n=50$, one breakpoint at $m=25$ with different slopes on each side ($\beta_x^{(0)}=-2$, $\beta_x^{(1)}=3$) and a jump ($\delta_x=30$), $\sigma_x^2=1$, $1000$ replicates, $L$ up to $24$ (the spacing limit given $m=25$, $n-m=25$).

```{r pwlin-L-sweep}
#| echo: false
#| cache: true
#| dependson: ["setup"]
n <- 50
m <- 25
b1 <- -2
b2 <- 3
delta <- 30
sigma <- 1
reps <- 1000

theta <- numeric(n)
theta[1:m] <- b1 * (1:m)
theta[(m + 1):n] <- theta[m] + delta + b2 * (1:(n - m))

L_grid <- c(4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)
fit_labs_L <- c("Quadratic (M=2)", "Cubic (M=3)")
fit_cols_L <- c("Quadratic (M=2)" = "steelblue", "Cubic (M=3)" = "firebrick")

sweep_L <- purrr::map_dfr(L_grid, function(Lv) {
  set.seed(1)
  ests <- t(sapply(1:reps, function(r) {
    X <- theta + rnorm(n, sd = sigma)
    Tk <- sapply(1:Lv, function(k) lag_diff(X, k = k))
    c(fit_poly(Tk, Lv, 2, n), fit_poly(Tk, Lv, 3, n))
  }))
  colnames(ests) <- fit_labs_L
  tibble::tibble(L = Lv, fit = fit_labs_L, est = colMeans(ests), mse = colMeans((ests - sigma^2)^2))
})
```

```{r pwlin-L-mean-noise}
#| echo: false
#| dependson: ["pwlin-L-sweep"]
set.seed(1)
X_example <- theta + rnorm(n, sd = sigma)

tibble::tibble(i = 1:n, theta = theta, x = X_example) |>
  ggplot2::ggplot(ggplot2::aes(x = i)) +
  ggplot2::geom_line(ggplot2::aes(y = x), color = "grey80") +
  ggplot2::geom_line(ggplot2::aes(y = theta), color = "steelblue", linewidth = 1) +
  ggplot2::labs(x = "Index i", y = "Value") +
  ggplot2::theme_minimal()
```

```{r pwlin-L-sweep-plot}
#| echo: false
#| dependson: ["pwlin-L-sweep"]
p1 <- ggplot2::ggplot(sweep_L, ggplot2::aes(x = L, y = est, color = fit)) +
  ggplot2::geom_hline(yintercept = sigma^2, color = "red", linetype = "dashed") +
  ggplot2::geom_line() + ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(breaks = L_grid) +
  ggplot2::scale_color_manual(values = fit_cols_L, name = NULL) +
  ggplot2::labs(x = "Lag window L", y = expression(hat(sigma)[x]^2)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

p2 <- ggplot2::ggplot(sweep_L, ggplot2::aes(x = L, y = mse, color = fit)) +
  ggplot2::geom_line() + ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(breaks = L_grid) +
  ggplot2::scale_y_continuous(trans = "log10") +
  ggplot2::scale_color_manual(values = fit_cols_L, name = NULL) +
  ggplot2::labs(x = "Lag window L", y = "MSE (log scale)") +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

patchwork::wrap_plots(p1, p2, guides = "collect") & ggplot2::theme(legend.position = "bottom")
```

Larger $L$ resolves the tradeoff in Cubic's favor. Cubic's bias stays pinned near the truth throughout (guaranteed by the Theorem); its MSE falls as $L$ grows past $4$, gaining residual degrees of freedom. Quadratic goes the other way: its bias runs from about $0.2$ at $L=4$ to about $-72$ at $L=24$, and its MSE grows by roughly three orders of magnitude over the same range, because the missing $k^3$ term costs more exactly where the true cubic curvature is most pronounced. The MSE crossover is around $L=7$ — past that point Cubic dominates on both bias and MSE.

## Many Breakpoints: Choosing a Good L

A single breakpoint gives Cubic unlimited room to grow $L$ (up to $\min(m,n-m)$). With breakpoints packed every $20$ points, $L$ has a hard ceiling: the spacing condition needs $L\le20$, so the "grow $L$" strategy above can only be pushed so far before it breaks the Theorem's own assumption. $n=400$, breakpoints at $20,40,\dots,380$ ($19$ of them, every segment length $20$), a random slope per segment and a random jump per breakpoint (seed fixed for reproducibility), $\sigma_x^2=1$, $1000$ replicates.

```{r pwlin-many-bp-setup}
#| echo: false
#| cache: true
#| dependson: ["setup"]
n <- 400
breakpoints <- seq(20, 380, by = 20)
J <- length(breakpoints)
set.seed(42)
slopes <- runif(J + 1, -1, 1)
jumps <- runif(J, -8, 8)
sigma <- 1
reps <- 1000

build_theta <- function(n, breakpoints, slopes, jumps) {
  step_idx <- findInterval(1:(n - 1), c(0, breakpoints, n), rightmost.closed = TRUE)
  theta <- c(0, cumsum(slopes[step_idx]))
  for (j in seq_along(breakpoints)) theta[(breakpoints[j] + 1):n] <- theta[(breakpoints[j] + 1):n] + jumps[j]
  theta
}
theta <- build_theta(n, breakpoints, slopes, jumps)

L_grid <- c(4, 6, 8, 10, 12, 14, 16, 18, 19, 20, 21)
fit_labs_L2 <- c("Quadratic (M=2)", "Cubic (M=3)")
fit_cols_L2 <- c("Quadratic (M=2)" = "steelblue", "Cubic (M=3)" = "firebrick")

sweep_L2 <- purrr::map_dfr(L_grid, function(Lv) {
  set.seed(1)
  ests <- t(sapply(1:reps, function(r) {
    X <- theta + rnorm(n, sd = sigma)
    Tk <- sapply(1:Lv, function(k) lag_diff(X, k = k))
    c(fit_poly(Tk, Lv, 2, n), fit_poly(Tk, Lv, 3, n))
  }))
  colnames(ests) <- fit_labs_L2
  tibble::tibble(L = Lv, fit = fit_labs_L2, est = colMeans(ests), mse = colMeans((ests - sigma^2)^2))
})
```

```{r pwlin-many-bp-mean-noise}
#| echo: false
#| dependson: ["pwlin-many-bp-setup"]
set.seed(1)
X_example <- theta + rnorm(n, sd = sigma)

tibble::tibble(i = 1:n, theta = theta, x = X_example) |>
  ggplot2::ggplot(ggplot2::aes(x = i)) +
  ggplot2::geom_line(ggplot2::aes(y = x), color = "grey80") +
  ggplot2::geom_line(ggplot2::aes(y = theta), color = "steelblue", linewidth = 1) +
  ggplot2::labs(x = "Index i", y = "Value") +
  ggplot2::theme_minimal()
```

```{r pwlin-many-bp-plot}
#| echo: false
#| dependson: ["pwlin-many-bp-setup"]
p1 <- ggplot2::ggplot(sweep_L2, ggplot2::aes(x = L, y = est, color = fit)) +
  ggplot2::geom_hline(yintercept = sigma^2, color = "red", linetype = "dashed") +
  ggplot2::geom_vline(xintercept = 20, color = "grey50", linetype = "dotted") +
  ggplot2::geom_line() + ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(breaks = L_grid) +
  ggplot2::scale_color_manual(values = fit_cols_L2, name = NULL) +
  ggplot2::labs(x = "Lag window L", y = expression(hat(sigma)[x]^2)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

p2 <- ggplot2::ggplot(sweep_L2, ggplot2::aes(x = L, y = mse, color = fit)) +
  ggplot2::geom_vline(xintercept = 20, color = "grey50", linetype = "dotted") +
  ggplot2::geom_line() + ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(breaks = L_grid) +
  ggplot2::scale_y_continuous(trans = "log10") +
  ggplot2::scale_color_manual(values = fit_cols_L2, name = NULL) +
  ggplot2::labs(x = "Lag window L", y = "MSE (log scale)") +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

patchwork::wrap_plots(p1, p2, guides = "collect") & ggplot2::theme(legend.position = "bottom")
```

The dotted line at $L=20$ marks the spacing limit. Inside it, Cubic's bias stays flat at the truth and its MSE keeps falling with $L$, same story as the single-breakpoint sweep. At $L=21$ the window starts crossing two breakpoints at once — outside every guarantee in this chapter — and Cubic's own bias reappears (jumping to about $1.8$, MSE almost tripling); this isn't a bug, it's the assumption being violated on schedule. Quadratic's MSE is actually *lower* than Cubic's for small $L$ ($L\lesssim10$ here) — the same finite-sample story as the $L=4$ single-breakpoint case, since Cubic still has little room to average at small $L$ — but it crosses over by $L\approx12$ and gets steadily worse as $L$ grows, since it's misspecified for this varying-slope mean regardless of breakpoint spacing.

```{r pwlin-many-bp-histogram}
#| echo: false
#| cache: true
#| dependson: ["pwlin-many-bp-setup"]
Ls_hist <- c(4, 16)
hist_data <- purrr::map_dfr(Ls_hist, function(Lv) {
  set.seed(1)
  ests <- t(sapply(1:reps, function(r) {
    X <- theta + rnorm(n, sd = sigma)
    Tk <- sapply(1:Lv, function(k) lag_diff(X, k = k))
    c(fit_poly(Tk, Lv, 2, n), fit_poly(Tk, Lv, 3, n))
  }))
  colnames(ests) <- fit_labs_L2
  tibble::tibble(L = Lv, fit = rep(fit_labs_L2, each = nrow(ests)), val = as.vector(ests))
})
means_df <- hist_data |> dplyr::summarise(mean_val = mean(val), .by = c(L, fit))

ggplot2::ggplot(hist_data, ggplot2::aes(x = val, fill = fit)) +
  ggplot2::geom_histogram(ggplot2::aes(y = ggplot2::after_stat(density)), bins = 40, color = "white", alpha = 0.6, position = "identity") +
  ggplot2::geom_vline(xintercept = sigma^2, color = "red") +
  ggplot2::geom_vline(data = means_df, ggplot2::aes(xintercept = mean_val, color = fit), linetype = "dashed") +
  ggplot2::scale_fill_manual(values = fit_cols_L2, name = NULL) +
  ggplot2::scale_color_manual(values = fit_cols_L2, guide = "none") +
  ggplot2::facet_wrap(~ paste0("L = ", L), scales = "free_x") +
  ggplot2::labs(x = expression(hat(sigma)[x]^2), y = "density") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

At $L=4$, Cubic's histogram is visibly wider than Quadratic's (SD $\approx1.74$ vs.\ $0.73$, roughly $2.4\times$; range about $-4.6$ to $6.5$ over $1000$ replicates) — the interpolation-extrapolation instability from the weight-vector argument earlier, seen directly rather than summarized as a single MSE number. By $L=16$, Cubic's spread has narrowed to about the same width as Quadratic's (SD $\approx0.68$), but stays centered on $\sigma_x^2=1$; Quadratic's histogram is narrower ($\text{SD}\approx0.61$) yet centered near $0.2$, visibly off the truth. This is the same crossover the MSE plot shows, just seen directly in the sampling distributions rather than summarized as a single number.

**A good choice here is $L=16$.** It sits comfortably past the MSE crossover (so Cubic already beats Quadratic), it's close to the flat part of Cubic's MSE curve (little left to gain by pushing further), and it leaves a $4$-point margin below the hard spacing limit at $L=20$ — enough room that being off by a breakpoint or two in practice (spacing isn't always exactly $20$) doesn't immediately break unbiasedness. The general rule this example illustrates: pick $L$ as large as the *minimum* breakpoint spacing comfortably allows, not the largest $L$ that happens to still validate — the MSE gains from larger $L$ taper off well before the validity boundary, so there's little reason to run right up against it.
