---
title: "Power"
---

```{R}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
devtools::load_all()
results <- read_csv("../data/results.csv")

method_colors <- c(
  "z.test.gaussian" = "#E41A1C",
  "z.test.kappa"    = "#377EB8",
  "bs.multiplier"   = "#4DAF4A",
  "bs.parametric"   = "#FF7F00"
)
```

## Z-Test: Gaussian vs. Kappa

We compare `z.test.gaussian` and `z.test.kappa` at $n = 200$, $p = 2$, across scenarios and error distributions.

```{r}
#| echo: false
#| results: asis
dat <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.gaussian", "z.test.kappa"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range <- range(dat$metric_val)

plots <- dat |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors) +
      scale_y_continuous(limits = y_range) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

## Z-Test vs. Bootstrap Methods

We compare all four methods at $n = 200$, $p = 2$, across scenarios and error distributions.

```{r}
#| echo: false
#| results: asis
dat2 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.gaussian", "z.test.kappa", "bs.multiplier", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range2 <- range(dat2$metric_val)

plots2 <- dat2 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors) +
      scale_y_continuous(limits = y_range2) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat2$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots2, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

## Variance Difference: General vs. Gaussian

Recall the asymptotic variance under the general and Gaussian assumptions:
$$
V = \frac{1}{n}\left[(1-\rho^2)\!\left(\frac{\omega_x^2}{\sigma_x^2} + \frac{\omega_y^2}{\sigma_y^2} - \frac{2\rho\,\omega_{xy}}{\sigma_x\sigma_y}\right) + \frac{\rho^2}{4}(\kappa_{40}+2\kappa_{22}+\kappa_{04}) - \rho(\kappa_{31}+\kappa_{13}) + \frac{5}{2}(1-\rho^2)^2 + \kappa_{22}\right],
$$
$$
V_G = \frac{1}{n}(1-\rho^2)\!\left(\frac{\omega_x^2}{\sigma_x^2} + \frac{\omega_y^2}{\sigma_y^2} - \frac{2\rho\,\omega_{xy}}{\sigma_x\sigma_y} + \frac{7}{2}(1-\rho^2)\right).
$$
The jump-noise terms cancel, giving
$$
V - V_G = \frac{1}{n}\left[\frac{\rho^2}{4}(\kappa_{40}+2\kappa_{22}+\kappa_{04}) - \rho(\kappa_{31}+\kappa_{13}) + \kappa_{22} - (1-\rho^2)^2\right].
$$
This is zero under the Gaussian assumption and measures the excess variance due to non-Gaussianity.

## Cumulants by Distribution

The magnitude of $V - V_G$ depends on how far the joint cumulants depart from their Gaussian values. The tables below report $(\kappa_{40}, \kappa_{31}, \kappa_{22}, \kappa_{13}, \kappa_{04})$ and $n(V - V_G)$ for each supported error distribution, estimated via $10^7$ samples.

**$\rho = 0.3$**

```{r}
#| message: false
#| echo: false
#| cache: true
devtools::load_all("~/Documents/Research/Code/ECE")

err_types <- c("normal", "exponential", "t", "lognormal")

make_kappa_table <- function(rho) {
  base <- scenario(1, n = 100, r = rho, err_type = "normal")
  rows <- lapply(err_types, \(et) {
    p <- modifyList(base, list(err_type = et))
    p$kappa <- kappa_from_params(p)
    k <- p$kappa
    diff <- rho^2/4 * (k$k40 + 2*k$k22 + k$k04) - rho*(k$k31 + k$k13) +
            k$k22 - (1 - rho^2)^2
    data.frame(
      Distribution = et,
      k40 = k$k40, k31 = k$k31, k22 = k$k22, k13 = k$k13, k04 = k$k04,
      diff = diff
    )
  })
  do.call(rbind, rows)
}

knitr::kable(make_kappa_table(0.3), digits = 3,
  col.names = c("Distribution", "$\\kappa_{40}$", "$\\kappa_{31}$",
                "$\\kappa_{22}$", "$\\kappa_{13}$", "$\\kappa_{04}$",
                "$n(V - V_G)$"))
```

**$\rho = 0$**

```{r}
#| message: false
#| echo: false
#| cache: true
knitr::kable(make_kappa_table(0), digits = 3,
  col.names = c("Distribution", "$\\kappa_{40}$", "$\\kappa_{31}$",
                "$\\kappa_{22}$", "$\\kappa_{13}$", "$\\kappa_{04}$",
                "$n(V - V_G)$"))
```

::: {.callout-note collapse="true"}
## Computational verification

```{r}
#| message: false
devtools::load_all("~/Documents/Research/Code/ECE")

params <- scenario(2, n = 1000, L = 4, r = 0.3, err_type = "exponential")

n   <- params$n
sx  <- sqrt(params$S[1, 1])
sy  <- sqrt(params$S[2, 2])
sxy <- params$S[1, 2]
rho <- sxy / (sx * sy)
hx  <- params$h[, 1]
hy  <- params$h[, 2]
k   <- params$kappa

omega <- (1 - rho^2) * (
  lag_diff(hx) / (n * sx^2) +
  lag_diff(hy) / (n * sy^2) -
  2 * rho * lag_diff(hx, hy) / (n * sx * sy)
)

V_general <- (1/n) * (
  omega +
  rho^2/4 * (k$k40 + 2*k$k22 + k$k04) - rho*(k$k31 + k$k13) +
  5/2 * (1 - rho^2)^2 + k$k22
)
V_gaussian <- (1/n) * (omega + 7/2 * (1 - rho^2)^2)

data.frame(
  V_diff_direct  = V_general - V_gaussian,
  V_diff_formula = (1/n) * (
    rho^2/4 * (k$k40 + 2*k$k22 + k$k04) - rho*(k$k31 + k$k13) +
    k$k22 - (1 - rho^2)^2
  )
)
```
:::

