---
title: "Power"
---

```{R}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
devtools::load_all()
results <- read_csv("../data/results.csv")

method_colors <- c(
  "z.test.gaussian"             = "#E41A1C",
  "z.test.oracle"               = "#377EB8",
  "bs.multiplier"               = "#4DAF4A",
  "bs.parametric"               = "#FF7F00",
  "bs.parametric.psd"           = "#0072B2",
  "z.test.null"                 = "#984EA3",
  "z.test.null.gaussian"        = "#A65628",
  "z.test.null.oracle"          = "#F781BF",
  "z.test"                      = "#999999",
  "z.test.null.matrix"          = "#66C2A5",
  "z.test.null.regress"         = "#8DA0CB",
  "bs.parametric.lag1"          = "#E6AB02",
  "bs.parametric.lag1.psd"      = "#A6761D",
  "bs.parametric.lag1.mask"     = "#66A61E",
  "bs.parametric.lag1.mask.psd" = "#1B9E77",
  "bs.parametric.mask"          = "#D95F02",
  "bs.parametric.mask.psd"      = "#7570B3"
)

method_labels <- c(
  "z.test.gaussian"             = "Z-Test (Gaussian)",
  "z.test.oracle"               = "Z-Test (Oracle)",
  "bs.multiplier"               = "Bootstrap (Multiplier)",
  "bs.parametric"               = "Bootstrap (Parametric)",
  "bs.parametric.psd"           = "Bootstrap (Parametric, PSD)",
  "z.test.null"                 = "Z-Test (Null)",
  "z.test.null.gaussian"        = "Z-Test (Null, Gaussian)",
  "z.test.null.oracle"          = "Z-Test (Null, Oracle)",
  "z.test"                      = "Z-Test",
  "z.test.null.matrix"          = "Z-Test (Null, Matrix)",
  "z.test.null.regress"         = "Z-Test (Null, Regress)",
  "bs.parametric.lag1"          = "Bootstrap (Parametric, Lag1)",
  "bs.parametric.lag1.psd"      = "Bootstrap (Parametric, Lag1, PSD)",
  "bs.parametric.lag1.mask"     = "Bootstrap (Parametric, Lag1, Mask)",
  "bs.parametric.lag1.mask.psd" = "Bootstrap (Parametric, Lag1, Mask, PSD)",
  "bs.parametric.mask"          = "Bootstrap (Parametric, Mask)",
  "bs.parametric.mask.psd"      = "Bootstrap (Parametric, Mask, PSD)"
)
```

## Summary

- **Overall ranking ($p = 2$):** Z-Test (Oracle) > Z-Test (Gaussian) > Z-Test (Null) $\approx$ Bootstrap (Parametric) > Bootstrap (Multiplier)
- **Oracle vs Gaussian:** Z-Test (Oracle) has perfect knowledge of $\kappa$, which is a function of the true correlation — making it a super upper bound, not an achievable target. The gap is largest near $\rho = 0$ (where $\hat\rho$ is least precise) and closes at large $\rho$ and as $n$ grows.
- **Wald vs score:** Z-Test (Gaussian) outperforms Z-Test (Null, Gaussian) because estimating $\hat{\rho}$ in the SE (Wald-type) is more efficient than plugging in $\rho = 0$ (score-type)
- **Shared variance estimator:** Z-Test (Null) and Bootstrap (Parametric) track each other closely, as both rely on $\widehat{\text{Cov}}(S)$
- **$p > 2$:** only bootstrap methods available; Bootstrap (Parametric) dominates Bootstrap (Multiplier)

## $p = 2$

::: {.callout-note collapse="true" title="Z-Test (Null, Oracle) and Z-Test (Null, Gaussian) are identical under normal errors"}

$n = 200$, $p = 2$.

Under normal errors, oracle $\kappa$ at $\rho = 0$ equals the Gaussian $\kappa$ at $\rho = 0$ exactly ($\kappa_{22} = 1$, $\kappa_{31} = \kappa_{13} = 0$, $\kappa_{40} = \kappa_{04} = 3$). Both null tests therefore use the same SE, and the curves coincide. Under exponential errors, oracle $\kappa$ departs from the Gaussian values, producing a gap.

```{r}
#| echo: false
#| warning: false
#| results: asis
dat_null_oracle <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null.gaussian", "z.test.null.oracle"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range_no <- range(dat_null_oracle$metric_val)

plots_no <- dat_null_oracle |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range_no) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat_null_oracle$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots_no, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Oracle) is a super upper bound: perfect κ implicitly encodes true correlation"}

$n = 200$, $p = 2$.

- **Normal errors:** Z-Test (Oracle) has a visible advantage near $\rho = 0$, where $\hat\rho$ is least precise and estimation error in $\hat\kappa$ is largest. The gap closes at large $\rho$ (where $\text{Var}(\hat\rho) \propto (1-\rho^2)^2/n$ shrinks) and diminishes as $n$ grows.
- **Exponential errors:** Z-Test (Oracle) has a larger advantage throughout — Z-Test (Gaussian) incurs both estimation error in $\hat\rho$ and misspecification of the Gaussian $\kappa$ formula.

```{r}
#| echo: false
#| warning: false
#| results: asis
dat <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.gaussian", "z.test.oracle"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range <- range(dat$metric_val)

plots <- dat |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Gaussian) is more powerful than bootstrap methods, regardless of error distribution"}

$n = 200$, $p = 2$.

- **Power ranking:** Z-Test (Gaussian) > Bootstrap (Parametric) > Bootstrap (Multiplier)
- **Error distribution:** power curves are similar across normal and exponential errors

```{r}
#| echo: false
#| warning: false
#| results: asis
dat2 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.gaussian", "z.test.oracle", "bs.multiplier", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range2 <- range(dat2$metric_val)

plots2 <- dat2 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range2) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat2$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots2, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Null) is less powerful than Z-Test (Null, Gaussian) because κ₂₂ must be estimated"}

$n = 200$, $p = 2$.

- **Z-Test (Null, Gaussian)** assumes Gaussian errors and avoids estimating $\kappa_{22}$
- **Z-Test (Null)** estimates $\kappa_{22}$ from data, incurring a power cost

```{r}
#| echo: false
#| warning: false
#| results: asis
dat4 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null", "z.test.null.gaussian"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range4 <- range(dat4$metric_val)

plots4 <- dat4 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range4) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat4$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots4, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Gaussian) is more powerful than Z-Test (Null, Gaussian): estimating ρ improves the SE"}

$n = 200$, $p = 2$.

- **Z-Test (Gaussian)** uses the unrestricted SE — the denominator includes $\hat{\rho}$
- **Z-Test (Null, Gaussian)** plugs in $\rho = 0$ under $H_0$, giving a less efficient SE
- This mirrors classical theory: the Wald statistic (unrestricted SE) is generally more powerful than the score statistic (null SE) at alternatives away from zero

```{r}
#| echo: false
#| warning: false
#| results: asis
dat5 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null.gaussian", "z.test.gaussian"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range5 <- range(dat5$metric_val)

plots5 <- dat5 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range5) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat5$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots5, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Null) and Bootstrap (Parametric) have nearly identical power"}

$n = 200$, $p = 2$.

- The two methods have similar power across scenarios and error distributions
- This is unsurprising: the Z-Test (Null) estimates $\kappa_{22}$ via $\widehat{\text{Cov}}(S)$, which is the same quantity the parametric bootstrap variance estimate is based on

```{r}
#| echo: false
#| warning: false
#| results: asis
dat6 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range6 <- range(dat6$metric_val)

plots6 <- dat6 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range6) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat6$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots6, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

## $p > 2$

::: {.callout-note collapse="true" title="Bootstrap (Parametric) outperforms Bootstrap (Multiplier) at p = 3; higher dimensions unavailable"}

$n = 200$, $p \in \{3, 5, 10\}$.

- **p = 3:** Bootstrap (Parametric) is more powerful than Bootstrap (Multiplier)
- **p = 5, 10:** no results — the estimated covariance matrix $\hat{W}$ is sometimes not positive semi-definite, causing simulation errors

```{r}
#| echo: false
#| warning: false
#| results: asis
dat3 <- results |>
  filter(p %in% c(3, 5, 10), n == 200, method %in% c("bs.multiplier", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors)),
    p        = factor(p, levels = c(3, 5, 10), labels = c("p = 3", "p = 5", "p = 10")),
    cov_type = factor(cov_type, levels = c("ar1", "compound"))
  )

y_range3 <- range(dat3$metric_val)

plots3 <- dat3 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method, linetype = cov_type)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_grid(p ~ scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_linetype_manual(values = c("ar1" = "dashed", "compound" = "solid")) +
      scale_y_continuous(limits = y_range3) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power",
           color = "Method", linetype = "Covariance type") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat3$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots3, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Bootstrap (Parametric, PSD) outperforms Bootstrap (Multiplier) across all dimensions"}

$n = 200$, $p \in \{3, 5, 10\}$.

- **p = 3:** Bootstrap (Parametric, PSD) is more powerful than Bootstrap (Multiplier) — matching the uncorrected parametric bootstrap
- **p = 5, 10:** Bootstrap (Parametric, PSD) remains available (unlike the uncorrected version, which fails when $\hat{W}$ is not PSD) and continues to dominate Bootstrap (Multiplier)

```{r}
#| echo: false
#| warning: false
#| results: asis
dat3 <- results |>
  filter(p %in% c(3, 5, 10), n == 200, method %in% c("bs.multiplier", "bs.parametric", "bs.parametric.psd"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors)),
    p        = factor(p, levels = c(3, 5, 10), labels = c("p = 3", "p = 5", "p = 10")),
    cov_type = factor(cov_type, levels = c("ar1", "compound"))
  )

y_range3 <- range(dat3$metric_val)

plots3 <- dat3 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method, linetype = cov_type)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_grid(p ~ scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_linetype_manual(values = c("ar1" = "dashed", "compound" = "solid")) +
      scale_y_continuous(limits = y_range3) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power",
           color = "Method", linetype = "Covariance type") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat3$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots3, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Parametric bootstrap power is recovered with PSD eigenvalue correction"}

$\widehat{\text{Cov}}(S)$ may not always produce a PSD covariance matrix; consequently, we cannot sample from a non-PSD covariance matrix. The eigenvalue correction clips any negative eigenvalues of $\hat{W}$ to zero before sampling.

```{R}
#| echo: false
#| warning: false
#| message: false
#results_psd <- read_csv("../data/raw/results-2026-05-27_18-35.csv")
results_psd <- read_csv("../data/results.csv")
err_df <- results_psd |>
  filter(method %in% c("bs.parametric", "bs.parametric.psd")) |>
  pivot_wider(names_from = metric, values_from = metric_val) |>
  mutate(err_pct = n_errors / n_sim)
```

::: {.panel-tabset}

## Low (~6–7% failure)

```{R}
#| echo: false
#| warning: false
err_df |>
  filter(n == 500, scenario == 1, p == 10, cov_type == "compound", err_type == "normal") |>
  ggplot(aes(x = r, y = power, color = method)) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
  geom_line() +
  geom_point() +
  scale_color_manual(values = method_colors, labels = method_labels) +
  labs(
    title = "n = 500, p = 10 — compound covariance, normal errors, no change points",
    x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Under normal errors with $n = 500$, roughly 6–7% of replicates produce a non-PSD $\hat{W}$. The corrected curve is nearly indistinguishable from the uncorrected — the correction is harmless at low failure rates.

## Moderate (~30% failure)

```{R}
#| echo: false
#| warning: false
err_df |>
  filter(n == 500, scenario == 1, p == 10, cov_type == "compound", err_type == "exponential") |>
  ggplot(aes(x = r, y = power, color = method)) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
  geom_line() +
  geom_point() +
  scale_color_manual(values = method_colors, labels = method_labels) +
  labs(
    title = "n = 500, p = 10 — compound covariance, exponential errors, no change points",
    x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Under exponential errors with $n = 500$, roughly 30% of replicates produce a non-PSD $\hat{W}$. Even at this failure rate, the corrected curve continues to track the uncorrected closely — the correction is robust.

## Extreme (100% failure)

```{R}
#| echo: false
#| warning: false
err_df |>
  filter(n == 200, scenario == 1, p == 10, cov_type == "compound", err_type == "exponential") |>
  ggplot(aes(x = r, y = power, color = method)) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
  geom_line() +
  geom_point() +
  scale_color_manual(values = method_colors, labels = method_labels) +
  labs(
    title = "n = 200, p = 10 — compound covariance, exponential errors, no change points",
    x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Under exponential errors with $n = 200$, every replicate fails to produce a PSD $\hat{W}$ — no uncorrected curve exists. The corrected curve is the only available estimate. The Type I error is deflated rather than inflated — conservative behavior, which is preferable to anti-conservative.

:::


:::

::: {.callout-note collapse="true" title="Masking worsens the non-PSD error rate; lag-1 truncation improves it; combined, they roughly cancel out"}

$n = 200$, compound covariance, normal errors, Scenario 1 (no change points), $\rho = 0$, $\text{amp} = 1$. Dimension is cut off at $p = 15$, beyond which every method reaches a 100% error rate.

- **Mask:** masking the disjoint-pair covariance terms makes the error rate worse than the baseline parametric bootstrap
- **Lag-1:** truncating to lag-1 (ignoring lag-2 and beyond) makes the error rate better than the baseline
- **Lag-1 + Mask:** combining both roughly cancels out, tracking close to the baseline

```{R}
#| echo: false
#| warning: false
#| results: asis
plot_dat <- results |>
  pivot_wider(names_from = metric, values_from = metric_val) |>
  filter(
    n == 200,
    cov_type == "compound",
    r == 0, amp == 1,
    err_type == "normal",
    scenario == 1,
    p <= 15,
  ) |>
  mutate(error_rate = n_errors / n_sim) |>
  select(method, n, p, error_rate)

make_err_plot <- function(methods) {
  plot_dat |>
    filter(method %in% methods) |>
    ggplot(aes(x = p, y = error_rate, color = method)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = method_colors, labels = method_labels) +
    labs(x = "Dimension (p)", y = "Error Rate", color = "Method") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

tabs <- list(
  "Mask"         = c("bs.parametric", "bs.parametric.mask"),
  "Lag-1"        = c("bs.parametric", "bs.parametric.lag1"),
  "Lag-1 + Mask" = c("bs.parametric", "bs.parametric.lag1.mask")
)

cat("::: {.panel-tabset}\n\n")
iwalk(tabs, \(methods, label) {
  cat(paste0("## ", label, "\n\n"))
  print(make_err_plot(methods))
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-warning collapse="true" title="Parametric Bootstrap power is nearly identical to Z-Test (Null)"}
## Parametric Bootstrap vs Z-Test (Null)
When correlation is non-trivial, `bs.parametric` and `z.test.null` track each other closely. Both methods rely on $\widehat{\text{Cov}}(S)$: the parametric bootstrap uses it to characterize the sampling distribution, while Z-Test (Null) uses it to estimate $\kappa_{22}$.

```{R}
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 5

results_nm <- read_csv("../data/results.csv") |>
  pivot_wider(names_from = metric, values_from = metric_val)

null_ref <- results_nm |>
  filter(p == 2, n == 200, method == "z.test.null") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  select(scenario, err_type, r, null_power = power)

dat_diff <- results_nm |>
  filter(p == 2, n == 200) |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  filter(!method %in% c("bs.parametric", "bs.multiplier.nosplit")) |>
  left_join(null_ref, by = c("scenario", "err_type", "r")) |>
  mutate(
    power_diff = power - null_power,
    err_type   = factor(err_type, levels = c("normal", "exponential")),
    method     = factor(method, levels = names(method_colors))
  )

ggplot(dat_diff, aes(x = r, y = power_diff, color = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line() +
  geom_point() +
  facet_grid(
    err_type ~ scenario,
    labeller = labeller(
      scenario = c("1" = "No Change Points", "2" = "Change Points (L = 4)"),
      err_type = c("normal" = "Normal", "exponential" = "Exponential")
    )
  ) +
  scale_color_manual(values = method_colors, labels = method_labels) +
  labs(
    x     = expression(paste("Correlation (", rho, ")")),
    y     = "Power − Power(Z-Test Null)",
    color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
:::

