---
title: "Power"
---

```{R}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
devtools::load_all()
results <- read_csv("../data/results.csv")

method_colors <- c(
  "z.test.gaussian"     = "#E41A1C",
  "z.test.oracle.kappa" = "#377EB8",
  "bs.multiplier"       = "#4DAF4A",
  "bs.parametric"       = "#FF7F00",
  "z.test.null"         = "#984EA3",
  "z.test.null.gaussian" = "#A65628"
)

method_labels <- c(
  "z.test.gaussian"     = "Z-Test (Gaussian)",
  "z.test.oracle.kappa" = "Z-Test (Oracle κ₂₂)",
  "bs.multiplier"       = "Bootstrap (Multiplier)",
  "bs.parametric"       = "Bootstrap (Parametric)",
  "z.test.null"         = "Z-Test (Null)",
  "z.test.null.gaussian" = "Z-Test (Null, Gaussian)"
)
```

::: {.callout-note collapse="true" title="The Gaussian z-test is nearly as powerful as knowing the true κ₂₂"}

$n = 200$, $p = 2$.

- **Normal errors:** Z-Test (Gaussian) and Z-Test (Oracle κ₂₂) are indistinguishable
- **Exponential errors:** Z-Test (Oracle κ₂₂) has a slight power advantage over Z-Test (Gaussian)

```{r}
#| echo: false
#| warning: false
#| results: asis
dat <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.gaussian", "z.test.oracle.kappa"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range <- range(dat$metric_val)

plots <- dat |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Gaussian) is more powerful than bootstrap methods, regardless of error distribution"}

$n = 200$, $p = 2$.

- **Power ranking:** Z-Test (Gaussian) > Bootstrap (Parametric) > Bootstrap (Multiplier)
- **Error distribution:** power curves are similar across normal and exponential errors

```{r}
#| echo: false
#| warning: false
#| results: asis
dat2 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.gaussian", "z.test.oracle.kappa", "bs.multiplier", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range2 <- range(dat2$metric_val)

plots2 <- dat2 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range2) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat2$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots2, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Null) is less powerful than Z-Test (Null, Gaussian) because κ₂₂ must be estimated"}

$n = 200$, $p = 2$.

- **Z-Test (Null, Gaussian)** assumes Gaussian errors and avoids estimating $\kappa_{22}$
- **Z-Test (Null)** estimates $\kappa_{22}$ from data, incurring a power cost

```{r}
#| echo: false
#| warning: false
#| results: asis
dat4 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null", "z.test.null.gaussian"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range4 <- range(dat4$metric_val)

plots4 <- dat4 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range4) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat4$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots4, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Gaussian) is more powerful than Z-Test (Null, Gaussian): estimating ρ improves the SE"}

$n = 200$, $p = 2$.

- **Z-Test (Gaussian)** uses the unrestricted SE — the denominator includes $\hat{\rho}$
- **Z-Test (Null, Gaussian)** plugs in $\rho = 0$ under $H_0$, giving a less efficient SE
- This mirrors classical theory: the Wald statistic (unrestricted SE) is generally more powerful than the score statistic (null SE) at alternatives away from zero

```{r}
#| echo: false
#| warning: false
#| results: asis
dat5 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null.gaussian", "z.test.gaussian"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range5 <- range(dat5$metric_val)

plots5 <- dat5 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range5) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat5$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots5, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Bootstrap (Parametric) outperforms Bootstrap (Multiplier) at p = 3; higher dimensions unavailable"}

$n = 200$, $p \in \{3, 5, 10\}$.

- **p = 3:** Bootstrap (Parametric) is more powerful than Bootstrap (Multiplier)
- **p = 5, 10:** no results — the estimated covariance matrix $\hat{W}$ is sometimes not positive semi-definite, causing simulation errors

```{r}
#| echo: false
#| warning: false
#| results: asis
dat3 <- results |>
  filter(p %in% c(3, 5, 10), n == 200, method %in% c("bs.multiplier", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors)),
    p        = factor(p, levels = c(3, 5, 10), labels = c("p = 3", "p = 5", "p = 10")),
    cov_type = factor(cov_type, levels = c("ar1", "compound"))
  )

y_range3 <- range(dat3$metric_val)

plots3 <- dat3 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method, linetype = cov_type)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_grid(p ~ scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_linetype_manual(values = c("ar1" = "dashed", "compound" = "solid")) +
      scale_y_continuous(limits = y_range3) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power",
           color = "Method", linetype = "Covariance type") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat3$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots3, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::


