---
title: "Power"
---

```{R}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
devtools::load_all()
results <- read_csv("../data/results.csv")

method_colors <- c(
  "z.test.gaussian"     = "#E41A1C",
  "z.test.oracle.kappa" = "#377EB8",
  "bs.multiplier"       = "#4DAF4A",
  "bs.parametric"       = "#FF7F00",
  "bs.parametric.psd"   = "#0072B2",
  "z.test.null"         = "#984EA3",
  "z.test.null.gaussian" = "#A65628"
)

method_labels <- c(
  "z.test.gaussian"     = "Z-Test (Gaussian)",
  "z.test.oracle.kappa" = "Z-Test (Oracle κ₂₂)",
  "bs.multiplier"       = "Bootstrap (Multiplier)",
  "bs.parametric"       = "Bootstrap (Parametric)",
  "bs.parametric.psd"   = "Bootstrap (Parametric, PSD)",
  "z.test.null"         = "Z-Test (Null)",
  "z.test.null.gaussian" = "Z-Test (Null, Gaussian)"
)
```

## Summary

- **Overall ranking ($p = 2$):** Z-Test (Gaussian) > Z-Test (Oracle $\kappa_{22}$) $\approx$ Z-Test (Gaussian) > Z-Test (Null) $\approx$ Bootstrap (Parametric) > Bootstrap (Multiplier)
- **Gaussian assumption vs oracle $\kappa_{22}$:** little difference in power; the Gaussian SE is nearly as good as knowing the true $\kappa_{22}$
- **Wald vs score:** Z-Test (Gaussian) outperforms Z-Test (Null, Gaussian) because estimating $\hat{\rho}$ in the SE (Wald-type) is more efficient than plugging in $\rho = 0$ (score-type)
- **Shared variance estimator:** Z-Test (Null) and Bootstrap (Parametric) track each other closely, as both rely on $\widehat{\text{Cov}}(S)$
- **$p > 2$:** only bootstrap methods available; Bootstrap (Parametric) dominates Bootstrap (Multiplier)

## $p = 2$

::: {.callout-note collapse="true" title="The Gaussian z-test is nearly as powerful as knowing the true κ₂₂"}

$n = 200$, $p = 2$.

- **Normal errors:** Z-Test (Gaussian) and Z-Test (Oracle κ₂₂) are indistinguishable
- **Exponential errors:** Z-Test (Oracle κ₂₂) has a slight power advantage over Z-Test (Gaussian)

```{r}
#| echo: false
#| warning: false
#| results: asis
dat <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.gaussian", "z.test.oracle.kappa"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range <- range(dat$metric_val)

plots <- dat |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Gaussian) is more powerful than bootstrap methods, regardless of error distribution"}

$n = 200$, $p = 2$.

- **Power ranking:** Z-Test (Gaussian) > Bootstrap (Parametric) > Bootstrap (Multiplier)
- **Error distribution:** power curves are similar across normal and exponential errors

```{r}
#| echo: false
#| warning: false
#| results: asis
dat2 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.gaussian", "z.test.oracle.kappa", "bs.multiplier", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range2 <- range(dat2$metric_val)

plots2 <- dat2 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range2) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat2$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots2, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Null) is less powerful than Z-Test (Null, Gaussian) because κ₂₂ must be estimated"}

$n = 200$, $p = 2$.

- **Z-Test (Null, Gaussian)** assumes Gaussian errors and avoids estimating $\kappa_{22}$
- **Z-Test (Null)** estimates $\kappa_{22}$ from data, incurring a power cost

```{r}
#| echo: false
#| warning: false
#| results: asis
dat4 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null", "z.test.null.gaussian"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range4 <- range(dat4$metric_val)

plots4 <- dat4 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range4) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat4$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots4, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Gaussian) is more powerful than Z-Test (Null, Gaussian): estimating ρ improves the SE"}

$n = 200$, $p = 2$.

- **Z-Test (Gaussian)** uses the unrestricted SE — the denominator includes $\hat{\rho}$
- **Z-Test (Null, Gaussian)** plugs in $\rho = 0$ under $H_0$, giving a less efficient SE
- This mirrors classical theory: the Wald statistic (unrestricted SE) is generally more powerful than the score statistic (null SE) at alternatives away from zero

```{r}
#| echo: false
#| warning: false
#| results: asis
dat5 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null.gaussian", "z.test.gaussian"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range5 <- range(dat5$metric_val)

plots5 <- dat5 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range5) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat5$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots5, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Z-Test (Null) and Bootstrap (Parametric) have nearly identical power"}

$n = 200$, $p = 2$.

- The two methods have similar power across scenarios and error distributions
- This is unsurprising: the Z-Test (Null) estimates $\kappa_{22}$ via $\widehat{\text{Cov}}(S)$, which is the same quantity the parametric bootstrap variance estimate is based on

```{r}
#| echo: false
#| warning: false
#| results: asis
dat6 <- results |>
  filter(p == 2, n == 200, method %in% c("z.test.null", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors))
  )

y_range6 <- range(dat6$metric_val)

plots6 <- dat6 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_wrap(~scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_y_continuous(limits = y_range6) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat6$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots6, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

## $p > 2$

::: {.callout-note collapse="true" title="Bootstrap (Parametric) outperforms Bootstrap (Multiplier) at p = 3; higher dimensions unavailable"}

$n = 200$, $p \in \{3, 5, 10\}$.

- **p = 3:** Bootstrap (Parametric) is more powerful than Bootstrap (Multiplier)
- **p = 5, 10:** no results — the estimated covariance matrix $\hat{W}$ is sometimes not positive semi-definite, causing simulation errors

```{r}
#| echo: false
#| warning: false
#| results: asis
dat3 <- results |>
  filter(p %in% c(3, 5, 10), n == 200, method %in% c("bs.multiplier", "bs.parametric"), metric == "power") |>
  filter(scenario == 1 | (scenario == 2 & scenario_params == "{\\L\\:[4]}")) |>
  mutate(
    err_type = factor(err_type, levels = c("normal", "exponential")),
    method   = factor(method, levels = names(method_colors)),
    p        = factor(p, levels = c(3, 5, 10), labels = c("p = 3", "p = 5", "p = 10")),
    cov_type = factor(cov_type, levels = c("ar1", "compound"))
  )

y_range3 <- range(dat3$metric_val)

plots3 <- dat3 |>
  group_by(err_type) |>
  group_map(\(d, g) {
    ggplot(d, aes(x = r, y = metric_val, color = method, linetype = cov_type)) +
      geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
      geom_line() +
      geom_point() +
      facet_grid(p ~ scenario, labeller = labeller(scenario = c(
        "1" = "No Change Points",
        "2" = "Change Points (L = 4)"
      ))) +
      scale_color_manual(values = method_colors, labels = method_labels) +
      scale_linetype_manual(values = c("ar1" = "dashed", "compound" = "solid")) +
      scale_y_continuous(limits = y_range3) +
      labs(x = expression(paste("Correlation (", rho, ")")), y = "Power",
           color = "Method", linetype = "Covariance type") +
      theme_minimal() +
      theme(legend.position = "bottom")
  }) |>
  setNames(levels(dat3$err_type))

cat("::: {.panel-tabset}\n\n")
iwalk(plots3, \(plot, err_type) {
  cat(paste0("## ", err_type, "\n\n"))
  print(plot)
  cat("\n\n")
})
cat(":::\n")
```

:::

::: {.callout-note collapse="true" title="Parametric bootstrap power is recovered with PSD eigenvalue correction"}

$\widehat{\text{Cov}}(S)$ may not always produce a PSD covariance matrix; consequently, we cannot sample from a non-PSD covariance matrix. The eigenvalue correction clips any negative eigenvalues of $\hat{W}$ to zero before sampling.

```{R}
#| echo: false
#| warning: false
#| message: false
results_psd <- read_csv("../data/raw/results-2026-05-27_18-35.csv")
err_df <- results_psd |>
  filter(method %in% c("bs.parametric", "bs.parametric.psd")) |>
  pivot_wider(names_from = metric, values_from = metric_val) |>
  mutate(err_pct = n_errors / n_sim)
```

::: {.panel-tabset}

## Low (~6–7% failure)

```{R}
#| echo: false
#| warning: false
err_df |>
  filter(n == 500, scenario == 1, p == 10, cov_type == "compound", err_type == "normal") |>
  ggplot(aes(x = r, y = power, color = method)) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
  geom_line() +
  geom_point() +
  scale_color_manual(values = method_colors, labels = method_labels) +
  labs(
    title = "n = 500, p = 10 — compound covariance, normal errors, no change points",
    x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Under normal errors with $n = 500$, roughly 6–7% of replicates produce a non-PSD $\hat{W}$. The corrected curve is nearly indistinguishable from the uncorrected — the correction is harmless at low failure rates.

## Moderate (~30% failure)

```{R}
#| echo: false
#| warning: false
err_df |>
  filter(n == 500, scenario == 1, p == 10, cov_type == "compound", err_type == "exponential") |>
  ggplot(aes(x = r, y = power, color = method)) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
  geom_line() +
  geom_point() +
  scale_color_manual(values = method_colors, labels = method_labels) +
  labs(
    title = "n = 500, p = 10 — compound covariance, exponential errors, no change points",
    x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Under exponential errors with $n = 500$, roughly 30% of replicates produce a non-PSD $\hat{W}$. Even at this failure rate, the corrected curve continues to track the uncorrected closely — the correction is robust.

## Extreme (100% failure)

```{R}
#| echo: false
#| warning: false
err_df |>
  filter(n == 200, scenario == 1, p == 10, cov_type == "compound", err_type == "exponential") |>
  ggplot(aes(x = r, y = power, color = method)) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray50") +
  geom_line() +
  geom_point() +
  scale_color_manual(values = method_colors, labels = method_labels) +
  labs(
    title = "n = 200, p = 10 — compound covariance, exponential errors, no change points",
    x = expression(paste("Correlation (", rho, ")")), y = "Power", color = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Under exponential errors with $n = 200$, every replicate fails to produce a PSD $\hat{W}$ — no uncorrected curve exists. The corrected curve is the only available estimate. The Type I error is deflated rather than inflated — conservative behavior, which is preferable to anti-conservative.

:::


:::

::: {.callout-warning collapse="true" title="Parametric Bootstrap Power is suspiciously similar to Z-Test (Null) "}
## Parametric Bootstrap vs Z-Test (Null) 
When there is non-trivial correlation, z-test (null) and `bs.parametric` are very similar.  

My guess as to why this is, is because both `z-test (null)` and `bs.parametric `use `Cov(S)`. `bs.parametric` uses it to characterize the parametric distribution to sample from. `z.test (null)` uses it to produce an estimator for `kappa22`. 

```{R} 
#| message: false
#| echo: false

# Read Data
results <- read_csv("../data/results.csv") |> 
  pivot_wider(names_from=metric, values_from=metric_val) 
results <- read_csv("../data/raw/results-new-methods.csv") |> 
  pivot_wider(names_from=metric, values_from=metric_val) 

# Figure: z.test.null - [METHOD]
# - Hypothesis: z.test.null and bs.parametric are basically the same
scenario_num <- 2
sample_size <- 1000
error_type <- "normal"
rho <- 0.4

# Get Reference
ztest_power <- results |> 
  filter(
    scenario == scenario_num, 
    n == sample_size, 
    err_type == error_type
  ) |> 
  filter(method == "z.test.null") |>
  filter(r == rho) |> pull(power)

# Compare against other methods
results |> 
  filter(
    scenario == scenario_num, 
    n == sample_size, 
    err_type == error_type
  ) |> 
  filter(r == rho) |> 
  mutate(diff = power - ztest_power) |> 
  select(method, power, diff) |> 
  arrange(abs(diff))
```
:::

