---
title: "Smooth Mean"
---

## Motivation

Every prior chapter assumes piecewise-constant $\theta$: the bias formula in [Lagged Differences](lagged-differences.qmd), the [Omega Estimation](omega-estimation.qmd) parameter, and the ECE construction in [Bilinear Form](bilinear-form.qmd) all depend on it. Real trends are often smooth instead — slow drift, gradual regime change. This chapter asks what changes if $\theta$ is smooth, and whether the same lag-based machinery generalizes.

## Bias Under a Smooth Mean

Let $\theta_x(t), \theta_y(t)$ be differentiable, sampled at $t = i/n$. For lag $k \ll n$, a Taylor expansion gives

$$
\theta_{x,i} - \theta_{x,i+k} \approx -\theta_x'(t)\cdot\frac{k}{n}.
$$

Substituting into $T_k(\theta_x,\theta_y) = \sum_i (\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k})$ and treating the sum as a Riemann sum,

\begin{align}
T_k(\theta_x,\theta_y) &= \sum_i (\theta_{x,i}-\theta_{x,i+k})(\theta_{y,i}-\theta_{y,i+k}) \\
&\approx \frac{k^2}{n^2}\sum_i \theta_x'(t)\theta_y'(t) \\
&\approx \frac{k^2}{n}\int_0^1 \theta_x'(t)\theta_y'(t)\,dt.
\end{align}

Piecewise-constant contamination is **linear** in $k$ — each extra lag can straddle one more change point. Smooth contamination is **quadratic** in $k$ — it comes from squaring a derivative times a step of size $k/n$.

::: {.callout-note collapse="true"}
## Test: Quadratic-in-$k$ bias for a sinusoidal $\theta$

For $\theta_x(t) = \sin(2\pi f t)$, $\theta_y(t) = A\sin(2\pi f t + \phi)$, the predicted constant is $\int_0^1 \theta_x'\theta_y'\,dt = A(2\pi f)^2\cdot\tfrac12\cos(\phi)$. The chunk below checks $T_k(\theta_x,\theta_y) \approx \frac{k^2}{n}\cdot A(2\pi f)^2\cdot\tfrac12\cos(\phi)$ across several lags.

```{r sine-quadratic-test}
#| message: false
#| warning: false
devtools::load_all("..")
library(testthat)

n <- 2000
f <- 3
A <- 1.5
phi <- 0.7
t <- (1:n) / n
theta_x <- sin(2 * pi * f * t)
theta_y <- A * sin(2 * pi * f * t + phi)

ks <- c(1, 2, 3, 5, 8)
Tk_numerical <- sapply(ks, function(k) lag_diff(theta_x, theta_y, k = k))
Tk_predicted <- sapply(ks, function(k) k^2 / n * A * (2 * pi * f)^2 * 0.5 * cos(phi))

test_that("quadratic-in-k bias formula holds for a sinusoidal theta", {
  expect_equal(Tk_numerical, Tk_predicted, tolerance = 1e-3)
})

plot(ks, Tk_numerical,
  type = "b", pch = 19, col = "steelblue",
  xlab = "Lag k", ylab = expression(T[k](theta[x], theta[y])),
  main = "Quadratic growth in k for a smooth (sinusoidal) mean"
)
lines(ks, Tk_predicted, col = "firebrick", lty = 2)
legend("topleft", legend = c("numerical", "predicted (quadratic)"),
       col = c("steelblue", "firebrick"), lty = c(1, 2), pch = c(19, NA))
```
:::

## Polynomial Order of Contamination

Contamination in $T_k$ is a polynomial in $k$; its degree depends on $\theta$'s shape.

| Model for $\theta$ | Bias in $\mathbb{E}[T_k(X,Y)]$ | Leading term |
|---|---|---|
| Piecewise constant | linear in $k$ | $k\,\omega_{xy}^{(1)}$, $\omega_{xy}^{(1)} = \omega_{xy}$ (see [Omega Estimation](omega-estimation.qmd)) |
| Smooth (differentiable) | quadratic in $k$ | $k^2\,\omega_{xy}^{(2)}$, $\omega_{xy}^{(2)} = \tfrac1n\int\theta_x'\theta_y'\,dt$ |

General model:
$$
\mathbb{E}[T_k(X,Y)] = 2n\sigma_{xy} + \sum_{m=1}^M k^m\,\omega_{xy}^{(m)}.
$$
$M$ indexes how many shapes of contamination are protected against at once. $\omega_{xy}^{(1)}$: jump synchrony. $\omega_{xy}^{(2)}$: curvature synchrony — whether $\theta_x, \theta_y$ bend together, independent of shared slow drift (see [Omega Estimation](omega-estimation.qmd)'s frequency-weighting discussion).

## Two Mechanisms, Not One Taylor Series

The $k$ and $k^2$ terms are not successive orders of one Taylor expansion. An exact identity holds for any $\theta$, no Taylor needed:
$$
\mathbb{E}[T_k(X,Y)] = T_k(\theta_x,\theta_y) + 2n\sigma_{xy}
$$
(bilinear algebra, see [Bilinear Form](bilinear-form.qmd)). The two polynomial terms come from unrelated mechanisms for modeling $T_k(\theta,\theta)$:

- **$k\,\omega_{xy}^{(1)}$ (jumps).** Exact — counts boundary crossings in a lag-$k$ window. No differentiability involved; $\theta$ isn't differentiable at a jump.
- **$k^2\,\omega_{xy}^{(2)}$ (curvature).** From the Taylor expansion above; exact iff $\theta''=0$ (piecewise linear). Leading-order only for general smooth $\theta$.

A shared regression on $(1,k,k^2)$ absorbs both only if they don't co-occur in the same window. A jump and slope change at the same boundary produce a cross term in $T_k$ proportional to neither $k$ nor $k^2$; the regression misattributes it to both coefficients — the failure mode in "Piecewise-Linear, Discontinuous, Segments = L" below.

## Does This Work at All?

Before generalizing anything further, a direct check: for a fast sinusoidal $\theta$ (period 25, $n=200$), does fitting a line through $T_k(X,Y)$ actually look wrong, and does adding a $k^2$ term fix it? Both fits use $k=1,\ldots,L=4$ — the only points shown — and are extended back to $k=0$, where the true $2n\sigma_{xy}$ is marked.

```{r smooth-fit-comparison}
#| echo: false
#| cache: true
#| message: false
n <- 200
L <- 4
reps <- 1000

K1 <- cbind(1, 1:L)
K2 <- cbind(1, 1:L, (1:L)^2)

fit_panel <- function(Tk_avg, true_val, ylab) {
  beta1 <- solve(t(K1) %*% K1) %*% t(K1) %*% Tk_avg
  beta2 <- solve(t(K2) %*% K2) %*% t(K2) %*% Tk_avg

  points_df <- tibble::tibble(k = 1:L, Tk = Tk_avg / (2 * n))
  kseq <- seq(0, L, length.out = 200)
  lines_df <- tibble::tibble(
    k = rep(kseq, 2),
    fit = rep(c("Linear (M=1)", "Quadratic (M=2)"), each = length(kseq)),
    val = c(
      beta1[1] + beta1[2] * kseq,
      beta2[1] + beta2[2] * kseq + beta2[3] * kseq^2
    ) / (2 * n)
  )

  ggplot2::ggplot() +
    ggplot2::geom_hline(yintercept = 0, color = "grey40") +
    ggplot2::geom_vline(xintercept = 0, color = "grey40") +
    ggplot2::geom_point(data = points_df, ggplot2::aes(x = k, y = Tk), size = 2) +
    ggplot2::geom_line(data = lines_df, ggplot2::aes(x = k, y = val, color = fit)) +
    ggplot2::geom_point(ggplot2::aes(x = 0, y = true_val), color = "red", size = 3) +
    ggplot2::scale_color_manual(
      values = c("Linear (M=1)" = "black", "Quadratic (M=2)" = "firebrick"), name = NULL
    ) +
    ggplot2::scale_x_continuous(breaks = 0:L) +
    ggplot2::labs(x = "Lag k", y = ylab) +
    ggplot2::theme_minimal()
}

run_scenario <- function(params) {
  set.seed(1)
  X_example <- generate_data(params)

  Tk_mat <- lapply(1:reps, function(i) {
    X <- generate_data(params)
    list(
      xx = sapply(1:L, function(k) lag_diff(X[, 1], X[, 1], k = k)),
      yy = sapply(1:L, function(k) lag_diff(X[, 2], X[, 2], k = k)),
      xy = sapply(1:L, function(k) lag_diff(X[, 1], X[, 2], k = k))
    )
  })

  list(
    params = params,
    X_example = X_example,
    Tk_avg_xx = Reduce(`+`, lapply(Tk_mat, `[[`, "xx")) / reps,
    Tk_avg_yy = Reduce(`+`, lapply(Tk_mat, `[[`, "yy")) / reps,
    Tk_avg_xy = Reduce(`+`, lapply(Tk_mat, `[[`, "xy")) / reps
  )
}

mean_noise_plot <- function(sc) {
  tibble::tibble(i = 1:n, theta = sc$params$h[, 1], x = sc$X_example[, 1]) |>
    ggplot2::ggplot(ggplot2::aes(x = i)) +
    ggplot2::geom_line(ggplot2::aes(y = x), color = "grey80") +
    ggplot2::geom_line(ggplot2::aes(y = theta), color = "steelblue", linewidth = 1) +
    ggplot2::labs(x = "Index i", y = "Value") +
    ggplot2::theme_minimal()
}

regression_matrix_plot <- function(sc) {
  p_xx <- fit_panel(sc$Tk_avg_xx, sc$params$S[1, 1], expression(T[k](X, X) / (2 * n)))
  p_yy <- fit_panel(sc$Tk_avg_yy, sc$params$S[2, 2], expression(T[k](Y, Y) / (2 * n)))
  p_xy <- fit_panel(sc$Tk_avg_xy, sc$params$S[1, 2], expression(T[k](X, Y) / (2 * n)))

  (p_xx | p_xy) / (patchwork::plot_spacer() | p_yy) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "bottom")
}

square_sc <- run_scenario(scenario(scenario_num = 2, n = n, L = 4, amp = 3, r = 0.5, err_type = "normal"))

set.seed(2)
L_seg <- 20
pwlin_slopes <- rep(rnorm(n / L_seg, sd = 0.2), each = L_seg)
pwlin_theta <- cumsum(pwlin_slopes)
pwlin_params <- scenario(scenario_num = 1, n = n, amp = 3, r = 0.5, err_type = "normal")
pwlin_params$h <- matrix(rep(pwlin_theta, 2), n, 2)
pwlin_sc <- run_scenario(pwlin_params)

sine_sc <- run_scenario(scenario(scenario_num = 3, n = n, period = 25, amp = 3, r = 0.5, err_type = "normal"))

saw_theta <- 3 * rep(0:(L - 1), n / L)
saw_params <- scenario(scenario_num = 1, n = n, amp = 3, r = 0.5, err_type = "normal")
saw_params$h <- matrix(rep(saw_theta, 2), n, 2)
saw_sc <- run_scenario(saw_params)

set.seed(4)
jump_a <- rnorm(n / L, sd = 3)
jump_b <- rnorm(n / L, sd = 1)
jump_theta <- rep(jump_a, each = L) + rep(jump_b, each = L) * rep(0:(L - 1), n / L)
jump_params <- scenario(scenario_num = 1, n = n, amp = 3, r = 0.5, err_type = "normal")
jump_params$h <- matrix(rep(jump_theta, 2), n, 2)
jump_sc <- run_scenario(jump_params)
```

:::: {.panel-tabset}

## Square Wave (scenario 2, L=4)

::: {.panel-tabset}

### Mean and Noise

```{r square-mean-noise}
#| echo: false
mean_noise_plot(square_sc)
```

### Regression

```{r square-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(square_sc)
```

:::

Here $\theta$ is exactly piecewise constant with segments of length $L=4$ — squarely the original model's home turf. Both fits track the points closely and both intercepts land on the true value: this is a check that the generalization doesn't cost anything when the piecewise-constant assumption already holds.

## Piecewise-Linear (random slopes)

::: {.panel-tabset}

### Mean and Noise

```{r pwlin-mean-noise}
#| echo: false
mean_noise_plot(pwlin_sc)
```

### Regression

```{r pwlin-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(pwlin_sc)
```

:::

Now a case squarely inside the $M=2$ model's class: $\theta$ is piecewise **linear**, slope drawn i.i.d. $N(0, 0.2^2)$ per block of length $L_{\text{seg}}=20$ — 5 times the lag window $L=4$, so most $k=1,\ldots,4$ windows stay within one segment and see a single constant slope. Within such a window, $\theta_i - \theta_{i+k}$ is *exactly* linear in $k$, so contamination is exactly quadratic in $k$ with no truncation error — unlike the sine case above. The linear fit is visibly biased ($\hat\sigma_{xy}\approx0.40$ against true $0.5$), since it can't absorb any curvature. The quadratic fit lands almost exactly on the truth ($\approx0.49$), confirming the generalization isn't just a leading-order patch for sinusoids — it recovers the truth (up to boundary-crossing effects at segments $\approx L$) whenever contamination truly has this shape.

## Fast Sine (scenario 3)

::: {.panel-tabset}

### Mean and Noise

```{r sine-mean-noise}
#| echo: false
mean_noise_plot(sine_sc)
```

### Regression

```{r sine-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(sine_sc)
```

:::

The points curve visibly upward — real quadratic contamination, sharp enough now that the linear fit's intercept lands *below zero*. The quadratic fit (red curve) still passes right through the true intercept. Reading off each fit's intercept: linear gives $\hat\sigma_{xy} \approx -0.13$ — wrong sign — while quadratic gives $\approx 0.44$, against a true $\sigma_{xy} = 0.5$. The original method doesn't just lose precision here; it fails qualitatively. The generalization still works.

## Sawtooth (constant slope, period = L)

::: {.panel-tabset}

### Mean and Noise

```{r saw-mean-noise}
#| echo: false
mean_noise_plot(saw_sc)
```

### Regression

```{r saw-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(saw_sc)
```

:::

The failure below came from randomness, not from combining a jump with a slope per se. Here $\theta$ is a deterministic sawtooth — $\theta_i = 3(i \bmod 4)$, a fixed slope that resets every $L=4$ steps, so every "segment" has the identical slope and the identical reset. This makes $T_k(\theta,\theta)$ an *exact* quadratic in $k$ with no residual at all: $T_k(\theta,\theta) = n(Lk-k^2)$ (amplitude-squared factor omitted), verified to machine precision. The linear fit is catastrophic ($\hat\sigma_{xy}\approx23$, true $0.5$) — a steady drift is the worst case for a $k$-only model. The quadratic fit lands almost exactly on the truth ($\approx0.50$). Unlike the next case, jump and slope here aren't independent per segment — they're the same fixed pair repeating, so their cross term is itself a clean function of $k$ that the $k^2$ coefficient absorbs. Entanglement only defeats the model when it varies unpredictably from segment to segment.

## Piecewise-Linear, Discontinuous, Segments = L

::: {.panel-tabset}

### Mean and Noise

```{r jump-mean-noise}
#| echo: false
mean_noise_plot(jump_sc)
```

### Regression

```{r jump-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(jump_sc)
```

:::

A more extreme version of the sawtooth above: segments still have length exactly $L=4$, but now each segment gets its own i.i.d. random slope ($N(0,1)$) *and* its own i.i.d. random level ($N(0,3^2)$) — no continuity constraint at block boundaries, so this is jumps and curvature superimposed at the shortest possible spacing. The hypothesis was that this should still work, since jumps load onto the $k$ coefficient and curvature onto $k^2$. It doesn't: both fits are badly wrong here. The linear fit overshoots substantially ($\hat\sigma_{xy}\approx1.5$, true $0.5$), and the quadratic fit undershoots to near zero ($\approx0.12$) — worse, in relative terms, than the linear fit in the fast-sine case above. Because every lag window straddles a boundary (segment length equals $L$, leaving no within-segment room), the jump and curvature contributions mix inside each $T_k$ term rather than adding cleanly — the cross term between them lands on neither the $k$ nor the $k^2$ coefficient, and the OLS fit misattributes it to the intercept. This is the "mixed, entangled" case flagged as an open problem above, made concrete: no amount of matching segment length to $L$ rescues it once jump and curvature vary together.

::::

## Generalized Regression Estimator (Sketch)

[Bilinear Form](bilinear-form.qmd)'s "Regression Approach" fits $R_k = T_k(X,Y)$ on $(1,k)$ across $k=1,\ldots,L$ to extract $\sigma_{xy}$. Adding a $k^2$ column:

$$
\mathbb{E}[R] = K\beta, \qquad K = \begin{bmatrix} 1 & 1 & 1 \\ 1 & 2 & 4 \\ \vdots & \vdots & \vdots \\ 1 & L & L^2 \end{bmatrix} \in \mathbb{R}^{L\times 3}, \qquad \beta = \begin{pmatrix} 2n\sigma_{xy} \\ \omega_{xy}^{(1)} \\ \omega_{xy}^{(2)} \end{pmatrix},
$$

requiring $L \ge 3$ lags to identify all three coefficients. This protects $\hat\sigma_{xy}$ against a pure jump (exactly) or pure smooth curvature (to leading order) — not necessarily both at once; see below.

Not yet solved in closed form. Next step: repeat `bilinear-form.qmd`'s linear-case derivation — solve the OLS system, express as $X^TAY$ with $A$ a symmetric circulant, check feasibility and minimum-norm optimality.

## What Is and Isn't Preserved

- **Pure jumps, spaced $\ge L$.** Exact, unchanged. The $k$ coefficient is the same algebraic identity as the original ECE; a jump contributes only there, nothing to $k^2$.
- **Pure smooth curvature.** Approximate only. The $k^2$ term is a truncated Taylor expansion — bias remains from third-order derivatives and beyond. No finite-sample exact guarantee, unlike the piecewise-constant case.
- **Mixed: jumps closer than $L$, or a jump entangled with curvature.** Uncovered. Contamination doesn't decompose cleanly into $k$ and $k^2$ terms. Open problem.

Net: exact unbiasedness survives only for the original piecewise-constant class. Smooth contamination gets leading-order correction, not exact removal. Mixed cases are unprotected.

## Simulation

No closed form is needed to test the $M=2$ estimator — it's a direct OLS fit. Fix $L=4$ throughout.

```{r sigma-hat-quad-def}
#| message: false
#| warning: false
devtools::load_all("..")
library(purrr)

sigma_hat_quad <- function(x, y, L = 4) {
  n <- length(x)
  R <- sapply(1:L, function(k) lag_diff(x, y, k))
  K <- cbind(1, 1:L, (1:L)^2)
  beta <- solve(t(K) %*% K) %*% t(K) %*% R
  as.numeric(beta[1]) / (2 * n)
}
```

### Pure Smooth: Period Sweep

For a sinusoidal $\theta$ (`scenario(scenario_num = 3)`, $n=200$, amplitude 3, $r=0.5$, Gaussian errors), compare the bias of the current linear ECE (`ece.cov`, $L=4$) against `sigma_hat_quad` ($L=4$) as the period shrinks from slow to fast — 300 replicates per period.

```{r smooth-sim-sweep}
#| cache: true
#| message: false
#| dependson: ["sigma-hat-quad-def"]
periods <- c(2000, 1000, 500, 200, 100, 50, 25, 10)
reps <- 300
L <- 4

sweep <- map_dfr(periods, function(P) {
  params <- scenario(scenario_num = 3, n = 200, period = P, amp = 3, r = 0.5, err_type = "normal")
  true_sigma <- params$S[1, 2]
  ests <- map_dfr(1:reps, function(i) {
    X <- generate_data(params)
    tibble::tibble(
      m1 = ece.cov(X[, 1], X[, 2], L = L),
      m2 = sigma_hat_quad(X[, 1], X[, 2], L = L)
    )
  })
  tibble::tibble(
    period = P,
    bias_linear = mean(ests$m1) - true_sigma,
    bias_quadratic = mean(ests$m2) - true_sigma
  )
})
```

```{r smooth-sim-sweep-plot}
#| echo: false
sweep |>
  tidyr::pivot_longer(c(bias_linear, bias_quadratic), names_to = "estimator", values_to = "bias") |>
  ggplot2::ggplot(ggplot2::aes(x = period, y = bias, color = estimator)) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  ggplot2::scale_x_log10() +
  ggplot2::labs(x = "Period (log scale, slow → fast going left)", y = expression(hat(sigma)[xy] - sigma[xy]),
                title = "Bias vs. oscillation speed: linear (M=1) vs. quadratic (M=2)") +
  ggplot2::theme_minimal()
```

The quadratic estimator keeps bias near zero much further into the fast-oscillation range than the linear one — it extends the robust range, it doesn't remove the breakdown. At the fastest period (10), $L=4$ lags no longer satisfy the Taylor-validity condition from the previous section, and both estimators fail together.

### Mixed: Smooth Trend Plus Closely-Spaced Jumps

Now overlay a slow sine (period 1000, amplitude 3) with jumps spaced $L_{\text{seg}}=2$ apart (amplitude 3) — closer than $L=4$, the case flagged as uncovered above.

```{r smooth-mixed-sim}
#| cache: true
#| message: false
#| dependson: ["sigma-hat-quad-def"]
n <- 200
params <- scenario(scenario_num = 1, n = n, r = 0.5, err_type = "normal")
true_sigma <- params$S[1, 2]

theta <- 3 * sin(2 * pi * (1:n) / 1000) + 3 * square_wave(n, L = 2)
params$h <- matrix(rep(theta, 2), n, 2)

mixed_ests <- map_dfr(1:300, function(i) {
  X <- generate_data(params)
  tibble::tibble(
    m1 = ece.cov(X[, 1], X[, 2], L = 4),
    m2 = sigma_hat_quad(X[, 1], X[, 2], L = 4)
  )
})

tibble::tibble(
  true_sigma = true_sigma,
  bias_linear = mean(mixed_ests$m1) - true_sigma,
  bias_quadratic = mean(mixed_ests$m2) - true_sigma
)
```

Both estimators are severely biased — neither the linear nor the quadratic model captures contamination from jumps spaced closer than $L$. This confirms the "mixed" case above is a real gap, not just a theoretical caveat: adding the $k^2$ term does not rescue this scenario.

## Connection to Difference-Based Nonparametric Regression

Choosing difference weights that vanish against polynomials up to degree $m$, to remove a smooth trend before estimating variance, is established in nonparametric regression: Rice (1984); Hall, Kay & Titterington (1990); Müller & Stadtmüller; Wang, Brown, Cai & Levine. This chapter's generalization is the bivariate cross-covariance analog, reusing the min-norm-subject-to-constraints framework already built for $L>2$ in [Bilinear Form](bilinear-form.qmd).

## Open Questions

- Closed-form $\hat\sigma_{xy}, \hat\omega_{xy}^{(1)}, \hat\omega_{xy}^{(2)}$ for the $L\ge3$ regression, and their joint asymptotic distribution (paralleling [Omega Estimation](omega-estimation.qmd)).
- Bias-variance tradeoff as a function of $M$ and $L$ — mirroring the $L>2$ noise-variance analysis in [Bilinear Form](bilinear-form.qmd).
- Handling the mixed case: jumps and curvature close together, or jumps spaced closer than $L$.
- Polynomial-in-$k$ vs. direct frequency-domain filter design (a high-pass filter with a specified rolloff) — more flexible, at the cost of a tuning parameter.
