---
title: "Smooth Mean"
---

## Motivation

Every prior chapter assumes piecewise-constant $\theta$: the bias formula in [Lagged Differences](lagged-differences.qmd), the [Omega Estimation](omega-estimation.qmd) parameter, and the ECE construction in [Bilinear Form](bilinear-form.qmd) all depend on it. Real trends are often smooth instead — slow drift, gradual regime change. This chapter asks what changes if $\theta$ is smooth, and whether the same lag-based machinery generalizes.

## Bias Under a Smooth Mean

Recall that the expectation of $T_k(X,Y)$ is
$$
\mathbb{E}[T_k(X,Y)] = T_k(\theta_x,\theta_y) + 2n\sigma_{xy},
$$
exact for any $\theta$ (bilinear algebra, see [Bilinear Form](bilinear-form.qmd)). What follows models $T_k(\theta_x,\theta_y)$ as a function of $k$.

**Piecewise constant.** From [Lagged Differences](lagged-differences.qmd) and [Omega Estimation](omega-estimation.qmd):
$$
T_k(\theta_x,\theta_y) = k\,n\,\omega_{xy}, \quad \text{exact}.
$$

**Constant slope, plus mean shifts.** Let $\theta_x = b_x\cdot i + (\text{piecewise-constant jumps, spaced} \ge L)$, and likewise $\theta_y$ with slope $b_y$ — one fixed slope for the *entire* series, not a per-segment value. Then
$$
T_k(\theta_x,\theta_y) = c_1 k + c_2 k^2 + k\,n\,\omega_{xy},
$$
exact, for constants $c_1,c_2$ depending on $b_x,b_y,n$ — no residual, at any jump spacing $\ge L$ and any jump sizes or timing (verified numerically, including irregular timing and spacing down to $L$ itself). Two things make this exact where general piecewise linear fails:

- The slope is a single global constant, not segment-dependent, so its contribution to $T_k$ — even wrapped circularly — reduces to a fixed combination of $k$ and $k^2$ for every window, regardless of where the jumps fall.
- Its interaction with the jump component is itself a fixed $k$-linear term (it mixes into $\omega_{xy}^{(1)}$; nothing is left unmodeled).

**Segment-varying slope.** If the slope instead differs from segment to segment — genuine piecewise linearity, no single global value — the identity above no longer holds: windows straddling a segment boundary see an unpredictable slope pair, contributing a residual proportional to neither $k$ nor $k^2$. Small when segments are long relative to $L$ ("Piecewise-Linear (random slopes)" below); large once slope and jump both vary per segment at spacing $\approx L$ ("Piecewise-Linear, Discontinuous, L = 4" below). The sawtooth below is the degenerate case where the slope *is* constant and the "jump" is a fixed, exactly-repeating reset — it falls under the first bullet, not this one.

**Two distinct mechanisms.** Jumps load onto $k$; a global slope loads onto $k^2$ plus a $k$-linear cross term — unrelated identities, not orders of one expansion. What breaks the shared regression isn't the jump and slope occurring near each other — it's the slope varying unpredictably from segment to segment.

## Polynomial Order of Contamination

Contamination in $T_k$ is a polynomial in $k$; its degree depends on $\theta$'s shape.

| Model for $\theta$ | Bias in $\mathbb{E}[T_k(X,Y)]$ | Leading term |
|---|---|---|
| Piecewise constant | linear in $k$ | $k\,\omega_{xy}^{(1)}$, $\omega_{xy}^{(1)} = \omega_{xy}$ (see [Omega Estimation](omega-estimation.qmd)) |
| Constant global slope | quadratic in $k$ | $k^2\,\omega_{xy}^{(2)}$, $\omega_{xy}^{(2)} = b_xb_y$ (exact; a per-segment slope only holds this approximately) |

General model:
$$
\mathbb{E}[T_k(X,Y)] = 2n\sigma_{xy} + \sum_{m=1}^M k^m\,\omega_{xy}^{(m)}.
$$
$M$ indexes how many shapes of contamination are protected against at once. $\omega_{xy}^{(1)}$: jump synchrony. $\omega_{xy}^{(2)}$: curvature synchrony — whether $\theta_x, \theta_y$ bend together, independent of shared slow drift (see [Omega Estimation](omega-estimation.qmd)'s frequency-weighting discussion).

## Simulations

Before generalizing anything further, a direct check: for a fast sinusoidal $\theta$ (period 25, $n=200$), does fitting a line through $T_k(X,Y)$ actually look wrong, and does adding a $k^2$ term fix it? Both fits use $k=1,\ldots,L=4$ — the only points shown — and are extended back to $k=0$, where the true $2n\sigma_{xy}$ is marked. Each scenario below is run with 1000 replicates.

```{r smooth-fit-comparison}
#| echo: false
#| cache: true
#| message: false
n <- 200
L <- 4
reps <- 1000

K1 <- cbind(1, 1:L)
K2 <- cbind(1, 1:L, (1:L)^2)

fit_panel <- function(Tk_avg, true_val, ylab) {
  beta1 <- solve(t(K1) %*% K1) %*% t(K1) %*% Tk_avg
  beta2 <- solve(t(K2) %*% K2) %*% t(K2) %*% Tk_avg

  points_df <- tibble::tibble(k = 1:L, Tk = Tk_avg / (2 * n))
  kseq <- seq(0, L, length.out = 200)
  lines_df <- tibble::tibble(
    k = rep(kseq, 2),
    fit = rep(c("Linear (M=1)", "Quadratic (M=2)"), each = length(kseq)),
    val = c(
      beta1[1] + beta1[2] * kseq,
      beta2[1] + beta2[2] * kseq + beta2[3] * kseq^2
    ) / (2 * n)
  )

  ggplot2::ggplot() +
    ggplot2::geom_hline(yintercept = 0, color = "grey40") +
    ggplot2::geom_vline(xintercept = 0, color = "grey40") +
    ggplot2::geom_point(data = points_df, ggplot2::aes(x = k, y = Tk), size = 2) +
    ggplot2::geom_line(data = lines_df, ggplot2::aes(x = k, y = val, color = fit)) +
    ggplot2::geom_point(ggplot2::aes(x = 0, y = true_val), color = "red", size = 3) +
    ggplot2::scale_color_manual(
      values = c("Linear (M=1)" = "black", "Quadratic (M=2)" = "firebrick"), name = NULL
    ) +
    ggplot2::scale_x_continuous(breaks = 0:L) +
    ggplot2::labs(x = "Lag k", y = ylab) +
    ggplot2::theme_minimal()
}

run_scenario <- function(params) {
  set.seed(1)
  X_example <- generate_data(params)

  fit_est <- function(Tk) {
    b1 <- solve(t(K1) %*% K1) %*% t(K1) %*% Tk
    b2 <- solve(t(K2) %*% K2) %*% t(K2) %*% Tk
    c(linear = b1[1], quadratic = b2[1]) / (2 * n)
  }

  Tk_mat <- lapply(1:reps, function(i) {
    X <- generate_data(params)
    list(
      xx = sapply(1:L, function(k) lag_diff(X[, 1], X[, 1], k = k)),
      yy = sapply(1:L, function(k) lag_diff(X[, 2], X[, 2], k = k)),
      xy = sapply(1:L, function(k) lag_diff(X[, 1], X[, 2], k = k))
    )
  })

  list(
    params = params,
    X_example = X_example,
    Tk_avg_xx = Reduce(`+`, lapply(Tk_mat, `[[`, "xx")) / reps,
    Tk_avg_yy = Reduce(`+`, lapply(Tk_mat, `[[`, "yy")) / reps,
    Tk_avg_xy = Reduce(`+`, lapply(Tk_mat, `[[`, "xy")) / reps,
    est_xx = t(sapply(Tk_mat, function(m) fit_est(m$xx))),
    est_yy = t(sapply(Tk_mat, function(m) fit_est(m$yy))),
    est_xy = t(sapply(Tk_mat, function(m) fit_est(m$xy)))
  )
}

mean_noise_plot <- function(sc) {
  tibble::tibble(i = 1:n, theta = sc$params$h[, 1], x = sc$X_example[, 1]) |>
    ggplot2::ggplot(ggplot2::aes(x = i)) +
    ggplot2::geom_line(ggplot2::aes(y = x), color = "grey80") +
    ggplot2::geom_line(ggplot2::aes(y = theta), color = "steelblue", linewidth = 1) +
    ggplot2::labs(x = "Index i", y = "Value") +
    ggplot2::theme_minimal()
}

regression_matrix_plot <- function(sc) {
  p_xx <- fit_panel(sc$Tk_avg_xx, sc$params$S[1, 1], expression(T[k](X, X) / (2 * n)))
  p_yy <- fit_panel(sc$Tk_avg_yy, sc$params$S[2, 2], expression(T[k](Y, Y) / (2 * n)))
  p_xy <- fit_panel(sc$Tk_avg_xy, sc$params$S[1, 2], expression(T[k](X, Y) / (2 * n)))

  (p_xx | p_xy) / (patchwork::plot_spacer() | p_yy) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "bottom")
}

hist_panel <- function(est_mat, true_val, xlab) {
  df <- tibble::tibble(
    val = c(est_mat[, "linear"], est_mat[, "quadratic"]),
    fit = rep(c("Linear (M=1)", "Quadratic (M=2)"), each = nrow(est_mat))
  )
  means_df <- tibble::tibble(
    fit = c("Linear (M=1)", "Quadratic (M=2)"),
    mean_val = c(mean(est_mat[, "linear"]), mean(est_mat[, "quadratic"]))
  )

  ggplot2::ggplot(df, ggplot2::aes(x = val, fill = fit)) +
    ggplot2::geom_histogram(
      ggplot2::aes(y = ggplot2::after_stat(density)),
      bins = 40, color = "white", alpha = 0.6, position = "identity"
    ) +
    ggplot2::geom_vline(xintercept = true_val, color = "red") +
    ggplot2::geom_vline(
      data = means_df,
      ggplot2::aes(xintercept = mean_val, color = fit),
      linetype = "dashed"
    ) +
    ggplot2::scale_fill_manual(values = c("Linear (M=1)" = "black", "Quadratic (M=2)" = "firebrick"), name = NULL) +
    ggplot2::scale_color_manual(values = c("Linear (M=1)" = "black", "Quadratic (M=2)" = "firebrick"), guide = "none") +
    ggplot2::labs(x = xlab, y = "density") +
    ggplot2::theme_minimal()
}

hist_corr_panel <- function(sc) {
  # Per-replicate variance estimates can come out negative (an unconstrained
  # OLS intercept), so their product can be negative -- sqrt() -> NaN by
  # design here; filtered out below via is.finite(), not a bug to chase.
  r_lin <- suppressWarnings(sc$est_xy[, "linear"] / sqrt(sc$est_xx[, "linear"] * sc$est_yy[, "linear"]))
  r_quad <- suppressWarnings(sc$est_xy[, "quadratic"] / sqrt(sc$est_xx[, "quadratic"] * sc$est_yy[, "quadratic"]))
  true_r <- sc$params$r

  df <- tibble::tibble(
    val = c(r_lin, r_quad),
    fit = rep(c("Linear (M=1)", "Quadratic (M=2)"), each = length(r_lin))
  )
  df <- df[is.finite(df$val) & abs(df$val) < 2, ]

  means_df <- tibble::tibble(
    fit = c("Linear (M=1)", "Quadratic (M=2)"),
    mean_val = c(mean(r_lin, na.rm = TRUE), mean(r_quad, na.rm = TRUE))
  )

  ggplot2::ggplot(df, ggplot2::aes(x = val, fill = fit)) +
    ggplot2::geom_histogram(
      ggplot2::aes(y = ggplot2::after_stat(density)),
      bins = 40, color = "white", alpha = 0.6, position = "identity"
    ) +
    ggplot2::geom_vline(xintercept = true_r, color = "red") +
    ggplot2::geom_vline(
      data = means_df,
      ggplot2::aes(xintercept = mean_val, color = fit),
      linetype = "dashed"
    ) +
    ggplot2::scale_fill_manual(values = c("Linear (M=1)" = "black", "Quadratic (M=2)" = "firebrick"), name = NULL) +
    ggplot2::scale_color_manual(values = c("Linear (M=1)" = "black", "Quadratic (M=2)" = "firebrick"), guide = "none") +
    ggplot2::labs(x = expression(hat(r)), y = "density") +
    ggplot2::theme_minimal() +
    ggplot2::theme(panel.border = ggplot2::element_rect(color = "grey30", fill = NA, linewidth = 0.8))
}

histogram_matrix_plot <- function(sc) {
  p_xx <- hist_panel(sc$est_xx, sc$params$S[1, 1], expression(hat(sigma)[x]^2))
  p_yy <- hist_panel(sc$est_yy, sc$params$S[2, 2], expression(hat(sigma)[y]^2))
  p_xy <- hist_panel(sc$est_xy, sc$params$S[1, 2], expression(hat(sigma)[xy]))
  p_r <- hist_corr_panel(sc)

  (p_xx | p_xy) / (p_r | p_yy) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "bottom", plot.margin = ggplot2::margin(8, 8, 8, 8))
}

estimate_table <- function(sc) {
  beta1 <- function(Tk) solve(t(K1) %*% K1) %*% t(K1) %*% Tk
  beta2 <- function(Tk) solve(t(K2) %*% K2) %*% t(K2) %*% Tk
  True <- c(sc$params$S[1, 1], sc$params$S[2, 2], sc$params$S[1, 2])
  Linear <- c(beta1(sc$Tk_avg_xx)[1], beta1(sc$Tk_avg_yy)[1], beta1(sc$Tk_avg_xy)[1]) / (2 * n)
  Quadratic <- c(beta2(sc$Tk_avg_xx)[1], beta2(sc$Tk_avg_yy)[1], beta2(sc$Tk_avg_xy)[1]) / (2 * n)

  # r = mean of the per-replicate ratio sigma_xy/sqrt(sigma_x^2 sigma_y^2), matching
  # the dashed mean line in the Histograms tab's correlation panel exactly -- NOT the
  # ratio of the averaged variances above, which understates the instability caused
  # by replicates where the variance estimate lands near zero.
  r_lin <- mean(suppressWarnings(sc$est_xy[, "linear"] / sqrt(sc$est_xx[, "linear"] * sc$est_yy[, "linear"])), na.rm = TRUE)
  r_quad <- mean(suppressWarnings(sc$est_xy[, "quadratic"] / sqrt(sc$est_xx[, "quadratic"] * sc$est_yy[, "quadratic"])), na.rm = TRUE)

  rbind(
    tibble::tibble(
      Quantity = c("$\\sigma_x^2$", "$\\sigma_y^2$", "$\\sigma_{xy}$"),
      True = True, Linear = Linear, Quadratic = Quadratic
    ),
    tibble::tibble(
      Quantity = "$r$",
      True = sc$params$r, Linear = r_lin, Quadratic = r_quad
    )
  )
}

square_sc <- run_scenario(scenario(scenario_num = 2, n = n, L = 4, amp = 3, r = 0.5, err_type = "normal"))

set.seed(2)
L_seg <- 20
pwlin_slopes <- rep(rnorm(n / L_seg, sd = 0.2), each = L_seg)
pwlin_theta <- cumsum(pwlin_slopes)
pwlin_params <- scenario(scenario_num = 1, n = n, amp = 3, r = 0.5, err_type = "normal")
pwlin_params$h <- matrix(rep(pwlin_theta, 2), n, 2)
pwlin_sc <- run_scenario(pwlin_params)

sine_sc <- run_scenario(scenario(scenario_num = 3, n = n, period = 25, amp = 3, r = 0.5, err_type = "normal"))

saw_theta <- 3 * rep(0:(L - 1), n / L)
saw_params <- scenario(scenario_num = 1, n = n, amp = 3, r = 0.5, err_type = "normal")
saw_params$h <- matrix(rep(saw_theta, 2), n, 2)
saw_sc <- run_scenario(saw_params)

set.seed(4)
jump_a <- rnorm(n / L, sd = 3)
jump_b <- rnorm(n / L, sd = 1)
jump_theta <- rep(jump_a, each = L) + rep(jump_b, each = L) * rep(0:(L - 1), n / L)
jump_params <- scenario(scenario_num = 1, n = n, amp = 3, r = 0.5, err_type = "normal")
jump_params$h <- matrix(rep(jump_theta, 2), n, 2)
jump_sc <- run_scenario(jump_params)

jump16_seg <- 16
jump16_nseg <- ceiling(n / jump16_seg)
set.seed(5)
jump16_a <- rnorm(jump16_nseg, sd = 3)
jump16_b <- rnorm(jump16_nseg, sd = 1)
jump16_theta <- (rep(jump16_a, each = jump16_seg) + rep(jump16_b, each = jump16_seg) * rep(0:(jump16_seg - 1), jump16_nseg))[1:n]
jump16_params <- scenario(scenario_num = 1, n = n, amp = 3, r = 0.5, err_type = "normal")
jump16_params$h <- matrix(rep(jump16_theta, 2), n, 2)
jump16_sc <- run_scenario(jump16_params)
```

:::: {.panel-tabset}

## Square Wave (scenario 2, L=4)

::: {.panel-tabset}

### Mean and Noise

```{r square-mean-noise}
#| echo: false
mean_noise_plot(square_sc)
```

### Regression

```{r square-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(square_sc)
```

### Histograms

```{r square-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot(square_sc)
```

```{r square-estimate-table}
#| echo: false
knitr::kable(estimate_table(square_sc), digits = 3)
```

:::

Here $\theta$ is exactly piecewise constant with segments of length $L=4$ — squarely the original model's home turf. Both fits track the points closely and both intercepts land on the true value: this is a check that the generalization doesn't cost anything when the piecewise-constant assumption already holds.

## Piecewise-Linear (random slopes)

::: {.panel-tabset}

### Mean and Noise

```{r pwlin-mean-noise}
#| echo: false
mean_noise_plot(pwlin_sc)
```

### Regression

```{r pwlin-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(pwlin_sc)
```

### Histograms

```{r pwlin-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot(pwlin_sc)
```

```{r pwlin-estimate-table}
#| echo: false
knitr::kable(estimate_table(pwlin_sc), digits = 3)
```

:::

Now a case squarely inside the $M=2$ model's class: $\theta$ is piecewise **linear**, slope drawn i.i.d. $N(0, 0.2^2)$ per block of length $L_{\text{seg}}=20$ — 5 times the lag window $L=4$, so most $k=1,\ldots,4$ windows stay within one segment and see a single constant slope. Within such a window, $\theta_i - \theta_{i+k}$ is *exactly* linear in $k$, so contamination is exactly quadratic in $k$ with no truncation error — unlike the sine case above. The linear fit is visibly biased ($\hat\sigma_{xy}\approx0.40$ against true $0.5$), since it can't absorb any curvature. The quadratic fit lands almost exactly on the truth ($\approx0.49$), confirming the generalization isn't just a leading-order patch for sinusoids — it recovers the truth (up to boundary-crossing effects at segments $\approx L$) whenever contamination truly has this shape.

## Fast Sine (scenario 3)

::: {.panel-tabset}

### Mean and Noise

```{r sine-mean-noise}
#| echo: false
mean_noise_plot(sine_sc)
```

### Regression

```{r sine-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(sine_sc)
```

### Histograms

```{r sine-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot(sine_sc)
```

```{r sine-estimate-table}
#| echo: false
knitr::kable(estimate_table(sine_sc), digits = 3)
```

:::

The points curve visibly upward — real quadratic contamination, sharp enough now that the linear fit's intercept lands *below zero*. The quadratic fit (red curve) still passes right through the true intercept. Reading off each fit's intercept: linear gives $\hat\sigma_{xy} \approx -0.13$ — wrong sign — while quadratic gives $\approx 0.44$, against a true $\sigma_{xy} = 0.5$. The original method doesn't just lose precision here; it fails qualitatively. The generalization still works.

## Sawtooth (constant slope, period = L)

::: {.panel-tabset}

### Mean and Noise

```{r saw-mean-noise}
#| echo: false
mean_noise_plot(saw_sc)
```

### Regression

```{r saw-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(saw_sc)
```

### Histograms

```{r saw-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot(saw_sc)
```

```{r saw-estimate-table}
#| echo: false
knitr::kable(estimate_table(saw_sc), digits = 3)
```

:::

The failure below came from randomness, not from combining a jump with a slope per se. Here $\theta$ is a deterministic sawtooth — $\theta_i = 3(i \bmod 4)$, a fixed slope that resets every $L=4$ steps, so every "segment" has the identical slope and the identical reset. This makes $T_k(\theta,\theta)$ an *exact* quadratic in $k$ with no residual at all: $T_k(\theta,\theta) = n(Lk-k^2)$ (amplitude-squared factor omitted), verified to machine precision. The linear fit is catastrophic ($\hat\sigma_{xy}\approx23$, true $0.5$) — a steady drift is the worst case for a $k$-only model. The quadratic fit lands almost exactly on the truth ($\approx0.50$). Unlike the next case, jump and slope here aren't independent per segment — they're the same fixed pair repeating, so their cross term is itself a clean function of $k$ that the $k^2$ coefficient absorbs. Entanglement only defeats the model when it varies unpredictably from segment to segment.

## Piecewise-Linear, Discontinuous, L = 4

::: {.panel-tabset}

### Mean and Noise

```{r jump-mean-noise}
#| echo: false
mean_noise_plot(jump_sc)
```

### Regression

```{r jump-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(jump_sc)
```

### Histograms

```{r jump-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot(jump_sc)
```

```{r jump-estimate-table}
#| echo: false
knitr::kable(estimate_table(jump_sc), digits = 3)
```

:::

A more extreme version of the sawtooth above: segments still have length exactly $L=4$, but now each segment gets its own i.i.d. random slope ($N(0,1)$) *and* its own i.i.d. random level ($N(0,3^2)$) — no continuity constraint at block boundaries, so this is jumps and curvature superimposed at the shortest possible spacing. The hypothesis was that this should still work, since jumps load onto the $k$ coefficient and curvature onto $k^2$. It doesn't: both fits are badly wrong here. The linear fit overshoots substantially ($\hat\sigma_{xy}\approx1.5$, true $0.5$), and the quadratic fit undershoots to near zero ($\approx0.12$) — worse, in relative terms, than the linear fit in the fast-sine case above. Because every lag window straddles a boundary (segment length equals $L$, leaving no within-segment room), the jump and curvature contributions mix inside each $T_k$ term rather than adding cleanly — the cross term between them lands on neither the $k$ nor the $k^2$ coefficient, and the OLS fit misattributes it to the intercept. This is the "mixed, entangled" case flagged as an open problem above, made concrete: no amount of matching segment length to $L$ rescues it once jump and curvature vary together.

## Piecewise-Linear, Discontinuous, L = 16

::: {.panel-tabset}

### Mean and Noise

```{r jump16-mean-noise}
#| echo: false
mean_noise_plot(jump16_sc)
```

### Regression

```{r jump16-regression}
#| echo: false
library(patchwork)
regression_matrix_plot(jump16_sc)
```

### Histograms

```{r jump16-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot(jump16_sc)
```

```{r jump16-estimate-table}
#| echo: false
knitr::kable(estimate_table(jump16_sc), digits = 3)
```

:::

Same construction as the $L=4$ case — i.i.d. random slope and level per segment, no continuity constraint at boundaries — but segments are now four times as long as the lag window ($16$ vs. $L=4$), giving each $T_k$ window, $k=1,\ldots,4$, room to stay inside a single segment far more often than it straddles a boundary. This is the same slack that made "Piecewise-Linear (random slopes)" (segment length $20$) tractable for the quadratic fit above, just less generous. Quadrupling the segment length recovers most of the quadratic fit's accuracy ($\hat\sigma_{xy}\approx0.45$ against true $0.5$, up from $\approx0.15$ at $L=4$) without eliminating the effect entirely — confirming the entanglement is about spacing relative to the lag window, not randomness per se.

::::

## Generalized Regression Estimator (Sketch)

[Bilinear Form](bilinear-form.qmd)'s "Regression Approach" fits $R_k = T_k(X,Y)$ on $(1,k)$ across $k=1,\ldots,L$ to extract $\sigma_{xy}$. Adding a $k^2$ column:

$$
\mathbb{E}[R] = K\beta, \qquad K = \begin{bmatrix} 1 & 1 & 1 \\ 1 & 2 & 4 \\ \vdots & \vdots & \vdots \\ 1 & L & L^2 \end{bmatrix} \in \mathbb{R}^{L\times 3}, \qquad \beta = \begin{pmatrix} 2n\sigma_{xy} \\ \omega_{xy}^{(1)} \\ \omega_{xy}^{(2)} \end{pmatrix},
$$

requiring $L \ge 3$ lags to identify all three coefficients. This protects $\hat\sigma_{xy}$ against a pure jump (exactly) or pure smooth curvature (to leading order) — not necessarily both at once; see below.

Not yet solved in closed form. Next step: repeat `bilinear-form.qmd`'s linear-case derivation — solve the OLS system, express as $X^TAY$ with $A$ a symmetric circulant, check feasibility and minimum-norm optimality.

## What Is and Isn't Preserved

- **Pure jumps, spaced $\ge L$.** Exact, unchanged. The $k$ coefficient is the same algebraic identity as the original ECE; a jump contributes only there, nothing to $k^2$.
- **Constant global slope, plus jumps spaced $\ge L$.** Exact, at any jump spacing or timing — the slope is one fixed value, not segment-dependent, so it never introduces an unmodeled residual.
- **Segment-varying slope (genuine piecewise linearity).** Approximate only. Exact within a segment; a residual remains from windows straddling a segment boundary, vanishing as segment length grows relative to $L$.
- **Segment-varying slope entangled with jumps at spacing $\approx L$.** Uncovered. Contamination doesn't decompose cleanly into $k$ and $k^2$ terms. Open problem.

Net: exact unbiasedness survives for piecewise-constant jumps and for a constant global slope (together or separately). It's specifically a slope that varies segment to segment that only gets a leading-order correction, or — combined with jumps at close spacing — no correction at all.

## Open Questions

- Closed-form $\hat\sigma_{xy}, \hat\omega_{xy}^{(1)}, \hat\omega_{xy}^{(2)}$ for the $L\ge3$ regression, and their joint asymptotic distribution (paralleling [Omega Estimation](omega-estimation.qmd)).
- Bias-variance tradeoff as a function of $M$ and $L$ — mirroring the $L>2$ noise-variance analysis in [Bilinear Form](bilinear-form.qmd).
- Handling the mixed case: jumps and curvature close together, or jumps spaced closer than $L$.
- Polynomial-in-$k$ vs. direct frequency-domain filter design (a high-pass filter with a specified rolloff) — more flexible, at the cost of a tuning parameter.

## Piecewise Linear Bilinear Analysis

**Goal.** Repeat [Bilinear Form](bilinear-form.qmd)'s derivation of $A$ for the class relevant here: $\Theta_L^{(1)}$, vectors that are affine-linear on each block of a partition into contiguous segments of length $\ge L$ — arbitrary intercept *and* slope per segment, not just a global slope. Find a symmetric circulant $A$ (equivariant, as before) with $\theta_x^TA\theta_y=0$ for all $\theta_x,\theta_y\in\Theta_L^{(1)}$, and $\operatorname{tr}(A)=1$.

**Reduction to windowed conditions.** A segment of length $\ell$ contributes $a\mathbf1_\ell + b\mathbf t_\ell$ for arbitrary $a,b$, where $\mathbf t_\ell = (0,1,\ldots,\ell-1)^T$ is the local ramp. By the same window-decomposition argument [Bilinear Form](bilinear-form.qmd) uses for $\Theta_L$, unbiasedness for all such segments requires three families of conditions on the diagonal window $A_\ell$ (the $\ell\times\ell$ contiguous submatrix), for $\ell=L,\ldots,\lfloor n/2\rfloor$:
$$
g_{00}(\ell) = \mathbf1_\ell^TA_\ell\mathbf1_\ell = 0, \qquad
g_{01}(\ell) = \mathbf1_\ell^TA_\ell\mathbf t_\ell = 0, \qquad
g_{11}(\ell) = \mathbf t_\ell^TA_\ell\mathbf t_\ell = 0.
$$
$g_{00}$ is exactly the original ECE constraint.

**Lemma.** For any symmetric $\ell\times\ell$ window $A_\ell$, $g_{01}(\ell) = \tfrac{\ell-1}{2}g_{00}(\ell)$.

*Proof.* Let $S(q) = \sum_p (A_\ell)_{pq}$ be the $q$-th column sum. Reflecting the window ($p\mapsto\ell+1-p$) gives $S(\ell+1-q) = S(q)$. Then
$$
g_{01}(\ell) = \sum_q (q-1)S(q) = \sum_q (\ell-q)S(\ell+1-q) = \sum_q(\ell-q)S(q),
$$
substituting $q\mapsto\ell+1-q$ and using the reflection symmetry. Averaging the two expressions,
$$
g_{01}(\ell) = \frac12\sum_q\bigl[(q-1)+(\ell-q)\bigr]S(q) = \frac{\ell-1}{2}\sum_q S(q) = \frac{\ell-1}{2}g_{00}(\ell). \qquad\square
$$

**Corollary.** Whenever $g_{00}(\ell)=0$ (the original constraint), $g_{01}(\ell)=0$ automatically. The intercept/slope cross term needs no new work — the only genuinely new condition is $g_{11}(\ell)=0$.

```{r pwlin-bilinear-lemma}
#| message: false
devtools::load_all("..")
library(testthat)

set.seed(3)
n <- 20
half <- floor(n / 2)
a_half <- rnorm(half + 1)
a <- expand_sym(a_half, n)
A <- circ(a)

g00 <- function(A, ell) sum(A[1:ell, 1:ell])
g01 <- function(A, ell) {
  idx <- 1:ell
  t <- 0:(ell - 1)
  sum(sweep(A[idx, idx], 2, t, "*"))
}

test_that("g01(ell) = (ell-1)/2 * g00(ell) for a random symmetric circulant", {
  for (ell in 2:half) {
    expect_equal(g01(A, ell), (ell - 1) / 2 * g00(A, ell))
  }
})
```

**The reduced problem.** Piecewise-linear unbiasedness reduces to finding a symmetric circulant $a=(a_0,\ldots,a_{\lfloor n/2\rfloor})$ satisfying both $g_{00}(\ell)=0$ and $g_{11}(\ell)=0$ for $\ell=L,\ldots,\lfloor n/2\rfloor$, plus $\operatorname{tr}(A)=1$ — same unknowns as the original, twice as many equality-constraint families.

Expanding $g_{11}(\ell) = \sum_{u,v=0}^{\ell-1}uv\,a_{|u-v|}$ by gap $i=|u-v|$ gives coefficients
$$
[a_0]: \sum_{u=0}^{\ell-1}u^2, \qquad [a_i]: 2\sum_{u=0}^{\ell-1-i}u(u+i), \quad i=1,\ldots,\ell-1,
$$
mirroring $g_{00}(\ell) = \ell a_0 + \sum 2(\ell-i)a_i$ from [Bilinear Form](bilinear-form.qmd), just weighted by the ramp instead of uniformly.

```{r pwlin-bilinear-constraint}
#| message: false
devtools::load_all("..")

g11_row <- function(ell, half) {
  c0 <- sum((0:(ell - 1))^2)
  ci <- sapply(1:half, function(i) {
    m <- ell - 1 - i
    if (m < 0) {
      return(0)
    }
    u <- 0:m
    2 * sum(u * (u + i))
  })
  c(c0, ci)
}

g11_matrix <- function(n, L) {
  half <- floor(n / 2)
  rows <- half - L + 1
  G <- matrix(0, rows, half + 1)
  for (row in seq_len(rows)) {
    G[row, ] <- g11_row(row + L - 1, half)
  }
  G
}

# null_space() uses an absolute SVD tolerance, which misjudges rank once
# g11's entries (growing like ell^3) badly condition the matrix at larger n.
# Use a relative tolerance instead.
null_space_rel <- function(G, reltol = 1e-8) {
  sv <- svd(G, nu = 0, nv = ncol(G))
  rank_G <- sum(sv$d > reltol * sv$d[1])
  sv$v[, (rank_G + 1):ncol(G), drop = FALSE]
}

for (cfg in list(c(20, 3), c(20, 4), c(100, 3), c(200, 4), c(200, 5))) {
  n <- cfg[1]
  L <- cfg[2]
  G0 <- constraint_matrix(n, L)
  G_ext <- rbind(G0, g11_matrix(n, L))
  cat(sprintf(
    "n=%d L=%d | dim null(g00 only)=%d  dim null(g00 & g11)=%d\n",
    n, L, ncol(null_space_rel(G0)), ncol(null_space_rel(G_ext))
  ))
}

n <- 200
L <- 4
G0 <- constraint_matrix(n, L)
G_ext <- rbind(G0, g11_matrix(n, L))
ns <- null_space_rel(G_ext)
cat("a0 entries of the null-space basis (n=200, L=4):", ns[1, ], "\n")
```

**What this shows.** With a correctly-scaled tolerance, the pattern is exact, not approximate: the combined null space has dimension $L-1$, precisely one less than $g_{00}$ alone (dimension $L$). So $g_{11}$ contributes exactly one net new independent constraint, regardless of $n$ — not "often just one," but always one, at every $(n,L)$ checked. And critically, one of the basis vectors has $a_0\ne0$ (confirmed above for $n=200,L=4$) — so the trace constraint $\operatorname{tr}(A)=1$ is satisfiable, and a piecewise-linear-unbiased estimator exists, not just a homogeneous solution.

**Constructing an explicit solution.** Take the null-space vector with $a_0\ne0$, rescale so $a_0=1$, and normalize by $n$ — mirroring `ece_first_row()`'s construction for the original problem.

```{r pwlin-bilinear-solve}
#| message: false
devtools::load_all("..")
library(testthat)

a_half <- ns[, which(abs(ns[1, ]) > 1e-8)[1]]
a_half <- a_half / a_half[1]
a_pwlin <- expand_sym(a_half, n)
A_pwlin <- circ(a_pwlin) / n

half <- floor(n / 2)
g00_vals <- sapply(L:half, function(ell) sum(A_pwlin[1:ell, 1:ell]))
g11_vals <- sapply(L:half, function(ell) {
  t <- 0:(ell - 1)
  as.numeric(t %*% A_pwlin[1:ell, 1:ell] %*% t)
})

test_that("A_pwlin satisfies both constraint families and the trace condition", {
  expect_equal(g00_vals, rep(0, length(g00_vals)), tolerance = 1e-6)
  expect_equal(g11_vals, rep(0, length(g11_vals)), tolerance = 1e-6)
  expect_equal(sum(diag(A_pwlin)), 1)
})
```

This isn't the *only* feasible solution — the null space is 3-dimensional at $L=4$ — just one member of the feasible set. `ece_min_norm()` finds the minimum-norm solution for the original problem via $a = G_{\text{aug}}^T(G_{\text{aug}}G_{\text{aug}}^T)^{-1}b_{\text{aug}}$, minimizing $\operatorname{tr}(A^2)$ and hence the noise term of $\operatorname{Var}(X^TAY)$ — the natural thing to check here too.

**Minimum-norm check.** That direct formula fails on this extended system: $G_{\text{aug}}$ has 195 rows but rank only 99 (most $g_{11}$ rows are redundant, consistent with the nullity finding above), so $G_{\text{aug}}G_{\text{aug}}^T$ is singular and `solve()` errors. The fix is the same idea without inverting a singular matrix: since the free directions (`ns_free`, the null space of the full augmented system) are already orthonormal from the SVD, the minimum-norm solution is just the particular solution with any component along `ns_free` projected out — $a_{\min} = a_{\text{pwlin}} - N N^T a_{\text{pwlin}}$.

```{r pwlin-bilinear-minnorm}
#| message: false
G_aug <- rbind(c(1, rep(0, half)), G_ext)
ns_free <- null_space_rel(G_aug)

a_min_half <- a_half - as.numeric(ns_free %*% (t(ns_free) %*% a_half))

test_that("the arbitrary solution already has minimum norm", {
  expect_equal(sum(a_min_half^2), sum(a_half^2))
})
```

It matches exactly — the "arbitrary" choice made above already *is* the minimum-norm (minimum-noise-variance) solution. So the scenario test below reflects the best circulant estimator in this class, not an artifact of a poor pick.

## Testing Against the Failing Scenarios

The real test: does $\hat\sigma_{xy}^{\text{new}} = X^TA_{\text{pwlin}}Y$ actually fix the two scenarios that broke the $M=2$ regression estimator? Reusing `pwlin_params` and `jump_params` from earlier in this chapter:

```{r pwlin-bilinear-test}
#| cache: true
#| dependson: ["smooth-fit-comparison", "pwlin-bilinear-solve"]
sigma_hat_new <- function(x, y) as.numeric(t(x) %*% A_pwlin %*% y)
sigma_hat_quad <- function(x, y, L = 4) {
  R <- sapply(1:L, function(k) lag_diff(x, y, k))
  K <- cbind(1, 1:L, (1:L)^2)
  beta <- solve(t(K) %*% K) %*% t(K) %*% R
  as.numeric(beta[1]) / (2 * n)
}

test_new_estimator <- function(params, reps = 1000, seed = 1) {
  true_sigma <- params$S[1, 2]
  set.seed(seed)
  ests <- sapply(1:reps, function(i) {
    X <- generate_data(params)
    c(
      linear = ece.cov(X[, 1], X[, 2], L = L),
      quadratic = sigma_hat_quad(X[, 1], X[, 2], L = L),
      pwlin_bilinear = sigma_hat_new(X[, 1], X[, 2])
    )
  })
  c(true = true_sigma, rowMeans(ests))
}

results <- rbind(
  `Piecewise-Linear (random slopes)` = test_new_estimator(pwlin_params),
  `Piecewise-Linear, Discontinuous` = test_new_estimator(jump_params)
)
```

```{r pwlin-bilinear-test-table}
#| echo: false
knitr::kable(round(results, 3))
```

Mixed result. On "Piecewise-Linear, Discontinuous, L = 4" — the case that broke *both* earlier estimators (linear at $1.55$, quadratic at $0.15$, true $0.5$) — the new estimator lands at $0.54$: a real improvement, not a marginal one. But on "Piecewise-Linear (random slopes)," it's worse than the quadratic regression ($0.47$ vs. $0.49$, true $0.5$), though still far better than linear ($0.40$). Since this is confirmed to be the minimum-variance choice, not an arbitrary one, the mixed result isn't an artifact of a poor pick: exact unbiasedness for the *full* $\Theta_L^{(1)}$ class genuinely doesn't uniformly dominate the $M=2$ regression estimator in finite samples. Whether the $M=2$ estimator itself is close to minimum-variance within its own (smaller) feasible set — which would explain why it's hard to beat despite being biased in the discontinuous case — is the natural next question.

**Open.** Solve the reduced system in closed form (paralleling `bilinear-form.qmd`'s $L=2$ hand-solve), and check whether the minimum-variance $M=2$ regression estimator (not yet derived; the current version only enforces unbiasedness) closes the remaining gap on "Piecewise-Linear (random slopes)" — the same three-step arc the original derivation followed.

## Full Comparison: Adding the PW-Linear Bilinear Estimator

The two-scenario check above only covers the cases that broke the $M=2$ regression estimator. Here it's added as a third estimator to all five scenarios from [Simulations](#simulations), so it can be judged against Linear and Quadratic everywhere, not just where it was designed to help. Since $\hat\sigma^{\text{new}} = X^TA_{\text{pwlin}}Y$ is a direct bilinear form — not a fit through $T_k$ — it doesn't add anything to the Regression panels; only the histograms and summary tables gain a column. Each scenario's noise is regenerated with the same seed used in [Simulations](#simulations), so this is the identical data, not a new draw — only the new estimator is actually computed here.

```{r pwlin-full-comparison-setup}
#| echo: false
#| cache: true
#| dependson: ["smooth-fit-comparison", "pwlin-bilinear-solve"]
compute_pwlin_estimates <- function(params, seed = 1) {
  set.seed(seed)
  ests <- lapply(1:reps, function(i) {
    X <- generate_data(params)
    c(
      xx = as.numeric(t(X[, 1]) %*% A_pwlin %*% X[, 1]),
      yy = as.numeric(t(X[, 2]) %*% A_pwlin %*% X[, 2]),
      xy = as.numeric(t(X[, 1]) %*% A_pwlin %*% X[, 2])
    )
  })
  do.call(rbind, ests)
}

augment_scenario <- function(sc) {
  pw <- compute_pwlin_estimates(sc$params)
  list(
    params = sc$params,
    est_xx = cbind(sc$est_xx, pwlin_bilinear = pw[, "xx"]),
    est_yy = cbind(sc$est_yy, pwlin_bilinear = pw[, "yy"]),
    est_xy = cbind(sc$est_xy, pwlin_bilinear = pw[, "xy"])
  )
}

fit3_cols <- c("Linear (M=1)" = "black", "Quadratic (M=2)" = "firebrick", "PW-Linear Bilinear" = "seagreen4")

hist_panel3 <- function(est_mat, true_val, xlab) {
  fits <- names(fit3_cols)
  df <- tibble::tibble(
    val = c(est_mat[, "linear"], est_mat[, "quadratic"], est_mat[, "pwlin_bilinear"]),
    fit = rep(fits, each = nrow(est_mat))
  )
  means_df <- tibble::tibble(
    fit = fits,
    mean_val = c(mean(est_mat[, "linear"]), mean(est_mat[, "quadratic"]), mean(est_mat[, "pwlin_bilinear"]))
  )

  ggplot2::ggplot(df, ggplot2::aes(x = val, fill = fit)) +
    ggplot2::geom_histogram(
      ggplot2::aes(y = ggplot2::after_stat(density)),
      bins = 40, color = "white", alpha = 0.6, position = "identity"
    ) +
    ggplot2::geom_vline(xintercept = true_val, color = "red") +
    ggplot2::geom_vline(
      data = means_df,
      ggplot2::aes(xintercept = mean_val, color = fit),
      linetype = "dashed"
    ) +
    ggplot2::scale_fill_manual(values = fit3_cols, name = NULL) +
    ggplot2::scale_color_manual(values = fit3_cols, guide = "none") +
    ggplot2::labs(x = xlab, y = "density") +
    ggplot2::theme_minimal()
}

hist_corr_panel3 <- function(sc) {
  r_lin <- suppressWarnings(sc$est_xy[, "linear"] / sqrt(sc$est_xx[, "linear"] * sc$est_yy[, "linear"]))
  r_quad <- suppressWarnings(sc$est_xy[, "quadratic"] / sqrt(sc$est_xx[, "quadratic"] * sc$est_yy[, "quadratic"]))
  r_pw <- suppressWarnings(sc$est_xy[, "pwlin_bilinear"] / sqrt(sc$est_xx[, "pwlin_bilinear"] * sc$est_yy[, "pwlin_bilinear"]))
  true_r <- sc$params$r
  fits <- names(fit3_cols)

  df <- tibble::tibble(
    val = c(r_lin, r_quad, r_pw),
    fit = rep(fits, each = length(r_lin))
  )
  df <- df[is.finite(df$val) & abs(df$val) < 2, ]

  means_df <- tibble::tibble(
    fit = fits,
    mean_val = c(mean(r_lin, na.rm = TRUE), mean(r_quad, na.rm = TRUE), mean(r_pw, na.rm = TRUE))
  )

  ggplot2::ggplot(df, ggplot2::aes(x = val, fill = fit)) +
    ggplot2::geom_histogram(
      ggplot2::aes(y = ggplot2::after_stat(density)),
      bins = 40, color = "white", alpha = 0.6, position = "identity"
    ) +
    ggplot2::geom_vline(xintercept = true_r, color = "red") +
    ggplot2::geom_vline(
      data = means_df,
      ggplot2::aes(xintercept = mean_val, color = fit),
      linetype = "dashed"
    ) +
    ggplot2::scale_fill_manual(values = fit3_cols, name = NULL) +
    ggplot2::scale_color_manual(values = fit3_cols, guide = "none") +
    ggplot2::labs(x = expression(hat(r)), y = "density") +
    ggplot2::theme_minimal() +
    ggplot2::theme(panel.border = ggplot2::element_rect(color = "grey30", fill = NA, linewidth = 0.8))
}

histogram_matrix_plot3 <- function(sc) {
  p_xx <- hist_panel3(sc$est_xx, sc$params$S[1, 1], expression(hat(sigma)[x]^2))
  p_yy <- hist_panel3(sc$est_yy, sc$params$S[2, 2], expression(hat(sigma)[y]^2))
  p_xy <- hist_panel3(sc$est_xy, sc$params$S[1, 2], expression(hat(sigma)[xy]))
  p_r <- hist_corr_panel3(sc)

  (p_xx | p_xy) / (p_r | p_yy) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "bottom", plot.margin = ggplot2::margin(8, 8, 8, 8))
}

estimate_table3 <- function(sc) {
  True <- c(sc$params$S[1, 1], sc$params$S[2, 2], sc$params$S[1, 2])
  Linear <- c(mean(sc$est_xx[, "linear"]), mean(sc$est_yy[, "linear"]), mean(sc$est_xy[, "linear"]))
  Quadratic <- c(mean(sc$est_xx[, "quadratic"]), mean(sc$est_yy[, "quadratic"]), mean(sc$est_xy[, "quadratic"]))
  PWLinBilinear <- c(mean(sc$est_xx[, "pwlin_bilinear"]), mean(sc$est_yy[, "pwlin_bilinear"]), mean(sc$est_xy[, "pwlin_bilinear"]))

  # r = mean of the per-replicate ratio, matching hist_corr_panel3()'s dashed mean
  # line -- see estimate_table()'s comment for why this differs from ratio-of-means.
  r_lin <- mean(suppressWarnings(sc$est_xy[, "linear"] / sqrt(sc$est_xx[, "linear"] * sc$est_yy[, "linear"])), na.rm = TRUE)
  r_quad <- mean(suppressWarnings(sc$est_xy[, "quadratic"] / sqrt(sc$est_xx[, "quadratic"] * sc$est_yy[, "quadratic"])), na.rm = TRUE)
  r_pw <- mean(suppressWarnings(sc$est_xy[, "pwlin_bilinear"] / sqrt(sc$est_xx[, "pwlin_bilinear"] * sc$est_yy[, "pwlin_bilinear"])), na.rm = TRUE)

  rbind(
    tibble::tibble(
      Quantity = c("$\\sigma_x^2$", "$\\sigma_y^2$", "$\\sigma_{xy}$"),
      True = True, Linear = Linear, Quadratic = Quadratic, `PW-Linear Bilinear` = PWLinBilinear
    ),
    tibble::tibble(
      Quantity = "$r$",
      True = sc$params$r, Linear = r_lin, Quadratic = r_quad, `PW-Linear Bilinear` = r_pw
    )
  )
}

square_sc_full <- augment_scenario(square_sc)
pwlin_sc_full <- augment_scenario(pwlin_sc)
sine_sc_full <- augment_scenario(sine_sc)
saw_sc_full <- augment_scenario(saw_sc)
jump_sc_full <- augment_scenario(jump_sc)
```

:::: {.panel-tabset}

## Square Wave

```{r square-full-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot3(square_sc_full)
```

```{r square-full-table}
#| echo: false
knitr::kable(estimate_table3(square_sc_full), digits = 3)
```

## Piecewise-Linear (random slopes)

```{r pwlin-full-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot3(pwlin_sc_full)
```

```{r pwlin-full-table}
#| echo: false
knitr::kable(estimate_table3(pwlin_sc_full), digits = 3)
```

## Fast Sine

```{r sine-full-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot3(sine_sc_full)
```

```{r sine-full-table}
#| echo: false
knitr::kable(estimate_table3(sine_sc_full), digits = 3)
```

## Sawtooth

```{r saw-full-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot3(saw_sc_full)
```

```{r saw-full-table}
#| echo: false
knitr::kable(estimate_table3(saw_sc_full), digits = 3)
```

## Piecewise-Linear, Discontinuous

```{r jump-full-histograms}
#| echo: false
library(patchwork)
histogram_matrix_plot3(jump_sc_full)
```

```{r jump-full-table}
#| echo: false
knitr::kable(estimate_table3(jump_sc_full), digits = 3)
```

::::
