---
title: "Type I Error"
execute:
  echo: false
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
devtools::load_all()
results_t1 <- read_csv("../data/results-t1.csv")

n_sim <- 10000

method_colors <- c(
  "z.test.gaussian"      = "#E41A1C",
  "z.test.oracle.kappa"  = "#377EB8",
  "bs.multiplier"        = "#4DAF4A",
  "bs.parametric"        = "#FF7F00",
  "bs.parametric.psd"    = "#0072B2",
  "z.test.null"          = "#984EA3",
  "z.test.null.gaussian" = "#A65628"
)
```

```{R} 
sim_dependent_quadratic <- function(params, method, n_sims=10000){
  set.seed(1)
  map_dbl(1:n_sims, ~{
    hx <- params$h[,1]
    hy <- params$h[,2]
    ex <- rnorm(params$n)
    ey <- (ex^2 - 1) / sqrt(2)
    X <- cbind(hx + ex, hy + ey)

    test.diag(X, method=method)$p.value
  })
}
```

## $p = 2$: Z-Test Methods

```{r}
#| cache: true
params_p2_s1 <- scenario(1, n = 1000, err_type = "normal") 
params_p2    <- scenario(2, L = 4, n = 1000, err_type = "exponential")

pvals_z_s1 <- list(
  "Z-Test | Null, Oracle"   = pcurve(params_p2_s1, "z.test.null.oracle",   n_sim = n_sim),
  "Z-Test | Null, Gaussian" = pcurve(params_p2_s1, "z.test.null.gaussian", n_sim = n_sim),
  "Z-Test | Null"           = pcurve(params_p2_s1, "z.test.null",          n_sim = n_sim),
  "Z-Test | Oracle"         = pcurve(params_p2_s1, "z.test.oracle",        n_sim = n_sim),
  "Z-Test | Gaussian"       = pcurve(params_p2_s1, "z.test.gaussian",      n_sim = n_sim),
  "Z-Test"                  = pcurve(params_p2_s1, "z.test",               n_sim = n_sim)
)

pvals_z_s2 <- list(
  "Z-Test | Null, Oracle"   = pcurve(params_p2, "z.test.null.oracle",   n_sim = n_sim),
  "Z-Test | Null, Gaussian" = pcurve(params_p2, "z.test.null.gaussian", n_sim = n_sim),
  "Z-Test | Null"           = pcurve(params_p2, "z.test.null",          n_sim = n_sim),
  "Z-Test | Oracle"         = pcurve(params_p2, "z.test.oracle",        n_sim = n_sim),
  "Z-Test | Gaussian"       = pcurve(params_p2, "z.test.gaussian",      n_sim = n_sim),
  "Z-Test"                  = pcurve(params_p2, "z.test",               n_sim = n_sim)
)
```

::: {.panel-tabset}

## Scenario 1 (Gaussian)

```{r}
#| echo: false
plot_pcurve(pvals_z_s1, title = "p = 2 — Scenario 1, Gaussian errors, σ = (1, 1)", dir = "v", ncol = 2)
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_z_s2, title = "p = 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)", dir = "v", ncol = 2)
```

:::

## $p = 2$: Null Z-Test $\kappa_{22}$ Estimators

```{r pvals-null}
#| cache: true
#| warning: false
n_vals        <- c(200, 500, 1000)
methods_null  <- c("z.test.null", "z.test.null.regress", "z.test.null.matrix")
method_labels <- c("Z-Test | Null (MoM)", "Z-Test | Null (Regress)", "Z-Test | Null (Matrix)")

pvals_null <- map_dfr(n_vals, \(n) {
  s1 <- scenario(1, n = n, err_type = "normal")
  s2 <- scenario(2, L = 4, n = n, err_type = "exponential")
  map_dfr(seq_along(methods_null), \(j) {
    bind_rows(
      tibble(p_value = pcurve(s1, methods_null[j], n_sim = n_sim),
             method = method_labels[j], n = n, scenario = "Scenario 1"),
      tibble(p_value = pcurve(s2, methods_null[j], n_sim = n_sim),
             method = method_labels[j], n = n, scenario = "Scenario 2")
    )
  })
}) |>
  mutate(
    n      = factor(n, levels = n_vals),
    method = factor(method, levels = method_labels)
  )
```

```{r}
na_tbl <- pvals_null |>
  summarise(
    n_na = sum(is.na(p_value)),
    pct  = mean(is.na(p_value)),
    .by  = c(scenario, method, n)
  ) |>
  filter(n_na > 0) |>
  arrange(scenario, n, method)

if (nrow(na_tbl) > 0) {
  knitr::kable(na_tbl, digits = 3,
    col.names = c("Scenario", "Method", "n", "# NA", "Frac. NA"))
} else {
  cat("No NA p-values across all methods and sample sizes.")
}
```

::: {.panel-tabset}

## Scenario 1 (Gaussian)

```{r}
pvals_null |>
  filter(scenario == "Scenario 1") |>
  plot_pcurve(facet = facet_grid(n ~ method),
              title = "p = 2 — Scenario 1, Gaussian errors")
```

## Scenario 2, L=4 (Exponential)

```{r}
pvals_null |>
  filter(scenario == "Scenario 2") |>
  plot_pcurve(facet = facet_grid(n ~ method),
              title = "p = 2 — Scenario 2 (L = 4), exponential errors")
```

:::

## $p = 2$: Dependence Testing

Correlation dependence testing are closely related but subtly different. In the classical Pearson test, interpretation is dependent on on the assumptions made. 

Consider this canonical example:
$$
\varepsilon_y = \frac{\varepsilon_x^2 - 1}{\sqrt{2}}, \quad 
\varepsilon_x \sim N(0, 1).
$$
Clearly, $\varepsilon_x$ and $\varepsilon_y$ are dependent, but by construction $\operatorname{Cor}(\varepsilon_x, \varepsilon_y) = 0$. Specifically, $\kappa_{22} = 5$. 

Therefore, a dependence test will identify dependence between $\varepsilon_x$ and $\varepsilon_y$, but a correlation test will not. 

::: {.panel-tabset}

## Scenario 1 (Dependent)

```{R dep-test-quadratic-s1}
#| cache: true
#| echo: false
params <- scenario(1) 
p_cor <- sim_dependent_quadratic(params, method="z.test.null")
p_dep <- sim_dependent_quadratic(params, method="z.test.null.gaussian")
```

```{R} 
#| echo: false
plot_pcurve(list(
  "Correlation test" = p_cor,
  "Dependence test"  = p_dep
))
```

## Scenario 2 (Dependent)

```{R dep-test-quadratic-s2}
#| cache: true
#| echo: false
params <- scenario(2, L=4) 
p_cor <- sim_dependent_quadratic(params, method="z.test.null")
p_dep <- sim_dependent_quadratic(params, method="z.test.null.gaussian")
```

```{R} 
#| echo: false
plot_pcurve(list(
  "Correlation test" = p_cor,
  "Dependence test"  = p_dep
))
```

:::

## $p = 2$: Bootstrap Methods

```{r bs-multiplier-s1-p2}
#| cache: true
#| echo: false
n_sim <- 10
pvals_bs_s1 <- list(
  "Bootstrap | Multiplier"           = pcurve(params_p2_s1, "bs.multiplier",         n_sim = n_sim),
  "Bootstrap | Multiplier, No Split" = pcurve(params_p2_s1, "bs.multiplier.nosplit",  n_sim = n_sim)
)
```

```{r bs-multiplier-s2-p2}
#| cache: true
#| echo: false
n_sim <- 10
pvals_bs_s2 <- list(
  "Bootstrap | Multiplier"           = pcurve(params_p2, "bs.multiplier",         n_sim = n_sim),
  "Bootstrap | Multiplier, No Split" = pcurve(params_p2, "bs.multiplier.nosplit",  n_sim = n_sim)
)
n_sim <- 10000
```

::: {.panel-tabset}

## Scenario 1 (Normal)

```{r}
#| echo: false
plot_pcurve(pvals_bs_s1, title = "p = 2 — Scenario 1, normal errors, σ = (1, 1)", ncol = 2)
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_bs_s2, title = "p = 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)", ncol = 2)
```

:::

## $p = 2$: Parametric Bootstrap Methods

```{r bs-parametric-s1-p2}
#| cache: true
#| echo: false
pvals_bsp_s1 <- list(
  "Bootstrap | Parametric"      = pcurve(params_p2_s1, "bs.parametric",     n_sim = n_sim),
  "Bootstrap | Parametric, PSD" = pcurve(params_p2_s1, "bs.parametric.psd", n_sim = n_sim)
)
```

```{r bs-parametric-s2-p2}
#| cache: true
#| echo: false
pvals_bsp_s2 <- list(
  "Bootstrap | Parametric"      = pcurve(params_p2, "bs.parametric",     n_sim = n_sim),
  "Bootstrap | Parametric, PSD" = pcurve(params_p2, "bs.parametric.psd", n_sim = n_sim)
)
```

::: {.panel-tabset}

## Scenario 1 (Normal)

```{r}
#| echo: false
plot_pcurve(pvals_bsp_s1, title = "p = 2 — Scenario 1, normal errors, σ = (1, 1)", ncol = 2)
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_bsp_s2, title = "p = 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)", ncol = 2)
```

:::

## $p > 2$: Multiplier Bootstrap

```{r}
params_p3_s1  <- scenario(1, n = 1000, err_type = "normal",            p = 3)
params_p5_s1  <- scenario(1, n = 1000, err_type = "normal",            p = 5)
params_p10_s1 <- scenario(1, n = 1000, err_type = "normal",            p = 10)
params_p3_s2  <- scenario(2, L = 4, n = 1000, err_type = "exponential", p = 3)
params_p5_s2  <- scenario(2, L = 4, n = 1000, err_type = "exponential", p = 5)
params_p10_s2 <- scenario(2, L = 4, n = 1000, err_type = "exponential", p = 10)
```

```{r mult-p3}
#| cache: true
pvals_mult_p3_s1 <- pcurve(params_p3_s1, "bs.multiplier", n_sim = n_sim)
pvals_mult_p3_s2 <- pcurve(params_p3_s2, "bs.multiplier", n_sim = n_sim)
```

```{r mult-p5}
#| cache: true
pvals_mult_p5_s1 <- pcurve(params_p5_s1, "bs.multiplier", n_sim = n_sim)
pvals_mult_p5_s2 <- pcurve(params_p5_s2, "bs.multiplier", n_sim = n_sim)
```

```{r mult-p10}
#| cache: true
pvals_mult_p10_s1 <- pcurve(params_p10_s1, "bs.multiplier", n_sim = n_sim)
pvals_mult_p10_s2 <- pcurve(params_p10_s2, "bs.multiplier", n_sim = n_sim)
```

```{r}
pvals_mult_pg2_s1 <- bind_rows(
  tibble(p_value = pvals_mult_p3_s1,  p = factor(3,  levels = c(3, 5, 10))),
  tibble(p_value = pvals_mult_p5_s1,  p = factor(5,  levels = c(3, 5, 10))),
  tibble(p_value = pvals_mult_p10_s1, p = factor(10, levels = c(3, 5, 10)))
)

pvals_mult_pg2_s2 <- bind_rows(
  tibble(p_value = pvals_mult_p3_s2,  p = factor(3,  levels = c(3, 5, 10))),
  tibble(p_value = pvals_mult_p5_s2,  p = factor(5,  levels = c(3, 5, 10))),
  tibble(p_value = pvals_mult_p10_s2, p = factor(10, levels = c(3, 5, 10)))
)
```

::: {.panel-tabset}

## Scenario 1 (Normal)

```{r}
#| echo: false
plot_pcurve(pvals_mult_pg2_s1, facet = ~p, ncol = 3,
  title = "p > 2 — Scenario 1, normal errors, σ = (1, 1)")
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_mult_pg2_s2, facet = ~p, ncol = 3,
  title = "p > 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)")
```

:::

## $p > 2$: Parametric Bootstrap

```{r param-p3}
#| cache: true
pvals_param_p3_s1 <- bind_rows(
  tibble(p_value = pcurve(params_p3_s1, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p3_s1, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
pvals_param_p3_s2 <- bind_rows(
  tibble(p_value = pcurve(params_p3_s2, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p3_s2, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
```

```{r param-p5}
#| cache: true
pvals_param_p5_s1 <- bind_rows(
  tibble(p_value = pcurve(params_p5_s1, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p5_s1, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
pvals_param_p5_s2 <- bind_rows(
  tibble(p_value = pcurve(params_p5_s2, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p5_s2, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
```

```{r param-p10}
#| cache: true
pvals_param_p10_s1 <- bind_rows(
  tibble(p_value = pcurve(params_p10_s1, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p10_s1, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
pvals_param_p10_s2 <- bind_rows(
  tibble(p_value = pcurve(params_p10_s2, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p10_s2, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
```

```{r}
p_levels <- c(3, 5, 10)
m_levels <- c("Bootstrap | Parametric", "Bootstrap | Parametric, PSD")

pvals_param_pg2_s1 <- bind_rows(
  mutate(pvals_param_p3_s1,  p = 3),
  mutate(pvals_param_p5_s1,  p = 5),
  mutate(pvals_param_p10_s1, p = 10)
) |> mutate(p = factor(p, levels = p_levels), method = factor(method, levels = m_levels))

pvals_param_pg2_s2 <- bind_rows(
  mutate(pvals_param_p3_s2,  p = 3),
  mutate(pvals_param_p5_s2,  p = 5),
  mutate(pvals_param_p10_s2, p = 10)
) |> mutate(p = factor(p, levels = p_levels), method = factor(method, levels = m_levels))
```

::: {.panel-tabset}

## Scenario 1 (Normal)

```{r}
#| echo: false
plot_pcurve(pvals_param_pg2_s1, facet = facet_grid(p ~ method),
  title = "p > 2 — Scenario 1, normal errors, σ = (1, 1)")
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_param_pg2_s2, facet = facet_grid(p ~ method),
  title = "p > 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)")
```

:::

## Bootstrap Tests are Conservative

::: {.panel-tabset}

## Scenario 1

```{r bs-conservative-s1}
results_t1 |>
  filter(
    method   == "bs.multiplier",
    n        %in% c(1000, 2000, 3000, 5000),
    p        %in% c(15, 20, 25),
    scenario == 1,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(
    n = factor(n, levels = c(1000, 2000, 3000, 5000)),
    p = factor(p, levels = c(15, 20, 25))
  ) |>
  plot_pvals(facet = facet_grid(n ~ p, labeller = label_both),
             title = "Bootstrap Multiplier — Scenario 1, Normal errors")
```

## Scenario 2

```{r bs-conservative-s2}
results_t1 |>
  filter(
    method   == "bs.multiplier",
    n        %in% c(1000, 2000, 3000, 5000),
    p        %in% c(15, 20, 25),
    scenario == 2,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(
    n = factor(n, levels = c(1000, 2000, 3000, 5000)),
    p = factor(p, levels = c(15, 20, 25))
  ) |>
  plot_pvals(facet = facet_grid(n ~ p, labeller = label_both),
             title = "Bootstrap Multiplier — Scenario 2, Normal errors")
```

:::

## Parametric Bootstrap Tests are Conservative

::: {.panel-tabset}

## Scenario 1

```{r bs-psd-conservative-s1}
results_t1 |>
  filter(
    method   == "bs.parametric.psd",
    n        %in% c(1000, 2000, 3000, 5000),
    p        %in% c(15, 20, 25),
    scenario == 1,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(
    n = factor(n, levels = c(1000, 2000, 3000, 5000)),
    p = factor(p, levels = c(15, 20, 25))
  ) |>
  plot_pvals(facet = facet_grid(n ~ p, labeller = label_both),
             title = "Parametric Bootstrap (PSD) — Scenario 1, Normal errors")
```

## Scenario 2

```{r bs-psd-conservative-s2}
results_t1 |>
  filter(
    method   == "bs.parametric.psd",
    n        %in% c(1000, 2000, 3000, 5000),
    p        %in% c(15, 20, 25),
    scenario == 2,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(
    n = factor(n, levels = c(1000, 2000, 3000, 5000)),
    p = factor(p, levels = c(15, 20, 25))
  ) |>
  plot_pvals(facet = facet_grid(n ~ p, labeller = label_both),
             title = "Parametric Bootstrap (PSD) — Scenario 2, Normal errors")
```

:::
