---
title: "Type I Error"
execute:
  echo: false
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
devtools::load_all()
results_t1 <- read_csv("../data/results-t1.csv")

n_sim <- 10000

method_colors <- c(
  "z.test.gaussian"      = "#E41A1C",
  "z.test.oracle.kappa"  = "#377EB8",
  "bs.multiplier"        = "#4DAF4A",
  "bs.parametric"        = "#FF7F00",
  "bs.parametric.psd"    = "#0072B2",
  "z.test.null"          = "#984EA3",
  "z.test.null.gaussian" = "#A65628"
)
```

```{R} 
sim_dependent_quadratic <- function(params, method, n_sims=10000){
  set.seed(1)
  map_dbl(1:n_sims, ~{
    hx <- params$h[,1]
    hy <- params$h[,2]
    ex <- rnorm(params$n)
    ey <- (ex^2 - 1) / sqrt(2)
    X <- cbind(hx + ex, hy + ey)

    test.diag(X, method=method)$p.value
  })
}

sim_dependent_circular <- function(params, method, n_sims=10000){
  set.seed(1)
  map_dbl(1:n_sims, ~{
    hx <- params$h[,1]
    hy <- params$h[,2]
    U <- runif(params$n, 0, 2 * pi)
    ex <- cos(U) * sqrt(2)
    ey <- sin(U) * sqrt(2)
    X <- cbind(hx + ex, hy + ey)

    test.diag(X, method=method)$p.value
  })
}
```

```{R}
#| echo: false

get_oracle_cov <- function(params, n_sims=1000, ...){
  sim_W <- map(1:n_sims, ~{
    X <- generate_data(params)
    S <- ece_terms(X)
    (1 / params$n^2) * (
      t(S) %*% S +
        t(rotate(S)) %*% S + t(S) %*% rotate(S) +
        t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
    )
  }, ...)

  # Average Results
  reduce(sim_W, `+`) / n_sims
}

bs_parametric_oracle_test <- function(X, W, B = 1000) {
  # Calculate Variance of ECE Terms
  n <- nrow(X)
  S <- ece_terms(X)

  # Boostrap Test
  S_bs <- MASS::mvrnorm(n = B, mu = rep(0, ncol(W)), Sigma = W)
  Q_bs <- abs(S_bs) |> apply(1, max)
  Q_obs <- max(abs(colMeans(S)))

  # Return Object
  list(
    p.value = mean(Q_bs > Q_obs)
  )
}
```

## $p = 2$: Z-Test Methods

```{r z-test-pcurve}
#| cache: true
params_p2_s1 <- scenario(1, n = 1000, err_type = "normal") 
params_p2    <- scenario(2, L = 4, n = 1000, err_type = "exponential")

pvals_z_s1 <- list(
  "Z-Test | Null, Oracle"   = pcurve(params_p2_s1, "z.test.null.oracle",   n_sim = n_sim),
  "Z-Test | Null, Gaussian" = pcurve(params_p2_s1, "z.test.null.gaussian", n_sim = n_sim),
  "Z-Test | Null"           = pcurve(params_p2_s1, "z.test.null",          n_sim = n_sim),
  "Z-Test | Oracle"         = pcurve(params_p2_s1, "z.test.oracle",        n_sim = n_sim),
  "Z-Test | Gaussian"       = pcurve(params_p2_s1, "z.test.gaussian",      n_sim = n_sim),
  "Z-Test"                  = pcurve(params_p2_s1, "z.test",               n_sim = n_sim)
)

pvals_z_s2 <- list(
  "Z-Test | Null, Oracle"   = pcurve(params_p2, "z.test.null.oracle",   n_sim = n_sim),
  "Z-Test | Null, Gaussian" = pcurve(params_p2, "z.test.null.gaussian", n_sim = n_sim),
  "Z-Test | Null"           = pcurve(params_p2, "z.test.null",          n_sim = n_sim),
  "Z-Test | Oracle"         = pcurve(params_p2, "z.test.oracle",        n_sim = n_sim),
  "Z-Test | Gaussian"       = pcurve(params_p2, "z.test.gaussian",      n_sim = n_sim),
  "Z-Test"                  = pcurve(params_p2, "z.test",               n_sim = n_sim)
)
```

::: {.panel-tabset}

## Scenario 1 (Gaussian)

```{r}
#| echo: false
plot_pcurve(pvals_z_s1, title = "p = 2 — Scenario 1, Gaussian errors, σ = (1, 1)", dir = "v", ncol = 2)
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_z_s2, title = "p = 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)", dir = "v", ncol = 2)
```

:::

## $p = 2$: Null Z-Test $\kappa_{22}$ Estimators

```{r pvals-null}
#| cache: true
#| warning: false
n_vals        <- c(200, 500, 1000)
methods_null  <- c("z.test.null", "z.test.null.regress", "z.test.null.matrix")
method_labels <- c("Z-Test | Null (MoM)", "Z-Test | Null (Regress)", "Z-Test | Null (Matrix)")

pvals_null <- map_dfr(n_vals, \(n) {
  s1 <- scenario(1, n = n, err_type = "normal")
  s2 <- scenario(2, L = 4, n = n, err_type = "exponential")
  map_dfr(seq_along(methods_null), \(j) {
    bind_rows(
      tibble(p_value = pcurve(s1, methods_null[j], n_sim = n_sim),
             method = method_labels[j], n = n, scenario = "Scenario 1"),
      tibble(p_value = pcurve(s2, methods_null[j], n_sim = n_sim),
             method = method_labels[j], n = n, scenario = "Scenario 2")
    )
  })
}) |>
  mutate(
    n      = factor(n, levels = n_vals),
    method = factor(method, levels = method_labels)
  )
```

```{r}
na_tbl <- pvals_null |>
  summarise(
    n_na = sum(is.na(p_value)),
    pct  = mean(is.na(p_value)),
    .by  = c(scenario, method, n)
  ) |>
  filter(n_na > 0) |>
  arrange(scenario, n, method)

if (nrow(na_tbl) > 0) {
  knitr::kable(na_tbl, digits = 3,
    col.names = c("Scenario", "Method", "n", "# NA", "Frac. NA"))
} else {
  cat("No NA p-values across all methods and sample sizes.")
}
```

::: {.panel-tabset}

## Scenario 1 (Gaussian)

```{r}
pvals_null |>
  filter(scenario == "Scenario 1") |>
  plot_pcurve(facet = facet_grid(n ~ method),
              title = "p = 2 — Scenario 1, Gaussian errors")
```

## Scenario 2, L=4 (Exponential)

```{r}
pvals_null |>
  filter(scenario == "Scenario 2") |>
  plot_pcurve(facet = facet_grid(n ~ method),
              title = "p = 2 — Scenario 2 (L = 4), exponential errors")
```

:::

## $p = 2$: Dependence Testing (Quadratic)

Correlation dependence testing are closely related but subtly different. In the classical Pearson test, interpretation is dependent on on the assumptions made. 

Consider this canonical example:
$$
\varepsilon_y = \frac{\varepsilon_x^2 - 1}{\sqrt{2}}, \quad 
\varepsilon_x \sim N(0, 1).
$$
Clearly, $\varepsilon_x$ and $\varepsilon_y$ are dependent, but by construction $\operatorname{Cor}(\varepsilon_x, \varepsilon_y) = 0$. Specifically, $\kappa_{22} = 5$. 

Therefore, a dependence test will identify dependence between $\varepsilon_x$ and $\varepsilon_y$, but a correlation test will not. 

::: {.panel-tabset}

## Scenario 1 (Dependent)

```{R dep-test-quadratic-s1}
#| cache: true
#| echo: false
params <- scenario(1) 
p_cor <- sim_dependent_quadratic(params, method="z.test.null")
p_dep <- sim_dependent_quadratic(params, method="z.test.null.gaussian")
```
```{r bs-dep-quadratic-s1}
#| cache: true
#| echo: false
params <- scenario(1)
p_bs_psd_s1  <- sim_dependent_quadratic(params, method = "bs.parametric.psd", n_sims = 10000)
p_bs_mult_s1 <- sim_dependent_quadratic(params, method = "bs.multiplier",      n_sims = 10000)
```

```{R}
#| echo: false
plot_pcurve(list(
  "Correlation test (κ₂₂ estimated)\nv² = (ωx² + ωy² + κ₂₂ + 5/2)/n" = p_cor,
  "Dependence test (κ₂₂ = 1 assumed)\nv² = (ωx² + ωy² + 7/2)/n"              = p_dep
))
```

```{r}
#| echo: false
plot_pcurve(list(
  "Parametric Bootstrap (PSD)" = p_bs_psd_s1,
  "Multiplier Bootstrap"       = p_bs_mult_s1
), ncol = 2)
```

## Scenario 2 (Dependent)

```{R dep-test-quadratic-s2}
#| cache: true
#| echo: false
params <- scenario(2, L=4)
p_cor <- sim_dependent_quadratic(params, method="z.test.null")
p_dep <- sim_dependent_quadratic(params, method="z.test.null.gaussian")
```

```{r bs-dep-quadratic-s2}
#| cache: true
#| echo: false
params <- scenario(2, L = 4)
p_bs_psd_s2  <- sim_dependent_quadratic(params, method = "bs.parametric.psd", n_sims = 10000)
p_bs_mult_s2 <- sim_dependent_quadratic(params, method = "bs.multiplier",      n_sims = 10000)
```

```{R}
#| echo: false
plot_pcurve(list(
  "Correlation test (κ₂₂ estimated)\nv² = (ωx² + ωy² + κ₂₂ + 5/2)/n" = p_cor,
  "Dependence test (κ₂₂ = 1 assumed)\nv² = (ωx² + ωy² + 7/2)/n"              = p_dep
))
```

```{r}
#| echo: false
plot_pcurve(list(
  "Parametric Bootstrap (PSD)" = p_bs_psd_s2,
  "Multiplier Bootstrap"       = p_bs_mult_s2
), ncol = 2)
```

:::


## $p = 2:$ Dependence Testing ($\kappa_{22} = 0.5$)

The case $\kappa_{22} < 1$ is a bit unnatural. For bivariate Gaussian errors, 
$$\kappa_{22} = 1 + 2\rho^2 \geq 1 \text{ for all } \rho \in [-1, 1],$$ 
so the Gaussian family can never produce it.

Since 
$$\kappa_{22} = 1 + \frac{\operatorname{Cov}(\varepsilon_x^2, \varepsilon_y^2)}{\sigma_x^2 \sigma_y^2},$$ 
having $\kappa_{22} < 1$ requires $\operatorname{Cov}(\varepsilon_x^2, \varepsilon_y^2) < 0$: large squared values of one component must tend to coincide with small squared values of the other.

The circular distribution provides a concrete construction:
$$
U \sim \mathrm{Uniform}(0, 2\pi), \quad
\varepsilon_x = \sqrt{2}\cos(U), \quad
\varepsilon_y = \sqrt{2}\sin(U).
$$
Because $\varepsilon_x^2 + \varepsilon_y^2 = 2$ almost surely, whenever $\varepsilon_x^2$ is large $\varepsilon_y^2$ must be small, giving exactly the required negative covariance: $\operatorname{Cov}(\varepsilon_x^2, \varepsilon_y^2) = \mathbb{E}[\sin^2(2U)] - 1 = \tfrac{1}{2} - 1 = -\tfrac{1}{2}$. With $\sigma_x^2 = \sigma_y^2 = 1$, this yields $\kappa_{22} = 0.5$.

At the same time, $\operatorname{Cov}(\varepsilon_x, \varepsilon_y) = \mathbb{E}[\sin(2U)] = 0$, so the errors are uncorrelated and a Pearson correlation test has no power — yet the pair is strongly dependent.

::: {.panel-tabset}

## Scenario 1 

```{R sim-dep-circular-s1}
#| echo: false
#| cache: true
params <- scenario(1) 
p_cor <- sim_dependent_circular(params, method="z.test.null")
p_dep <- sim_dependent_circular(params, method="z.test.null.gaussian")
```

```{R}
#| echo: false
plot_pcurve(list(
  "Correlation test (κ₂₂ estimated)\nv² = (ωx² + ωy² + κ₂₂ + 5/2)/n" = p_cor,
  "Dependence test (κ₂₂ = 1 assumed)\nv² = (ωx² + ωy² + 7/2)/n"              = p_dep
))
```

```{r bs-dep-circular-s1}
#| cache: true
#| echo: false
params <- scenario(1)
p_bs_psd_s1  <- sim_dependent_circular(params, method = "bs.parametric.psd", n_sims = 10000)
p_bs_mult_s1 <- sim_dependent_circular(params, method = "bs.multiplier",      n_sims = 10000)
```

```{r}
#| echo: false
plot_pcurve(list(
  "Parametric Bootstrap (PSD)" = p_bs_psd_s1,
  "Multiplier Bootstrap"       = p_bs_mult_s1
), ncol = 2)
```

## Scenario 2

```{R sim-dep-circular-s2}
#| echo: false
#| cache: true
params <- scenario(2, L=4) 
p_cor <- sim_dependent_circular(params, method="z.test.null")
p_dep <- sim_dependent_circular(params, method="z.test.null.gaussian")
```

```{R}
#| echo: false
plot_pcurve(list(
  "Correlation test (κ₂₂ estimated)\nv² = (ωx² + ωy² + κ₂₂ + 5/2)/n" = p_cor,
  "Dependence test (κ₂₂ = 1 assumed)\nv² = (ωx² + ωy² + 7/2)/n"              = p_dep
))
```

```{r bs-dep-circular-s2}
#| cache: true
#| echo: false
params <- scenario(2, L = 4)
p_bs_psd_s2  <- sim_dependent_circular(params, method = "bs.parametric.psd", n_sims = 10000)
p_bs_mult_s2 <- sim_dependent_circular(params, method = "bs.multiplier",      n_sims = 10000)
```

```{r}
#| echo: false
plot_pcurve(list(
  "Parametric Bootstrap (PSD)" = p_bs_psd_s2,
  "Multiplier Bootstrap"       = p_bs_mult_s2
), ncol = 2)
```

:::


## $p = 2$: Bootstrap Methods

::: {.panel-tabset}

## Scenario 1 (Normal)

```{r}
#| echo: false
results_t1 |>
  filter(
    method   == "bs.multiplier",
    n        %in% c(200, 500, 1000),
    p        == 2,
    scenario == 1,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(n = factor(n, levels = c(200, 500, 1000))) |>
  plot_pvals(facet = ~n, ncol = 3,
             title = "p = 2 — Scenario 1, normal errors, σ = (1, 1)")
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
results_t1 |>
  filter(
    method   == "bs.multiplier",
    n        %in% c(200, 500, 1000),
    p        == 2,
    scenario == 2,
    cov_type == "compound",
    err_type == "exponential"
  ) |>
  mutate(n = factor(n, levels = c(200, 500, 1000))) |>
  plot_pvals(facet = ~n, ncol = 3,
             title = "p = 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)")
```

:::

## $p = 2$: Parametric Bootstrap Methods

```{r bs-parametric-s1-p2}
#| cache: true
#| echo: false
pvals_bsp_s1 <- list(
  "Bootstrap | Parametric"      = pcurve(params_p2_s1, "bs.parametric",     n_sim = n_sim),
  "Bootstrap | Parametric, PSD" = pcurve(params_p2_s1, "bs.parametric.psd", n_sim = n_sim)
)
```

```{r bs-parametric-s2-p2}
#| cache: true
#| echo: false
pvals_bsp_s2 <- list(
  "Bootstrap | Parametric"      = pcurve(params_p2, "bs.parametric",     n_sim = n_sim),
  "Bootstrap | Parametric, PSD" = pcurve(params_p2, "bs.parametric.psd", n_sim = n_sim)
)
```

::: {.panel-tabset}

## Scenario 1 (Normal)

```{r}
#| echo: false
plot_pcurve(pvals_bsp_s1, title = "p = 2 — Scenario 1, normal errors, σ = (1, 1)", ncol = 2)
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_bsp_s2, title = "p = 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)", ncol = 2)
```

:::

## $p > 2$: Multiplier Bootstrap

```{r}
params_p3_s1  <- scenario(1, n = 1000, err_type = "normal",            p = 3)
params_p5_s1  <- scenario(1, n = 1000, err_type = "normal",            p = 5)
params_p10_s1 <- scenario(1, n = 1000, err_type = "normal",            p = 10)
params_p3_s2  <- scenario(2, L = 4, n = 1000, err_type = "exponential", p = 3)
params_p5_s2  <- scenario(2, L = 4, n = 1000, err_type = "exponential", p = 5)
params_p10_s2 <- scenario(2, L = 4, n = 1000, err_type = "exponential", p = 10)
```

```{r mult-p3}
#| cache: true
pvals_mult_p3_s1 <- pcurve(params_p3_s1, "bs.multiplier", n_sim = n_sim)
pvals_mult_p3_s2 <- pcurve(params_p3_s2, "bs.multiplier", n_sim = n_sim)
```

```{r mult-p5}
#| cache: true
pvals_mult_p5_s1 <- pcurve(params_p5_s1, "bs.multiplier", n_sim = n_sim)
pvals_mult_p5_s2 <- pcurve(params_p5_s2, "bs.multiplier", n_sim = n_sim)
```

```{r mult-p10}
#| cache: true
pvals_mult_p10_s1 <- pcurve(params_p10_s1, "bs.multiplier", n_sim = n_sim)
pvals_mult_p10_s2 <- pcurve(params_p10_s2, "bs.multiplier", n_sim = n_sim)
```

```{r}
pvals_mult_pg2_s1 <- bind_rows(
  tibble(p_value = pvals_mult_p3_s1,  p = factor(3,  levels = c(3, 5, 10))),
  tibble(p_value = pvals_mult_p5_s1,  p = factor(5,  levels = c(3, 5, 10))),
  tibble(p_value = pvals_mult_p10_s1, p = factor(10, levels = c(3, 5, 10)))
)

pvals_mult_pg2_s2 <- bind_rows(
  tibble(p_value = pvals_mult_p3_s2,  p = factor(3,  levels = c(3, 5, 10))),
  tibble(p_value = pvals_mult_p5_s2,  p = factor(5,  levels = c(3, 5, 10))),
  tibble(p_value = pvals_mult_p10_s2, p = factor(10, levels = c(3, 5, 10)))
)
```

::: {.panel-tabset}

## Scenario 1 (Normal)

```{r}
#| echo: false
plot_pcurve(pvals_mult_pg2_s1, facet = ~p, ncol = 3,
  title = "p > 2 — Scenario 1, normal errors, σ = (1, 1)")
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_mult_pg2_s2, facet = ~p, ncol = 3,
  title = "p > 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)")
```

:::

## $p > 2$: Parametric Bootstrap

```{r param-p3}
#| cache: true
pvals_param_p3_s1 <- bind_rows(
  tibble(p_value = pcurve(params_p3_s1, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p3_s1, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
pvals_param_p3_s2 <- bind_rows(
  tibble(p_value = pcurve(params_p3_s2, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p3_s2, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
```

```{r param-p5}
#| cache: true
pvals_param_p5_s1 <- bind_rows(
  tibble(p_value = pcurve(params_p5_s1, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p5_s1, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
pvals_param_p5_s2 <- bind_rows(
  tibble(p_value = pcurve(params_p5_s2, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p5_s2, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
```

```{r param-p10}
#| cache: true
pvals_param_p10_s1 <- bind_rows(
  tibble(p_value = pcurve(params_p10_s1, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p10_s1, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
pvals_param_p10_s2 <- bind_rows(
  tibble(p_value = pcurve(params_p10_s2, "bs.parametric",     n_sim = n_sim), method = "Bootstrap | Parametric"),
  tibble(p_value = pcurve(params_p10_s2, "bs.parametric.psd", n_sim = n_sim), method = "Bootstrap | Parametric, PSD")
)
```

```{r}
p_levels <- c(3, 5, 10)
m_levels <- c("Bootstrap | Parametric", "Bootstrap | Parametric, PSD")

pvals_param_pg2_s1 <- bind_rows(
  mutate(pvals_param_p3_s1,  p = 3),
  mutate(pvals_param_p5_s1,  p = 5),
  mutate(pvals_param_p10_s1, p = 10)
) |> mutate(p = factor(p, levels = p_levels), method = factor(method, levels = m_levels))

pvals_param_pg2_s2 <- bind_rows(
  mutate(pvals_param_p3_s2,  p = 3),
  mutate(pvals_param_p5_s2,  p = 5),
  mutate(pvals_param_p10_s2, p = 10)
) |> mutate(p = factor(p, levels = p_levels), method = factor(method, levels = m_levels))
```

::: {.panel-tabset}

## Scenario 1 (Normal)

```{r}
#| echo: false
plot_pcurve(pvals_param_pg2_s1, facet = facet_grid(p ~ method),
  title = "p > 2 — Scenario 1, normal errors, σ = (1, 1)")
```

## Scenario 2, L=4 (Exponential)

```{r}
#| echo: false
plot_pcurve(pvals_param_pg2_s2, facet = facet_grid(p ~ method),
  title = "p > 2 — Scenario 2 (L = 4), exponential errors, σ = (1, 1)")
```

:::

## Bootstrap Tests are Conservative

### Bootstrap Multiplier is Conservative

::: {.panel-tabset}

## Scenario 1

```{r bs-conservative-s1}
results_t1 |>
  filter(
    method   == "bs.multiplier",
    n        %in% c(1000, 2000, 3000, 5000),
    p        %in% c(15, 20, 25),
    scenario == 1,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(
    n = factor(n, levels = c(1000, 2000, 3000, 5000)),
    p = factor(p, levels = c(15, 20, 25))
  ) |>
  plot_pvals(facet = facet_grid(n ~ p, labeller = label_both),
             title = "Bootstrap Multiplier — Scenario 1, Normal errors")
```

## Scenario 2

```{r bs-conservative-s2}
results_t1 |>
  filter(
    method   == "bs.multiplier",
    n        %in% c(1000, 2000, 3000, 5000),
    p        %in% c(15, 20, 25),
    scenario == 2,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(
    n = factor(n, levels = c(1000, 2000, 3000, 5000)),
    p = factor(p, levels = c(15, 20, 25))
  ) |>
  plot_pvals(facet = facet_grid(n ~ p, labeller = label_both),
             title = "Bootstrap Multiplier — Scenario 2, Normal errors")
```

:::

```{R bs-mult-split-pvals-n200-p15}
#| echo: false
#| cache: true
params <- scenario(1, n=200, p=15)
sims_pvals <- map(1:10000, ~{
  X <- generate_data(params)
  pvals <- bs_multiplier_test(X)
  list(
    p  = pvals$p.value,
    p1 = pvals$p.split[1],
    p2 = pvals$p.split[2],
    p3 = pvals$p.split[3]
  )
})
p  <- map_dbl(sims_pvals, "p")
p1 <- map_dbl(sims_pvals, "p1")
p2 <- map_dbl(sims_pvals, "p2")
p3 <- map_dbl(sims_pvals, "p3")
```

```{r}
#| echo: false
p_n200 <- p; p1_n200 <- p1; p2_n200 <- p2; p3_n200 <- p3
```

```{R bs-mult-split-pvals-n2000-p15}
#| echo: false
#| cache: true
params <- scenario(1, n=2000, p=15)
sims_pvals <- map(1:10000, ~{
  X <- generate_data(params)
  pvals <- bs_multiplier_test(X)
  list(
    p  = pvals$p.value,
    p1 = pvals$p.split[1],
    p2 = pvals$p.split[2],
    p3 = pvals$p.split[3]
  )
})
p  <- map_dbl(sims_pvals, "p")
p1 <- map_dbl(sims_pvals, "p1")
p2 <- map_dbl(sims_pvals, "p2")
p3 <- map_dbl(sims_pvals, "p3")
```

::: {.panel-tabset}

## n = 200

```{r}
#| echo: false
tibble(p_value = p_n200) |>
  ggplot(aes(x = p_value)) +
  geom_histogram(aes(y = after_stat(density)), breaks = seq(0, 1, by = 0.05), fill = "steelblue", color = "white") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(x = "p-value", y = "Density", title = "Combined p-value (n = 200, p = 15)") +
  theme_minimal()
```

```{r}
#| echo: false
bind_rows(
  tibble(p_value = p1_n200, split = "Split 1"),
  tibble(p_value = p2_n200, split = "Split 2"),
  tibble(p_value = p3_n200, split = "Split 3")
) |>
  mutate(split = factor(split, levels = c("Split 1", "Split 2", "Split 3"))) |>
  ggplot(aes(x = p_value)) +
  geom_histogram(aes(y = after_stat(density)), breaks = seq(0, 1, by = 0.05), fill = "steelblue", color = "white") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  facet_wrap(~ split, ncol = 3) +
  labs(x = "p-value", y = "Density", title = "Split p-values (n = 200, p = 15)") +
  theme_minimal()
```

## n = 2000

```{r}
#| echo: false
tibble(p_value = p) |>
  ggplot(aes(x = p_value)) +
  geom_histogram(aes(y = after_stat(density)), breaks = seq(0, 1, by = 0.05), fill = "steelblue", color = "white") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(x = "p-value", y = "Density", title = "Combined p-value (n = 2000, p = 15)") +
  theme_minimal()
```

```{r}
#| echo: false
bind_rows(
  tibble(p_value = p1, split = "Split 1"),
  tibble(p_value = p2, split = "Split 2"),
  tibble(p_value = p3, split = "Split 3")
) |>
  mutate(split = factor(split, levels = c("Split 1", "Split 2", "Split 3"))) |>
  ggplot(aes(x = p_value)) +
  geom_histogram(aes(y = after_stat(density)), breaks = seq(0, 1, by = 0.05), fill = "steelblue", color = "white") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  facet_wrap(~ split, ncol = 3) +
  labs(x = "p-value", y = "Density", title = "Split p-values (n = 2000, p = 15)") +
  theme_minimal()
```

:::


### Parametric Bootstrap is Conservative

::: {.panel-tabset}

## Scenario 1

```{r bs-psd-conservative-s1}
results_t1 |>
  filter(
    method   == "bs.parametric.psd",
    n        %in% c(1000, 2000, 3000, 5000),
    p        %in% c(15, 20, 25),
    scenario == 1,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(
    n = factor(n, levels = c(1000, 2000, 3000, 5000)),
    p = factor(p, levels = c(15, 20, 25))
  ) |>
  plot_pvals(facet = facet_grid(n ~ p, labeller = label_both),
             title = "Parametric Bootstrap (PSD) — Scenario 1, Normal errors")
```

## Scenario 2

```{r bs-psd-conservative-s2}
results_t1 |>
  filter(
    method   == "bs.parametric.psd",
    n        %in% c(1000, 2000, 3000, 5000),
    p        %in% c(15, 20, 25),
    scenario == 2,
    cov_type == "compound",
    err_type == "normal"
  ) |>
  mutate(
    n = factor(n, levels = c(1000, 2000, 3000, 5000)),
    p = factor(p, levels = c(15, 20, 25))
  ) |>
  plot_pvals(facet = facet_grid(n ~ p, labeller = label_both),
             title = "Parametric Bootstrap (PSD) — Scenario 2, Normal errors")
```

:::

### Mechanism: Not in Asymptotic CLT Regime
The u-shaped p-curves indicate that there are insufficient small and large p-values. This means the distribution of bootstrap statistics has less variance than the true distribution. 

To see why, we compare the distribution of the test statistic $Q = \max_j |\bar S_j|$ under two distributions. The **true distribution** is obtained by repeatedly drawing a fresh dataset and computing $Q$ from the observed $\bar S$. The **bootstrap distribution** is obtained by drawing a single dataset, estimating the covariance $W$ of $\bar S$ from that dataset, and then drawing one bootstrap replicate $Q^* = \max_j |Z_j|$ with $Z \sim N(0, W)$. When $n$ is small relative to $p$, $W$ underestimates the true variance of $\bar S$, so the bootstrap distribution is too narrow and the test is conservative.

::: {.panel-tabset}

## n = 200, p = 15

```{R simulate-test-statistic-distribution-small-n-large-p}
#| echo: false
#| cache: true

# Show Difference in the Distributions of the Test Statistic
params <- scenario(1, n=200, p=15)
n <- params$n

# True Distribution
sims_stat_true <- map_dbl(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  Q_obs <- max(abs(colMeans(S)))
  Q_obs
})

sims_stat_bs <- map(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  W <- (1 / n^2) * (
    t(S) %*% S +
    t(rotate(S)) %*% S + t(S) %*% rotate(S) +
    t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
  ) |> make_psd()
  MASS::mvrnorm(n = 1, mu = rep(0, ncol(W)), Sigma = W) |>
    abs() |> max() # apply(1, max)
}) |> unlist()
```

```{R simulate-test-statistic-distribution-small-n-large-p-oracle}
#| echo: false
#| cache: true

W_oracle <- get_oracle_cov(params, n_sims=10000)
sims_stat_bs_oracle <- map(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  MASS::mvrnorm(n = 1, mu = rep(0, ncol(W_oracle)), Sigma = W_oracle) |>
    abs() |> max() # apply(1, max)
}) |> unlist()
```

```{R}
#| echo: false
bind_rows(
  tibble(value = sims_stat_true, source = "True"),
  tibble(value = sims_stat_bs,   source = "Bootstrap")
) |>
  mutate(source = factor(source, levels = c("True", "Bootstrap"))) |>
  ggplot(aes(x = value, fill = source)) +
  geom_histogram(aes(y = after_stat(density)),
                 position = "identity", alpha = 0.5, bins = 40, color = "white") +
  scale_fill_manual(values = c(True = "steelblue", Bootstrap = "red")) +
  labs(x = "Q = max|S̅|", y = "Density",
       title = "Test statistic distribution: true vs. bootstrap (n = 200, p = 15)",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "top")
```

```{R}
#| echo: false
bind_rows(
  tibble(value = sims_stat_true, source = "True"),
  tibble(value = sims_stat_bs_oracle,   source = "Bootstrap (Oracle)")
) |>
  mutate(source = factor(source, levels = c("True", "Bootstrap (Oracle)"))) |>
  ggplot(aes(x = value, fill = source)) +
  geom_histogram(aes(y = after_stat(density)),
                 position = "identity", alpha = 0.5, bins = 40, color = "white") +
  scale_fill_manual(values = c(True = "steelblue", "Bootstrap (Oracle)" = "forestgreen")) +
  labs(x = "Q = max|S̅|", y = "Density",
       title = "Test statistic distribution: true vs. bootstrap (n = 200, p = 15)",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "top")
```


## n = 2000, p = 15

```{R simulate-test-statistic-distribution-large-n-large-p}
#| echo: false
#| cache: true

# Show Difference in the Distributions of the Test Statistic
params <- scenario(1, n=2000, p=15)
n <- params$n

# True Distribution
sims_stat_true <- map_dbl(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  Q_obs <- max(abs(colMeans(S)))
  Q_obs
})

sims_stat_bs <- map(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  W <- (1 / n^2) * (
    t(S) %*% S +
    t(rotate(S)) %*% S + t(S) %*% rotate(S) +
    t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
  ) |> make_psd()
  MASS::mvrnorm(n = 1, mu = rep(0, ncol(W)), Sigma = W) |>
    abs() |> max() # apply(1, max)
}) |> unlist()
```

```{R simulate-test-statistic-distribution-large-n-large-p-oracle}
#| echo: false
#| cache: true

W_oracle <- get_oracle_cov(params, n_sims=10000)
sims_stat_bs_oracle <- map(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  MASS::mvrnorm(n = 1, mu = rep(0, ncol(W_oracle)), Sigma = W_oracle) |>
    abs() |> max() # apply(1, max)
}) |> unlist()
```

```{R}
#| echo: false
bind_rows(
  tibble(value = sims_stat_true, source = "True"),
  tibble(value = sims_stat_bs,   source = "Bootstrap")
) |>
  mutate(source = factor(source, levels = c("True", "Bootstrap"))) |>
  ggplot(aes(x = value, fill = source)) +
  geom_histogram(aes(y = after_stat(density)),
                 position = "identity", alpha = 0.5, bins = 40, color = "white") +
  scale_fill_manual(values = c(True = "steelblue", Bootstrap = "red")) +
  labs(x = "Q = max|S̅|", y = "Density",
       title = "Test statistic distribution: true vs. bootstrap (n = 2000, p = 15)",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "top")
```

```{R}
#| echo: false
bind_rows(
  tibble(value = sims_stat_true, source = "True"),
  tibble(value = sims_stat_bs_oracle,   source = "Bootstrap (Oracle)")
) |>
  mutate(source = factor(source, levels = c("True", "Bootstrap (Oracle)"))) |>
  ggplot(aes(x = value, fill = source)) +
  geom_histogram(aes(y = after_stat(density)),
                 position = "identity", alpha = 0.5, bins = 40, color = "white") +
  scale_fill_manual(values = c(True = "steelblue", "Bootstrap (Oracle)" = "forestgreen")) +
  labs(x = "Q = max|S̅|", y = "Density",
       title = "Test statistic distribution: true vs. bootstrap (n = 2000, p = 15)",
       fill = NULL) +
  theme_minimal() +
  theme(legend.position = "top")
```

:::



```{R bs-param-benchmark s1-n200-p15}
#| cache: true
#| echo: false

# Set Parameters
params <- scenario(1, n=200, p=15)
W <- get_oracle_cov(params, n_sims=10000)

# Run Simulations
pval_oracle <- map_dbl(1:10000, ~{
  X <- generate_data(params)
  bs_parametric_oracle_test(X, W)$p.value
})
pval_param <- map_dbl(1:10000, ~{
  X <- generate_data(params)
  bs_parametric_test(X, force_psd=TRUE)$p.value
})

```

```{R bs-param-benchmark s1-n2000-p15}
#| cache: true
#| echo: false

# Set Parameters
params <- scenario(1, n=2000, p=15)
W_oracle <- get_oracle_cov(params, n_sims=10000)

# Run Simulations
pval_oracle_n2000 <- map_dbl(1:10000, ~{
  X <- generate_data(params)
  bs_parametric_oracle_test(X, W_oracle)$p.value
})
pval_param_n2000 <- map_dbl(1:10000, ~{
  X <- generate_data(params)
  bs_parametric_test(X, force_psd=TRUE)$p.value
})

```

::: {.panel-tabset}

## n = 200, p = 15
```{R}
#| echo: false

# Show P-Curve
plot_pcurve(list(
  "Parametric Bootstrap (PSD)" = pval_param,
  "Parametric Bootstrap (Oracle)" = pval_oracle
))
```

## n = 2000, p = 15

```{R}
#| echo: false

# Show P-Curve
plot_pcurve(list(
  "Parametric Bootstrap (PSD)" = pval_param_n2000,
  "Parametric Bootstrap (Oracle)" = pval_oracle_n2000
))
```
:::


::: {.panel-tabset}

## n = 200, p = 15

```{R params-n200-p15}
#| echo: false
#| cache: true
params_n200 <- scenario(1, n=200, p=15)
W_n200 <- get_oracle_cov(params_n200, n_sims=10000)
```

```{R simulate-sample-distribution-small-n-large-p-true}
#| echo: false
#| cache: true
#| warning: false
sims_sample_true_n200 <- map(1:10000, ~{
  X <- generate_data(params_n200)
  S <- ece_terms(X)
  colMeans(S) |> t() |> as_tibble()
}) |> list_rbind()
```

```{R simulate-sample-distribution-small-n-large-p-bs}
#| echo: false
#| cache: true
#| warning: false
sims_sample_bs_n200 <- map(1:10000, ~{
  X <- generate_data(params_n200)
  S <- ece_terms(X)
  W <- (1 / params_n200$n^2) * (
    t(S) %*% S +
    t(rotate(S)) %*% S + t(S) %*% rotate(S) +
    t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
  ) |> make_psd()
  MASS::mvrnorm(n = 1, mu = rep(0, ncol(W)), Sigma = W) |> t() |> as_tibble()
}) |> list_rbind()
```

```{R simulate-sample-distribution-small-n-large-p-oracle}
#| echo: false
#| cache: true
#| warning: false
sims_sample_bs_oracle_n200 <- map(1:10000, ~{
  X <- generate_data(params_n200)
  S <- ece_terms(X)
  MASS::mvrnorm(n = 1, mu = rep(0, ncol(W_n200)), Sigma = W_n200) |>
    t() |> as_tibble()
}) |> list_rbind()
```

```{R sample-spaghetti-plot-n200}
#| echo: false
var_true      <- apply(sims_sample_true_n200, 2, var)
var_bs_oracle <- apply(sims_sample_bs_oracle_n200, 2, var)
var_bs        <- apply(sims_sample_bs_n200, 2, var)

q <- length(var_true)
data.frame(
  term   = factor(seq_len(q)),
  bs     = var_bs,
  true   = var_true,
  oracle = var_bs_oracle
) |>
  pivot_longer(-term, names_to = "method", values_to = "variance") |>
  mutate(method = factor(method,
                          levels = c("bs", "true", "oracle"),
                          labels = c("Bootstrap", "True", "Bootstrap (Oracle)"))) |>
  ggplot(aes(x = method, y = variance, group = term)) +
  geom_line(alpha = 0.25, color = "gray40") +
  geom_point(alpha = 0.5, size = 1) +
  labs(x = NULL, y = "Variance") +
  theme_minimal()
```

```{R sample-histogram-n200}
#| echo: false
data.frame(
  variance = c(var_true, var_bs_oracle, var_bs),
  source   = rep(c("True", "Bootstrap (Oracle)", "Bootstrap"),
                  each = length(var_true))
) |>
  mutate(source = factor(source, levels = c("True", "Bootstrap (Oracle)", "Bootstrap"))) |>
  ggplot(aes(x = variance, fill = source)) +
  geom_histogram(aes(y = after_stat(density)),
                  position = "identity", alpha = 0.5,
                  bins = 15, color = "white") +
  scale_fill_manual(values = c(
    "True"               = "steelblue",
    "Bootstrap (Oracle)" = "forestgreen",
    "Bootstrap"          = "tomato"
  )) +
  labs(x = "Variance", y = "Density", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "top")
```

## n = 2000, p = 15

```{R simulate-sample-distribution-large-n-large-p-true}
#| echo: false
#| cache: true
#| warning: false
sims_sample_true <- map(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  colMeans(S) |> t() |> as_tibble()
}) |> list_rbind()
```

```{R simulate-sample-distribution-large-n-large-p-bs}
#| echo: false
#| cache: true
#| warning: false
sims_sample_bs <- map(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  W <- (1 / params$n^2) * (
    t(S) %*% S +
    t(rotate(S)) %*% S + t(S) %*% rotate(S) +
    t(rotate(S, 2)) %*% S + t(S) %*% rotate(S, 2)
  ) |> make_psd()
  MASS::mvrnorm(n = 1, mu = rep(0, ncol(W)), Sigma = W) |> t() |> as_tibble()
}) |> list_rbind()
```

```{R simulate-sample-distribution-large-n-large-p-oracle}
#| echo: false
#| cache: true
#| warning: false
sims_sample_bs_oracle <- map(1:10000, ~{
  X <- generate_data(params)
  S <- ece_terms(X)
  MASS::mvrnorm(n = 1, mu = rep(0, ncol(W_oracle)), Sigma = W_oracle) |>
    t() |> as_tibble()
}) |> list_rbind()
```

```{R sample-spaghetti-plot-n2000}
#| echo: false
var_true      <- apply(sims_sample_true, 2, var)
var_bs_oracle <- apply(sims_sample_bs_oracle, 2, var)
var_bs        <- apply(sims_sample_bs, 2, var)

q <- length(var_true)
data.frame(
  term   = factor(seq_len(q)),
  bs     = var_bs,
  true   = var_true,
  oracle = var_bs_oracle
) |>
  pivot_longer(-term, names_to = "method", values_to = "variance") |>
  mutate(method = factor(method,
                          levels = c("bs", "true", "oracle"),
                          labels = c("Bootstrap", "True", "Bootstrap (Oracle)"))) |>
  ggplot(aes(x = method, y = variance, group = term)) +
  geom_line(alpha = 0.25, color = "gray40") +
  geom_point(alpha = 0.5, size = 1) +
  labs(x = NULL, y = "Variance") +
  theme_minimal()
```

```{R sample-histogram-n2000}
#| echo: false
data.frame(
  variance = c(var_true, var_bs_oracle, var_bs),
  source   = rep(c("True", "Bootstrap (Oracle)", "Bootstrap"),
                  each = length(var_true))
) |>
  mutate(source = factor(source, levels = c("True", "Bootstrap (Oracle)", "Bootstrap"))) |>
  ggplot(aes(x = variance, fill = source)) +
  geom_histogram(aes(y = after_stat(density)),
                  position = "identity", alpha = 0.5,
                  bins = 15, color = "white") +
  scale_fill_manual(values = c(
    "True"               = "steelblue",
    "Bootstrap (Oracle)" = "forestgreen",
    "Bootstrap"          = "tomato"
  )) +
  labs(x = "Variance", y = "Density", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "top")
```
:::


