---
title: "Z-Test"
---

Define $Z = (Z_1, ..., Z_6)^T = (T_1(X, X), T_2(X, X), T_1(Y, Y), T_2(Y, Y), T_1(X, Y), T_2(X, Y))^T$.

## Asymptotic Normality

**Proposition.** Under the model assumptions, if $\max_j\{(\mu_{x,j} - \mu_{x,j+1})^2\} = o(n)$ and $\max_j\{(\mu_{y,j} - \mu_{y,j+1})^2\} = o(n)$, and errors have finite 4th moments, then
$$
\frac{1}{\sqrt{n}}(Z - \mathbb{E}[Z]) \xrightarrow{d} N(\mathbf{0}, \Sigma_Z).
$$

**Strategy.** Each component of $Z$ is of the form $T_k(X,Y)$ for some lag $k$ and choice of series. It therefore suffices to show $T_k(X,Y)$ is asymptotically normal, since any linear combination $a^T Z$ inherits this structure. The Cramér-Wold device then promotes univariate normality to joint normality of the full vector.

**Proof.** By the Cramér-Wold device, it suffices to show every linear combination $a^T Z$ is asymptotically normal. We handle the general cross-product component $T_k(X,Y)$; the diagonal components $T_k(X,X)$ and $T_k(Y,Y)$ follow as special cases.

Substituting $X_i = \theta_{x,i} + \varepsilon_{x,i}$ and $Y_i = \theta_{y,i} + \varepsilon_{y,i}$ and expanding gives four terms:

\begin{align}
T_k(X,Y) &= \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\theta_{y,i} - \theta_{y,i+k}) \\
&\quad + \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k}) + \sum_{i=1}^n (\theta_{y,i} - \theta_{y,i+k})(\varepsilon_{x,i} - \varepsilon_{x,i+k}) \\
&\quad + \sum_{i=1}^n (\varepsilon_{x,i} - \varepsilon_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k}).
\end{align}

The first term equals $kW_{xy}$ and is deterministic. The last term has expectation $2n\sigma_{xy}$, so adding and subtracting this expectation gives

\begin{align}
T_k(X,Y) &= \underbrace{kW_{xy} + 2n\sigma_{xy}}_{\mathbb{E}[T_k(X,Y)]} \\
&\quad + \underbrace{\sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k}) + \sum_{i=1}^n (\theta_{y,i} - \theta_{y,i+k})(\varepsilon_{x,i} - \varepsilon_{x,i+k})}_{\text{(I) mixed}} \\
&\quad + \underbrace{\sum_{i=1}^n \bigl[(\varepsilon_{x,i} - \varepsilon_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k}) - 2\sigma_{xy}\bigr]}_{\text{(II) purely random}},
\end{align}

where $W_{xy} = T_1(\theta_x, \theta_y)$.

**Term (II)** is a sum of mean-zero, $k$-dependent, identically distributed random variables (since errors are i.i.d.), so the CLT for $m$-dependent sequences applies directly.

**Term (I).** Consider the first sum; the second is identical in structure with $x$ and $y$ exchanged. Splitting the error differences,
$$
\sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k})
= \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})\varepsilon_{y,i} - \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})\varepsilon_{y,i+k},
$$
each of which is a weighted sum of independent errors with deterministic coefficients $c_i = \theta_{x,i} - \theta_{x,i+k}$. Since $\theta_x$ is piecewise constant, the nonzero coefficients take values $\{\mu_{x,j} - \mu_{x,j+1}\}$, giving $\sum_i c_i^2 = kW_x(\theta)$ and $\max_i c_i^2 = \max_j (\mu_{x,j} - \mu_{x,j+1})^2$.

The proposition condition $\max_j\{(\mu_{x,j} - \mu_{x,j+1})^2\} = o(n)$ implies one of two cases:

**Case 1:** $W_x(\theta) = o(n)$. Then Term (I) has standard deviation $O(\sqrt{W_x(\theta)}) = o(\sqrt{n})$, so Term (I)$/\sqrt{n} \xrightarrow{p} 0$. By Slutsky's theorem, $[{\rm Term\;(I)} + {\rm Term\;(II)}]/\sqrt{n}$ has the same limiting distribution as Term (II)$/\sqrt{n}$, which is normal by the CLT for $m$-dependent sequences.

**Case 2:** $W_x(\theta) = \Omega(n)$. Since $\max_j(\mu_{x,j} - \mu_{x,j+1})^2 = o(n)$ and $W_x(\theta) = \Omega(n)$, we have
$$
\frac{\max_j (\mu_{x,j} - \mu_{x,j+1})^2}{W_x(\theta)} \to 0,
$$
which is equivalent to $\max_i c_i^2 / \sum_i c_i^2 \to 0$. This is the Lindeberg condition for the weighted sum $\sum_i c_i \varepsilon_{y,i}$, which ensures it is asymptotically normal. Term (I) is therefore asymptotically normal, and together with Term (II), the sum is asymptotically normal.

Since any linear combination $a^T Z$ is a sum of terms of the form (I) and (II) across components and lags, the same argument applies in both cases, and the Cramér-Wold device gives joint normality of $Z$. $\square$

## Diagnostics

To verify the formulas, we simulate scenario 2 ($n = 1000$, $L = 4$, $r = 0.3$) under two error distributions and compare the empirical covariance matrix of $Z$ to the theoretical $\Sigma_Z$.

For each error type, we run 500 outer replicates. In each replicate, we draw 10 independent datasets, compute $Z$ for each, and take their sample covariance as one estimate of $\Sigma_Z$. Each panel in the grid below shows the distribution of residuals (estimated minus theoretical) for one entry of $\Sigma_Z$ across the 500 replicates. The red line marks zero (no bias) and the orange line marks the sample mean. The p-value is from a one-sample t-test of whether the mean residual is zero.

```{r}
#| message: false
#| echo: false
devtools::load_all("~/Documents/Research/Code/ECE")
library(tidyverse)
```

```{r}
#| echo: false
cov_Z <- function(params) {
  n   <- params$n
  sx  <- sqrt(params$S[1, 1])
  sy  <- sqrt(params$S[2, 2])
  sxy <- params$S[1, 2]
  hx  <- params$h[, 1]
  hy  <- params$h[, 2]
  wx  <- lag_diff(hx) / n
  wy  <- lag_diff(hy) / n
  wxy <- lag_diff(hx, hy) / n
  rxy <- sxy / (sx * sy)

  k <- kappa_from_params(params)
  k22 <- k$k22; k40 <- k$k40; k04 <- k$k04; k31 <- k$k31; k13 <- k$k13

  S11 <- var_Tk_est(n, sx, wx, k40, k = 1)
  S12 <- cov_Th_Tk_est(n, sx, wx, k40)
  S13 <- cov_Tk_Rk_est(n, sx, sy, sxy, wxy, k22, k = 1)
  S14 <- cov_Th_Rk_est(n, sx, sy, sxy, wxy, k22)
  S15 <- cov_Tk_Qk_est(n, sx, sy, sxy, wx, wxy, k31, k = 1)
  S16 <- cov_Th_Qk_est(n, sx, sy, sxy, wx, wxy, k31)
  S22 <- var_Tk_est(n, sx, wx, k40, k = 2)
  S23 <- cov_Th_Rk_est(n, sx, sy, sxy, wxy, k22)
  S24 <- cov_Tk_Rk_est(n, sx, sy, sxy, wxy, k22, k = 2)
  S25 <- cov_Th_Qk_est(n, sx, sy, sxy, wx, wxy, k31)
  S26 <- cov_Tk_Qk_est(n, sx, sy, sxy, wx, wxy, k31, k = 2)
  S33 <- var_Tk_est(n, sy, wy, k04, k = 1)
  S34 <- cov_Th_Tk_est(n, sy, wy, k04)
  S35 <- cov_Tk_Qk_est(n, sy, sx, sxy, wy, wxy, k13, k = 1)
  S36 <- cov_Th_Qk_est(n, sy, sx, sxy, wy, wxy, k13)
  S44 <- var_Tk_est(n, sy, wy, k04, k = 2)
  S45 <- cov_Th_Qk_est(n, sy, sx, sxy, wy, wxy, k13)
  S46 <- cov_Tk_Qk_est(n, sy, sx, sxy, wy, wxy, k13, k = 2)
  S55 <- var_Qh_est(n, sx, sy, sxy, wx, wy, wxy, k22, k = 1)
  S56 <- cov_Qh_Qk_est(n, sx, sy, sxy, wx, wy, wxy, k22)
  S66 <- var_Qh_est(n, sx, sy, sxy, wx, wy, wxy, k22, k = 2)

  S_u_upper <- (1 / (4 * n^2)) * c(
    S11,
    S12, S22,
    S13, S23, S33,
    S14, S24, S34, S44,
    S15, S25, S35, S45, S55,
    S16, S26, S36, S46, S56, S66
  )
  S_u <- matrix(0, nrow = 6, ncol = 6)
  S_u[upper.tri(S_u, diag = TRUE)] <- S_u_upper
  S_u + t(S_u) - diag(diag(S_u))
}

sim_Z_once <- function(params) {
  map_dfr(1:10, ~{
    X <- generate_data(params)
    matrix(c(
      lag_diff(X[,1], X[,1], k = 1),
      lag_diff(X[,1], X[,1], k = 2),
      lag_diff(X[,2], X[,2], k = 1),
      lag_diff(X[,2], X[,2], k = 2),
      lag_diff(X[,1], X[,2], k = 1),
      lag_diff(X[,1], X[,2], k = 2)
    ), nrow = 1) |> as.data.frame()
  }) |> cov()
}
```

```{r}
#| echo: false
#| cache: true
err_types <- c("normal", "exponential")

sim_results <- map(err_types, \(et) {
  params <- scenario(2, n = 1000, L = 4, r = 0.3, err_type = et)
  list(
    params        = params,
    cov_estimates = map(1:500, ~sim_Z_once(params)),
    S_z           = 4 * params$n^2 * cov_Z(params)
  )
}) |> setNames(err_types)
```

```{r}
#| echo: false
z_labels <- c("T1(X,X)", "T2(X,X)", "T1(Y,Y)", "T2(Y,Y)", "T1(X,Y)", "T2(X,Y)")
idx <- expand.grid(i = 1:6, j = 1:6) |> filter(i <= j)

diag_plots <- imap(sim_results, \(res, et) {
  cov_long <- imap_dfr(res$cov_estimates, ~{
    mutate(idx, rep = .y, value = .x[cbind(i, j)] - res$S_z[cbind(i, j)])
  }) |> mutate(
    row_label = factor(z_labels[i], levels = z_labels),
    col_label = factor(z_labels[j], levels = z_labels)
  )

  panel_stats <- cov_long |>
    summarise(
      mean  = mean(value),
      p_val = t.test(value)$p.value,
      .by   = c(row_label, col_label)
    ) |>
    mutate(label = paste0("p=", formatC(p_val, format = "g", digits = 2)))

  cov_long |>
    ggplot(aes(x = value)) +
    geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
    geom_vline(data = distinct(cov_long, row_label, col_label),
               aes(xintercept = 0), color = "firebrick", linewidth = 0.8) +
    geom_vline(data = panel_stats, aes(xintercept = mean),
               color = "orange", linewidth = 0.8) +
    geom_text(data = panel_stats, aes(x = Inf, y = Inf, label = label),
              hjust = 1.1, vjust = 1.5, size = 2.5, color = "black") +
    facet_grid(row_label ~ col_label, scales = "free") +
    labs(x = NULL, y = NULL,
         title = "Residuals: estimated - theoretical (n_inner=10, n_outer=500)",
         subtitle = "Red = theoretical (0), orange = sample mean, p = t-test for bias") +
    theme_minimal(base_size = 9) +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
})
```

```{r}
#| echo: false
#| results: asis
cat("::: {.panel-tabset}\n\n")
iwalk(diag_plots, \(p, et) {
  cat(paste0("## ", et, "\n\n"))
  print(p)
  cat("\n\n")
})
cat(":::\n")
```
