---
title: "Z-Test"
---

::: callout-important
## Theorem: Asymptotic Distribution of $\hat\rho$

If $\max_j\{(\mu_{x,j} - \mu_{x,j+1})^2\} = o(n)$, $\max_j\{(\mu_{y,j} - \mu_{y,j+1})^2\} = o(n)$, and the errors have finite 4th moments, then the ECE correlation estimator satisfies
$$
\sqrt{n}(\hat\rho - \rho) \xrightarrow{d} N(0,\, v^2),
$$
where
$$
v^2 = \frac{1}{n}\left[
(1-\rho^2)\!\left(\frac{\omega_x^2}{\sigma_x^2} + \frac{\omega_y^2}{\sigma_y^2} - \frac{2\rho\,\omega_{xy}}{\sigma_x\sigma_y}\right)
+ \frac{\rho^2}{4}(\kappa_{40} + 2\kappa_{22} + \kappa_{04})
- \rho(\kappa_{31} + \kappa_{13})
+ \frac{5}{2}(1-\rho^2)^2 + \kappa_{22}
\right].
$$

::: {.callout-note collapse="true"}
## Remark: Asymptotic Variance Under Gaussian Errors

Under the Gaussian assumption ($\kappa_{40} = \kappa_{04} = 3$, $\kappa_{31} = \kappa_{13} = 3\rho$, $\kappa_{22} = 1 + 2\rho^2$), $v^2$ reduces to
$$
v^2 = \frac{1}{n}(1-\rho^2)\!\left(\frac{\omega_x^2}{\sigma_x^2} + \frac{\omega_y^2}{\sigma_y^2} - \frac{2\rho\,\omega_{xy}}{\sigma_x\sigma_y} + \frac{7}{2}(1-\rho^2)\right).
$$
:::

::: {.callout-note collapse="true"}
## Remark: Asymptotic Variance for $\rho = 0$ Under Gaussian Errors

Under the Gaussian assumption $\kappa_{22} = 1 + 2\rho^2 = 1$, so
$$
v^2 = \frac{1}{n}\!\left(\frac{\omega_x^2}{\sigma_x^2} + \frac{\omega_y^2}{\sigma_y^2} + \frac{7}{2}\right).
$$
:::

::: {.callout-note collapse="true"}
## Remark: Asymptotic Variance for $\rho = 0$

Setting $\rho = 0$ in the general formula,
$$
v^2 = \frac{1}{n}\!\left(\frac{\omega_x^2}{\sigma_x^2} + \frac{\omega_y^2}{\sigma_y^2} + \frac{5}{2} + \kappa_{22}\right).
$$
:::
:::

Define the lag-$k$ cross-product statistic $T_k(X, Y) = \sum_{i=1}^{n} (X_i - X_{i+k})(Y_i - Y_{i+k})$, and let
$$
Z = \begin{pmatrix} Z_1 \\ Z_2 \\ Z_3 \\ Z_4 \\ Z_5 \\ Z_6 \end{pmatrix} = \begin{pmatrix} T_1(X,X) \\ T_2(X,X) \\ T_1(Y,Y) \\ T_2(Y,Y) \\ T_1(X,Y) \\ T_2(X,Y) \end{pmatrix}.
$$

The proof applies the **Multivariate Delta Method** in three steps:

**Step 1.** Show that
$$
\frac{1}{\sqrt{n}}(Z - \mu_Z) \xrightarrow{d} N(\mathbf{0}, \Sigma_Z), \quad \mu_Z = \mathbb{E}[Z].
$$

::: {.callout-important collapse="true"}
## Proposition: Asymptotic Normality of $Z$

Under the model assumptions, if $\max_j\{(\mu_{x,j} - \mu_{x,j+1})^2\} = o(n)$ and $\max_j\{(\mu_{y,j} - \mu_{y,j+1})^2\} = o(n)$, and errors have finite 4th moments, then
$$
\frac{1}{\sqrt{n}}(Z - \mathbb{E}[Z]) \xrightarrow{d} N(\mathbf{0}, \Sigma_Z).
$$

::: {.callout-note collapse="true"}
## Proof: Asymptotic Normality of $Z$

**Strategy.** Each component of $Z$ is of the form $T_k(X,Y)$ for some lag $k$ and choice of series. It therefore suffices to show $T_k(X,Y)$ is asymptotically normal, since any linear combination $a^T Z$ inherits this structure. The Cramér-Wold device then promotes univariate normality to joint normality of the full vector.

By the Cramér-Wold device, it suffices to show every linear combination $a^T Z$ is asymptotically normal. We handle the general cross-product component $T_k(X,Y)$; the diagonal components $T_k(X,X)$ and $T_k(Y,Y)$ follow as special cases.

Substituting $X_i = \theta_{x,i} + \varepsilon_{x,i}$ and $Y_i = \theta_{y,i} + \varepsilon_{y,i}$ and expanding gives four terms:

\begin{align}
T_k(X,Y) &= \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\theta_{y,i} - \theta_{y,i+k}) \\
&\quad + \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k}) + \sum_{i=1}^n (\theta_{y,i} - \theta_{y,i+k})(\varepsilon_{x,i} - \varepsilon_{x,i+k}) \\
&\quad + \sum_{i=1}^n (\varepsilon_{x,i} - \varepsilon_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k}).
\end{align}

The first term equals $kW_{xy}$ and is deterministic. The last term has expectation $2n\sigma_{xy}$, so adding and subtracting this expectation gives

\begin{align}
T_k(X,Y) &= \underbrace{kW_{xy} + 2n\sigma_{xy}}_{\mathbb{E}[T_k(X,Y)]} \\
&\quad + \underbrace{\sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k}) + \sum_{i=1}^n (\theta_{y,i} - \theta_{y,i+k})(\varepsilon_{x,i} - \varepsilon_{x,i+k})}_{\text{(I) mixed}} \\
&\quad + \underbrace{\sum_{i=1}^n \bigl[(\varepsilon_{x,i} - \varepsilon_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k}) - 2\sigma_{xy}\bigr]}_{\text{(II) purely random}},
\end{align}

where $W_{xy} = T_1(\theta_x, \theta_y)$.

**Term (II)** is a sum of mean-zero, $k$-dependent, identically distributed random variables (since errors are i.i.d.), so the CLT for $m$-dependent sequences applies directly.

**Term (I).** Consider the first sum; the second is identical in structure with $x$ and $y$ exchanged. Splitting the error differences,
$$
\sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})(\varepsilon_{y,i} - \varepsilon_{y,i+k})
= \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})\varepsilon_{y,i} - \sum_{i=1}^n (\theta_{x,i} - \theta_{x,i+k})\varepsilon_{y,i+k},
$$
each of which is a weighted sum of independent errors with deterministic coefficients $c_i = \theta_{x,i} - \theta_{x,i+k}$. Since $\theta_x$ is piecewise constant, the nonzero coefficients take values $\{\mu_{x,j} - \mu_{x,j+1}\}$, giving $\sum_i c_i^2 = kW_x(\theta)$ and $\max_i c_i^2 = \max_j (\mu_{x,j} - \mu_{x,j+1})^2$.

The proposition condition $\max_j\{(\mu_{x,j} - \mu_{x,j+1})^2\} = o(n)$ implies one of two cases:

**Case 1:** $W_x(\theta) = o(n)$. Then Term (I) has standard deviation $O(\sqrt{W_x(\theta)}) = o(\sqrt{n})$, so Term (I)$/\sqrt{n} \xrightarrow{p} 0$. By Slutsky's theorem, $[{\rm Term\;(I)} + {\rm Term\;(II)}]/\sqrt{n}$ has the same limiting distribution as Term (II)$/\sqrt{n}$, which is normal by the CLT for $m$-dependent sequences.

**Case 2:** $W_x(\theta) = \Omega(n)$. Since $\max_j(\mu_{x,j} - \mu_{x,j+1})^2 = o(n)$ and $W_x(\theta) = \Omega(n)$, we have
$$
\frac{\max_j (\mu_{x,j} - \mu_{x,j+1})^2}{W_x(\theta)} \to 0,
$$
which is equivalent to $\max_i c_i^2 / \sum_i c_i^2 \to 0$. This is the Lindeberg condition for the weighted sum $\sum_i c_i \varepsilon_{y,i}$, which ensures it is asymptotically normal. Term (I) is therefore asymptotically normal, and together with Term (II), the sum is asymptotically normal.

Since any linear combination $a^T Z$ is a sum of terms of the form (I) and (II) across components and lags, the same argument applies in both cases, and the Cramér-Wold device gives joint normality of $Z$. $\square$
:::

:::

**Step 2.** Compute $\Sigma_Z = \mathrm{Cov}(Z)$.

::: {.callout-note collapse="true"}
## $\Sigma_Z$ Entries

Let
$$
\omega_x^2 = \frac{T_1(\theta_x, \theta_x)}{n}, \qquad
\omega_y^2 = \frac{T_1(\theta_y, \theta_y)}{n}, \qquad
\omega_{xy} = \frac{T_1(\theta_x, \theta_y)}{n},
$$
and let $\kappa_{rs} = \kappa_{rs}(\varepsilon_x, \varepsilon_y)$ be the joint cumulants of the error distribution. The covariance matrix $\Sigma_Z = \mathrm{Cov}(Z)$ has the following entries (upper triangle; the matrix is symmetric).

| Entry $(i,j)$ | $[\Sigma_Z]_{ij}$ |
|---|---|
| $(T_1(X,X),\, T_1(X,X))$ | $4n\kappa_{40}\sigma_x^4 + 8n\sigma_x^2\omega_x^2$ |
| $(T_1(X,X),\, T_2(X,X))$ | $4n(\kappa_{40}-1)\sigma_x^4 + 8n\sigma_x^2\omega_x^2$ |
| $(T_2(X,X),\, T_2(X,X))$ | $4n\kappa_{40}\sigma_x^4 + 16n\sigma_x^2\omega_x^2$ |
| $(T_1(Y,Y),\, T_1(Y,Y))$ | $4n\kappa_{04}\sigma_y^4 + 8n\sigma_y^2\omega_y^2$ |
| $(T_1(Y,Y),\, T_2(Y,Y))$ | $4n(\kappa_{04}-1)\sigma_y^4 + 8n\sigma_y^2\omega_y^2$ |
| $(T_2(Y,Y),\, T_2(Y,Y))$ | $4n\kappa_{04}\sigma_y^4 + 16n\sigma_y^2\omega_y^2$ |
| $(T_1(X,X),\, T_1(Y,Y))$ | $4n\sigma_x^2\sigma_y^2(\kappa_{22}-1) + 4n\sigma_{xy}^2 + 8n\sigma_{xy}\omega_{xy}$ |
| $(T_1(X,X),\, T_2(Y,Y))$ | $4n\sigma_x^2\sigma_y^2(\kappa_{22}-1) + 8n\sigma_{xy}\omega_{xy}$ |
| $(T_2(X,X),\, T_1(Y,Y))$ | $4n\sigma_x^2\sigma_y^2(\kappa_{22}-1) + 8n\sigma_{xy}\omega_{xy}$ |
| $(T_2(X,X),\, T_2(Y,Y))$ | $4n\sigma_x^2\sigma_y^2(\kappa_{22}-1) + 4n\sigma_{xy}^2 + 16n\sigma_{xy}\omega_{xy}$ |
| $(T_1(X,X),\, T_1(X,Y))$ | $4n\sigma_x^3\sigma_y\kappa_{31} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy}$ |
| $(T_1(X,X),\, T_2(X,Y))$ | $4n\sigma_x^3\sigma_y\kappa_{31} - 4n\sigma_x^2\sigma_{xy} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy}$ |
| $(T_2(X,X),\, T_1(X,Y))$ | $4n\sigma_x^3\sigma_y\kappa_{31} - 4n\sigma_x^2\sigma_{xy} + 4n\sigma_{xy}\omega_x^2 + 4n\sigma_x^2\omega_{xy}$ |
| $(T_2(X,X),\, T_2(X,Y))$ | $4n\sigma_x^3\sigma_y\kappa_{31} + 8n\sigma_{xy}\omega_x^2 + 8n\sigma_x^2\omega_{xy}$ |
| $(T_1(Y,Y),\, T_1(X,Y))$ | $4n\sigma_y^3\sigma_x\kappa_{13} + 4n\sigma_{xy}\omega_y^2 + 4n\sigma_y^2\omega_{xy}$ |
| $(T_1(Y,Y),\, T_2(X,Y))$ | $4n\sigma_y^3\sigma_x\kappa_{13} - 4n\sigma_y^2\sigma_{xy} + 4n\sigma_{xy}\omega_y^2 + 4n\sigma_y^2\omega_{xy}$ |
| $(T_2(Y,Y),\, T_1(X,Y))$ | $4n\sigma_y^3\sigma_x\kappa_{13} - 4n\sigma_y^2\sigma_{xy} + 4n\sigma_{xy}\omega_y^2 + 4n\sigma_y^2\omega_{xy}$ |
| $(T_2(Y,Y),\, T_2(X,Y))$ | $4n\sigma_y^3\sigma_x\kappa_{13} + 8n\sigma_{xy}\omega_y^2 + 8n\sigma_y^2\omega_{xy}$ |
| $(T_1(X,Y),\, T_1(X,Y))$ | $2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) - 2n\sigma_{xy}^2 + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2 + 4n\sigma_{xy}\omega_{xy}$ |
| $(T_1(X,Y),\, T_2(X,Y))$ | $4n(\sigma_x^2\sigma_y^2\kappa_{22} - \sigma_{xy}^2) + 2n\sigma_y^2\omega_x^2 + 2n\sigma_x^2\omega_y^2 + 4n\sigma_{xy}\omega_{xy}$ |
| $(T_2(X,Y),\, T_2(X,Y))$ | $2n\sigma_x^2\sigma_y^2(2\kappa_{22}+1) - 2n\sigma_{xy}^2 + 4n\sigma_y^2\omega_x^2 + 4n\sigma_x^2\omega_y^2 + 8n\sigma_{xy}\omega_{xy}$ |

:::

**Step 3.** Define
$$
g(Z) = \frac{2Z_5 - Z_6}{\sqrt{(2Z_1 - Z_2)(2Z_3 - Z_4)}} = \hat\rho.
$$
Since $g(\mu_Z) = \rho$, the multivariate delta method gives
$$
\sqrt{n}(\hat\rho - \rho) \xrightarrow{d} N(0,\, v^2), \quad v^2 = \nabla g(\mu_Z)^\top \Sigma_Z\, \nabla g(\mu_Z).
$$

::: {.callout-note collapse="true"}
## Gradient: $\nabla g(\mu_Z)$

The gradient of $g$ evaluated at $\boldsymbol{\mu}_Z$ is

$$
\nabla g(\boldsymbol{\mu}_Z) =
\begin{pmatrix}
-\dfrac{g(\mathbf{Z})}{2Z_1 - Z_2} \\[6pt]
\dfrac{1}{2}\dfrac{g(\mathbf{Z})}{2Z_1 - Z_2} \\[6pt]
-\dfrac{g(\mathbf{Z})}{2Z_3 - Z_4} \\[6pt]
\dfrac{1}{2}\dfrac{g(\mathbf{Z})}{2Z_3 - Z_4} \\[6pt]
\dfrac{2}{\sqrt{(2Z_1-Z_2)(2Z_3-Z_4)}} \\[6pt]
-\dfrac{1}{\sqrt{(2Z_1-Z_2)(2Z_3-Z_4)}}
\end{pmatrix}_{\mathbf{Z} = \boldsymbol{\mu}_Z}
= \frac{1}{2n}
\begin{pmatrix}
-\rho/\sigma_x^2 \\ \rho/(2\sigma_x^2) \\ -\rho/\sigma_y^2 \\ \rho/(2\sigma_y^2) \\ 2/(\sigma_x\sigma_y) \\ -1/(\sigma_x\sigma_y)
\end{pmatrix},
$$

using $2\mu_{Z_1} - \mu_{Z_2} = 2n\sigma_x^2$, $2\mu_{Z_3} - \mu_{Z_4} = 2n\sigma_y^2$, $2\mu_{Z_5} - \mu_{Z_6} = 2n\sigma_{xy}$, and $g(\boldsymbol{\mu}_Z) = \rho$.
:::

## Diagnostics

To verify the formulas, we simulate scenario 2 ($n = 1000$, $L = 4$, $r = 0.3$) under two error distributions and compare the empirical covariance matrix of $Z$ to the theoretical $\Sigma_Z$.

::: {.callout-note collapse="true"}
## Remark: $v^2$ is scale-invariant

$v^2$ is invariant under $X \to aX$, $Y \to bY$, consistent with $\widehat\rho(aX, bY) = \widehat\rho(X, Y)$. Each piece transforms as follows.

**$\omega$ terms.** Scaling $X$ by $a$ scales $\theta_x \to a\theta_x$, so $\omega_x^2 \to a^2\omega_x^2$ and $\omega_x \to a\omega_x$. Similarly $\omega_y \to b\omega_y$ and $\omega_{xy} \to ab\,\omega_{xy}$. The ratios
$$
\frac{\omega_x^2}{\sigma_x^2} \to \frac{a^2\omega_x^2}{a^2\sigma_x^2} = \frac{\omega_x^2}{\sigma_x^2}, \qquad
\frac{\rho\,\omega_{xy}}{\sigma_x\sigma_y} \to \frac{\rho\cdot ab\,\omega_{xy}}{a\sigma_x\cdot b\sigma_y} = \frac{\rho\,\omega_{xy}}{\sigma_x\sigma_y}
$$
are therefore unchanged.

**$\kappa$ terms.** The cumulants $\kappa_{rs}$ are moments of the standardized errors $\varepsilon_x/\sigma_x$ and $\varepsilon_y/\sigma_y$. Since scaling $X$ by $a$ scales both $\varepsilon_x$ and $\sigma_x$ by $a$, the ratio is unchanged, and all $\kappa_{rs}$ are invariant.

**$\rho$.** Invariant by definition of correlation.
:::

::: {.callout-note collapse="true"}
## Test: $\nabla g(\mu_Z)^\top \Sigma_Z \nabla g(\mu_Z)$ matches closed-form $v^2$

We verify the proposition by checking that the closed-form $V$ equals $\nabla g(\boldsymbol{\mu}_Z)^\top \Sigma_Z \nabla g(\boldsymbol{\mu}_Z)$ computed explicitly as a matrix product, mirroring the structure of `ece.cor.se` but with oracle inputs.

```{r}
#| message: false
devtools::load_all("~/Documents/Research/Code/ECE")

V_formula <- function(params) {
  n   <- params$n
  sx  <- sqrt(params$S[1, 1])
  sy  <- sqrt(params$S[2, 2])
  sxy <- params$S[1, 2]
  rho <- sxy / (sx * sy)
  hx  <- params$h[, 1]
  hy  <- params$h[, 2]

  wx  <- lag_diff(hx) / n
  wy  <- lag_diff(hy) / n
  wxy <- lag_diff(hx, hy) / n
  k   <- params$kappa

  (1 / n) * (
    (1 - rho^2) * (wx / sx^2 + wy / sy^2 - 2 * rho * wxy / (sx * sy)) +
    rho^2 / 4 * (k$k40 + 2 * k$k22 + k$k04) - rho * (k$k31 + k$k13) +
    5 / 2 * (1 - rho^2)^2 + k$k22
  )
}

V_matrix <- function(params) {
  n   <- params$n
  sx  <- sqrt(params$S[1, 1])
  sy  <- sqrt(params$S[2, 2])
  sxy <- params$S[1, 2]
  rho <- sxy / (sx * sy)
  hx  <- params$h[, 1]
  hy  <- params$h[, 2]

  k   <- params$kappa
  k40 <- k$k40; k04 <- k$k04; k31 <- k$k31; k13 <- k$k13; k22 <- k$k22

  # Theoretical Sigma_Z / (4n^2), using non-_est functions (take full hx/hy)
  # to avoid the normalization ambiguity in the _est counterparts
  S_u_upper <- (1 / (4 * n^2)) * c(
    var_Tk(n, sx, hx, k40, k = 1),
    cov_Th_Tk(n, sx, hx, k40),              var_Tk(n, sx, hx, k40, k = 2),
    cov_Tk_Rk(n, sx, sy, sxy, hx, hy, k22, k = 1),
    cov_Th_Rk(n, sx, sy, sxy, hx, hy, k22), var_Tk(n, sy, hy, k04, k = 1),
    cov_Th_Rk(n, sx, sy, sxy, hx, hy, k22),
    cov_Tk_Rk(n, sx, sy, sxy, hx, hy, k22, k = 2), cov_Th_Tk(n, sy, hy, k04),
    var_Tk(n, sy, hy, k04, k = 2),
    cov_Tk_Qk(n, sx, sy, sxy, hx, hy, k31, k = 1),
    cov_Th_Qk(n, sx, sy, sxy, hx, hy, k31),
    cov_Tk_Qk(n, sy, sx, sxy, hy, hx, k13, k = 1),
    cov_Th_Qk(n, sy, sx, sxy, hy, hx, k13),  var_Qh(n, sx, sy, sxy, hx, hy, k22, k = 1),
    cov_Th_Qk(n, sx, sy, sxy, hx, hy, k31),
    cov_Tk_Qk(n, sx, sy, sxy, hx, hy, k31, k = 2),
    cov_Th_Qk(n, sy, sx, sxy, hy, hx, k13),
    cov_Tk_Qk(n, sy, sx, sxy, hy, hx, k13, k = 2),
    cov_Qh_Qk(n, sx, sy, sxy, hx, hy, k22),  var_Qh(n, sx, sy, sxy, hx, hy, k22, k = 2)
  )
  S_u <- matrix(0, 6, 6)
  S_u[upper.tri(S_u, diag = TRUE)] <- S_u_upper
  S_u <- S_u + t(S_u) - diag(diag(S_u))

  # Gradient at mu_Z (1/(2n) factor absorbed into S_u scaling)
  dg <- c(-rho / sx^2, 0.5 * rho / sx^2,
          -rho / sy^2, 0.5 * rho / sy^2,
           2 / (sx * sy), -1 / (sx * sy))

  as.numeric(t(dg) %*% S_u %*% dg)
}

# scenario() fixes S = diag(1,1); patch manually to test sigma != 1
set_sigma <- function(params, sx, sy) {
  rho <- params$S[1, 2] / (sqrt(params$S[1, 1]) * sqrt(params$S[2, 2]))
  params$S <- matrix(c(sx^2, rho * sx * sy, rho * sx * sy, sy^2), 2, 2)
  params
}

base <- scenario(2, n = 1000, L = 4, r = 0.3, err_type = "normal")
scenarios <- list(
  set_sigma(base, sx = 1, sy = 1),
  set_sigma(base, sx = 2, sy = 1),
  set_sigma(base, sx = 1, sy = 3),
  set_sigma(base, sx = 2, sy = 2)
)

data.frame(
  sx        = c(1, 2, 1, 2),
  sy        = c(1, 1, 3, 2),
  V_formula = sapply(scenarios, V_formula),
  V_matrix  = sapply(scenarios, V_matrix)
)
```
:::

For each error type, we run 500 outer replicates. In each replicate, we draw 10 independent datasets, compute $Z$ for each, and take their sample covariance as one estimate of $\Sigma_Z$. Each panel in the grid below shows the distribution of residuals (estimated minus theoretical) for one entry of $\Sigma_Z$ across the 500 replicates. The red line marks zero (no bias) and the orange line marks the sample mean. The p-value is from a one-sample t-test of whether the mean residual is zero.

```{r}
#| message: false
#| echo: false
devtools::load_all("~/Documents/Research/Code/ECE")
library(tidyverse)
```

```{r}
#| echo: false
cov_Z <- function(params) {
  n   <- params$n
  sx  <- sqrt(params$S[1, 1])
  sy  <- sqrt(params$S[2, 2])
  sxy <- params$S[1, 2]
  hx  <- params$h[, 1]
  hy  <- params$h[, 2]
  wx  <- lag_diff(hx) / n
  wy  <- lag_diff(hy) / n
  wxy <- lag_diff(hx, hy) / n
  rxy <- sxy / (sx * sy)

  k <- kappa_from_params(params)
  k22 <- k$k22; k40 <- k$k40; k04 <- k$k04; k31 <- k$k31; k13 <- k$k13

  S11 <- var_Tk_est(n, sx, wx, k40, k = 1)
  S12 <- cov_Th_Tk_est(n, sx, wx, k40)
  S13 <- cov_Tk_Rk_est(n, sx, sy, sxy, wxy, k22, k = 1)
  S14 <- cov_Th_Rk_est(n, sx, sy, sxy, wxy, k22)
  S15 <- cov_Tk_Qk_est(n, sx, sy, sxy, wx, wxy, k31, k = 1)
  S16 <- cov_Th_Qk_est(n, sx, sy, sxy, wx, wxy, k31)
  S22 <- var_Tk_est(n, sx, wx, k40, k = 2)
  S23 <- cov_Th_Rk_est(n, sx, sy, sxy, wxy, k22)
  S24 <- cov_Tk_Rk_est(n, sx, sy, sxy, wxy, k22, k = 2)
  S25 <- cov_Th_Qk_est(n, sx, sy, sxy, wx, wxy, k31)
  S26 <- cov_Tk_Qk_est(n, sx, sy, sxy, wx, wxy, k31, k = 2)
  S33 <- var_Tk_est(n, sy, wy, k04, k = 1)
  S34 <- cov_Th_Tk_est(n, sy, wy, k04)
  S35 <- cov_Tk_Qk_est(n, sy, sx, sxy, wy, wxy, k13, k = 1)
  S36 <- cov_Th_Qk_est(n, sy, sx, sxy, wy, wxy, k13)
  S44 <- var_Tk_est(n, sy, wy, k04, k = 2)
  S45 <- cov_Th_Qk_est(n, sy, sx, sxy, wy, wxy, k13)
  S46 <- cov_Tk_Qk_est(n, sy, sx, sxy, wy, wxy, k13, k = 2)
  S55 <- var_Qh_est(n, sx, sy, sxy, wx, wy, wxy, k22, k = 1)
  S56 <- cov_Qh_Qk_est(n, sx, sy, sxy, wx, wy, wxy, k22)
  S66 <- var_Qh_est(n, sx, sy, sxy, wx, wy, wxy, k22, k = 2)

  S_u_upper <- (1 / (4 * n^2)) * c(
    S11,
    S12, S22,
    S13, S23, S33,
    S14, S24, S34, S44,
    S15, S25, S35, S45, S55,
    S16, S26, S36, S46, S56, S66
  )
  S_u <- matrix(0, nrow = 6, ncol = 6)
  S_u[upper.tri(S_u, diag = TRUE)] <- S_u_upper
  S_u + t(S_u) - diag(diag(S_u))
}

sim_Z_once <- function(params) {
  map_dfr(1:10, ~{
    X <- generate_data(params)
    matrix(c(
      lag_diff(X[,1], X[,1], k = 1),
      lag_diff(X[,1], X[,1], k = 2),
      lag_diff(X[,2], X[,2], k = 1),
      lag_diff(X[,2], X[,2], k = 2),
      lag_diff(X[,1], X[,2], k = 1),
      lag_diff(X[,1], X[,2], k = 2)
    ), nrow = 1) |> as.data.frame()
  }) |> cov()
}
```

```{r}
#| echo: false
#| cache: true
err_types <- c("normal", "exponential")

sim_results <- map(err_types, \(et) {
  params <- scenario(2, n = 1000, L = 4, r = 0.3, err_type = et)
  list(
    params        = params,
    cov_estimates = map(1:500, ~sim_Z_once(params)),
    S_z           = 4 * params$n^2 * cov_Z(params)
  )
}) |> setNames(err_types)
```

```{r}
#| echo: false
z_labels <- c("T1(X,X)", "T2(X,X)", "T1(Y,Y)", "T2(Y,Y)", "T1(X,Y)", "T2(X,Y)")
idx <- expand.grid(i = 1:6, j = 1:6) |> filter(i <= j)

diag_plots <- imap(sim_results, \(res, et) {
  cov_long <- imap_dfr(res$cov_estimates, ~{
    mutate(idx, rep = .y, value = .x[cbind(i, j)] - res$S_z[cbind(i, j)])
  }) |> mutate(
    row_label = factor(z_labels[i], levels = z_labels),
    col_label = factor(z_labels[j], levels = z_labels)
  )

  panel_stats <- cov_long |>
    summarise(
      mean  = mean(value),
      p_val = t.test(value)$p.value,
      .by   = c(row_label, col_label)
    ) |>
    mutate(label = paste0("p=", formatC(p_val, format = "g", digits = 2)))

  cov_long |>
    ggplot(aes(x = value)) +
    geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
    geom_vline(data = distinct(cov_long, row_label, col_label),
               aes(xintercept = 0), color = "firebrick", linewidth = 0.8) +
    geom_vline(data = panel_stats, aes(xintercept = mean),
               color = "orange", linewidth = 0.8) +
    geom_text(data = panel_stats, aes(x = Inf, y = Inf, label = label),
              hjust = 1.1, vjust = 1.5, size = 2.5, color = "black") +
    facet_grid(row_label ~ col_label, scales = "free") +
    labs(x = NULL, y = NULL,
         title = "Residuals: estimated - theoretical (n_inner=10, n_outer=500)",
         subtitle = "Red = theoretical (0), orange = sample mean, p = t-test for bias") +
    theme_minimal(base_size = 9) +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
})
```

```{r}
#| echo: false
#| results: asis
cat("::: {.panel-tabset}\n\n")
iwalk(diag_plots, \(p, et) {
  cat(paste0("## ", et, "\n\n"))
  print(p)
  cat("\n\n")
})
cat(":::\n")
```
